### 2021-07-23 codebook_df_anal.Rmd
First draft of analyses

---
title: "Codebook created on 2021-05-12 at 2021-05-12 17:31:08"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis')
```

A codebook contains documentation and metadata describing the contents, structure, and layout of a data file.

## Dataset description
The data contains 495 cases and 98 variables.

## Codebook

```{r}
summaries <- read.csv("./codebook_df_anal.csv", stringsAsFactors = FALSE)
summaries <- summaries[, !colSums(is.na(summaries)) == nrow(summaries)]
options(knitr.kable.NA = '')
knitr::kable(summaries, row.names = FALSE, digits = 2)
```

### Legend

* __Name__: Variable name
* __type__: Data type of the variable
* __missing__: Proportion of missing values for this variable
* __unique__: Number of unique values
* __mean__: Mean value
* __median__: Median value
* __mode__: Most common value (for categorical variables, this shows the frequency of the most common category)
* **mode_value**: For categorical variables, the value of the most common category
* __sd__: Standard deviation (measure of dispersion for numerical variables
* __v__: Agresti's V (measure of dispersion for categorical variables)
* __min__: Minimum value
* __max__: Maximum value
* __range__: Range between minimum and maximum value
* __skew__: Skewness of the variable
* __skew_2se__: Skewness of the variable divided by 2*SE of the skewness. If this is greater than abs(1), skewness is significant
* __kurt__: Kurtosis (peakedness) of the variable
* __kurt_2se__: Kurtosis of the variable divided by 2*SE of the kurtosis. If this is greater than abs(1), kurtosis is significant.

This codebook was generated using the [Workflow for Open Reproducible Code in Science (WORCS)](https://osf.io/zcvbs/)

### 2021-07-23 manuscript/manuscript.Rmd
First draft of analyses

---
title             : "The title"
shorttitle        : "Title"

author: 
  - name          : "First Author"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "my@email.com"
  - name          : "Ernst-August Doelle"
    affiliation   : "1,2"

affiliation:
  - id            : "1"
    institution   : "Wilhelm-Wundt-University"
  - id            : "2"
    institution   : "Konstanz Business School"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.

  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.

  One sentence clearly stating the **general problem** being addressed by  this particular study.

  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).

  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.

  One or two sentences to put the results into a more **general context**.

  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "keywords"
wordcount         : "X"

bibliography      : ["r-references.bib", "references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
knit              : worcs::cite_all
---

```{r setup, include = FALSE}
library("papaja")
library("worcs")
# We recommend that you prepare your raw data for analysis in 'prepare_data.R',
# and end that file with either open_data(yourdata), or closed_data(yourdata).
# Then, uncomment the line below to load the original or synthetic data
# (whichever is available), to allow anyone to reproduce your code:
# load_data()
```

This manuscript uses the Workflow for Open Reproducible Code in Science [@vanlissaWORCSWorkflowOpen2020] to ensure reproducibility and transparency. All code <!--and data--> are available at <https://github.com/cjvanlissa/veni_forest.git>.

This is an example of a non-essential citation [@@vanlissaWORCSWorkflowOpen2020]. If you change the rendering function to `worcs::cite_essential`, it will be removed.

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

# Methods
We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

## Participants

## Material

## Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

# Results

# Discussion

\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

### 2021-07-23 scripts/DependencePlots.Rmd
First draft of analyses

---
title: "DependencePlots"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(semtree)
library(metaforest)
library(gridExtra)
library(grid)
```

```{r loading objects}
forestdata <- readRDS("data/forestfile.RData")
vim <- readRDS("data/forestvim.RData")
```

```{r extracting importance values}
sort.values = T 
aggregate = "mean" 
horiz = T
las = 1 
convergence = F 
scale = "absolute" 
xlim = NULL 
head = NULL 
tail = NULL 
na.omit = FALSE

vimp <- vim
x <- semtree:::aggregateVarimp(vimp, aggregate, scale, na.omit)

vnames <- vimp$var.names
low <- min(x, na.rm = T) - 1
filt <- is.na(x)
x[filt] <- low
srt <- sort(x, index.return = T)
x <- x[srt$ix]
vnames <- vnames[srt$ix]
x[x <= (low + 0.5)] <- NA

selection <- 1:length(x)

impmat <- x[selection]
```

```{r partialDependence residual}
indxvec <- matrix(c(1:5, (length(impmat)-4):length(impmat)), nrow=1)

pDres <- readRDS("data/pDres.RData")
```

```{r partialDependence meani}
pDmi <- readRDS("data/pDmi.RData")
```

```{r partialDependence means}
pDms <- readRDS("data/pDms.RData")
```

```{r partialDependence meanq}
pDmq <- readRDS("data/pDmq.RData")
```

```{r plots,  results="asis"}
texttempl <- "This is variable of importance rank %i
Variable: %s
"
plotres <- list()
plotmi <- list()
plotms <- list()
plotmq <- list()

for(i in 1:length(indxvec)){
  plotres[[i]] <- ggplot() +
  geom_line(mapping=aes(x=as.matrix(pDres[[i]]$xgrid), y=as.matrix(unlist(pDres[[i]]$dict))), col="red") +
  geom_point(mapping=aes(x=as.matrix(pDres[[i]]$xgrid), y=as.matrix(unlist(pDres[[i]]$dict))), col="black", 
             shape=3, size=3) +
  labs(x=names(impmat)[indxvec[i]], y="residual") +
  theme_minimal()

  plotmi[[i]] <- ggplot() +
  geom_line(mapping=aes(x=as.matrix(pDmi[[i]]$xgrid), y=as.matrix(unlist(pDmi[[i]]$dict))), col="red") +
  geom_point(mapping=aes(x=as.matrix(pDmi[[i]]$xgrid), y=as.matrix(unlist(pDmi[[i]]$dict))), col="black", 
             shape=3, size=3) +
  labs(x=names(impmat)[indxvec[i]], y="mean i") +
  theme_minimal()

  plotms[[i]] <- ggplot() +
  geom_line(mapping=aes(x=as.matrix(pDms[[i]]$xgrid), y=as.matrix(unlist(pDms[[i]]$dict))), col="red") +
  geom_point(mapping=aes(x=as.matrix(pDms[[i]]$xgrid), y=as.matrix(unlist(pDms[[i]]$dict))), col="black", 
             shape=3, size=3) +
  labs(x=names(impmat)[indxvec[i]], y="mean s") +
  theme_minimal()

  plotmq[[i]] <- ggplot() +
  geom_line(mapping=aes(x=as.matrix(pDmq[[i]]$xgrid), y=as.matrix(unlist(pDmq[[i]]$dict))), col="red") +
  geom_point(mapping=aes(x=as.matrix(pDmq[[i]]$xgrid), y=as.matrix(unlist(pDmq[[i]]$dict))), col="black", 
             shape=3, size=3) +
  labs(x=names(impmat)[indxvec[i]], y="mean q") +
  theme_minimal()
}

for(i in 1:length(indxvec)){
  grid.arrange(plotres[[i]], plotmi[[i]], plotms[[i]], plotmq[[i]], nrow=2,
               top=textGrob(sprintf(texttempl, (length(impmat)-indxvec[i]+1), names(impmat)[indxvec[i]])))
}
```

### 2021-07-23 growth_trajectories.R

@@ -1,0 +1,149 @@
+dotsize=1
+scores <- data.frame(restsne$Y)
+p <- ggplot(data = scores, aes(x = X1, y = X2)) + geom_point(colour = "skyblue") + theme_bw()
+
+respc <- princomp(M_dist)
+
+p <- ggplot(data = as.data.frame(respc$scores[, 1:2]), aes(x = Comp.1, y = Comp.2)) + geom_point(colour = "skyblue") + theme_bw()
+
+tmp <- M_dist
+class(tmp) <- "matrix"
+psych::fa.parallel(tmp)
+# 3 classes
+res_k <- kmeans(M_dist, 3)
+table(res_k$cluster)
+
+mmg <- m0
+mmg$data <- NULL
+mmg <- lapply(1:max(res_k$cluster), function(i){
+  mod <- mmg
+  mod$S$labels[which(mod$S$labels == "vres")] <- paste0("vres", i)
+  mod$M$labels[which(grepl("^mean", mod$M$labels))] <- paste0(mod$M$labels[which(grepl("^mean", mod$M$labels))], i)
+  mxModel(name = paste0("group ", i), mod,
+          mxData(df_anal[res_k$cluster == i, paste0("de", 2:6)], type = "raw"),
+          mxFitFunctionML())
+})
+mmg <- do.call(mxModel, c(list(model = "mg_starts", mxFitFunctionMultigroup(paste0("group ", 1:max(res_k$cluster))), mmg)))
+mxRun(mmg) -> res_mmg
+summary(res_mmg)
+
+
+
+library(semtree)
+#> Loading required package: OpenMx
+#> OpenMx may run faster if it is compiled to take advantage of multiple cores.
+#> Registered S3 method overwritten by 'sets':
+#>   method        from   
+#>   print.element ggplot2
+set.seed(123)
+N <- 1000
+grp1 <- factor(sample(x = c(0,1), size=N, replace=TRUE))
+grp2 <- factor(sample(x = c(0,1), size=N, replace=TRUE))
+noise <- factor(sample(x = c(0,1),size=N, replace=TRUE))
+Sigma <- matrix(byrow=TRUE,
+                nrow=2,c(2,0.2,
+                         0.2,1))
+obs <- MASS::mvrnorm(N,mu=c(0,0),
+                     Sigma=Sigma)
+obs[,1] <- obs[,1] + ifelse(grp1==1,3,0)
[…]

### 2021-07-23 plot_growth.R

@@ -1,0 +1,115 @@
+
+library(semtree)
+#> Loading required package: OpenMx
+#> OpenMx may run faster if it is compiled to take advantage of multiple cores.
+#> Registered S3 method overwritten by 'sets':
+#>   method        from   
+#>   print.element ggplot2
+set.seed(123)
+N <- 1000
+grp1 <- factor(sample(x = c(0,1), size=N, replace=TRUE))
+grp2 <- factor(sample(x = c(0,1), size=N, replace=TRUE))
+noise <- factor(sample(x = c(0,1),size=N, replace=TRUE))
+Sigma <- matrix(byrow=TRUE,
+                nrow=2,c(2,0.2,
+                         0.2,1))
+obs <- MASS::mvrnorm(N,mu=c(0,0),
+                     Sigma=Sigma)
+obs[,1] <- obs[,1] + ifelse(grp1==1,3,0)
+obs[,2] <- obs[,2] + ifelse(grp2==1,3,0)
+df.biv <- data.frame(obs, grp1, grp2, noise)
+names(df.biv)[1:2] <- paste0("x",1:2)
+manifests<-c("x1","x2")
+model.biv <- mxModel("Bivariate_Model", 
+                     type="RAM",
+                     manifestVars = manifests,
+                     latentVars = c(),
+                     mxPath(from="x1",to=c("x1","x2"), 
+                            free=c(TRUE,TRUE), value=c(1.0,.2) , 
+                            arrows=2, label=c("VAR_x1","COV_x1_x2") ),
+                     mxPath(from="x2",to=c("x2"), free=c(TRUE), 
+                            value=c(1.0) , arrows=2, label=c("VAR_x2") ),
+                     mxPath(from="one",to=c("x1","x2"), label=c("mu1","mu2"),
+                            free=TRUE, value=0, arrows=1),
+                     mxData(df.biv, type = "raw")
+);
+result <- mxRun(model.biv)
+
+
+fp <- "mu2" # predicted by grp2
+#fp <- "mu1" # predicted by grp1
+
+tree.biv <- semtree(model.biv, data=df.biv, constraints = list(focus.parameters=fp))
+#> ✔ Tree construction finished [took 1s].
+
+semtree:::plot.semtree(tree.biv)
+
+fp <- "mu2" # predicted by grp2
+forest <- semforest(model.biv, data=df.biv,
[…]

### 2021-07-23 prepare_data.R

@@ -1,0 +1,42 @@
+# In this file, write the R-code necessary to load your original data file
+# (e.g., an SPSS, Excel, or SAS-file), and convert it to a data.frame. Then,
+# use the function open_data(your_data_frame) or closed_data(your_data_frame)
+# to store the data.
+
+library(worcs)
+library(foreign)
+imp_fun <- function(x){
+  UseMethod("imp_fun", x)
+}
+imp_fun.data.frame <- function(x){
+  data.frame(lapply(x, imp_fun))
+}
+imp_fun.default <- function(x){
+  out <- x
+  if(inherits(x, "numeric")){
+    out[is.na(out)] <- mean(x[!is.na(out)])
+  } else {
+    out[is.na(out)] <- names(sort(table(out), decreasing = TRUE))[1]
+    if(!class(out) == class(x)){
+      tryCatch({
+        out <- do.call(paste0("as.", class(x)[1]), list(out))
+      }, error = function(e){
+        stop("Could not impute missing values.")
+      })
+    }
+  }
+  out
+}
+
+df <- read.spss("DADradar-27-07-2020-Van_Lissa_2.sav", to.data.frame = TRUE, use.value.labels = FALSE)
+names(df) <- tolower(names(df))
+# Pseudonymize
+df[c("rpp", "bg41aa14b3", grep("^id", names(df), value = TRUE))] <- NULL
+
+# Check descriptives
+desc <- descriptives(df)
+desc[which(desc$unique == 1), ]
+# table(df$rpp)
+
+# Store data
+closed_data(df, codebook = NULL, synthetic = FALSE)

### 2021-07-23 run_me.R

@@ -1,0 +1,14 @@
+library(drake)
+
+plan <- drake_plan(
+  raw_data = readxl::read_excel(file_in("raw_data.xlsx")),
+  data = raw_data %>%
+    mutate(Ozone = replace_na(Ozone, mean(Ozone, na.rm = TRUE))),
+  hist = create_plot(data),
+  fit = lm(Ozone ~ Wind + Temp, data),
+  report = rmarkdown::render(
+    knitr_in("report.Rmd"),
+    output_file = file_out("report.html"),
+    quiet = TRUE
+  )
+)

### 2021-07-23 scripts/assessingforest.R

@@ -1,0 +1,41 @@
+### Emotional Dysregulation # Assessing SEM-forest results ###
+
+#############################
+# Lukas Beinhauer
+# 27/06/20
+#############################
+
+########################################################################################################
+# Over the following lines, code can be found to assess the results from the SEM-forest                #
+# Important: Run <missRanger.R>, <creatingscales.R > and < plantingforest.R > first!                                  #
+########################################################################################################
+
+forestdata <- readRDS("data/forestfile.RData")
+
+library(semtree)
+library(metaforest)
+
+
+# system.time(# time of growing forest consisting of 10 trees: 238.52s; forest with 50 trees 1167.93s
+#   vim <- varimp(predgrowthforest)
+# ) #assessing variabel importance of that forest: 93.4 s; forest w/ 50 trees: 343.36s
+
+system.time(
+  vim <- varimp(forestdata)
+)
+
+plot(vim) 
+#result from 50 trees, top 5:
+#   1.: ra11aa: 14.666
+#   2.: as11aa: 14.122
+#   3.: ys11aa: 11.441
+#   4.: sc11aa: 10.262
+#   5.: pp11av: 9.700
+
+print(vim)
+# vim$importance
+
+# forestdata$forest[[1]]$param_names
+
+saveRDS(vim, "data/forestvim.RData")
+

### 2021-07-23 scripts/cj.R

@@ -1,0 +1,179 @@
+library(foreign)
+imp_fun <- function(x){
+  UseMethod("imp_fun", x)
+}
+imp_fun.data.frame <- function(x){
+  data.frame(lapply(x, imp_fun))
+}
+imp_fun.default <- function(x){
+  out <- x
+  if(inherits(x, "numeric")){
+    out[is.na(out)] <- mean(x[!is.na(out)])
+  } else {
+    out[is.na(out)] <- names(sort(table(out), decreasing = TRUE))[1]
+    if(!class(out) == class(x)){
+      tryCatch({
+        out <- do.call(paste0("as.", class(x)[1]), list(out))
+      }, error = function(e){
+        stop("Could not impute missing values.")
+      })
+    }
+  }
+  out
+}
+
+df <- read.spss("c:/git_repositories/veni_forest/DADradar-27-07-2020-Van_Lissa_2.sav", to.data.frame = TRUE, use.value.labels = FALSE)
+names(df) <- tolower(names(df))
+names(df)[match("rir11aa19r", names(df))] <- "rir11aa19"
+
+# Compare to requested scales
+scalesrx <- read.csv("c:/git_repositories/veni_forest/selected_scales.csv", stringsAsFactors = FALSE)
+scalesrx <- unlist(sapply(scalesrx$item.names, function(x){strsplit(x, " ")[[1]]}, USE.NAMES = FALSE))
+# Only parent sex and age missing, but we have other variables for this
+#scalesrx[!scalesrx %in% names(df)]
+
+# Check what these are
+write.table(names(df)[!names(df) %in% scalesrx], "clipboard", sep = "\n", row.names = F, col.names = F)
+
+# Recode missings
+is_mis <- df == 888|df == 999
+df[is_mis] <- NA
+is_mis <- sapply(df, trimws)
+is_mis <- is_mis == ""
+df[is_mis] <- NA
+
+#which(colSums(is_mis, na.rm = TRUE) > 0)
+
+# Retain only relevant variables
+dropthese <- c("id_gezin", "id_gezin_jonglabel", "idp_tar", "idp_moe", "radar_lb",
+               "moeder_wisseling", "sib_wisseling", "vader_wisseling", "age.difference.target.between.w6.and.w8",
[…]

### 2021-07-23 scripts/creatingscales.R

@@ -1,0 +1,123 @@
+### Emotional Dysregulation # Creating Scales from Radar ###
+
+#############################
+# Lukas Beinhauer
+# 08/07/20
+#############################
+
+########################################################################################################
+# Over the following lines, code can be found in order to create scales from the synthetic RADAR data. #
+# Important: run <missRanger.R> first!                                                                 #
+# Scale scores are averages of each individual's item scores per scale/subscale                        #
+########################################################################################################
+
+library(tidySEM)
+library(lavaan)
+library(psych)
+library(ggplot2)
+library(semTools)
+library(worcs)
+
+data <- readRDS("ImputedData.RData") #load data
+
+scales_list <- readRDS("scales_list.RData")
+
+
+# Dependent variable ------------------------------------------------------
+
+m_config <- character()
+for(i in 2:6){
+  m_config <- paste0(m_config,
+                     paste0(
+                       "de", i, " =~ ", 
+                       paste0(grep(paste0("^de", i), names(data), value = TRUE), collapse = " + ")
+                       ),
+                     sep = "\n"
+                     )
+}
+itemcors <- paste0(unlist(lapply(gsub(".{6}", "", grep("^de", names(data), value = TRUE)), function(i){
+  #i = "03"
+  m <- matrix(apply(expand.grid(paste0("de", 2:6, "1aa", i), "~~", paste0("de", 2:6, "1aa", i)), 1, paste0, collapse = " "), ncol = 5, byrow = TRUE)
+  as.vector(m[lower.tri(m)])
+})), collapse = "\n")
+m_config <- paste0(m_config, itemcors)
+
+r_config <- sem(m_config, data)
+
+df_plot <- parameterestimates(r_config)
+df_plot <- df_plot[df_plot$lhs %in% paste0("de", 2:6) & !df_plot$rhs %in% paste0("de", 2:6), ]
+df_plot$rhs <- ordered(as.integer(gsub("^.{6}", "", df_plot$rhs)))
+
+ggplot(df_plot, aes(y = rhs, x = est)) +
+  geom_point() +
+  geom_errorbarh(aes(xmin = ci.lower, xmax = ci.upper)) +
+  facet_wrap(~lhs)
+
+ggplot(df_plot, aes(y = rhs, x = est, shape = lhs)) +
[…]

### 2021-07-23 scripts/missRanger.R

@@ -1,0 +1,184 @@
+### Emotional Dysregulation # Imputing data using missForest algorithm ###
+
+# (missRanger is faster implementation of missForest)
+
+########################################################################################################
+# Over the following lines, code can be found in order to implement the missForest approach to data    #
+# imputation. A random forest will be grown to build an imputation model.                              #
+########################################################################################################
+
+library(missRanger)
+library(worcs)
+try_numeric <- function(x){
+  miss_x <- is.na(x)
+  y <- tryCatch({as.numeric(x)},
+                error = function(e){
+                  x
+                })
+  miss_y <- is.na(y)
+  if(!all(miss_y == miss_x)){
+    return(y)
+  }
+  x
+}
+
+drop_floats <- function(x){
+  tmp <- table(x)
+  is_int <- as.numeric(names(tmp)) %% 1 == 0
+  if(any(!is_int)){
+    not_int <- which(x %% 1 != 0)
+    x[not_int] <- NA
+  }
+  x
+}
+# Loading the data (important: First export the data in SPSS to a .csv file. The initial .sav data is 
+#   password-protected. Save the file as <radardata.csv> in the subdirectory of the R-project)
+df <- load_data(to_envir = FALSE)$df
+
+
+# Fix problem with names --------------------------------------------------
+
+names(df)[names(df) == "rir11aa19r"] <- "rir11aa19"
+
+# Drop multiwave data, except for DERS
+df[grepl("^.{2}[2-9]", names(df))& !grepl("^de", names(df))] <- NULL
+
+# it would appear that some variables are recognized as characters/strings when read by R
+#   however, all relevant variables should be numerical (mostly categorical, but numerical anyways)
+indx1 <- sapply(df, is.character)
+if(any(indx1)) browser()
+#df[indx1] <- as.data.frame(lapply(df[indx1], function(x) as.numeric(x))) # Therefore those are changed to numeric
+
+# Some errors appear to sit in the df (non-integer values for scales where only integers are allowed)
+#   This is most likely due to imputation measures undertaken by the RADAR group (only for participants
[…]

### 2021-07-23 scripts/partDependence.R

@@ -1,0 +1,75 @@
+### Emotional Dysregulation # Generating partial dependence ###
+
+#############################
+# Lukas Beinhauer
+# 24/08/20
+#############################
+
+########################################################################################################
+# Over the following lines, code can be found in order to generate partial dependence for the most and #
+# least relevant variables as identified using the variable importance. Run the scripts <missRanger.R>,#
+# <creatingscales.R>, <plantingforest.R>, and <assessingforest.R> first.!                              #
+########################################################################################################
+
+library(semtree)
+library(metaforest)
+library(gridExtra)
+
+# load data and variable importance
+forestdata <- readRDS("data/forestfile.RData")
+vim <- readRDS("data/forestvim.RData")
+
+
+######   extracting variable importance values and sorting variables by importance (ascending)   ######
+# code stems from github.com/brandmaier/semtree (can also be found in semtree package)
+sort.values = T 
+aggregate = "mean" 
+horiz = T
+las = 1 
+convergence = F 
+scale = "absolute" 
+xlim = NULL 
+head = NULL 
+tail = NULL 
+na.omit = FALSE
+
+vimp <- vim
+x <- semtree:::aggregateVarimp(vimp, aggregate, scale, na.omit)
+
+vnames <- vimp$var.names
+low <- min(x, na.rm = T) - 1
+filt <- is.na(x)
+x[filt] <- low
+srt <- sort(x, index.return = T)
+x <- x[srt$ix]
+vnames <- vnames[srt$ix]
+x[x <= (low + 0.5)] <- NA
+
+selection <- 1:length(x)
+
+impmat <- x[selection] #returns a matrix containing variable importance of all variables supplied to the semforest
+
+
+
+# select top 5 and bottom 5 of most relevant variables
+indxvec <- matrix(c(1:5, (length(impmat)-4):length(impmat)), nrow=1)
+
+# calculate partial dependence for selected variables concerning "residual" parameter
+pDres <- apply(indxvec, 2, FUN=function(x) partialDependence(forestdata, reference.var=names(impmat[x]), 
+                                                             reference.param="residual"))
+saveRDS(pDres, file="data/pDres.RData")
+
+# calculate partial dependence for selected variables concerning "mean intercept" parameter
+pDmi <- apply(indxvec, 2, FUN=function(x) partialDependence(forestdata, reference.var=names(impmat[x]), 
+                                                            reference.param="meani"))
+saveRDS(pDmi, file="data/pDmi.RData")
+
[…]

### 2021-07-23 scripts/plantingforest.R

@@ -1,0 +1,122 @@
+### Emotional Dysregulation # SEM-trees ###
+
+########################################################################################################
+# Over the following lines, code can be found in order grow SEM-trees or a SEM-forest                  #
+# Important: Run  <creatingscales.R > first!                                                           #
+########################################################################################################
+library(worcs)
+library(semtree)
+library(semPlot)
+library(lavaan)
+library(tidySEM)
+library(semtree)
+library(parallel)
+library(metaforest)
+library(future)
+
+df_anal <- load_data(to_envir = FALSE)$df_anal
+
+dich_vars <- c("cigarettes", "alcohol", "drugs")
+df_anal[dich_vars] <- lapply(df_anal[dich_vars], function(i){as.factor(!(i == min(i, na.rm = TRUE)))})
+
+facs <- c("geslacht", "brpmoe_lmh", "brpvad_lmh", "sesgez_low", "bg11aa04", "reli")
+df_anal[facs] <- lapply(df_anal[facs], as.factor)
+
+# Specifying model linear growth curve model using the openMX code (should be the same model as the lavaan model)
+
+dersvar <- c(paste0("de", 2:6))
+predvar <- names(df_anal)[!names(df_anal) %in% dersvar]
+df_anal <- as.data.frame(df_anal)
+df_anal <- df_anal[, !names(df_anal) %in% c("awareness", "clarity", "impulsivity", 
+                        "goals", "accept", "strategies")]
+
+m0 <- as_ram("i =~ 1*de2 + 1*de3 + 1*de4 + 1*de5 + 1*de6
+s =~ 0*de2 + 1*de3 + 2*de4 + 3*de5 + 4*de6
+q =~ 0*de2 + 1*de3 + 4*de4 + 9*de5 + 16*de6
+i ~ meani*1
+s ~ means*1
+q ~ meanq*1
+de2 ~~ vres*de2
+de3 ~~ vres*de3
+de4 ~~ vres*de4
+de5 ~~ vres*de5
+de6 ~~ vres*de6
+de2 ~ 0*1
+de3 ~ 0*1
+de4 ~ 0*1
+de5 ~ 0*1
+de6 ~ 0*1
+i ~~ 0*s
+i ~~ 0*q
+s ~~ 0*q
+i ~~ 0*i
+s ~~ 0*s
+q ~~ 0*q")
+m0 <- run_mx(m0, data = df_anal)
+table_results(m0)
+
+# fitting a single tree using the model
+
[…]

### 2022-02-13 manuscript/manuscript.Rmd
papr writing

@@
 ---
-title             : "The title"
+  title             : "Complementing theory on adolescent emotion regulation using data-driven insights from machine learning"
 shorttitle        : "Title"
@@
 author:
-  - name          : "First Author"
-    affiliation   : "1"
-    corresponding : yes    # Define only one corresponding author
-    address       : "Postal address"
-    email         : "my@email.com"
-  - name          : "Ernst-August Doelle"
-    affiliation   : "1,2"
+  - name: "Caspar J. Van Lissa"
+    affiliation: "1,2"
+    corresponding: yes
+    address: "Padualaan 14, 3584CH Utrecht, The Netherlands"
+    email: "c.j.vanlissa@uu.nl"
+  - name: "Lukas Beinhauer"
+    affiliation: "3"
+  - name: "Susan Branje"
+    affiliation: "4"
+  - name: "Wim H. J. Meeus"
+    affiliation: "4"
 
 affiliation:
-  - id            : "1"
-    institution   : "Wilhelm-Wundt-University"
-  - id            : "2"
-    institution   : "Konstanz Business School"
+  - id: "1"
+    institution: "Utrecht University, department of Methodology & Statistics"
+  - id: "2"
+    institution: "Open Science Community Utrecht"
+  - id: "3"
+    institution: "Helmut-Schmidt-Universität, Department of Methodology and Statistics for Psychology"
+  - id: "4"
+    institution: "Utrecht University, department of Youth and Family"
 
 authornote: |
-  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
-
-  Enter author note here.
+  Data of the RADAR (Research on Adolescent Development And Relationships) study were used (https://doi.org/10.17026/dans-zrb-v5wp).
+RADAR has been financially supported by main grants from the Netherlands Organisation for Scientific Research (GB-MAGW 480-03-005, GB-MAGW 480-08-006), Stichting Achmea Slachtoffer en Samenleving (SASS), the Netherlands Organisation for Scientific Research to the Consortium Individual Development (CID; 024.001.003), a grant of the European Research Council (ERC-2017-CoG - 773023 INTRANSITION), and various other grants from the Netherlands Organisation for Scientific Research, VU University Amsterdam, and Utrecht University.
+The funders had no role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript.
+If the regulations of the journal allow it, the participating children, parents and schools are to be thanked as well.
 
 abstract: |
+  https://www.srcd.org/news/child-development-special-section-formalizing-theories-child-development
   One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
@@
 
-This manuscript uses the Workflow for Open Reproducible Code in Science [@vanlissaWORCSWorkflowOpen2020] to ensure reproducibility and transparency.
-All code <!--and data--> are available at <https://github.com/cjvanlissa/veni_forest.git>.
+<!-- TS: Emotion regulation development is a critical challenge in adolescence. -->
+<!-- TS: Emotion dysregulation has severe consequences for individual adolescents, and society at large. -->
+TS: Therefore, substantial research has addressed this topic.
+TS: Most preceding research has been limited in scope, focusing on only a few theoretically-driven predictors.
+TS: Machine learning makes it possible to cast a wider net, and explore all (available) potentially relevant predictors.
+TS: Confirmatory versus exploratory research: Define the distinction.
+TS: Some longitudinal research falls between these categories.
+For example, complex longitudinal studies, where prior research is sufficient to lead us to hypothesize *some* associations between two constructs, but not sufficiently specific to let us hypothesize which parameters will be significant.
+TS: Instead, us
 
-This is an example of a non-essential citation [@@vanlissaWORCSWorkflowOpen2020]. If you change the rendering function to `worcs::cite_essential`, it will be removed.
+A key challenge in adolescence is developing mature emotion regulation abilities [@zimmermannEmotionRegulationEarly2014]. 
+Adolescents experience biological, cognitive, and social changes that create an opportunity for positive emotion regulation development [@steinbergAgeOpportunityLessons2014].
+Still, as many as 20% of adolescents instead experience severe emotion regulation difficulties [@leeAdolescentMentalHealth2014],
+with potentially lifelong implications for mental health and well-being [@aldaoEmotionregulationStrategiesPsychopathology2010].
+These difficulties in emotion regulation can cause substantial distress to the individual,
+and incur a sizable societal cost.
+Researchers thus seek to understand what risk factors predispose some adolescents to emotion dysregulation,
+so that effective screening and interventions can be developed.
+Despite abundant empirical research, however,
+an overarching understanding is still lacking because
+the literature is fragmented across subdisciplines 
+and lacks consistent terminology and guiding theory [@vanlissaMappingPhenomenaRelevant2021].
 
-```{r analysis-preferences}
-# Seed for random number generation
-set.seed(42)
-knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
+The present study seeks to gain a more comprehensive understanding of adolescent emotion regulation development by using machine learning methods for rigorous exploration.
+<!-- These predictors may include both risk factors that render some youth susceptible to emotion regulation difficulties, and correlates that can be used to identify at-risk youth. -->
+The SEM-forests algorithm was used to
+identify the most important early indicators of developmental trajectories
+of emotion regulation across adolescence, from a large set of potential candidates.
+The resulting insights can guide future research and interventions to focus on the most important predictors,
+and lays the foundation for formation and revision of overarching theory, such that essential phenomena are adequately represented.
+
+Recently, a text mining systematic review of 6305 publications on adolescents' emotion regulation has revealed what phenomena
+
+Theory cons methods for theory construction state that the first step in this endeavor is to identify well-established relevant *empirical phenomena*,
+defined as stable and general features of the world [@borsboomTheoryConstructionMethodology2020].
+The current study used a text mining systematic review to provide a comprehensive map of the empirical phenomena relevant to adolescent emotion regulation.
+
+Currently, few instruments exist for the identification of relevant phenomena; instead, theorists typically rely on expert opinion [@borsboomTheoryConstructionMethodology2020].
+This paper argues that systematic reviews are a suitable and relatively more objective method for phenomena detection.
+Traditional narrative reviews are limited, however, by small convenience samples,
+confirmation bias,
+and an undue emphasis on positive results [@littellEvidencebasedBiasedQuality2008].
+These limitations are overcome by a novel method:
+the text mining systematic review (TMSR).
+Relative to narrative reviews, text mining can cover arbitrarily greater corpora,
+and gleans insights from the literature by means of a relatively more transparent, objective, and reproducible process.
+<!--Where sentient readers tend to structure their reading of the literature around established ideas [@littellEvidencebasedBiasedQuality2008],
+the inherent "fairness" of text mining gives emerging themes a chance to come to the fore.-->
+This approach assumes the frequency with which phenomena recur in the literature to be indicative of their relevance.
+If it is additionally assumed that the frequency with which phenomena are studied together is indicative of the extent to which they are thought to be associated,
+then the co-occurrence of phenomena within publications can be interpreted as a rudimentary nomological network: a mapping of relationships between theoretically relevant phenomena [see @alaviAligningTheoryMethodology2018].
+Such a network can serve as a starting point for the formulation of a proto-theory,
+which would additionally involve abductive reasoning: specifying putative mechanisms to explain associations between the phenomena [@borsboomTheoryConstructionMethodology2020].
+Text mining systematic reviews constitute an inductive and exploratory approach:
+Their aim is phenomena detection based on generalization from observations (data) without appeal to theory [@haigExploratoryFactorAnalysis2005].
+By identifying and mapping the phenomena relevant to adolescent emotion regulation,
+such an inductive approach lays the groundwork for future theory development.
+
+<!--
+stifterEmotionRegulation2019
+* call for studies incorporating predictors of ER across multiple levels of analysis
+* different conceptual frameworks
+
+riedigerEmotionRegulationAdolescence2014
+* difficulty in deriving a cohesive picture about emotion regulation in adolescence
+* different conceptual frameworks
+
+* three themes (neurocognitive development, socialization, interplay between internal and external factors) currently represent relatively independent bodies of research
+* A more explicit integration of these various research perspectives will help us to arrive at a more integrated and interpretable picture of the development of emorion regulation during adolescence.
+
+@bussTheoriesEmotionalDevelopment2019
+
+* emotional development theory is still in its own state of development. 
+* empirical evidence can contribute to theory development.
+
+@bariolaChildAdolescentEmotion2011
+* Further research apply reﬁned theoretical conceptualizations of ER.
+
+Where sentient readers tend to structure their reading of the literature around established ideas [@littellEvidencebasedBiasedQuality2008],
+This potentially reveals phenomena and associations missed by narrative reviewers,
+thereby laying the groundwork for future empirical and theoretical work.
+
+we set out to map these relevant empirical phenomena using a systematic review.
+
+generate a proto-theoretical nomological network of the relevant empirical phenomena.
+We discuss how this nomological network relates to existing relevant theory, and conclude
+with recommendations for future theory-generating efforts.
+There have been calls for the  consolidation of these disparate insights into overarching theory would be an important step forward.
+An important first step 
+and intrinsic limitations of narrative reviews prohibit comprehensive coverage of the vast but diffuse literature .-->
+<!--
+Recent years have seen the publication of several excellent systematic reviews,
+but these have all been narrative reviews.-->
+
+ <!-- CJ To fill out gaps, following up on promising constructs and unify heretofore disparate regions of the lit -->
+<!-- SH: this implies that the main contribution of the paper is predominantly methodological is that the case?
+Is there some primary thesis/argument that will be advanced?
+It is really important to take a top-down approach to that messaging, beginning here in the first paragraph-->
+<!-- Ergens anders: Similarly, although
+there are several *relevant* theories, there is no specific theoretical
+framework to guide research on adolescents' emotion regulation. Although theory
+development is beyond the scope of this paper, we set out to generate a
+proto-theoretical nomological network of the potential risk factors,
+manifestations, and outcomes of emotion regulation in adolescence. We discuss
+how this nomological network relates to existing relevant theory, and conclude
+with recommendations for future theory-generating efforts.-->
+<!-- At the same time, the staggered development of motivational-emotional and regulatory brain circuits gives rise
+to a maturity gap [@croneUnderstandingAdolescencePeriod2012]. 
+This leads youngsters to pursue new experiences in life and love,
+without being fully prepared to cope with the emotional outcomes. -->
+
+<!--The literature thus paints a picture of
+adolescence as a chrysalis for emotional development: Children enter this stage
+with emotion regulation skills adapted to the challenges of childhood. During
+adolescence, emotional systems are rearranged substantially.-->
+
+<!--
+* intense negative emotions in daily life (Silk, Steinberg, & Morris, 2003) 
+* unstable peer or romantic relationships (Furman & Collins, 2009)
+* decrease in perceived support from parents (Furman & Buhrmester, 1992).
+* early adolescence is characterized by a higher rate of conflicts with parents (Laursen, Coy, & Collins, 1998)
+* higher variability of negative emotions compared to late adolescence (Larson, Moneta, Richards, & Wilson, 2002).
+* relative to other developmental stages, adolescence involves a particularly large number of transitions, novel situations, and new stimuli, physical, cognitive, emotional, and social development—particularly in early and mid-adolescence (Seiffge-Krenke, 2000)
+-->
+
+## Knowledge gap
+
+Substantial empirical research, but:
+
+* Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014]
+* No consistent terminology [@bariolaChildAdolescentEmotion2011]
+* No conceptual frameworks [@stifterEmotionRegulation2019]
+* No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]
+
+Knowledge must be consolidated into overarching theory
+
+## Relevant theory
+
+* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
+* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
+* Hall's notion of "storm and stress"
+* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
+* Tripartite model [@morrisRoleFamilyContext2007]
+* Internalization model [@holodynskiDevelopmentEmotionsEmotion2006]
+* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
+* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
+* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
+* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]
+
+## Limitations of relevant theory
+
+* Few explicitly address adolescence
+* Few comprehensively address predictors of emotion regulation
+* None directly guide contemporary research 
+* Theories vary widely in scope: Some are broad and non-specific; others describe a specific phenomenon in detail, but lack a broader perspective.
+    + Broad theories can frame any research, specific theories generate hypotheses.
+    + It would be beneficial to bridge these levels of analysis.
+
+## Towards integrative theory
+
+First step: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
+
+We conducted a text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]
+
+  * Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
+  * TMSR: Unlimited sample size, transparent, objective, reproducible
+
+6653 papers on Addresses **emotion regulation** in population overlapping with adolescence [10-24]
+
+<doi.org/10.1007/s40894-021-00160-7>
+
+## Co-occurrence graph
+
+```{r networks1, eval = TRUE, results='asis', out.width = "90%"}
+knitr::include_graphics("tmnetworks_small.png")
 ```
 
-# Methods
-We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study.
-<!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->
+Author keywords (a) and abstracts (b)
 
-## Participants
+## Next step
 
-## Material
+```{r, out.height="50%"}
+knitr::include_graphics("groups.png")
+```
 
-## Procedure
+* Predictors of latent growth model?
+  + Only linear differences, few predictors (power)
+* Multigroup latent growth model?
+  + Non-linear differences, but only one moderator
+* Latent class growth model with auxilliary variables?
+  + Groups with regard to trajectory, not predictors
 
-## Data analysis
-We used `r cite_r("r-references.bib")` for all our analyses.
+## Solution: Machine learning
+
+* Allows non-linear differences between trajectories
+* Performs variable selection
+  + Cast wide net among potentially relevant predictors
+* Checks & balances ensure generalizability
+* Exploratory: No hypothesis
+
+## Method illustrated
+
+```{r}
+knitr::include_graphics("tree.png")
+```
+
+## Method
+
+SEM-forest [@brandmaierTheoryguidedExplorationStructural2016]
+
+1. Bootstrap sample
+2. On each sample, estimate a SEM-model
+3. Consider $k$ candidate predictors
+4. Identify predictor and value that maximizes $LR$ of post-split multi-group model
+5. Average predictions across bootstrap samples
+
+We used 1000 bootstrap samples, $k = 9$
+
+<!-- TS: Confirmatory versus exploratory research: Define the distinction. -->
+<!-- TS: Some longitudinal research falls between these categories.
+For example, complex longitudinal studies, where prior research is sufficient to lead us to hypothesize *some* associations between two constructs, but not sufficiently specific to let us hypothesize which parameters will be significant.
+-->
+
+<!-- ## Results -->
+
+<!-- * Both analyses reflected some constructs from theoretical literature -->
+<!--     + neurodevelopment, socialization -->
+<!-- * Undertheorized themes -->
+<!--     + Developmental disorders -->
+<!--     + Physical health (sic) -->
+<!--     + External stressors -->
+<!--     + Structural disadvantage -->
+<!--     + Addictive behavior -->
+<!--     + Identity and moral development -->
+<!--     + Sexual development -->
+
+<!-- * This offers guidance for phenomena to consider as relevant predictors -->
+
+## Present study
+
+RQ1: What are the most important predictors of adolescents' trajectories of emotion regulation development?
+
+RQ2: What is the nature of the association of the predictors with the trajectories?
+
+**Participants**
+
+* RADAR data (https://www.uu.nl/en/research/radar)
+* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46)
+* Collected between 2006-2011
+* Most families were  medium- to high-SES (10% low-SES)
+
+## Candidate predictors
+
+Predictors: 87
+
+Categories:
+
+* Demographics (e.g., SES)
+* Biological (age, sex, pubertal development)
+* Individual differences (big5, BIS/BAS, empathy)
+* Risk behavior (in/externalizing symptoms, substance use, delinquency, school performance, victimization)
+* Relationship quality (parents/peers)
+* Parenting (perceived and parent-reported)
+* Conflict resolution styles
+
+## Dependent variable
+
+Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
+
+* 24 items (omitted 'emotional awareness' subscale)
+* E.g.: “When I’m upset, I have difficulty thinking about anything else”
+* 1 (“Almost never”) to 5 (“Almost always”)
+* Excellent reliability, $(.94 > \alpha > .95)$.
+* Metric invariance across waves
+* Quadratic growth model had good fit
+  + CFI = .996, TLI = .993, RMSEA = .047, SRMR = 0.020
+
+## Open science
+
+Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]
+
+All code and synthetic data available at
+
+<https://github.com/cjvanlissa/veni_forest>
 
@@
 
-# Discussion
+## Variable importance
+
+```{r, out.height="100%"}
+vim <- readRDS("varimp_comb_21-08-2021.RData")
+vim
+```
+
+## Marginal association
+
+```{r, out.width="40%"}
+knitr::include_graphics("plots/pdp1-12.png")
+```
+
+## Marginal association 2
+
+```{r, out.width="40%"}
+knitr::include_graphics("plots/pdp13-24.png")
+```
+
+## Marginal association 3
+
+```{r, out.width="40%"}
+knitr::include_graphics("plots/pdp25-36.png")
+```
+
+## Marginal association 4
+
+```{r, out.width="40%"}
+knitr::include_graphics("plots/pdp37-44.png")
+```
+
+## Summary
+
+Important:
+
+* Best predictors: Personality and related constructs
+* Best parental predictor is autonomy support (bal. rel.)
+  + Negative/intrusive parenting more predictive than other positive practices
+* Conflict frequency and behavior
+* Empathy, particularly personal distress
+
+Less important than expected:
+
+* SES
+* Bullying/victimization
+* Delinquency
+* Substance use
+* Monitoring
+
+## Reflection content
+
+* Most important predictors are routinely assessed (big 5) or overt (conflict)
+  + Prime candidates for early risk assessment
+  + Conflict resolution behavior can be taught
+    - Target for intervention if association is causal
+* Emphasis on parenting (in literature and theory) might not be justified
+  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
+* Order of predictors mirrors bioecological model
+
+## {.fullslide}
+
+![](polarplot2.png)
+
+## Reflections form
+
+* Many predictors show non-linear effects: Emotion dysregulation only for +/-1SD
+* Some predictors show almost no marginal effect (e.g., father's age, drug use)
+  + This suggests they might be important in **interactions**
+
+## Limitations
+
+* Selective sample
+  + WEIRD, Dutch, mostly mid- high-SES (restriction of range)
+* Panel study, so few biological / cognitive / neurological measures
+* Relatively small sample size
+  + If you have access to 1000+ sample, please get in touch
+* Computational constraints prevented us from tuning the model beyond default settings, and from exploring marginal effects for higher order interactions
+
+## Conclusions
+
+* Proximal factors more predictive of ER development than distal predictors
+* Parenting may be less relevant than previously thought
+* Personality and conflict behavior are candidates for early diagnosis
+* Conflict skills training might be avenue for intervention
 

### 2022-02-13 presentation_sra2022.Rmd
papr writing

--- 2021-07-23 manuscript/manuscript.Rmd
@@
 ---
-title             : "The title"
-shorttitle        : "Title"
-
-author: 
-  - name          : "First Author"
-    affiliation   : "1"
-    corresponding : yes    # Define only one corresponding author
-    address       : "Postal address"
-    email         : "my@email.com"
-  - name          : "Ernst-August Doelle"
-    affiliation   : "1,2"
-
-affiliation:
-  - id            : "1"
-    institution   : "Wilhelm-Wundt-University"
-  - id            : "2"
-    institution   : "Konstanz Business School"
-
-authornote: |
-  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
-
-  Enter author note here.
-
-abstract: |
-  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
-
-  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
-
-  One sentence clearly stating the **general problem** being addressed by  this particular study.
-
-  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
-
-  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
-
-  One or two sentences to put the results into a more **general context**.
-
-  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
-
-  <!-- https://tinyurl.com/ybremelq -->
-
-keywords          : "keywords"
-wordcount         : "X"
-
-bibliography      : ["r-references.bib", "references.bib"]
-
-floatsintext      : no
-figurelist        : no
-tablelist         : no
-footnotelist      : no
-linenumbers       : yes
-mask              : no
-draft             : no
-
-documentclass     : "apa6"
-classoption       : "man"
-output            : papaja::apa6_pdf
-knit              : worcs::cite_all
+title: "Complementing theory on adolescent emotion regulation using  machine learning"
+author: "Caspar J. Van Lissa<sup>1</sup>, Lukas Beinhauer, Susan Branje, Wim H. J. Meeus"
+date: "<p><sup>1</sup>: Utrecht University, dept. Methodology & Statistics | Open Science Community Utrecht<br>Funded by NWO Veni grant  VI.Veni.191G.090"
+knit: (function(inputFile, encoding) {
+  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
+output: 
+  ioslides_presentation:
+    widescreen: true
+    css: slide.css
+bibliography      : ["manuscript/references.bib"]
 ---
 
-```{r setup, include = FALSE}
-library("papaja")
-library("worcs")
-# We recommend that you prepare your raw data for analysis in 'prepare_data.R',
-# and end that file with either open_data(yourdata), or closed_data(yourdata).
-# Then, uncomment the line below to load the original or synthetic data
-# (whichever is available), to allow anyone to reproduce your code:
-# load_data()
+```{r setup, include=FALSE}
+#css: https://maxcdn.bootstrapcdn.com/bootstrap/3.3.7/css/bootstrap.min.css
+knitr::opts_chunk$set(echo = FALSE)
 ```
 
-This manuscript uses the Workflow for Open Reproducible Code in Science [@vanlissaWORCSWorkflowOpen2020] to ensure reproducibility and transparency.
-All code <!--and data--> are available at <https://github.com/cjvanlissa/veni_forest.git>.
+## Emotion regulation in adolescence
 
-This is an example of a non-essential citation [@@vanlissaWORCSWorkflowOpen2020]. If you change the rendering function to `worcs::cite_essential`, it will be removed.
+Developmentally sensitive period [@zimmermannEmotionRegulationEarly2014]
 
-```{r analysis-preferences}
-# Seed for random number generation
-set.seed(42)
-knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
+20% develop psychopathology [@leeAdolescentMentalHealth2014]
+
+Potentially lifelong implications for mental health and well-being
+
+```{r, out.height="20%"}
+knitr::include_graphics("trajectories.jpg")
 ```
 
-# Methods
-We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study.
-<!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->
+## Knowledge gap
 
-## Participants
+Substantial empirical research, but:
 
-## Material
+* Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014]
+* No consistent terminology [@bariolaChildAdolescentEmotion2011]
+* No conceptual frameworks [@stifterEmotionRegulation2019]
+* No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]
 
-## Procedure
+Knowledge must be consolidated into overarching theory
 
-## Data analysis
-We used `r cite_r("r-references.bib")` for all our analyses.
+## Relevant theory
+
+* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
+* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
+* Hall's notion of "storm and stress"
+* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
+* Tripartite model [@morrisRoleFamilyContext2007]
+* Internalization model [@holodynskiDevelopmentEmotionsEmotion2006]
+* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
+* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
+* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
+* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]
+
+## Limitations of relevant theory
+
+* Few explicitly address adolescence
+* Few comprehensively address predictors of emotion regulation
+* None directly guide contemporary research 
+* Theories vary widely in scope: Some are broad and non-specific; others describe a specific phenomenon in detail, but lack a broader perspective.
+    + Broad theories can frame any research, specific theories generate hypotheses.
+    + It would be beneficial to bridge these levels of analysis.
+
+## Towards integrative theory
+
+First step: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
+
+We conducted a text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]
+
+  * Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
+  * TMSR: Unlimited sample size, transparent, objective, reproducible
+
+6653 papers on Addresses **emotion regulation** in population overlapping with adolescence [10-24]
+
+<doi.org/10.1007/s40894-021-00160-7>
+
+## Co-occurrence graph
+
+```{r networks1, eval = TRUE, results='asis', out.width = "90%"}
+knitr::include_graphics("tmnetworks_small.png")
+```
+
+Author keywords (a) and abstracts (b)
+
+## Next step
+
+```{r, out.height="50%"}
+knitr::include_graphics("groups.png")
+```
+
+* Predictors of latent growth model?
+  + Only linear differences, few predictors (power)
+* Multigroup latent growth model?
+  + Non-linear differences, but only one moderator
+* Latent class growth model with auxilliary variables?
+  + Groups with regard to trajectory, not predictors
+
+## Solution: Machine learning
+
+* Allows non-linear differences between trajectories
+* Performs variable selection
+  + Cast wide net among potentially relevant predictors
+* Checks & balances ensure generalizability
+* Exploratory: No hypothesis
+
+## Method illustrated
+
+```{r}
+knitr::include_graphics("tree.png")
+```
+
+## Method
+
+SEM-forest [@brandmaierTheoryguidedExplorationStructural2016]
+
+1. Bootstrap sample
+2. On each sample, estimate a SEM-model
+3. Consider $k$ candidate predictors
+4. Identify predictor and value that maximizes $LR$ of post-split multi-group model
+5. Average predictions across bootstrap samples
+
+We used 1000 bootstrap samples, $k = 9$
+
+<!-- TS: Confirmatory versus exploratory research: Define the distinction. -->
+<!-- TS: Some longitudinal research falls between these categories.
+For example, complex longitudinal studies, where prior research is sufficient to lead us to hypothesize *some* associations between two constructs, but not sufficiently specific to let us hypothesize which parameters will be significant.
+-->
+
+<!-- ## Results -->
+
+<!-- * Both analyses reflected some constructs from theoretical literature -->
+<!--     + neurodevelopment, socialization -->
+<!-- * Undertheorized themes -->
+<!--     + Developmental disorders -->
+<!--     + Physical health (sic) -->
+<!--     + External stressors -->
+<!--     + Structural disadvantage -->
+<!--     + Addictive behavior -->
+<!--     + Identity and moral development -->
+<!--     + Sexual development -->
+
+<!-- * This offers guidance for phenomena to consider as relevant predictors -->
+
+## Present study
+
+RQ1: What are the most important predictors of adolescents' trajectories of emotion regulation development?
+
+RQ2: What is the nature of the association of the predictors with the trajectories?
+
+**Participants**
+
+* RADAR data (https://www.uu.nl/en/research/radar)
+* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46)
+* Collected between 2006-2011
+* Most families were  medium- to high-SES (10% low-SES)
+
+## Candidate predictors
+
+Predictors: 87
+
+Categories:
+
+* Demographics (e.g., SES)
+* Biological (age, sex, pubertal development)
+* Individual differences (big5, BIS/BAS, empathy)
+* Risk behavior (in/externalizing symptoms, substance use, delinquency, school performance, victimization)
+* Relationship quality (parents/peers)
+* Parenting (perceived and parent-reported)
+* Conflict resolution styles
+
+## Dependent variable
+
+Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
+
+* 24 items (omitted 'emotional awareness' subscale)
+* E.g.: “When I’m upset, I have difficulty thinking about anything else”
+* 1 (“Almost never”) to 5 (“Almost always”)
+* Excellent reliability, $(.94 > \alpha > .95)$.
+* Metric invariance across waves
+* Quadratic growth model had good fit
+  + CFI = .996, TLI = .993, RMSEA = .047, SRMR = 0.020
+
+## Open science
+
+Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]
+
+All code and synthetic data available at
+
+<https://github.com/cjvanlissa/veni_forest>
 
@@
 
-# Discussion
+## Variable importance
 
-\newpage
-
-# References
-```{r create_r-references}
-r_refs(file = "r-references.bib")
+```{r, out.height="100%"}
+vim <- readRDS("varimp_comb_21-08-2021.RData")
+vim
 ```
 
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
+## Marginal association
 
-<div id="refs" custom-style="Bibliography"></div>
-\endgroup
+```{r, out.width="40%"}
+knitr::include_graphics("plots/pdp1-12.png")
+```
+
+## Marginal association 2
+
+```{r, out.width="40%"}
+knitr::include_graphics("plots/pdp13-24.png")
+```
+
+## Marginal association 3
+
+```{r, out.width="40%"}
+knitr::include_graphics("plots/pdp25-36.png")
+```
+
+## Marginal association 4
+
+```{r, out.width="40%"}
+knitr::include_graphics("plots/pdp37-44.png")
+```
+
+## Summary
+
+Important:
+
+* Best predictors: Personality and related constructs
+* Best parental predictor is autonomy support (bal. rel.)
+  + Negative/intrusive parenting more predictive than other positive practices
+* Conflict frequency and behavior
+* Empathy, particularly personal distress
+
+Less important than expected:
+
+* SES
+* Bullying/victimization
+* Delinquency
+* Substance use
+* Monitoring
+
+## Reflection content
+
+* Most important predictors are routinely assessed (big 5) or overt (conflict)
+  + Prime candidates for early risk assessment
+  + Conflict resolution behavior can be taught
+    - Target for intervention if association is causal
+* Emphasis on parenting (in literature and theory) might not be justified
+  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
+* Order of predictors mirrors bioecological model
+
+## {.fullslide}
+
+![](polarplot2.png)
+
+## Reflections form
+
+* Many predictors show non-linear effects: Emotion dysregulation only for +/-1SD
+* Some predictors show almost no marginal effect (e.g., father's age, drug use)
+  + This suggests they might be important in **interactions**
+
+## Limitations
+
+* Selective sample
+  + WEIRD, Dutch, mostly mid- high-SES (restriction of range)
+* Panel study, so few biological / cognitive / neurological measures
+* Relatively small sample size
+  + If you have access to 1000+ sample, please get in touch
+* Computational constraints prevented us from tuning the model beyond default settings, and from exploring marginal effects for higher order interactions
+
+## Conclusions
+
+* Proximal factors more predictive of ER development than distal predictors
+* Parenting may be less relevant than previously thought
+* Personality and conflict behavior are candidates for early diagnosis
+* Conflict skills training might be avenue for intervention
+
+## References

### 2022-02-13 polarplot.R

@@ -1,0 +1,14 @@
+df_plot <- data.frame(Variable = names(VI$variable.importance), Value = VI$variable.importance)
+df_plot$Variable <- gsub("^\\d+\\. ", "", df_plot$Variable)
+df_plot$Level <- ordered(ren$V4, levels = c("Individual", "Microsystem", "Mesosystem", "Macrosystem"))
+df_plot$x <- runif(nrow(df_plot))
+df_plot$y <- df_plot$Level
+levels(df_plot$y) <- c(1, 7, 9, 10)
+df_plot$y <- as.numeric(as.character(df_plot$y))
+df_plot$y[df_plot$y == 1] <- runif(length(df_plot$y[df_plot$y == 1]), 2, 6)
+ggplot(df_plot, aes(x=x, y = y, size = Value, color = Level, label = Variable)) + #geom_point()+
+  geom_text()+
+  coord_polar()+
+  theme_minimal()+
+  theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank())+
+  scale_size(guide = 'none')

### 2022-02-13 scripts/assessingforest.R

@@ -35,8 +35,21 @@ fac_labs <- list(
 #saveRDS(vim, paste0("results/vim_all_2", gsub("[: ]", "_", Sys.time()), ".RData"))
 vim <- readRDS("results/vim_all_22021-08-21_17_23_27.RData")
 VI <- list(variable.importance = semtree:::aggregateVarimp(vim, aggregate = "median", scale = "absolute", TRUE))
+ren <- read.csv("scale_rename.csv", stringsAsFactors = F, header = FALSE)
+names(VI$variable.importance)[names(VI$variable.importance) %in% ren$V1] <- ren$V2[match(names(VI$variable.importance)[names(VI$variable.importance) %in% ren$V1], ren$V1)]
+VI$variable.importance <- sort(VI$variable.importance, decreasing = TRUE)
+names(VI$variable.importance) <- paste0(1:length(VI$variable.importance), ". ", names(VI$variable.importance))
 class(VI) <- "ranger"
-p <- metaforest::VarImpPlot(VI, 92) 
+v1 <- v2 <- VI
+v1$variable.importance <- v1$variable.importance[1:44]
+p1 <- metaforest::VarImpPlot(v1, 44)+theme(axis.text.y = element_text(hjust=0))+xlab(NULL)+scale_x_continuous(limits = c(0, max(VI$variable.importance)))
+v2$variable.importance <- v2$variable.importance[45:87]
+p2 <- metaforest::VarImpPlot(v2, 43)+theme(axis.text.y = element_text(hjust=0))+xlab(NULL)+scale_x_continuous(limits = c(0, max(VI$variable.importance)))
+library("cowplot")
+pcomb <- plot_grid(p1,p2,
+          ncol = 2, nrow = 1)
+saveRDS(pcomb, "varimp_comb_21-08-2021.RData")
+p <- metaforest::VarImpPlot(VI, length(VI$variable.importance))
 saveRDS(p, "varimp_21-08-2021.RData")
 
 

### 2022-02-13 scripts/plantingforest.R

@@ -12,6 +12,7 @@ library(tidySEM)
 library(semtree)
 library(metaforest)
 library(future.apply)
+library(yaml)
 
 df_anal <- load_data(to_envir = FALSE)$df_anal
 
@@ -59,7 +60,10 @@ p <- p +
 
 
 # SEM forest --------------------------------------------------------------
-
+basicgrowth <- lavaan::growth("i =~ 1*de2 + 1*de3 + 1*de4 + 1*de5 + 1*de6
+s =~ 0*de2 + 1*de3 + 2*de4 + 3*de5 + 4*de6
+q =~ 0*de2 + 1*de3 + 4*de4 + 9*de5 + 16*de6" , data = df_anal)
+write_yaml(as.list(fitmeasures(basicgrowth)), "fitmeasures_growth.yml")
 m0 <- as_ram("i =~ 1*de2 + 1*de3 + 1*de4 + 1*de5 + 1*de6
 s =~ 0*de2 + 1*de3 + 2*de4 + 3*de5 + 4*de6
 q =~ 0*de2 + 1*de3 + 4*de4 + 9*de5 + 16*de6
@@ -84,7 +88,7 @@ s ~~ 0*s
 q ~~ 0*q")
 m0 <- run_mx(m0, data = df_anal)
 table_results(m0)
-
+table_fit(m0)
 # fitting a single tree using the model
 
 # system.time(

### 2022-02-22 manuscript/manuscript.Rmd
update 2022-02-22 15:25:11

@@
 <!-- TS: Emotion dysregulation has severe consequences for individual adolescents, and society at large. -->
-TS: Therefore, substantial research has addressed this topic.
-TS: Most preceding research has been limited in scope, focusing on only a few theoretically-driven predictors.
-TS: Machine learning makes it possible to cast a wider net, and explore all (available) potentially relevant predictors.
-TS: Confirmatory versus exploratory research: Define the distinction.
-TS: Some longitudinal research falls between these categories.
+<!-- TS: Therefore, substantial research has addressed this topic. -->
+<!-- TS: Most preceding research has been limited in scope, focusing on only a few theoretically-driven predictors. -->
+<!-- TS: Machine learning makes it possible to cast a wider net, and explore all (available) potentially relevant predictors. -->
+<!-- TS: Confirmatory versus exploratory research: Define the distinction. -->
+<!-- TS: Some longitudinal research falls between these categories.
 For example, complex longitudinal studies, where prior research is sufficient to lead us to hypothesize *some* associations between two constructs, but not sufficiently specific to let us hypothesize which parameters will be significant.
-TS: Instead, us
+-->
+<!-- TS: Instead, us -->
 
@@
 and incur a sizable societal cost.
-Researchers thus seek to understand what risk factors predispose some adolescents to emotion dysregulation,
-so that effective screening and interventions can be developed.
-Despite abundant empirical research, however,
-an overarching understanding is still lacking because
-the literature is fragmented across subdisciplines 
-and lacks consistent terminology and guiding theory [@vanlissaMappingPhenomenaRelevant2021].
+A comprehensive understanding of the risk factors that predispose some adolescents to emotion dysregulation could aid in the development of effective screening and interventions.
+However, despite an abundance of empirical research,
+many studies are narrow in scope, considering only a few predictors,
+the literature is fragmented across subdisciplines, and lacks a unifying theoretical framework
+<!-- that is both specific to adolescence and comprehensive in scope  -->
+[@vanlissaMappingPhenomenaRelevant2021].
 
-The present study seeks to gain a more comprehensive understanding of adolescent emotion regulation development by using machine learning methods for rigorous exploration.
+The present study seeks to overcome these limitations
+by using data-driven insights to complement and reflect on existing theory.
+To this end, a broad range of potential risk factors were analyzed using machine learning,
+and ranked in terms of predictive importance.
 <!-- These predictors may include both risk factors that render some youth susceptible to emotion regulation difficulties, and correlates that can be used to identify at-risk youth. -->
-The SEM-forests algorithm was used to
+Specifically,
+the SEM-forests algorithm was used to
 identify the most important early indicators of developmental trajectories
-of emotion regulation across adolescence, from a large set of potential candidates.
-The resulting insights can guide future research and interventions to focus on the most important predictors,
-and lays the foundation for formation and revision of overarching theory, such that essential phenomena are adequately represented.
+of emotion regulation across adolescence.
+The resulting insights indicate which factors are most predictive of adolescents' diverging destinies in terms of emotion regulation development.
+Armed with this knowledge, future research and interventions can focus on the most important predictors.
+Moreover, these insights can guide the formation of overarching theory that adequately represents essential phenomena.
+<!-- complement existing theory on adolescent emotion regulation using data-driven insights from machine learning. -->
 
-Recently, a text mining systematic review of 6305 publications on adolescents' emotion regulation has revealed what phenomena
+##
+
+## Emotion dysregulation in adolescence
+
+Emotion regulation,
+defined as the ability to manage emotional experience and expression,
+is critical for mental health4–6 and social competence2 throughout life. There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions1. 
+According to Crone and Dahl's neurodevelopmental model1,
+staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence:
+A period wherein the emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind8,21,22.
+This imbalance motivates most adolescents to engage in adaptive exploration:
+Seeking out experiences required to develop mature socio-emotional competences.
+At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation1,10,
+<!-- exposes youth to a risk of emotion dysregulation, -->
+for example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
+or persevere in pursuing adaptive goals when facing obstacles.
+Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions2,9,11,14,
+which has been linked to the emergence of several mental disorders5,10.
+<!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
+<!-- but renders some of them vulnerable to emotion dysregulation1,10.
+Unfortunately, there is little theory about which risk factors and environmental hazards render adolescents susceptible to emotional difficulties11.
+-->
+<!-- Acquiring adaptive emotion regulation skills in adolescence plays a key role in this regard7,20, because t -->
+
+Empirical support for this notion of diverging destinies11 is evident from studies that examine developmental trajectories of emotion regulation in adolescence: 
+Such studies consistently reveal large between-individual heterogeneity 2,7,9.
+An important next step is to identify which risk factors render some adolescents susceptible to emotion dysregulation.
+This involves three important challenges:
+The first is that the number of potentially relevant risk factors is large,
+the second is that there is a paucity of relevant theory to narrow down this number,
+and the third is that methods commonly used in the field are not well suited for large numbers of predictors.
+We argue that machine learning methods that perform variable selection overcome these challenges.
+
+<!-- This goal fits within the paradigm shift that has gripped the field of pedagogy, towards “person-centered” research, which focuses on individual developmental trajectories and within-family predictive processes, rather than average patterns across samples16,23.
+Seminal studies have applied “clustering” methods to identify youth with similar developmental trajectories14,24–26.
+Adolescents may follow similar trajectories for different reasons, however, and person-centered methods are too complex to accommodate more than a handful of theoretically-derived predictors.
+Unfortunately, relevant theory on adolescent emotion regulation development is limited27–29.
+The field is therefore at an impasse: We know that some adolescents are more susceptible to emotion dysregulation than others, but lack tools and theory to identify important predictors of individual development30.
+-->
+
+## Potential predictors
+
+<!-- Maybe review theory briefly? -->
+A recent literature review set out to identify relevant phenomena in relation to adolescents' emotion regulation [@vanlissaMappingPhenomenaRelevant2021].
+It used a new technique, text mining systematic review,
+to identify relevant phenomena based on frequently occurring terms in the author keywords and abstracts of 6305 publications.
+Whereas narrative reviews are typically restricted to small corpora and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008],
+text mining systematic reviews are more comprehensive and derive insights from the literature through a relatively transparent and objective procedure - at the expense of reflection and meaning-making.
+
+The findings of that systematic review indicated that internalizing and externalizing problems and mental health-related outcomes were most frequently studied in conjunction with emotion regulation. 
+Other relevant phenomena reflected themes discussed in prior narrative reviews,
+including cognitive development, parenting, and external stressors [cf. @coe-odessEmergentEmotionsAdolescence2019].
+Other relevant phenomena were underrepresented in prior theory and reviews,
+such as substance use, structural disadvantage,
+identity, morality, sex and sexual development.
+This relatively comprehensive overview offers a starting point for empirical investigations of predictors of adolescents' emotion regulation development.
+<!-- To this end, the present study sought to identify operationalizations of these relevant phenomena in a large-scale panel study. -->
+
+## Complementing theory using machine learning
+
+As mentioned before, the field of adolescent emotion regulation is prolific, but lacks overarching theory.
+For guidance on how to remedy this shortcoming, we turn to philosophy of science.
+De Groot’s "empirical cycle" constitutes a model of the production of knowledge through scientific research [@degrootMethodologieGrondslagenVan1961].
+It consists of five phases: *observation* of phenomena to be explained; *induction* of hypotheses; *deduction* or derivation of testable predictions; *testing* these predictions in new data; and finally *evaluation*: reflecting back on the theories and hypotheses that gave rise to the tested predictions.
+The model is cyclical because *evaluation* offers new interpretative data for the researcher to observe - such as rejected hypotheses and serendipitous findings.
+
+<!-- The empirical cycle is merely one model, -->
+<!-- and there is a pluriformity of other epistemological approaches to knowledge production. -->
+<!-- Nonetheless, u -->
+Unsubstantiated or undisclosed deviations from the empirical cycle may signal practices that limit scientific progress.
+Consider three examples:
+First, conducting many analyses and then writing a paper that presents significant findings as tests of a-priori hypotheses, or "hypothesizing after the results are known", constitutes a questionable research practice that leads to false-positive results [@wichertsDegreesFreedomPlanning2016].
+Second, failure to engage in evaluation and reflect back on theory in light of empirical findings leads to flexible, "invulnerable" theories that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
+Third, and particularly relevant to the present study,
+is failure to complete the cycle.
+In a productive field of study, the empirical cycle is regularly completed;
+evaluation of theories in the light of empirical findings allows scholars to hone those theories,
+which in turn yield increasingly precise predictions.
+We argue that, the paucity of overarching theory in our field can be remedied by completing the empirical cycle.
+
+The phases of the empirical cycle map onto two modes of conducting research:
+Confirmatory research, which involves the phases of deduction, testing, and evaluation, and exploratory research, which involves the phases of observation and induction  [@wagenmakersCreativityVerificationCyclePsychological2018].
+Although developmental psychology is dominated by quantitative positivism,
+relatively few studies are explicitly exploratory [@demuthQualitativeMethodologyDevelopmental2015].
+
+and devoting greater attention to exploratory research and theory formation.
+
+<!-- In this model, the hypothesis plays an important role. -->
+<!-- It is a testable prediction that can be supported or falsified; depending on the results, we reflect on old knowledge – which includes theory – and make necessary adjustments that lead to new hypotheses.
+-->
+<!-- A hypothesis can thus be a specific prediction that delivers new knowledge when it is false; -->
+<!-- this drives a scientific field forward. -->
+
+Prior theory and empirical research thus suggests many phenomena potentially related to heterogeneity in adolescents' emotion regulation development.
+What is missing is an understanding of the relative utility of each of these phenomena in predicting developmental outcomes.
+Individual studies do not suffice for this purpose as
+they tend to sample from different populations,
+use different research designs,
+different control variables and statistical models.
+Relatedly, many studies tend to examine only a handful of predictors,
+often guided by a theory with narrow scope - 
+, it is hard to tell how important the variables are, relative to other variables considered by other theories (or variables not considered at all).
+Machine learning is a more holistic methodology, as it can assess and compare a large number of potential predictors simultaneously, including theoretically relevant ones, and identify which predictors ultimately explain the most variance in the outcome measure of interest
+
+  Machine learning is a more holistic methodology, as it can assess and compare a large number of potential predictors simultaneously, including theoretically relevant ones, and identify which predictors ultimately explain the most variance in the outcome measure of interest.
 
@@
 
-```{r networks1, eval = TRUE, results='asis', out.width = "90%"}
-knitr::include_graphics("tmnetworks_small.png")
-```
-
 Author keywords (a) and abstracts (b)
@@
 ## Next step
-
-```{r, out.height="50%"}
-knitr::include_graphics("groups.png")
-```
 
@@
 ## Method illustrated
-
-```{r}
-knitr::include_graphics("tree.png")
-```
 
@@
 
-```{r, out.height="100%"}
-vim <- readRDS("varimp_comb_21-08-2021.RData")
-vim
-```
-
 ## Marginal association
-
-```{r, out.width="40%"}
-knitr::include_graphics("plots/pdp1-12.png")
-```
 
@@
 
-```{r, out.width="40%"}
-knitr::include_graphics("plots/pdp13-24.png")
-```
-
 ## Marginal association 3
 
-```{r, out.width="40%"}
-knitr::include_graphics("plots/pdp25-36.png")
-```
-
 ## Marginal association 4
-
-```{r, out.width="40%"}
-knitr::include_graphics("plots/pdp37-44.png")
-```
 
@@
 * Order of predictors mirrors bioecological model
-
-## {.fullslide}
-
-![](polarplot2.png)
 

### 2022-02-22 scripts/creatingscales.R

@@ -113,7 +113,8 @@ df_anal <- data.frame(scores,
 
 res <- lavaan::growth("# intercept and slope with fixed coefficients 
                        i =~ 1*de2 + 1*de3 + 1*de4 + 1*de5 + 1*de6
-                       s =~ 0*de2 + 1*de3 + 2*de4 + 3*de5 + 4*de6",
+                       s =~ 0*de2 + 1*de3 + 2*de4 + 3*de5 + 4*de6
+                       q =~ 0*de2 + 1*de3 + 4*de4 + 9*de5 + 16*de6",
                df_anal)
 
 fitMeasures(res)

### 2022-02-23 manuscript/manuscript.Rmd
update 2022-02-23 15:16:07

@@
 
-##
-
 ## Emotion dysregulation in adolescence
@@
 defined as the ability to manage emotional experience and expression,
-is critical for mental health4–6 and social competence2 throughout life. There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions1. 
-According to Crone and Dahl's neurodevelopmental model1,
-staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence:
-A period wherein the emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind8,21,22.
+is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010, @schaferEmotionRegulationStrategies2017, @braetEmotionRegulationChildren2014]
+and social competence [@zimmermannEmotionRegulationEarly2014] throughout life.
+There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions: 
+staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
+The emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
 This imbalance motivates most adolescents to engage in adaptive exploration:
 Seeking out experiences required to develop mature socio-emotional competences.
-At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation1,10,
+At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014],
 <!-- exposes youth to a risk of emotion dysregulation, -->
@@
 or persevere in pursuing adaptive goals when facing obstacles.
-Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions2,9,11,14,
-which has been linked to the emergence of several mental disorders5,10.
+Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions [@craccoEmotionRegulationChildhood2017],
+which has been linked to the emergence of several mental disorders [@schaferEmotionRegulationStrategies2017].
 <!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
@@
 
-Empirical support for this notion of diverging destinies11 is evident from studies that examine developmental trajectories of emotion regulation in adolescence: 
-Such studies consistently reveal large between-individual heterogeneity 2,7,9.
+Empirical support for this notion of diverging destinies is evident from studies that examine developmental trajectories of emotion regulation in adolescence: 
+Such studies consistently reveal large between-individual heterogeneity [see @craccoEmotionRegulationChildhood2017; @vanlissaRoleFathersMothers2019; @zimmermannEmotionRegulationEarly2014; @reindlSocializationEmotionRegulation2016].
 An important next step is to identify which risk factors render some adolescents susceptible to emotion dysregulation.
-This involves three important challenges:
-The first is that the number of potentially relevant risk factors is large,
-the second is that there is a paucity of relevant theory to narrow down this number,
-and the third is that methods commonly used in the field are not well suited for large numbers of predictors.
-We argue that machine learning methods that perform variable selection overcome these challenges.
-
 <!-- This goal fits within the paradigm shift that has gripped the field of pedagogy, towards “person-centered” research, which focuses on individual developmental trajectories and within-family predictive processes, rather than average patterns across samples16,23.
@@
 
-## Potential predictors
+## Potential predictors of emotion regulation development
+
+Several theories are commonly cited in research on adolescent emotion regulation.
+Notably, there appears to be a trade-off between these theories' scope and specificity.
+Some theories are broad in scope, considering many potential risk factors, but are non-specific.
+This limits their usefulness in generating testable predictions, and renders them impervious to falsification.
+Other theories are narrow in scope, and are sufficiently specific to derive testable predictions, but lack broader context.
+Two examples of macro-theories are the bioecological model [@bronfenbrennerBioecologicalModelHuman2007] and the transactional model [@sameroffUnifiedTheoryDevelopment2010].
+The bioecological model describes how individual development is shaped by factors at different levels of analysis:
+The individual level, which includes biological make-up and individual differences;
+the microsystem, composed of primary caregivers and close relationships;
+the mesosystem, including school and neighborhood factors;
+the exosystem, including media, political- and economic influences;
+and the macrosystem, consisting of cultural norms and values.
+The transactional model is compatible with the bioecological model;
+it describes development as an interplay between the individual child and the environment.
+It distinguishes proximal environmental influences, roughly corresponding to the microsystem,
+from more distal influences.
+With increasing age, distal influences are thought to become more influential. 
+More specific theories primarily focus on cognitive maturation and the role of primary caregivers,
+including the theory of normative emotional development [@sroufeAttachmentDyadicRegulation1996]
+and the tripartite model of emotion regulation socialization by parents [@morrisRoleFamilyContext2007] and peers [@reindlSocializationEmotionRegulation2016].
+Note that the tripartite model is one of the few theories that explicitly addresses the life phase of adolescence.
+It would be beneficial to work towards a more comprehensive theory of emotion regulation development,
+specific to adolescence,
+that is broad in scope and still specific enough to derive testable predictions.
+
+<!-- * Bioecological model [@bronfenbrennerBioecologicalModelHuman2007] -->
+<!-- * Transactional model [@sameroffUnifiedTheoryDevelopment2010] -->
+<!-- * Hall's notion of "storm and stress" -->
+<!-- * Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995] -->
+<!-- * Tripartite model [@morrisRoleFamilyContext2007] -->
+<!-- * Internalization model [@holodynskiDevelopmentEmotionsEmotion2006] -->
+<!-- * Polyvagal theory [@porgesOrientingDefensiveWorld1995]. -->
+<!-- * Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012] -->
+<!-- * Process model of emotion regulation [@grossHandbookEmotionRegulation2013] -->
+<!-- * Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000] -->
 
 <!-- Maybe review theory briefly? -->
-A recent literature review set out to identify relevant phenomena in relation to adolescents' emotion regulation [@vanlissaMappingPhenomenaRelevant2021].
-It used a new technique, text mining systematic review,
-to identify relevant phenomena based on frequently occurring terms in the author keywords and abstracts of 6305 publications.
-Whereas narrative reviews are typically restricted to small corpora and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008],
-text mining systematic reviews are more comprehensive and derive insights from the literature through a relatively transparent and objective procedure - at the expense of reflection and meaning-making.
+The first step towards theory formation or revision is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
+Thus, offering a starting point for this endeavor,<!-- empirical investigations of predictors of adolescents' emotion regulation development -->
+a recent literature review 
+used a new method, the text mining systematic review, to identify relevant phenomena in relation to adolescents' emotion regulation based on the analysis of author keywords and abstracts of 6305 publications  [@vanlissaMappingPhenomenaRelevant2021].
+Typically, phenomena are identified using narrative literature reviews.
+However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008]
+Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
+<!-- reflection and meaning-making-->
 
-The findings of that systematic review indicated that internalizing and externalizing problems and mental health-related outcomes were most frequently studied in conjunction with emotion regulation. 
-Other relevant phenomena reflected themes discussed in prior narrative reviews,
+The results of that review indicated that internalizing and externalizing problems and mental health-related outcomes were most frequently studied in conjunction with emotion regulation. 
+Some relevant phenomena reflected themes also identified in prior narrative reviews,
 including cognitive development, parenting, and external stressors [cf. @coe-odessEmergentEmotionsAdolescence2019].
@@
 identity, morality, sex and sexual development.
-This relatively comprehensive overview offers a starting point for empirical investigations of predictors of adolescents' emotion regulation development.
+This relatively comprehensive overview offers a starting point for empirical investigations of predictors of adolescents' emotion regulation development, such as the present study.
 <!-- To this end, the present study sought to identify operationalizations of these relevant phenomena in a large-scale panel study. -->
@@
 
-As mentioned before, the field of adolescent emotion regulation is prolific, but lacks overarching theory.
-For guidance on how to remedy this shortcoming, we turn to philosophy of science.
-De Groot’s "empirical cycle" constitutes a model of the production of knowledge through scientific research [@degrootMethodologieGrondslagenVan1961].
-It consists of five phases: *observation* of phenomena to be explained; *induction* of hypotheses; *deduction* or derivation of testable predictions; *testing* these predictions in new data; and finally *evaluation*: reflecting back on the theories and hypotheses that gave rise to the tested predictions.
-The model is cyclical because *evaluation* offers new interpretative data for the researcher to observe - such as rejected hypotheses and serendipitous findings.
+According to models of knowledge production within the quantitative positivist paradigm,
+a field progresses through the cyclical alternation of exploratory and confirmatory research, also known as the "empirical cycle" [@degrootMethodologieGrondslagenVan1961, @wagenmakersCreativityVerificationCyclePsychological2018].
+Exploratory research gives rise to testable predictions,
+and confirmatory research evaluates testable predictions on new data.
+Crucially, these phases are linked through evaluation of existing theory in light of empirical findings [@degrootMethodologieGrondslagenVan1961].
+The fact that our field still lacks overarching theory despite decades of productive empirical research
+suggests that the empirical cycle is not routinely completed.
+In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
+More exploratory research is thus needed to complete the cycle.
+Secondly, and paradoxically given the first argument,
+the link between confirmatory research and theory is often tenuous.
+For instance, many studies derive their predictions from prior empirical research, not from theory [e.g., @vanlissaCostEmpathyParentadolescent2017].
+Thirdly, theories are rarely falsified or revised in light of empirical findings.
+This failure to evaluate theory leads to flexible, "invulnerable" frameworks that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
+To advance theory formation, our field might benefit from "closing the empirical cycle" through increased exploratory research and reflection on theory.
+
+There are preliminary indications of a move towards exploratory research:
+Developmental scientists are turning to increasingly complex structural equation models.
+Consider, for example, the burgeoning interest in disentangling between-family differences from within-family processes using the innovative random-intercept cross-lagged panel model [@hamakerCritiqueCrosslaggedPanel2015] and dynamic structural equation modeling [@asparouhovDynamicStructuralEquation2018].
+Many authors using these methods have realized that existing theory is not sufficiently precise to derive distinct hypotheses at these two levels [see @nevilleBidirectionalAssociationsScreen2021; @vanlissaMothersFathersQuantitative2020].
+By necessity, these models are thus used in at least a partly exploratory capacity.
+This approach has two pitfalls.
+Firstly, complex models are by definition more flexible and fit data better then simpler models, at the expense of decreased generalizability [@hastieElementsStatisticalLearning2009].
+Flexibility is a hallmark of exploratory research [@wagenmakersCreativityVerificationCyclePsychological2018],
+but increases the risk of false-positive results and questionable research practices in confirmatory research [@wichertsDegreesFreedomPlanning2016, @simmonsFalsePositivePsychologyUndisclosed2011].
+It is therefore essential that flexibility is disclosed, and that exploratory and confirmatory research are clearly demarcated.
+Secondly, structural equation models are inherently a confirmatory method,
+and they can only include a limited number of predictors due to concerns of statistical power [@klinePrinciplesPracticeStructural2005].
+Their use for exploration is limited, effortful, and potentially fraught (e.g., when using p-values for discovery instead of inference).
+We argue that machine learning offers a suitable tool for rigorous exploration:
+These methods discover patterns in data while maximizing generalizability,
+and incorporate checks and balances to curtail false-positive results [@hastieElementsStatisticalLearning2009].
+In the context of the present study, we use a specific machine learning algorithm uniquely suited for theory-guided exploration: SEM-forests [@brandmaierTheoryguidedExplorationStructural2016].
+
+<!-- * Predictors of latent growth model? -->
+<!--   + Only linear differences, few predictors (power) -->
+<!-- * Multigroup latent growth model? -->
+<!--   + Non-linear differences, but only one moderator -->
+<!-- * Latent class growth model with auxilliary variables? -->
+<!--   + Groups with regard to trajectory, not predictors -->
+
+<!-- ## Solution: Machine learning -->
+
+<!-- * Allows non-linear differences between trajectories -->
+<!-- * Performs variable selection -->
+<!--   + Cast wide net among potentially relevant predictors -->
+<!-- * Checks & balances ensure generalizability -->
+<!-- * Exploratory: No hypothesis -->
+
+## The present study
+
+The present study set out to identify the most important early indicators of developmental trajectories
+of emotion regulation across adolescence.
+This involves three important challenges:
+First, the number of potentially relevant risk factors, as identified in a prior text mining systematic review, is large.
+Second, there is a paucity of relevant theory to narrow down the candidate predictors.
+Third, methods commonly used in the field are not well suited for large numbers of predictors.
+Machine learning methods that perform variable selection can overcome these challenges.
+We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence.
+The primary aim is to identify the most important predictors of adolescents' trajectories of emotion regulation development.
+The secondary aim is to probe the nature of the marginal association of the predictors with these trajectories.
+
+<!-- Deductive research, or hypothesis-testing, is the predominant focus of contemporary behavioral research.
+It tends to focus on a relatively narrow set of theoretically-derived variables, and the results revolve around statistical inference: Whether the theoretical hypotheses are supported by significant or reliable effects.
+In deductive research, less emphasis is placed on comprehensiveness or breadth of candidate predictors.
+Relatedly, the relative importance of different predictors is often of secondary importance, as is the model’s predictive performance.
+Thus, although an advantage of deductive approaches is that they can be used to draw inferences about theoretical hypotheses, they also have specific limitations.
+-->
+<!-- To allocate scientific resources effectively, it is important to cast a wide net among potential predictors, across different theories, and even include under-theorized factors to unearth potential blind spots in the extant literature.
+Inductive research – that is, rigorous exploratory work that identifies reliable patterns in data, is more suited to these demands.
+-->
+<!-- In recent years, inductive research has been gaining traction as a technique to complement existing theories by identifying important omissions[2].
+In particular, machine learning offers powerful new tools for systematic exploration that can identify relevant predictors and complex relationships that have eluded theoreticians[5].
+Machine learning is an approach to data analysis that focuses on maximizing predictive performance.
+This involves the use of flexible models to find reliable patterns in data.
+Machine learning models can distill a large set of candidate variables down to the ones that are most important in predicting the outcome of interest, and also indicate the direction and shape of the marginal association between those predictors and the outcome.
+In a context where predictor variables are likely to be related to each other, machine learning is better suited to manage these complex relationships than, e.g., multiple regressions.
+Moreover, it incorporates checks and balances to prevent spurious findings [3].
+-->
+
+<!-- The field of adolescent emotion regulation is prolific, but lacks overarching theory. -->
+<!-- For guidance on how to remedy this shortcoming, we turn to philosophy of science. -->
+<!-- De Groot’s "empirical cycle" constitutes a model of the production of knowledge through scientific research [@degrootMethodologieGrondslagenVan1961]. -->
+<!-- It consists of five phases: *observation* of phenomena to be explained; *induction* of hypotheses; *deduction* or derivation of testable predictions; *testing* these predictions in new data; and finally *evaluation*: reflecting back on the theories and hypotheses that gave rise to the tested predictions.
+-->
+<!-- The model is cyclical because *evaluation* offers new interpretative data for the researcher to observe - such as rejected hypotheses and serendipitous findings. -->
 
@@
 <!-- Nonetheless, u -->
-Unsubstantiated or undisclosed deviations from the empirical cycle may signal practices that limit scientific progress.
-Consider three examples:
-First, conducting many analyses and then writing a paper that presents significant findings as tests of a-priori hypotheses, or "hypothesizing after the results are known", constitutes a questionable research practice that leads to false-positive results [@wichertsDegreesFreedomPlanning2016].
-Second, failure to engage in evaluation and reflect back on theory in light of empirical findings leads to flexible, "invulnerable" theories that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
-Third, and particularly relevant to the present study,
-is failure to complete the cycle.
-In a productive field of study, the empirical cycle is regularly completed;
-evaluation of theories in the light of empirical findings allows scholars to hone those theories,
-which in turn yield increasingly precise predictions.
-We argue that, the paucity of overarching theory in our field can be remedied by completing the empirical cycle.
+<!-- Unsubstantiated or undisclosed deviations from the empirical cycle may signal practices that limit scientific progress. -->
+<!-- Consider three examples: -->
+<!-- First, conducting many analyses and then writing a paper that presents significant findings as tests of a-priori hypotheses, or "hypothesizing after the results are known", constitutes a questionable research practice that leads to false-positive results [@wichertsDegreesFreedomPlanning2016].
+-->
+<!-- Second, failure to engage in evaluation and reflect back on theory in light of empirical findings leads to flexible, "invulnerable" theories that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
+-->
+<!-- Third, and particularly relevant to the present study, -->
+<!-- is failure to complete the cycle. -->
+<!-- In a productive field of study, the empirical cycle is regularly completed; -->
+<!-- evaluation of theories in the light of empirical findings allows scholars to hone those theories, -->
+<!-- which in turn yield increasingly precise predictions. -->
+<!-- We argue that, the paucity of overarching theory in our field can be remedied by completing the empirical cycle. -->
 
-The phases of the empirical cycle map onto two modes of conducting research:
-Confirmatory research, which involves the phases of deduction, testing, and evaluation, and exploratory research, which involves the phases of observation and induction  [@wagenmakersCreativityVerificationCyclePsychological2018].
-Although developmental psychology is dominated by quantitative positivism,
-relatively few studies are explicitly exploratory [@demuthQualitativeMethodologyDevelopmental2015].
+<!-- The phases of the empirical cycle map onto two modes of conducting research: -->
+<!-- Confirmatory research, which involves the phases of deduction, testing, and evaluation, and exploratory research, which involves the phases of observation and induction  [@wagenmakersCreativityVerificationCyclePsychological2018].
+-->
+<!-- Although developmental psychology is dominated by quantitative positivism, -->
+<!-- relatively few studies are explicitly exploratory [@demuthQualitativeMethodologyDevelopmental2015]. -->
+<!-- There are two main indications that some studies are implicitly exploratory. -->
 
-and devoting greater attention to exploratory research and theory formation.
+<!-- The first is that some, apparently confirmatory, papers are not rooted in any theoretical framework, but instead based exclusively on prior empirical literature. -->
+
+<!-- The second indication is that an increasing number of studies use complex modeling techniques in the absence of theory specific enough to derive hypotheses for such models. -->
+<!-- For example, consider the recent literature using the random-intercept cross-lagged panel model, -->
+<!-- a complex structural equation model that separates stable differences between individuals from processes within  individuals (or families). -->
+<!-- Several authors have remarked that theory in their areas of expertise is insufficiently precise to derive specific and distinct predictions about effects at the between- and within level. -->
+<!-- At first glance, these publications fit the mold of confirmatory research, providing hypotheses and tests thereof. -->
+<!-- Upon further reflection, the hypotheses tested are not very specific (e.g., specifying a positive relation between two constructs, regardless of level of analysis), -->
+<!-- and they can be simultaneously supported and refuted by different parameters in the model without repercussions for theory. -->
+<!-- They are thus not truly "testable" in the sense that De Groot intended. -->
+<!-- Furthermore, the focus in these publications is on post-hoc interpretation of many parameter estimates - a hallmark of exploratory research. -->
+<!-- Implicitly exploratory research such as this suggests that others have perceived the need for more exploration, -->
+<!-- but that exploratory work is wrt as confirmatory the exploratory mode of research is not ingrained in the field. -->
+
+<!-- have noted that existing theory is insufficiently specific  -->
+<!-- ,  -->
+
+<!-- and devoting greater attention to exploratory research and theory formation. -->
 
@@
 
-Prior theory and empirical research thus suggests many phenomena potentially related to heterogeneity in adolescents' emotion regulation development.
-What is missing is an understanding of the relative utility of each of these phenomena in predicting developmental outcomes.
-Individual studies do not suffice for this purpose as
-they tend to sample from different populations,
-use different research designs,
-different control variables and statistical models.
-Relatedly, many studies tend to examine only a handful of predictors,
-often guided by a theory with narrow scope - 
-, it is hard to tell how important the variables are, relative to other variables considered by other theories (or variables not considered at all).
-Machine learning is a more holistic methodology, as it can assess and compare a large number of potential predictors simultaneously, including theoretically relevant ones, and identify which predictors ultimately explain the most variance in the outcome measure of interest
+<!-- Prior theory and empirical research thus suggests many phenomena potentially related to heterogeneity in adolescents' emotion regulation development. -->
+<!-- What is missing is an understanding of the relative utility of each of these phenomena in predicting developmental outcomes. -->
+<!-- Individual studies do not suffice for this purpose as -->
+<!-- they tend to sample from different populations, -->
+<!-- use different research designs, -->
+<!-- different control variables and statistical models. -->
+<!-- Relatedly, many studies tend to examine only a handful of predictors, -->
+<!-- often guided by a theory with narrow scope -  -->
+<!-- , it is hard to tell how important the variables are, relative to other variables considered by other theories (or variables not considered at all).
+Machine learning is a more holistic methodology, as it can assess and compare a large number of potential predictors simultaneously, including theoretically relevant ones, and identify which predictors ultimately explain the most variance in the outcome measure of interest -->
 
-  Machine learning is a more holistic methodology, as it can assess and compare a large number of potential predictors simultaneously, including theoretically relevant ones, and identify which predictors ultimately explain the most variance in the outcome measure of interest.
+<!--   Machine learning is a more holistic methodology, as it can assess and compare a large number of potential predictors simultaneously, including theoretically relevant ones, and identify which predictors ultimately explain the most variance in the outcome measure of interest.
+-->
 
-Theory cons methods for theory construction state that the first step in this endeavor is to identify well-established relevant *empirical phenomena*,
-defined as stable and general features of the world [@borsboomTheoryConstructionMethodology2020].
-The current study used a text mining systematic review to provide a comprehensive map of the empirical phenomena relevant to adolescent emotion regulation.
+<!-- Theory cons methods for theory construction state that the first step in this endeavor is to identify well-established relevant *empirical phenomena*, -->
+<!-- defined as stable and general features of the world [@borsboomTheoryConstructionMethodology2020]. -->
+<!-- The current study used a text mining systematic review to provide a comprehensive map of the empirical phenomena relevant to adolescent emotion regulation. -->
 
-Currently, few instruments exist for the identification of relevant phenomena; instead, theorists typically rely on expert opinion [@borsboomTheoryConstructionMethodology2020].
-This paper argues that systematic reviews are a suitable and relatively more objective method for phenomena detection.
-Traditional narrative reviews are limited, however, by small convenience samples,
-confirmation bias,
-and an undue emphasis on positive results [@littellEvidencebasedBiasedQuality2008].
-These limitations are overcome by a novel method:
-the text mining systematic review (TMSR).
-Relative to narrative reviews, text mining can cover arbitrarily greater corpora,
-and gleans insights from the literature by means of a relatively more transparent, objective, and reproducible process.
-<!--Where sentient readers tend to structure their reading of the literature around established ideas [@littellEvidencebasedBiasedQuality2008],
-the inherent "fairness" of text mining gives emerging themes a chance to come to the fore.-->
-This approach assumes the frequency with which phenomena recur in the literature to be indicative of their relevance.
-If it is additionally assumed that the frequency with which phenomena are studied together is indicative of the extent to which they are thought to be associated,
-then the co-occurrence of phenomena within publications can be interpreted as a rudimentary nomological network: a mapping of relationships between theoretically relevant phenomena [see @alaviAligningTheoryMethodology2018].
-Such a network can serve as a starting point for the formulation of a proto-theory,
-which would additionally involve abductive reasoning: specifying putative mechanisms to explain associations between the phenomena [@borsboomTheoryConstructionMethodology2020].
-Text mining systematic reviews constitute an inductive and exploratory approach:
-Their aim is phenomena detection based on generalization from observations (data) without appeal to theory [@haigExploratoryFactorAnalysis2005].
-By identifying and mapping the phenomena relevant to adolescent emotion regulation,
-such an inductive approach lays the groundwork for future theory development.
+<!-- Currently, few instruments exist for the identification of relevant phenomena; instead, theorists typically rely on expert opinion [@borsboomTheoryConstructionMethodology2020]. -->
+<!-- This paper argues that systematic reviews are a suitable and relatively more objective method for phenomena detection. -->
+<!-- Traditional narrative reviews are limited, however, by small convenience samples, -->
+<!-- confirmation bias, -->
+<!-- and an undue emphasis on positive results [@littellEvidencebasedBiasedQuality2008]. -->
+<!-- These limitations are overcome by a novel method: -->
+<!-- the text mining systematic review (TMSR). -->
+<!-- Relative to narrative reviews, text mining can cover arbitrarily greater corpora, -->
+<!-- and gleans insights from the literature by means of a relatively more transparent, objective, and reproducible process. -->
+<!-- <!--Where sentient readers tend to structure their reading of the literature around established ideas [@littellEvidencebasedBiasedQuality2008], -->
+<!-- the inherent "fairness" of text mining gives emerging themes a chance to come to the fore.--> -->
+<!-- This approach assumes the frequency with which phenomena recur in the literature to be indicative of their relevance. -->
+<!-- If it is additionally assumed that the frequency with which phenomena are studied together is indicative of the extent to which they are thought to be associated, -->
+<!-- then the co-occurrence of phenomena within publications can be interpreted as a rudimentary nomological network: a mapping of relationships between theoretically relevant phenomena [see @alaviAligningTheoryMethodology2018].
+-->
+<!-- Such a network can serve as a starting point for the formulation of a proto-theory, -->
+<!-- which would additionally involve abductive reasoning: specifying putative mechanisms to explain associations between the phenomena [@borsboomTheoryConstructionMethodology2020]. -->
+<!-- Text mining systematic reviews constitute an inductive and exploratory approach: -->
+<!-- Their aim is phenomena detection based on generalization from observations (data) without appeal to theory [@haigExploratoryFactorAnalysis2005]. -->
+<!-- By identifying and mapping the phenomena relevant to adolescent emotion regulation, -->
+<!-- such an inductive approach lays the groundwork for future theory development. -->
 
@@
 
-## Knowledge gap
+<!-- ## Knowledge gap -->
 
-Substantial empirical research, but:
+<!-- Substantial empirical research, but: -->
 
-* Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014]
-* No consistent terminology [@bariolaChildAdolescentEmotion2011]
-* No conceptual frameworks [@stifterEmotionRegulation2019]
-* No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]
+<!-- * Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014] -->
+<!-- * No consistent terminology [@bariolaChildAdolescentEmotion2011] -->
+<!-- * No conceptual frameworks [@stifterEmotionRegulation2019] -->
+<!-- * No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019] -->
 
-Knowledge must be consolidated into overarching theory
-
-## Relevant theory
-
-* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
-* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
-* Hall's notion of "storm and stress"
-* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
-* Tripartite model [@morrisRoleFamilyContext2007]
-* Internalization model [@holodynskiDevelopmentEmotionsEmotion2006]
-* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
-* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
-* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
-* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]
-
-## Limitations of relevant theory
-
-* Few explicitly address adolescence
-* Few comprehensively address predictors of emotion regulation
-* None directly guide contemporary research 
-* Theories vary widely in scope: Some are broad and non-specific; others describe a specific phenomenon in detail, but lack a broader perspective.
-    + Broad theories can frame any research, specific theories generate hypotheses.
-    + It would be beneficial to bridge these levels of analysis.
-
-## Towards integrative theory
-
-First step: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
-
-We conducted a text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]
-
-  * Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
-  * TMSR: Unlimited sample size, transparent, objective, reproducible
-
-6653 papers on Addresses **emotion regulation** in population overlapping with adolescence [10-24]
-
-<doi.org/10.1007/s40894-021-00160-7>
-
-## Co-occurrence graph
-
-Author keywords (a) and abstracts (b)
-
-## Next step
-
-* Predictors of latent growth model?
-  + Only linear differences, few predictors (power)
-* Multigroup latent growth model?
-  + Non-linear differences, but only one moderator
-* Latent class growth model with auxilliary variables?
-  + Groups with regard to trajectory, not predictors
-
-## Solution: Machine learning
-
-* Allows non-linear differences between trajectories
-* Performs variable selection
-  + Cast wide net among potentially relevant predictors
-* Checks & balances ensure generalizability
-* Exploratory: No hypothesis
-
-## Method illustrated
+<!-- Knowledge must be consolidated into overarching theory -->
 
@@
 We used 1000 bootstrap samples, $k = 9$
-
-<!-- TS: Confirmatory versus exploratory research: Define the distinction. -->
-<!-- TS: Some longitudinal research falls between these categories.
-For example, complex longitudinal studies, where prior research is sufficient to lead us to hypothesize *some* associations between two constructs, but not sufficiently specific to let us hypothesize which parameters will be significant.
--->
-
-<!-- ## Results -->
-
-<!-- * Both analyses reflected some constructs from theoretical literature -->
-<!--     + neurodevelopment, socialization -->
-<!-- * Undertheorized themes -->
-<!--     + Developmental disorders -->
-<!--     + Physical health (sic) -->
-<!--     + External stressors -->
-<!--     + Structural disadvantage -->
-<!--     + Addictive behavior -->
-<!--     + Identity and moral development -->
-<!--     + Sexual development -->
-
-<!-- * This offers guidance for phenomena to consider as relevant predictors -->
-
-## Present study
-
-RQ1: What are the most important predictors of adolescents' trajectories of emotion regulation development?
-
-RQ2: What is the nature of the association of the predictors with the trajectories?
 

### 2022-02-24 manuscript/manuscript.Rmd
update 2022-02-24 12:38:04

@@
 title             : "Complementing theory on adolescent emotion regulation using data-driven insights from machine learning"
-shorttitle        : "Title"
+shorttitle        : "Emotion Regulation and Machine Learning"
 
@@
 library("worcs")
-# We recommend that you prepare your raw data for analysis in 'prepare_data.R',
-# and end that file with either open_data(yourdata), or closed_data(yourdata).
-# Then, uncomment the line below to load the original or synthetic data
-# (whichever is available), to allow anyone to reproduce your code:
-# load_data()
+library(tidySEM)
+options(knitr.kable.NA = '')
+
 ```
@@
 An important next step is to identify which risk factors render some adolescents susceptible to emotion dysregulation.
-<!-- This goal fits within the paradigm shift that has gripped the field of pedagogy, towards “person-centered” research, which focuses on individual developmental trajectories and within-family predictive processes, rather than average patterns across samples16,23.
-Seminal studies have applied “clustering” methods to identify youth with similar developmental trajectories14,24–26.
+<!-- This goal fits within the paradigm shift that has gripped the field of pedagogy, towards "person-centered" research, which focuses on individual developmental trajectories and within-family predictive processes, rather than average patterns across samples16,23.
+Seminal studies have applied "clustering" methods to identify youth with similar developmental trajectories14,24–26.
 Adolescents may follow similar trajectories for different reasons, however, and person-centered methods are too complex to accommodate more than a handful of theoretically-derived predictors.
@@
 The first step towards theory formation or revision is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
-Thus, offering a starting point for this endeavor,<!-- empirical investigations of predictors of adolescents' emotion regulation development -->
-a recent literature review 
-used a new method, the text mining systematic review, to identify relevant phenomena in relation to adolescents' emotion regulation based on the analysis of author keywords and abstracts of 6305 publications  [@vanlissaMappingPhenomenaRelevant2021].
+<!--Thus, offering a starting point for this endeavor,<!-- empirical investigations of predictors of adolescents' emotion regulation development -->
+A recent literature review did so
+using a text mining systematic review of author keywords and abstracts of 6305 publications on  adolescents’ emotion regulation [@vanlissaMappingPhenomenaRelevant2021].
 Typically, phenomena are identified using narrative literature reviews.
@@
 
-<!-- Deductive research, or hypothesis-testing, is the predominant focus of contemporary behavioral research.
-It tends to focus on a relatively narrow set of theoretically-derived variables, and the results revolve around statistical inference: Whether the theoretical hypotheses are supported by significant or reliable effects.
-In deductive research, less emphasis is placed on comprehensiveness or breadth of candidate predictors.
-Relatedly, the relative importance of different predictors is often of secondary importance, as is the model’s predictive performance.
-Thus, although an advantage of deductive approaches is that they can be used to draw inferences about theoretical hypotheses, they also have specific limitations.
+# Method
+
+The Workflow for Open Reproducible Code in Science (WORCS) was used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020].
+All analysis code, supplemental materials and synthetic data are available at <https://github.com/cjvanlissa/veni_forest>.
+The original data (restricted) and all study documentation are available via <https://doi.org/10.17026/DANS-ZRB-V5WP>.
+
+## Participants
+
+```{r}
+desc_num <- read.csv("../desc_num.csv", stringsAsFactors = FALSE)
+desc_cat <- read.csv("../desc_cat.csv", stringsAsFactors = FALSE)
+```
+Participants were 497 Dutch adolescents and their parents enrolled in the longitudinal RADAR study  [@branjeResearchAdolescentDevelopment2018].
+The RADAR sample was recruited from randomly selected elementary schools in the province of Utrecht, and four main cities in The Netherlands.
+Of 1,544 randomly selected families, 497 produced informed consent and were included in the study.
+From 2006 to 2012, trained interviewers conducted annual home visits to collect questionnaire data from adolescents and parents.
+All independent variables were assessed in the first measurement wave.
+The dependent variable was measured from the second to the sixth wave.
+Adolescents received financial compensation for their participation in annual measurements (approximately \$17).
+
+## Measures
+
+## Dependent variable
+
+```{r}
+ders_desc <- readRDS("../ders_desc.RData")
+alphs <- sapply(ders_desc, function(x){x$alpha$total$raw_alpha})
+```
+
+From wave 2-6, adolescents completed the difficulty in emotion regulation scale [DERS, @gratzMultidimensionalAssessmentEmotion2004].
+Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2018],
+we used 24 items of this 32-item scale
+(e.g., "When I’m upset, I have difficulty thinking
+about anything else").
+Responses ranged
+from 1 (Almost never) to 5 (Almost always).
+Reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
+According to confirmatory factor analyses,
+a model that imposed metric measurement invariance across waves ($BIC = 140913.27$) was preferable over a model with configural invariance ($BIC = 141247.46$).
+The RMSEA and SRMR of this model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
+However, as reliability and standardized factor loadings ($\lambda \in [0.44, 0.84]$) were high across all waves, we extracted individual factor scores for further analysis.
+As quadratic growth model fit to the full sample had excellent fit, $\text{RMSEA} < .05, \text{SRMR} = 0.02, \text{CFI} = 1.00$.
+
+## Candidate predictors
+
+We set out to identify operationalizations of all relevant phenomena identified in @vanlissaMappingPhenomenaRelevant2021 that were available in the RADAR data.
+This yielded 87 candidate predictors, of which `r nrow(desc_cat)` categorical (see Table \@ref(tab:tabcat)), `r sum(is.na(desc_num$Items))` numerical, and `r sum(!is.na(desc_num$Items))` multi-item scales, see \@ref(tab:tabcont).
+We used principal components analysis for dimension reduction of multi-item scales.
+Across all scales, the explained variance of the first component ranged from $R^2 \in [0.18, 0.90]$.
+Supplementary Table S1 additionally reports skew and kurtosis,
+the minimum and maximum factor loading of a CFA fit to each scale,
+sample items,
+and references to the scale validation papers.
+
+```{r tabcat, results='asis'}
+papaja::apa_table(desc_cat, caption = "Sample descriptive statistics for categorical predictors.", format.args = list(na_string = ""))
+```
+```{r tabcont, results='asis'}
+ options(knitr.kable.NA = '')
+papaja::apa_table(desc_num[,c("name", "n", "Items", "mean", "sd", "min", "max", "Reliability", "PC1")
+], caption = "Sample descriptive statistics for continuous variables and scales.", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
+```
+
+## Strategy of analyses
+
+<!-- Our analyses were based on an extension of cross-lagged panel modeling (Selig & Little, 2012), with staggered measurement occasions for one of the variables (mood variability).
+Mood variability was measured each time during the year between two measurement waves.
+Thus, for mood variability, no within-wave correlations were estimated, and all links with conflict frequency and difficulties in emotion regulation were specified as (predictive) regression paths.
+We used this staggered cross-lagged panel model to investigate over-time predictive effects between conflict frequency, day-to-day mood variability, and difficulties in emotion regulation.
+First, we estimated a single-group path model with autoregressive and cross-lagged paths, as well as within-time correlations for concurrent measurements (Model 1, see Table 1).
+Then we introduced empathy (high, average, low) as a moderator in a multi-group model, with all parameters free to vary between groups (Model 2).
+To derive the best fitting, most parsimonious model, we constrained parameters that were not significantly different.
+We used Wald \chi^2 tests to evaluate which parameters did not differ significantly, first over time and then between groups.
+We used the resulting final model to test our hypotheses (Model 3, see Figure 2).
+For hypotheses involving moderation, we used Wald tests to investigate whether parameters differed significantly between empathy groups.
+To test hypotheses involving mediation, we inspected indirect effects with bootstrapped standard errors (10,000 resamples).
+Furthermore, although posterior predictive probabilities were high, we conducted additional analyses, using the approach described by Asparouhov and Muthén (2014) to account for classification uncertainty.
+As can be seen in Appendix A, results were in line with the multi-group model reported here.
 -->
-<!-- To allocate scientific resources effectively, it is important to cast a wide net among potential predictors, across different theories, and even include under-theorized factors to unearth potential blind spots in the extant literature.
-Inductive research – that is, rigorous exploratory work that identifies reliable patterns in data, is more suited to these demands.
+<!-- Attrition for the annual measurements ranged from 0.40% at age 13 to 14.50% at age 18, and attrition for the daily mood diaries ranged from 3.40% to 23.40%.
+These data were missing completely at random, Little’s (1988) MCAR test \chi^2(2179) = 2124.07, p = .80.
+Therefore, full information maximum likelihood estimation (FIML) was warranted to make use of all available information without estimating missing data.
+All analyses were conducted in Mplus Version 7 (Muthén & Muthén, 1998–2012).
+Per the developers’ recommendation, we used robust maximum-likelihood estimation, which yields a Satorra-Bentler scaled \chi^2 value to account for potential non-normality (Satorra, 2000).
+Model fit was evaluated using the Comparative Fit Index and Tucker–Lewis Index (CFI and TLI, acceptable fit = 0.90-0.95, Bentler & Bonett, 1980), and Root Mean Square Error of Approximation (RMSEA, close fit = 0.01-0.06, acceptable fit = 0.06-0.08, Browne & Cudeck, 1993).
+According to these criteria, the fit of the final model was acceptable-to-good (Table 1).
+In addition to these measures of objective fit, we also provide two comparative fit indices, the Akaike Information Criterion (AIC) and sample-size adjusted Bayesian Information Criterion (saBIC).
+Lower values on these comparative fit indices indicate a better fitting model.
+According to both relative and absolute fit indices, the final model had the best fit.
 -->
-<!-- In recent years, inductive research has been gaining traction as a technique to complement existing theories by identifying important omissions[2].
-In particular, machine learning offers powerful new tools for systematic exploration that can identify relevant predictors and complex relationships that have eluded theoreticians[5].
-Machine learning is an approach to data analysis that focuses on maximizing predictive performance.
-This involves the use of flexible models to find reliable patterns in data.
-Machine learning models can distill a large set of candidate variables down to the ones that are most important in predicting the outcome of interest, and also indicate the direction and shape of the marginal association between those predictors and the outcome.
-In a context where predictor variables are likely to be related to each other, machine learning is better suited to manage these complex relationships than, e.g., multiple regressions.
-Moreover, it incorporates checks and balances to prevent spurious findings [3].
--->
-
-<!-- The field of adolescent emotion regulation is prolific, but lacks overarching theory. -->
-<!-- For guidance on how to remedy this shortcoming, we turn to philosophy of science. -->
-<!-- De Groot’s "empirical cycle" constitutes a model of the production of knowledge through scientific research [@degrootMethodologieGrondslagenVan1961]. -->
-<!-- It consists of five phases: *observation* of phenomena to be explained; *induction* of hypotheses; *deduction* or derivation of testable predictions; *testing* these predictions in new data; and finally *evaluation*: reflecting back on the theories and hypotheses that gave rise to the tested predictions.
--->
-<!-- The model is cyclical because *evaluation* offers new interpretative data for the researcher to observe - such as rejected hypotheses and serendipitous findings. -->
-
-<!-- The empirical cycle is merely one model, -->
-<!-- and there is a pluriformity of other epistemological approaches to knowledge production. -->
-<!-- Nonetheless, u -->
-<!-- Unsubstantiated or undisclosed deviations from the empirical cycle may signal practices that limit scientific progress. -->
-<!-- Consider three examples: -->
-<!-- First, conducting many analyses and then writing a paper that presents significant findings as tests of a-priori hypotheses, or "hypothesizing after the results are known", constitutes a questionable research practice that leads to false-positive results [@wichertsDegreesFreedomPlanning2016].
--->
-<!-- Second, failure to engage in evaluation and reflect back on theory in light of empirical findings leads to flexible, "invulnerable" theories that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
--->
-<!-- Third, and particularly relevant to the present study, -->
-<!-- is failure to complete the cycle. -->
-<!-- In a productive field of study, the empirical cycle is regularly completed; -->
-<!-- evaluation of theories in the light of empirical findings allows scholars to hone those theories, -->
-<!-- which in turn yield increasingly precise predictions. -->
-<!-- We argue that, the paucity of overarching theory in our field can be remedied by completing the empirical cycle. -->
-
-<!-- The phases of the empirical cycle map onto two modes of conducting research: -->
-<!-- Confirmatory research, which involves the phases of deduction, testing, and evaluation, and exploratory research, which involves the phases of observation and induction  [@wagenmakersCreativityVerificationCyclePsychological2018].
--->
-<!-- Although developmental psychology is dominated by quantitative positivism, -->
-<!-- relatively few studies are explicitly exploratory [@demuthQualitativeMethodologyDevelopmental2015]. -->
-<!-- There are two main indications that some studies are implicitly exploratory. -->
-
-<!-- The first is that some, apparently confirmatory, papers are not rooted in any theoretical framework, but instead based exclusively on prior empirical literature. -->
-
-<!-- The second indication is that an increasing number of studies use complex modeling techniques in the absence of theory specific enough to derive hypotheses for such models. -->
-<!-- For example, consider the recent literature using the random-intercept cross-lagged panel model, -->
-<!-- a complex structural equation model that separates stable differences between individuals from processes within  individuals (or families). -->
-<!-- Several authors have remarked that theory in their areas of expertise is insufficiently precise to derive specific and distinct predictions about effects at the between- and within level. -->
-<!-- At first glance, these publications fit the mold of confirmatory research, providing hypotheses and tests thereof. -->
-<!-- Upon further reflection, the hypotheses tested are not very specific (e.g., specifying a positive relation between two constructs, regardless of level of analysis), -->
-<!-- and they can be simultaneously supported and refuted by different parameters in the model without repercussions for theory. -->
-<!-- They are thus not truly "testable" in the sense that De Groot intended. -->
-<!-- Furthermore, the focus in these publications is on post-hoc interpretation of many parameter estimates - a hallmark of exploratory research. -->
-<!-- Implicitly exploratory research such as this suggests that others have perceived the need for more exploration, -->
-<!-- but that exploratory work is wrt as confirmatory the exploratory mode of research is not ingrained in the field. -->
-
-<!-- have noted that existing theory is insufficiently specific  -->
-<!-- ,  -->
-
-<!-- and devoting greater attention to exploratory research and theory formation. -->
-
-<!-- In this model, the hypothesis plays an important role. -->
-<!-- It is a testable prediction that can be supported or falsified; depending on the results, we reflect on old knowledge – which includes theory – and make necessary adjustments that lead to new hypotheses.
--->
-<!-- A hypothesis can thus be a specific prediction that delivers new knowledge when it is false; -->
-<!-- this drives a scientific field forward. -->
-
-<!-- Prior theory and empirical research thus suggests many phenomena potentially related to heterogeneity in adolescents' emotion regulation development. -->
-<!-- What is missing is an understanding of the relative utility of each of these phenomena in predicting developmental outcomes. -->
-<!-- Individual studies do not suffice for this purpose as -->
-<!-- they tend to sample from different populations, -->
-<!-- use different research designs, -->
-<!-- different control variables and statistical models. -->
-<!-- Relatedly, many studies tend to examine only a handful of predictors, -->
-<!-- often guided by a theory with narrow scope -  -->
-<!-- , it is hard to tell how important the variables are, relative to other variables considered by other theories (or variables not considered at all).
-Machine learning is a more holistic methodology, as it can assess and compare a large number of potential predictors simultaneously, including theoretically relevant ones, and identify which predictors ultimately explain the most variance in the outcome measure of interest -->
-
-<!--   Machine learning is a more holistic methodology, as it can assess and compare a large number of potential predictors simultaneously, including theoretically relevant ones, and identify which predictors ultimately explain the most variance in the outcome measure of interest.
--->
-
-<!-- Theory cons methods for theory construction state that the first step in this endeavor is to identify well-established relevant *empirical phenomena*, -->
-<!-- defined as stable and general features of the world [@borsboomTheoryConstructionMethodology2020]. -->
-<!-- The current study used a text mining systematic review to provide a comprehensive map of the empirical phenomena relevant to adolescent emotion regulation. -->
-
-<!-- Currently, few instruments exist for the identification of relevant phenomena; instead, theorists typically rely on expert opinion [@borsboomTheoryConstructionMethodology2020]. -->
-<!-- This paper argues that systematic reviews are a suitable and relatively more objective method for phenomena detection. -->
-<!-- Traditional narrative reviews are limited, however, by small convenience samples, -->
-<!-- confirmation bias, -->
-<!-- and an undue emphasis on positive results [@littellEvidencebasedBiasedQuality2008]. -->
-<!-- These limitations are overcome by a novel method: -->
-<!-- the text mining systematic review (TMSR). -->
-<!-- Relative to narrative reviews, text mining can cover arbitrarily greater corpora, -->
-<!-- and gleans insights from the literature by means of a relatively more transparent, objective, and reproducible process. -->
-<!-- <!--Where sentient readers tend to structure their reading of the literature around established ideas [@littellEvidencebasedBiasedQuality2008], -->
-<!-- the inherent "fairness" of text mining gives emerging themes a chance to come to the fore.--> -->
-<!-- This approach assumes the frequency with which phenomena recur in the literature to be indicative of their relevance. -->
-<!-- If it is additionally assumed that the frequency with which phenomena are studied together is indicative of the extent to which they are thought to be associated, -->
-<!-- then the co-occurrence of phenomena within publications can be interpreted as a rudimentary nomological network: a mapping of relationships between theoretically relevant phenomena [see @alaviAligningTheoryMethodology2018].
--->
-<!-- Such a network can serve as a starting point for the formulation of a proto-theory, -->
-<!-- which would additionally involve abductive reasoning: specifying putative mechanisms to explain associations between the phenomena [@borsboomTheoryConstructionMethodology2020]. -->
-<!-- Text mining systematic reviews constitute an inductive and exploratory approach: -->
-<!-- Their aim is phenomena detection based on generalization from observations (data) without appeal to theory [@haigExploratoryFactorAnalysis2005]. -->
-<!-- By identifying and mapping the phenomena relevant to adolescent emotion regulation, -->
-<!-- such an inductive approach lays the groundwork for future theory development. -->
-
-<!--
-stifterEmotionRegulation2019
-* call for studies incorporating predictors of ER across multiple levels of analysis
-* different conceptual frameworks
-
-riedigerEmotionRegulationAdolescence2014
-* difficulty in deriving a cohesive picture about emotion regulation in adolescence
-* different conceptual frameworks
-
-* three themes (neurocognitive development, socialization, interplay between internal and external factors) currently represent relatively independent bodies of research
-* A more explicit integration of these various research perspectives will help us to arrive at a more integrated and interpretable picture of the development of emorion regulation during adolescence.
-
-@bussTheoriesEmotionalDevelopment2019
-
-* emotional development theory is still in its own state of development. 
-* empirical evidence can contribute to theory development.
-
-@bariolaChildAdolescentEmotion2011
-* Further research apply reﬁned theoretical conceptualizations of ER.
-
-Where sentient readers tend to structure their reading of the literature around established ideas [@littellEvidencebasedBiasedQuality2008],
-This potentially reveals phenomena and associations missed by narrative reviewers,
-thereby laying the groundwork for future empirical and theoretical work.
-
-we set out to map these relevant empirical phenomena using a systematic review.
-
-generate a proto-theoretical nomological network of the relevant empirical phenomena.
-We discuss how this nomological network relates to existing relevant theory, and conclude
-with recommendations for future theory-generating efforts.
-There have been calls for the  consolidation of these disparate insights into overarching theory would be an important step forward.
-An important first step 
-and intrinsic limitations of narrative reviews prohibit comprehensive coverage of the vast but diffuse literature .-->
-<!--
-Recent years have seen the publication of several excellent systematic reviews,
-but these have all been narrative reviews.-->
-
- <!-- CJ To fill out gaps, following up on promising constructs and unify heretofore disparate regions of the lit -->
-<!-- SH: this implies that the main contribution of the paper is predominantly methodological is that the case?
-Is there some primary thesis/argument that will be advanced?
-It is really important to take a top-down approach to that messaging, beginning here in the first paragraph-->
-<!-- Ergens anders: Similarly, although
-there are several *relevant* theories, there is no specific theoretical
-framework to guide research on adolescents' emotion regulation. Although theory
-development is beyond the scope of this paper, we set out to generate a
-proto-theoretical nomological network of the potential risk factors,
-manifestations, and outcomes of emotion regulation in adolescence. We discuss
-how this nomological network relates to existing relevant theory, and conclude
-with recommendations for future theory-generating efforts.-->
-<!-- At the same time, the staggered development of motivational-emotional and regulatory brain circuits gives rise
-to a maturity gap [@croneUnderstandingAdolescencePeriod2012]. 
-This leads youngsters to pursue new experiences in life and love,
-without being fully prepared to cope with the emotional outcomes. -->
-
-<!--The literature thus paints a picture of
-adolescence as a chrysalis for emotional development: Children enter this stage
-with emotion regulation skills adapted to the challenges of childhood. During
-adolescence, emotional systems are rearranged substantially.-->
-
-<!--
-* intense negative emotions in daily life (Silk, Steinberg, & Morris, 2003) 
-* unstable peer or romantic relationships (Furman & Collins, 2009)
-* decrease in perceived support from parents (Furman & Buhrmester, 1992).
-* early adolescence is characterized by a higher rate of conflicts with parents (Laursen, Coy, & Collins, 1998)
-* higher variability of negative emotions compared to late adolescence (Larson, Moneta, Richards, & Wilson, 2002).
-* relative to other developmental stages, adolescence involves a particularly large number of transitions, novel situations, and new stimuli, physical, cognitive, emotional, and social development—particularly in early and mid-adolescence (Seiffge-Krenke, 2000)
--->
-
-<!-- ## Knowledge gap -->
-
-<!-- Substantial empirical research, but: -->
-
-<!-- * Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014] -->
-<!-- * No consistent terminology [@bariolaChildAdolescentEmotion2011] -->
-<!-- * No conceptual frameworks [@stifterEmotionRegulation2019] -->
-<!-- * No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019] -->
-
-<!-- Knowledge must be consolidated into overarching theory -->
-
-## Method
 
@@
 
-## Candidate predictors
+# Results
 
-Predictors: 87
+<!-- Categories:  -->
 
-Categories:
-
-* Demographics (e.g., SES)
-* Biological (age, sex, pubertal development)
-* Individual differences (big5, BIS/BAS, empathy)
-* Risk behavior (in/externalizing symptoms, substance use, delinquency, school performance, victimization)
-* Relationship quality (parents/peers)
-* Parenting (perceived and parent-reported)
-* Conflict resolution styles
-
-## Dependent variable
-
-Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
-
-* 24 items (omitted 'emotional awareness' subscale)
-* E.g.: “When I’m upset, I have difficulty thinking about anything else”
-* 1 (“Almost never”) to 5 (“Almost always”)
-* Excellent reliability, $(.94 > \alpha > .95)$.
-* Metric invariance across waves
-* Quadratic growth model had good fit
-  + CFI = .996, TLI = .993, RMSEA = .047, SRMR = 0.020
-
-## Open science
-
-Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]
-
-All code and synthetic data available at
-
-<https://github.com/cjvanlissa/veni_forest>
-
-# Results
+<!-- * Demographics (e.g., SES) -->
+<!-- * Biological (age, sex, pubertal development) -->
+<!-- * Individual differences (big5, BIS/BAS, empathy) -->
+<!-- * Risk behavior (in/externalizing symptoms, substance use, delinquency, school performance, victimization) -->
+<!-- * Relationship quality (parents/peers) -->
+<!-- * Parenting (perceived and parent-reported) -->
+<!-- * Conflict resolution styles -->
 

### 2022-02-24 scripts/creatingscales.R

@@ -25,6 +25,15 @@ scales_list <- readRDS("scales_list.RData")
 
 # Dependent variable ------------------------------------------------------
 
+ders_desc <- lapply(2:6, function(i){
+  df_tmp <- data[grep(paste0("^de", i), names(data), value = TRUE)]
+  mod <- paste0("F =~ ", paste0(names(df_tmp), collapse = "+"))
+  res_cfa <- cfa(mod, data)
+  res_paral <- fa.parallel(df_tmp)
+  res_alpha <- psych::alpha(df_tmp)
+  list(cfa = res_cfa, paral = res_paral, alpha = res_alpha)
+})
+saveRDS(ders_desc, "ders_desc.RData")
 m_config <- character()
 for(i in 2:6){
   m_config <- paste0(m_config,
@@ -77,6 +86,7 @@ m_metric <- paste0(m_metric, itemcors)
 r_metric <- sem(m_metric, data)
 
 
+
 anova(r_config, r_metric)
 compareFit(r_config, r_metric)
 scores <- lavPredict(r_metric)
@@ -103,12 +113,68 @@ range(pc_mat[,1][!pc_mat[,1]==1])
 #summary(prcomp(data[scales_list$schoolprestatie_aa]))
 
 scores_pred <- data.frame(sapply(pred_list, function(x){x$x[, 1]}))
+reldata <- readRDS("reldata.RData")
 
+scales_list_desc <- lapply(scales_list, function(i){
+  if(any(!i %in% names(reldata))){
+    NULL
+  } else {
+    i
+  }
+})
+scales_list_desc <- scales_list_desc[!sapply(scales_list_desc, is.null)]
+# Descriptive table -------------------------------------------------------
+renm <- read.csv("scale_rename.csv", stringsAsFactors = FALSE, header = F)
+# Scales
+desc_scales <- tidySEM:::create_scales.data.frame(reldata, keys.list = scales_list_desc)$descriptives
+pc1 <- pc_mat[,1]
+names(pc1) <- rownames(pc_mat)
+desc_scales$PC1 <- formatC(pc1[desc_scales$Subscale], digits = 2, format = "f")
+desc_scales[c("n", "Items")] <- lapply(desc_scales[c("n", "Items")], as.integer)
+
+# Individual items
+data_noscale2 <- reldata[, which(!names(reldata) %in% unlist(scales_list))]
+data_noscale2$cigarettes <- as.numeric(reldata$su11aa04)
[…]

### 2022-02-24 scripts/missRanger.R

@@ -151,6 +151,7 @@ numvars <- c("leeftijd_target_11", "leeftijd_moeder_11", "leeftijd_sibling_11",
 intvars <- intvars[!intvars %in% numvars]
 reldata[intvars] <- lapply(reldata[intvars], drop_floats)
 table(drop_floats(reldata$de21aa03))
+saveRDS(reldata, "reldata.RData")
 #!
 # @Caspar - make sure to adjust the num.trees parameter! 
 #!

### 2022-02-26 manuscript/manuscript.Rmd
update 2022-02-26 13:35:02

@@
 knit              : worcs::cite_all
+header-includes:
+  - \raggedbottom
 ---
@@
 library(tidySEM)
+# this does nothing for apa_table:
 options(knitr.kable.NA = '')
-
+# To make svg work, use this and 
+# header-includes:
+#   - \usepackage{svg}
+#options(tinytex.engine_args = '--shell-escape')
 ```
@@
 
-## Participants
+## Participants and procedure
 
@@
 Adolescents received financial compensation for their participation in annual measurements (approximately \$17).
+
+<!-- **Participants** -->
+
+<!-- * RADAR data (https://www.uu.nl/en/research/radar) -->
+<!-- * 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) -->
+<!-- * Collected between 2006-2011 -->
+<!-- * Most families were  medium- to high-SES (10% low-SES) -->
 
@@
 Reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
-According to confirmatory factor analyses,
-a model that imposed metric measurement invariance across waves ($BIC = 140913.27$) was preferable over a model with configural invariance ($BIC = 141247.46$).
-The RMSEA and SRMR of this model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
-However, as reliability and standardized factor loadings ($\lambda \in [0.44, 0.84]$) were high across all waves, we extracted individual factor scores for further analysis.
-As quadratic growth model fit to the full sample had excellent fit, $\text{RMSEA} < .05, \text{SRMR} = 0.02, \text{CFI} = 1.00$.
+We used latent variable scores (instead of mean scores) for further analyses.
+To this end, we imposed a five-factor CFA (one for each wave) on the indicators of the dependent variable.
+Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the overall trajectory.
+As measurement invariance must be imposed in order to compare latent scores across waves,
+we compared a model with configural invariance ($BIC = 141247.46$)
+to a model that imposed metric invariance ($BIC = 140913.27$).
+The lower BIC indicates support for metric invariance.
+The RMSEA and SRMR of the invariant model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
+As reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
+we considered it acceptable to proceed with individual factor scores for further analysis.
 
@@
 We set out to identify operationalizations of all relevant phenomena identified in @vanlissaMappingPhenomenaRelevant2021 that were available in the RADAR data.
-This yielded 87 candidate predictors, of which `r nrow(desc_cat)` categorical (see Table \@ref(tab:tabcat)), `r sum(is.na(desc_num$Items))` numerical, and `r sum(!is.na(desc_num$Items))` multi-item scales, see \@ref(tab:tabcont).
-We used principal components analysis for dimension reduction of multi-item scales.
+This yielded 87 candidate predictors, of which `r nrow(desc_cat)` categorical (see Table \@ref(tab:tabcat)), `r sum(is.na(desc_num$Items))` numerical, and `r sum(!is.na(desc_num$Items))` multi-item scales, see \@ref(tab:tabnum).
+For multi-item scales, we used principal components analysis for dimension reduction.
 Across all scales, the explained variance of the first component ranged from $R^2 \in [0.18, 0.90]$.
-Supplementary Table S1 additionally reports skew and kurtosis,
-the minimum and maximum factor loading of a CFA fit to each scale,
+This first component represents the largest amount of shared variance among the scale items, and was used for further analysis.
+[Supplementary Table S1](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_1.csv) additionally reports skew and kurtosis,
+minimum and maximum factor loadings,
 sample items,
-and references to the scale validation papers.
+and references to the validation paper for each scale.
 
 ```{r tabcat, results='asis'}
-papaja::apa_table(desc_cat, caption = "Sample descriptive statistics for categorical predictors.", format.args = list(na_string = ""))
+vimp <- readRDS("../variable_importance.RData")
+ordr <- match(desc_cat$name, names(vimp))
+desc_cat$name <- paste0(ordr, ". ", desc_cat$name)
+desc_cat <- desc_cat[order(ordr, decreasing = FALSE), ]
+names(desc_cat) <- c("Variable", "n", "%", "Category")
+papaja::apa_table(desc_cat, caption = "Sample descriptive statistics for categorical predictors.", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
 ```
-```{r tabcont, results='asis'}
- options(knitr.kable.NA = '')
-papaja::apa_table(desc_num[,c("name", "n", "Items", "mean", "sd", "min", "max", "Reliability", "PC1")
-], caption = "Sample descriptive statistics for continuous variables and scales.", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
+```{r tabnum, results='asis'}
+ordr <- match(desc_num$name, names(vimp))
+desc_num$name <- paste0(ordr, ". ", desc_num$name)
+desc_num <- desc_num[order(ordr, decreasing = FALSE), ]
+desc_num <- desc_num[,c("name", "n", "Items", "mean", "sd", "min", "max", "Reliability", "PC1")]
+names(desc_num) <- c("Variable", "n", "i", "M", "SD", "min", "max", "alpha", "PC1")
+papaja::apa_table(desc_num, caption = "Sample descriptive statistics for continuous variables and scales.
+n: valid sample size, i: number of items, M: mean, SD: standard deviation, alpha: Cronbach's alpha", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
 ```
+
+## Missing data
+
+The SEM-forests algorithm requires complete data and cannot accommodate multiple imputed data.
+Consequently, we used single imputation using the missForest algorithm,
+a random forests based approach with comparable performance to multiple imputation and FIML estimation [@stekhovenMissForestNonparametricMissing2012].
+This method is valid when missingness is either random or contingent on observed data (MAR or MCAR).
+It does not make any distributional assumptions and does not extrapolate,
+meaning it returns realistic values within the range of the observed data.
+The percentage of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
+and from $`r report(min((495-desc_num[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for continuous data (see Table \@ref(tab:tabnum)).
+<!-- Original publication had up to 30% missing, Wolputte, E.
+V., & Blockeel, H.
+(2020, October).
+Missing value imputation with MERCS: a faster alternative to MissForest.
+In International Conference on Discovery Science (pp.
+502-516).
+Springer, Cham.
+looked at higher numbers.
+According to Mplus: minimum 10% -->
 
@@
 
+Trajectories of emotion regulation development were modeled using a curvilinear latent growth curve model with an intercept (I), linear slope (S), and quadratic slope (Q).
+The full sample fit was excellent, $\text{RMSEA} < .05, \text{SRMR} = 0.02, \text{CFI} = 1.00$.
+As this model is relatively complex for the subsequent SEM-forests analysis, residual variances were constrained over time, reflecting an assumption of homoscedasticity,
+and variances of growth parameters were fixed at zero,
+reflecting the assumption of homogeneity of trajectories within terminal nodes,
+with remaining heterogeneity attributed to error variance.
+
+To identify predictors of these latent growth trajectories, we used the SEM forests algorithm [@brandmaierTheoryguidedExplorationStructural2016].
+This algorithm proceeds as follows:
+
+1. Bootstrap the data with replacement; we used 1000 bootstrap replications
+2. For each bootstrap sample, construct a SEM tree:
+  i. Estimate the latent growth model on the bootstrap sample
+  ii. Consider a random selection of $K$ candidates out of all $p$ predictors; we used $k = 9$ based on the standard practice to use $K = \lfloor\sqrt{p}\rfloor$
+  iii. Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the likelihood ratio $LR$ of a post-split multigroup model
+  iv. Repeat steps 3 and 4 until all post-split groups reach a minimum size
+3. Aggregate predictions from all SEM-trees
+
+Each tree thus maps relationships between the predictors and outcome as a series of recursive splits of the data.
+This approach intrinsically accommodates non-linear effects and (complex) interactions, which is an advantage over parametric models that assume linearity.
+The algorithm additionally performs variable selection;
+unimportant variables are not selected in step 2.iii.
+Thus, the number of predictors can be large without risking model non-identification.
+
+Our primary outcome of interest, variable importance,
+is defined as the average absolute increase in model misfit due to randomization of each variable.
+For each tree, the log-likelihood of cases not in that tree's bootstrap sample is computed twice:
+Once using the true data for these cases, and once after randomizing one predictor.
+The variable importance is given by averaging the ratio of these two log-likelihoods across all trees.
+Our secondary outcome of interest, marginal associations,
+were derived using a custom function developed for this paper.
+For three values of each continuous predictor - namely the percentiles corresponding to the median, +/- 1SD of a standard normal distribution - 
+model-implied growth trajectories were predicted for all other values of all other predictors.
+To average across all other predictors, the median of these predicted growth trajectories was taken and plotted.
+
+<!-- $$ -->
+<!-- \begin{align} -->
+<!-- \bm{\lambda} &=  -->
+<!-- \begin{bmatrix} -->
+<!-- 1 & 0 & 0^2 \\ -->
+<!-- 1 & 1 & 1^2 \\ -->
+<!-- \vdots & \vdots & \vdots \\ -->
+<!-- 1 & 5 & 5^2 \\ -->
+
+<!-- \end{bmatrix} -->
+<!-- \\ -->
+<!-- \bm{\psi} &= -->
+<!-- \end{align} -->
+<!-- $$ -->
+<!-- Across all measurements, 13.3% of data were missing. -->
+<!-- Little’s (1988) Missing Completely at Random (MCAR) -->
+<!-- test produced a statistically significant value for the daily -->
+<!-- identity and self-concept clarity variables, x -->
+<!-- 2 -->
+<!-- (8592) ¼ -->
+<!-- 9228.07, p < .001, although the x -->
+<!-- 2 -->
+<!-- /df ratio was only 1.07. -->
+<!-- Similarly, for the annual assessments, the MCAR test -->
+<!-- produced a statistically significant value, x -->
+<!-- 2 -->
+<!-- (56) ¼ 75.19, -->
+<!-- p < .001, although the x -->
+<!-- 2 -->
+<!-- /df ratio was only 1.34. These -->
+<!-- findings suggest that data were likely missing at random, -->
+<!-- and that maximum likelihood techniques were appropriate -->
+<!-- for use. Moreover, even when the assumption of missing -->
+<!-- completely at random is not met, the Full Information -->
+<!-- Maximum Likelihood (FIML) procedure is still preferable to -->
+<!-- listwise deletion of cases with missing data (McKnight, -->
+<!-- McKnight, Sidani, & Figuredo, 2007). Listwise deletion -->
+<!-- results in a biased sample, and therefore biased results, -->
+<!-- by systematically excluding cases with missing data on -->
+<!-- any of the study variables. We therefore utilized FIML -->
+<!-- estimation using Mplus release 5.1 (Muthe´n & Muthe´n, -->
+<!-- 2007). -->
 <!-- Our analyses were based on an extension of cross-lagged panel modeling (Selig & Little, 2012), with staggered measurement occasions for one of the variables (mood variability).
@@
 
-SEM-forest [@brandmaierTheoryguidedExplorationStructural2016]
+# Results
 
-1. Bootstrap sample
-2. On each sample, estimate a SEM-model
-3. Consider $k$ candidate predictors
-4. Identify predictor and value that maximizes $LR$ of post-split multi-group model
-5. Average predictions across bootstrap samples
+## Variable importance
 
-We used 1000 bootstrap samples, $k = 9$
+Candidate predictors were rank-ordered by their relative variable importance,
+as illustrated in Figure 1.
+<!-- Table \@ref(tab:tablegend) provides a brief legend of the 30 most important variables, and  -->
+Supplementary Table S1 provides a full legend of all variables.
 
-**Participants**
+```{r figvarimp, fig.cap="Predictors ranked in order of relative importance. m: mother, f: father, b: best friend, s: sibling."}
+knitr::include_graphics("../pcomb.pdf")
+```
 
-* RADAR data (https://www.uu.nl/en/research/radar)
-* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46)
-* Collected between 2006-2011
-* Most families were  medium- to high-SES (10% low-SES)
+```{r legtab, results="asis", eval = FALSE}
+legtab <- read.csv("../legtab.csv", stringsAsFactors = FALSE)
+papaja::apa_table(legtab, caption = "Brief description of 30 most important predictors.", format.arg = list(na_string = ""))
+```
 
-# Results
+Examining the variable importance rankings,
+several themes emerged.
+Specifically, results indicated that several individual differences - personality and related constructs - were ranked among the most important predictors.
+For example, three of the big five personality dimensions rank among the top six predictors: Neuroticism (1), extraversion (4), and agreeableness (6).
+The Behavioral Inhibition Scale (BIS, 2), which reflects motivation to avoid aversive outcomes, ranked highly - although the BAS (73), its counterpart which measures motivation to approach goal-oriented outcomes,
+ranked very low.
+One dimension of empathy - personal distress (9), the tendency to experience aversive emotions in response to the misfortunes of others,
+also ranked highly, and substantially higher than other dimensions of empathy.
+Finally, self-concept clarity (11) also ranked highly; this construct represents the internal consistency of self-referential thought, and has close theoretical ties to identity [@schwartzDailyDynamicsPersonal2011].
+
+A second emerging theme is that most constructs related to conflict frequency and behavior ranked highly, and few ranked low.
+For instance, adolescent-perceived interparental conflict (8),
+adolescents' conflict engagement (escalation) with best friends (10),
+mothers (13) and fathers (20),
+frequency (14) and negative emotionality (28) of adolescents' conflict with both parents,
+and withdrawal from conflicts with best friends (18) and mothers (19).
+
+A third emerging theme is that parenting was relatively underrepresented among the most important predictors;
+the highest-ranking parental predictor was balanced relatedness (3),
+a construct that represents the extent to which parents support and respect the adolescent's autonomy.
+It further appeared that negative and intrusive aspects of parenting were more predictive than other positive practices:
+Prohibition and control of friendship choice (16),
+perceived intrusiveness and meddling (25),
+psychological control (27),
+and supportive criticism (33).
+Positively valenced parenting dimensions like maternal (54) and paternal (55) support were ranked low,
+and neutrally valenced aspects like parental monitoring (sollicitation 71, adolescent disclosure 72) were ranked very low.
+
+It is further of note that some themes that are considered to be theoretically relevant ranked low.
+For instance, indicators of maternal (86) and paternal (87) socio-economic status were ranked the lowest,
+as were other family demographic indicators, like religion (81, 82). 
+Aspects of adolescents' peer networks,
+such as bullying (60, 84) and victimization (51, 62, 83),
+and contact with delinquent peers (85) also ranked low.
+The same applied to substance use;
+the most commonly used substances of alcohol (78) and cigarettes (61) were ranked very low; drug use, which is less common, ranked a bit higher than this (37, but see direction of association).
+
+## Direction of association
+
+**Note that a number of predictors were reverse coded; re-computing the analyses that generate the Figures requires access to a supercomputer, which we have not been able to obtain at this point.
+These reverse-coded plots have been highlighted in red.
+Reverse the + and - 1SD lines to interpret these.**
+
+Figures \@ref(fig:figpdp1)-\@ref(fig:figpdp4) depict the marginal association of the predictors with the outcome,
+also in order of importance.
+This provides an indication of whether the effect is generally positive, negative, or other.
+Note that the dependent variable measures *difficulties* in emotion regulation;
+thus, higher values represent poorer emotion regulation.
+
+```{r figpdp1, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 1-12. Red highlights indicate reverse-coded predictors."}
+knitr::include_graphics("../plots/pdp1-12r.png")
+```
+```{r figpdp2, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 13-24. Red highlights indicate reverse-coded predictors."}
+knitr::include_graphics("../plots/pdp13-24r.png")
+```
+```{r figpdp3, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 25-36. Red highlights indicate reverse-coded predictors."}
+knitr::include_graphics("../plots/pdp25-36r.png")
+```
+```{r figpdp4, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 37-44. Red highlights indicate reverse-coded predictors."}
+knitr::include_graphics("../plots/pdp37-44r.png")
+```
+
+All associations were in a direction consistent with prior literature.
+Among these were several factors associated with less difficulties in emotion regulation (i.e., potential protective factors), including
+parents' autonomy support and tolerance,
+big five extraversion and agreeableness, 
+self-concept clarity,
+problem solving in conflict with both parents,
+parents' own emotion regulation ability,
+and maternal age (older mothers had better adjusted children).
+
+Other factors were associated with more difficulties in emotion regulation (i.e., potential risk factors or diagnostic aids),
+including big five neuroticism, 
+behavioral inhibition system,
+externalizing and internalizing (depression and anxiety) problem behavior,
+frequency and intensity of conflict with both parents,
+and use of the conflict resolutions styles engagement and withdrawal.
+Intrusive parenting behaviors were also negative predictors, including psychological control, control of friendships, paternal power, and constructive criticism.
+All empathy dimensions except perspective taking predicted greater emotion dysregulation,
+as did different daily negative emotions,
+parental stress and interparental conflict,
+age and pubertal development,
+and sex (girls experienced greater difficulties in emotion regulation).
 
@@
 
-## Variable importance
+# Discussion
 
-## Marginal association
+This study used the machine learning algorithm SEM-forests
+to identify the most important early indicators of emotion regulation development across adolescence,
+and to explore the nature of the association of those predictors with development. 
+The results identified several potential protective- and risk factors.
+Although all predictors attained a positive variable importance value,
+there were differences in relative importance.
+Individual differences in personality, motivation, self-concept clarity, and personal distress were among the most important predictors.
+Conflict frequency and behavior also ranked highly.
+Parenting ranked relatively low, with two notable exceptions:
+First, balanced relatedness, a measure of autonomy support, was an important protective factor, and second,
+among all remaining parenting constructs, negative and intrusive dimensions were more predictive (as risk factors) than positive or neutral parenting practices.
+Other themes that ranked very low were family demographics,
+bullying, victimization and delinquency, and substance use.
 
-## Marginal association 2
+Relating these results to prior theory,
+it is noteworthy that the order of importance of these themes appears to correspond to the levels of the bioecological model.
+Predictors at the individual level were most important,
+followed by those at the microsystem level, such as aspects of the relationship with parents,
+followed by the mesosystem, including aspects of the peer network (e.g., bullying/victimization and delinquency), school performance, and socio-economic status,
+and the macrosystem indicator religiosity was among the least important predictors.
 
-## Marginal association 3
+These findings have several implications for the early identification of youth at-risk for emotion regulation difficulties.
+The fact that individual-level predictors were most important
+suggests that individual assessments might be sufficient.
+Holistic approaches that involve the family system might not be necessary for preliminary risk assessment.
+Moreover, some of the most important predictors are already routinely assessed (for example, big five personality).
+Other important predictors are overt, and thus easily observed by teachers and parents - for example, externalizing problem behavior and conflict frequency and behavior.
+Future research might thus develop and validate risk assessment instruments
+focusing on these constructs.
 
-## Marginal association 4
+The present findings have important implications for intervention.
+However, it is important to consider the correlational nature of the study design.
+Our findings reveal the strength of the association of each predictor with emotion regulation development.
+Intervention can only be effective if this association can additionally be assumed to be causal, based on theory or experimental research.
+Two phenomena appear to be prime candidates for intervention,
+based on the variable importance rankings:
+conflict resolution behavior, and (negative) parenting.
 
-## Summary
+There is prior evidence for the efficacy of conflict resolution training in adolescents [@breunlinConflictResolutionTraining2002].
+The same evidence indicates that conflict resolution training caused a decrease in externalizing problems (another important predictor), suggesting potential compounding benefits.
 
-Important:
+Parenting behavior is another promising target for intervention,
+as indeed it has historically been.
+Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting behaviors.
+By contrast, promoting positive parenting practices may have limited potential for impact.
+With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
+Although one meta-analysis did report large effects of the Triple P-Positive Parenting Program on parental outcomes - including some negative parenting behaviors [@fletcherImpactBehaviouralParent2011],
+another meta-analysis called these findings into question,
+citing mixed evidence, a lack of adequate blinding, risk of bias, and the substantial cost of such programs [@wilsonHowEvidencebasedEvidencebased2012].
+<!-- Another study Even when negative parenting has been assessed as outcomes, -->
+<!-- significant effects were found for positive practices only [@marcynyszynGettingEvidencebasedProgram2011]. -->
+<!-- and that there have been prior concerns about the efficacy of such training [@wilsonHowEvidencebasedEvidencebased2012; @cottamDiscoursesUnderpinningParenting2014]. -->
+Also note that the efficacy of parenting training has been evaluated primarily in terms of parents' self-reported outcomes;
+the ultimate benefit for children,
+in particular for adolescents' emotion regulation,
+is less studied.
+The relatively higher importance rankings for negative versus parenting is in line with the well-established *negativity bias*,
+that negative stimuli are more psychologically impactful than positive ones,
+which has also been found in children [@vaishNotAllEmotions2008].
+Future parenting interventions might take this into consideration by devoting sufficient attention to the elimination of negative parenting practices.
 
-* Best predictors: Personality and related constructs
-* Best parental predictor is autonomy support (bal. rel.)
-  + Negative/intrusive parenting more predictive than other positive practices
-* Conflict frequency and behavior
-* Empathy, particularly personal distress
+As parents have a legal and moral imperative to care for their children,
+much of developmental research has understandably focused on parents' role in child and adolescent development.
+It is thus noteworthy that, with the exception of autonomy support and negative parenting,
+other parenting behaviors were rated low in terms of predictive importance.
+This finding seems paradoxical in light of prior theory and empirical research.
+At the same time, these findings are consistent with an increasing number of recent studies - particularly those employing innovative methods to study processes within families - that have found little evidence for parent-directed processes in adolescence.
+At this point, the evidence suggests that we must consider the possibility that parents' influence in adolescence is more limited than once was thought.
+Based on our findings, the best advice to give parents might be to support their individuating adolescents' autonomy,
+refrain from overly invasive or negative parenting practices,
+and model constructive problem solving behavior when conflicts arise.
 
-Less important than expected:
+Another paradoxical finding was that substance use and other "problem behaviors",
+like time spent with delinquent peers, had very low variable importance ratings in predicting emotion dysregulation.
+Even the highest ranked problem behavior, drug use, had no distinguishable marginal effect.
+This suggests that drug use derives its high importance from higher-order interactions with other predictors.
+It might be the case, for example, that for some youth, drug use contributes to self-exploration and social relationships, whereas for others, it may be a form of escapism, self-medication, or even self-harm.
+This is merely a hypothesis, and the present study cannot conclusively identify what interactions are at play.
+Future research should explore this issue further.
 
-* SES
-* Bullying/victimization
-* Delinquency
-* Substance use
-* Monitoring
+The present study set out to identify the most important early indicators of developmental trajectories
+of emotion regulation across adolescence.
+This involves three important challenges:
+First, the number of potentially relevant risk factors, as identified in a prior text mining systematic review, is large.
+Second, there is a paucity of relevant theory to narrow down the candidate predictors.
+Third, methods commonly used in the field are not well suited for large numbers of predictors.
+Machine learning methods that perform variable selection can overcome these challenges.
+We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence.
+The primary aim is to identify the most important predictors of adolescents' trajectories of emotion regulation development.
+The secondary aim is to probe the nature of the marginal association of the predictors with these trajectories.
 
@@
 
-* Most important predictors are routinely assessed (big 5) or overt (conflict)
-  + Prime candidates for early risk assessment
-  + Conflict resolution behavior can be taught
-    - Target for intervention if association is causal
 * Emphasis on parenting (in literature and theory) might not be justified
   + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
-* Order of predictors mirrors bioecological model
 

### 2022-02-26 scripts/assessingforest.R

@@ -36,9 +36,17 @@ fac_labs <- list(
 vim <- readRDS("results/vim_all_22021-08-21_17_23_27.RData")
 VI <- list(variable.importance = semtree:::aggregateVarimp(vim, aggregate = "median", scale = "absolute", TRUE))
 ren <- read.csv("scale_rename.csv", stringsAsFactors = F, header = FALSE)
-names(VI$variable.importance)[names(VI$variable.importance) %in% ren$V1] <- ren$V2[match(names(VI$variable.importance)[names(VI$variable.importance) %in% ren$V1], ren$V1)]
 VI$variable.importance <- sort(VI$variable.importance, decreasing = TRUE)
-names(VI$variable.importance) <- paste0(1:length(VI$variable.importance), ". ", names(VI$variable.importance))
+
+# Legend
+leg <- read.csv("supplemental_table_1.csv", stringsAsFactors = FALSE)
+legtab <- data.frame(Predictor = ren$V2[match(names(VI$variable.importance)[1:30], ren$V1)], Description = leg$description.en[match(names(VI$variable.importance)[1:30], leg$variable.name)])
+                                        
+write.csv(legtab, "legtab.csv", row.names = FALSE)
+names(VI$variable.importance)[names(VI$variable.importance) %in% ren$V1] <- ren$V2[match(names(VI$variable.importance)[names(VI$variable.importance) %in% ren$V1], ren$V1)]
+saveRDS(VI$variable.importance, "variable_importance.RData")
+
+names(VI$variable.importance) <- paste0(1:length(VI$variable.importance), ". ", c(rep("  ", 9), rep("", (length(VI$variable.importance)-9))), names(VI$variable.importance))
 class(VI) <- "ranger"
 v1 <- v2 <- VI
 v1$variable.importance <- v1$variable.importance[1:44]
@@ -48,6 +56,7 @@ p2 <- metaforest::VarImpPlot(v2, 43)+theme(axis.text.y = element_text(hjust=0))+
 library("cowplot")
 pcomb <- plot_grid(p1,p2,
           ncol = 2, nrow = 1)
+ggsave("pcomb.pdf", pcomb, device = "pdf")
 saveRDS(pcomb, "varimp_comb_21-08-2021.RData")
 p <- metaforest::VarImpPlot(VI, length(VI$variable.importance))
 saveRDS(p, "varimp_21-08-2021.RData")

### 2022-02-26 scripts/creatingscales.R

@@ -98,6 +98,17 @@ fitmeasures(r_config)[c("rmsea", "cfi", "tli")]
 # Prepare predictor scales ------------------------------------------------
 unlist(scales_list)[!unlist(scales_list) %in% names(data)]
 pred_list <- lapply(scales_list, function(x){ prcomp(data[x]) })
+revcod <- sapply(scales_list, function(x){ 
+  pc <- prcomp(data[x]) 
+  signs <- sign(pc$rotation[,1])
+  tb <- table(signs)
+  tb <- sort(tb, decreasing = TRUE)
+  names(tb)[1] == "-1"
+  })
+# Note: Neuroticism and self-concept clarity were already reverse-coded;
+# this reverse coding has flipped them the right way around. Thus,
+# remove them from this list.
+saveRDS(names(revcod)[revcod], "revcod.RData")
 
 pc_mat <- t(sapply(pred_list, function(x){
   tst <- summary(x)
@@ -155,7 +166,14 @@ desc_num$name <- renm$V2[match(desc_num$name, renm$V1)]
 write.csv(desc_num, "desc_num.csv", row.names = FALSE)
 
 desc_cat <- desc_noscale[!desc_noscale$type == "numeric", -2]
+desc_cat$unique <- sapply(desc_cat$name, function(i){length(table(reldata[[i]]))})
+desc_cat <- desc_cat[inverse.rle(list(lengths = desc_cat$unique, values = 1:nrow(desc_cat))), ]
 desc_cat <- desc_cat[, !sapply(desc_cat, function(i){all(is.na(i))})][, -3]
+desc_cat <- cbind(desc_cat, do.call(rbind, lapply(unique(desc_cat$name), function(n){
+  as.data.frame.table(prop.table(table(data_noscale2[[n]])))
+})))
+desc_cat <- desc_cat[, c("name", "n", "Freq", "Var1")]
+names(desc_cat) <- c("name", "n", "%", "Category")
 desc_cat$name <- renm$V2[match(desc_cat$name, renm$V1)]
 write.csv(desc_cat, "desc_cat.csv", row.names = FALSE)
 

### 2022-02-26 manuscript/manuscript.Rmd
update 2022-02-26 14:53:24

@@
 This is merely a hypothesis, and the present study cannot conclusively identify what interactions are at play.
-Future research should explore this issue further.
+Future research should explore this further.
 
@@
 
-## Reflections form
+Correspondence between research interest vs empirical importance.
+one of the most commonly studied correlates of emotion regulation in adolescence is parenting [@vanlissaMappingPhenomenaRelevant2021]
 
-* Many predictors show non-linear effects: Emotion dysregulation only for +/-1SD
-* Some predictors show almost no marginal effect (e.g., father's age, drug use)
-  + This suggests they might be important in **interactions**
+One emergent theme in the developmental literature is attention for fathers' role in child development [ref Keizer].
+Authors have lamented the lack of attention to fathers, and the equation of parenting with mothering.
+With this in mind, it is worth noteworthy that the predictive importance of mothers' parenting did not in any way outrank that of fathers' parenting.
+In fact, constructive conflict resolution with fathers was the highest-ranked construct related to conflict with parents;
+paternal power was the highest-ranked parenting construct,
+and maternal and paternal support and socio-economic status were both ranked approximately equally, despite both constructs being historically gendered.
+This highlights the importance of considering both parents as equally relevant in research and in daily life.
+
+Linear statistical models are the default in our field.
+Therefore, an interesting result is that most predictors had non-linear effects:
+Out of the three values contrasted in the marginal dependence plots,
+only the one corresponding to the least desirable conditions diverged from the others.
+<!-- For instanhigh neuroticism, high interparental conflict, low constructive conflict resolution with fathers. -->
+Our findings send an important signal that many of the associations between predictors and developmental outcomes might be better modeled non-linearly.
+In particular, they suggest a notion of "good-enough" preconditions for the development of mature emotion regulation.
+Most adolescents seem to follow trajectories characterized by low difficulties in emotion regulation - except those with the most disadvantage.
+Some prior research has considered this notion of a "good-enough" environment [@vanlissaCostEmpathyParentadolescent2017; @poonGoodEnoughParental2017];
+the consistency of our findings suggests that it may be more ubiquitous than previously thought.
+
+It is also noteworthy that some predictors had relatively high predictive importance,
+but almost no marginal effect (e.g., father's age, drug use, openness to experience).
+Such predictors likely derive their importance from interactions.
+For instance, drug use might be a vehicle for social status or self-harm;
+older fathers might be an asset if they are involved in children's lives but a liability if they prioritize professional success;
+openness to experience may be an asset if it leads to self-discovery but a liability if it leads to risk behavior.
+As the present method is not well-suited to identify what interaction drives a variable's importance,
+these results should be taken as indications for potentially fruitful areas of future research.
 

### 2022-02-27 manuscript/manuscript.Rmd
update 2022-02-27 11:00:23

@@
 
-Trajectories of emotion regulation development were modeled using a curvilinear latent growth curve model with an intercept (I), linear slope (S), and quadratic slope (Q).
-The full sample fit was excellent, $\text{RMSEA} < .05, \text{SRMR} = 0.02, \text{CFI} = 1.00$.
-As this model is relatively complex for the subsequent SEM-forests analysis, residual variances were constrained over time, reflecting an assumption of homoscedasticity,
-and variances of growth parameters were fixed at zero,
-reflecting the assumption of homogeneity of trajectories within terminal nodes,
+Preliminary analyses were conducted to ensure the appropriateness of growth curve modeling.
+To this end, trajectories of emotion regulation development were modeled using a curvilinear latent growth curve model with an intercept (I), linear slope (S), and quadratic slope (Q).
+The full sample fit of this model was excellent, $\text{RMSEA} < .05, \text{SRMR} = 0.02, \text{CFI} = 1.00$.
+Second, latent class growth analyses were conducted to determine the presence of youth with trajectories that indicated elevated risk of emotion regulation difficulties.
+We used the R-package `tidySEM` to estimate and compare 1-6 class solutions of the aforementioned curvilinear growth curve model;
+analysis results are displayed in [Supplemental Table 2]().
+To ensure model convergence in small classes, the latent variable covariance matrix was restricted to zero,
+reflecting the assumption of homogeneity of trajectories,
 with remaining heterogeneity attributed to error variance.
+Class solutions with more than two classes were excluded from consideration due to low class separability (entropy < .9).
+The LMR likelihood ratio test indicated that two classes significantly improved fit ($p < .001$); more classes did not ($p = .26$).
+As displayed in Figure \@ref(fig:figclass),
+this analysis indicated a similar proportion of youth at-risk for difficulties in emotion regulation to what was found in previous studies [cf @leeAdolescentMentalHealth2014].
 
-To identify predictors of these latent growth trajectories, we used the SEM forests algorithm [@brandmaierTheoryguidedExplorationStructural2016].
-This algorithm proceeds as follows:
+```{r figclass, fig.cap="Latent class analysis of risk profiles."}
+knitr::include_graphics("../classplot.pdf")
+```
+
+In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
+It thus answers the question: "Which adolescents have similar growth trajectories?"
+SEM-forests similarly aim to ensure maximum homogeneity of growth trajectories within clusters,
+but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
+They thus answer the question: "Why do some adolescents have similar growth trajectories"?
+
+For the SEM-forests analysis,
+we imposed an additional restriction on the diagonal of the residual covariance matrix, reflecting an assumption of homoscedastic residual variance over time.
+The SEM-forests algorithm operates as follows:
 
@@
     ii. Consider a random selection of $K$ candidates out of all $p$ predictors; we used $k = 9$ based on the standard practice to use $K = \lfloor\sqrt{p}\rfloor$
-  iii. Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the likelihood ratio $LR$ of a post-split multigroup model
+    iii.
+Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the likelihood ratio $LR$ of a post-split multigroup model
     iv. Repeat steps 3 and 4 until all post-split groups reach a minimum size
@@
 among all remaining parenting constructs, negative and intrusive dimensions were more predictive (as risk factors) than positive or neutral parenting practices.
-Other themes that ranked very low were family demographics,
+Themes that ranked notably low were family demographics (SES and religion),
 bullying, victimization and delinquency, and substance use.
 
+## Implications for theory
+
 Relating these results to prior theory,
-it is noteworthy that the order of importance of these themes appears to correspond to the levels of the bioecological model.
+it is noteworthy that predictive importance of different themes appears to correspond to the levels of the bioecological model:
 Predictors at the individual level were most important,
@@
 and the macrosystem indicator religiosity was among the least important predictors.
+
+* Emphasis on parenting (in literature and theory) might not be justified
+  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
+
+Correspondence between research interest vs empirical importance.
+one of the most commonly studied correlates of emotion regulation in adolescence is parenting [@vanlissaMappingPhenomenaRelevant2021]
+
+As parents have a legal and moral imperative to care for their children,
+much of developmental research has understandably focused on parents' role in child and adolescent development.
+It is thus noteworthy that, with the exception of autonomy support and negative parenting,
+other parenting behaviors were rated low in terms of predictive importance.
+This finding seems paradoxical in light of prior theory and empirical research.
+At the same time, these findings are consistent with an increasing number of recent studies - particularly those employing innovative methods to study processes within families - that have found little evidence for parent-directed processes in adolescence.
+Of course, an alternative explanation might be that parenting effects take place earlier in development,
+and are engrained in individual differences by the time youth reach adolescence.
+A recent study on the role of parents in younger children's emotion regulation development casts doubt on this hypothesis however, as it similarly found little evidence for parent-directed processes.
+<!-- aren’t individual differences the result of the environment where those people were growing up, hence parental style as well? -->
+We must thus consider the possibility that parents' influence, at least in adolescence, is less prominent than once was thought.
+Based on present findings, the best advice to give parents might be to support their individuating adolescents' autonomy,
+refrain from overly invasive or negative parenting practices,
+and model problem solving behavior when conflicts arise.
+
+Another paradoxical finding was that substance use and other "problem behaviors",
+like time spent with delinquent peers, had very low variable importance ratings in predicting emotion dysregulation.
+Even the highest ranked problem behavior, drug use, had no distinguishable marginal effect.
+This suggests that drug use derives its high importance from higher-order interactions with other predictors.
+It might be the case, for example, that for some youth, drug use contributes to self-exploration and social relationships, whereas for others, it may be a form of escapism, self-medication, or even self-harm.
+This is merely a hypothesis, and the present study cannot conclusively identify what interactions are at play.
+Future research should explore this further.
+
+One emergent theme in the developmental literature is attention for fathers' role in child development [ref Keizer].
+Authors have lamented the lack of attention to fathers, and the equation of parenting with mothering.
+With this in mind, it is worth noteworthy that the predictive importance of mothers' parenting did not in any way outrank that of fathers' parenting.
+In fact, constructive conflict resolution with fathers was the highest-ranked construct related to conflict with parents;
+paternal power was the highest-ranked parenting construct,
+and maternal and paternal support and socio-economic status were both ranked approximately equally, despite both constructs being historically gendered.
+This highlights the importance of considering both parents as equally relevant in research and in daily life.
+
+## Implications for practice
 
@@
 
-The present findings have important implications for intervention.
+The present findings also have important implications for intervention.
 However, it is important to consider the correlational nature of the study design.
@@
 The same evidence indicates that conflict resolution training caused a decrease in externalizing problems (another important predictor), suggesting potential compounding benefits.
-
 Parenting behavior is another promising target for intervention,
@@
 which has also been found in children [@vaishNotAllEmotions2008].
-Future parenting interventions might take this into consideration by devoting sufficient attention to the elimination of negative parenting practices.
+Future parenting interventions might take this into consideration by devoting sufficient attention to the elimination of negative parenting practices,
+and including children's outcomes, including emotion regulation, into consideration.
 
-As parents have a legal and moral imperative to care for their children,
-much of developmental research has understandably focused on parents' role in child and adolescent development.
-It is thus noteworthy that, with the exception of autonomy support and negative parenting,
-other parenting behaviors were rated low in terms of predictive importance.
-This finding seems paradoxical in light of prior theory and empirical research.
-At the same time, these findings are consistent with an increasing number of recent studies - particularly those employing innovative methods to study processes within families - that have found little evidence for parent-directed processes in adolescence.
-At this point, the evidence suggests that we must consider the possibility that parents' influence in adolescence is more limited than once was thought.
-Based on our findings, the best advice to give parents might be to support their individuating adolescents' autonomy,
-refrain from overly invasive or negative parenting practices,
-and model constructive problem solving behavior when conflicts arise.
-
-Another paradoxical finding was that substance use and other "problem behaviors",
-like time spent with delinquent peers, had very low variable importance ratings in predicting emotion dysregulation.
-Even the highest ranked problem behavior, drug use, had no distinguishable marginal effect.
-This suggests that drug use derives its high importance from higher-order interactions with other predictors.
-It might be the case, for example, that for some youth, drug use contributes to self-exploration and social relationships, whereas for others, it may be a form of escapism, self-medication, or even self-harm.
-This is merely a hypothesis, and the present study cannot conclusively identify what interactions are at play.
-Future research should explore this further.
-
-The present study set out to identify the most important early indicators of developmental trajectories
-of emotion regulation across adolescence.
-This involves three important challenges:
-First, the number of potentially relevant risk factors, as identified in a prior text mining systematic review, is large.
-Second, there is a paucity of relevant theory to narrow down the candidate predictors.
-Third, methods commonly used in the field are not well suited for large numbers of predictors.
-Machine learning methods that perform variable selection can overcome these challenges.
-We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence.
-The primary aim is to identify the most important predictors of adolescents' trajectories of emotion regulation development.
-The secondary aim is to probe the nature of the marginal association of the predictors with these trajectories.
-
-## Reflection content
-
-* Emphasis on parenting (in literature and theory) might not be justified
-  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
-
-Correspondence between research interest vs empirical importance.
-one of the most commonly studied correlates of emotion regulation in adolescence is parenting [@vanlissaMappingPhenomenaRelevant2021]
-
-One emergent theme in the developmental literature is attention for fathers' role in child development [ref Keizer].
-Authors have lamented the lack of attention to fathers, and the equation of parenting with mothering.
-With this in mind, it is worth noteworthy that the predictive importance of mothers' parenting did not in any way outrank that of fathers' parenting.
-In fact, constructive conflict resolution with fathers was the highest-ranked construct related to conflict with parents;
-paternal power was the highest-ranked parenting construct,
-and maternal and paternal support and socio-economic status were both ranked approximately equally, despite both constructs being historically gendered.
-This highlights the importance of considering both parents as equally relevant in research and in daily life.
+## Methodological implications
 
@@
 openness to experience may be an asset if it leads to self-discovery but a liability if it leads to risk behavior.
-As the present method is not well-suited to identify what interaction drives a variable's importance,
-these results should be taken as indications for potentially fruitful areas of future research.
+The present method is not well-suited to identify what interaction drives a variable's importance,
+but these results might indicate potentially fruitful areas of future research.
 
-## Limitations
+## Strengths and limitations
 
-* Selective sample
-  + WEIRD, Dutch, mostly mid- high-SES (restriction of range)
-* Panel study, so few biological / cognitive / neurological measures
-* Relatively small sample size
-  + If you have access to 1000+ sample, please get in touch
-* Computational constraints prevented us from tuning the model beyond default settings, and from exploring marginal effects for higher order interactions
+The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development, where prior work has taken more piecewise approaches, considering only a few predictors at a time.
+<!-- The present study set out to identify the most important early indicators of developmental trajectories -->
+<!-- of emotion regulation across adolescence. -->
+<!-- This involves three important challenges: -->
+<!-- First, the number of potentially relevant risk factors, as identified in a prior text mining systematic review, is large. -->
+<!-- Second, there is a paucity of relevant theory to narrow down the candidate predictors. -->
+<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
+<!-- Machine learning methods that perform variable selection can overcome these challenges. -->
+<!-- We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence. -->
+<!-- The primary aim is to identify the most important predictors of adolescents' trajectories of emotion regulation development. -->
+<!-- The secondary aim is to probe the nature of the marginal association of the predictors with these trajectories. -->
+
+The study also has some limitations, most notably that the sample was selective,
+which may limit the generalizability of the findings.
+Specifically, this was a non-random, ethnically and socially relatively homogenous  sample from the Netherlands.
+This raises some specific concerns, for example, that the low importance of SES might be due to a restriction of range.
+The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
+The only way to address these concerns is to replicate these findings in other samples.
+Note that all code is available in a reproducible archive, which lowers the threshold to replication.
+Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
+One final limitation related to the data is the relatively small sample size.
+Machine learning methods benefit from large samples.
+Although random forests are relatively robust to overfitting,
+larger samples will likely yield more generalizable results,
+and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
+Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
+Computational constraints thus prevented us from tuning the model beyond default settings,
+and from exploring marginal effects for higher order interactions.
+Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.
 
@@
 
-* Proximal factors more predictive of ER development than distal predictors
-* Parenting may be less relevant than previously thought
-* Personality and conflict behavior are candidates for early diagnosis
-* Conflict skills training might be avenue for intervention
+Empirical research on adolescent emotion regulation development has been fragmented,
+and lacks an overarching theoretical framework. 
+The present study took a holistic approach to identify the most important predictors of emotion regulation development using machine learning.
+Candidate predictors were selected based on a prior systematic review of 6305 relevant papers.
+The SEM-forest algorithm was applied to rank-order these predictors by their relative importance for explaining heterogeneity in adolescents' developmental trajectories,
+and to probe the nature of their association with emotion regulation development. 
+Results indicated that the importance of different predictors corresponded somewhat with the levels of the bioecological model:
+Proximal factors were more predictive of emotion regulation development than distal predictors; individual differences were most predictive, and mesosystem factors (e.g., SES, religion) were least predictive.
+The results also suggested that parenting may be less relevant for emotion regulation development than previously thought, and that negative parenting may be more impactful than positive parenting.
+The results have clear implications for early risk assessment:
+Regularly administered personality inventories, and readily observable indicators like externalizing problems and conflict behavior were among the most important predictors, and these are suitable indicators for diagnostic tools.
+The results further suggest that conflict skills training might be a promising avenue for adolescent-directed interventions.
+Parenting interventions might benefit more from addressing the notion of "good enough" parenting and remedying negative and invasive behaviors than from promoting positive parenting behaviors.
+The results also provide clear guidance for the ongoing effort of theory generation in this field:
+Such theory should focus on proximal processes over distal ones.
+Parents' roles might be less central than previously thought;
+their support of adolescents' autonomy and the way they navigate conflict might be more important than their parenting behaviors.
+Moreover, there was no evidence of differences between mothers' and fathers' parenting, indicating that both parents should be equally represented in future theory.
+Finally, common perceptions about risk factors should be reassessed: 
+delinquency, substance use, and bullying were all relatively less important predictors.
+Even drug use, the highest-ranking problem behavior, had no distinguishable marginal effect, suggesting that it is more problematic for some adolescents than others.
+These results illustrate how machine learning can provide a holistic view of a research area,
+complement existing knowledge,
+offer alternative perspectives,
+and guide future research, practice, and theory formation.
 

### 2022-02-27 mixture_model.R

@@ -1,0 +1,45 @@
+library(tidySEM)
+df <- df_anal[1:5]
+class(df) <- "data.frame"
+names(df) <- paste0("de", 1:5)
+res <- mx_growth_mixture(model = "i =~ 1*de1 + 1*de2 + 1*de3 +1*de4 +1*de5
+                           s =~ 0*de1 + 1*de2 + 2*de3 +3*de4 +4*de5
+                           q =~ 0*de1 + 1*de2 + 4*de3 +9*de4 +16*de5
+                           de1 ~~ vde1*de1
+                           de2 ~~ vde2*de2
+                           de3 ~~ vde3*de3
+                           de4 ~~ vde4*de4
+                           de5 ~~ vde5*de5
+                           i ~~ 0*i
+                           s ~~ 0*s
+                           q ~~ 0*q",
+                  classes = 1:6,
+                  data = df,
+                  auto.cov.lv.x = FALSE
+                  )
+saveRDS(res, "mixmod.RData")
+class(res) <- c("mixture_list", class(res))
+table_fit(res)
+# blrts <- BLRT(res, replications = 4)
+props <- apply(class_prob(res[[2]])$individual, 1, which.max)
+
+library(MplusAutomation)
+res_mp <- createMixtures(classes = 1:6,
+                         filename_stem = "mix",
+                         model_overall =
+"i s q | de1@0 de2@1 de3@2 de4@3 de5@4;
+i-q@0;",
+rdata = df,
+usevariables = names(df), run = 1L)
+saveRDS(res_mp, "mix_mplus.RData")
+desc <- mixtureSummaryTable(res_mp)
+write.csv(desc, "supplemental_table_2.csv", row.names = FALSE)
+classlabs <- paste0(c("Low", "High"), ": ", round(res_mp$mix_2_class.out$results$class_counts$modelEstimated$proportion*100), "%")
+p <- plotGrowthMixtures(res_mp[[2]], rawdata = TRUE, bw = TRUE)
+p <- p + scale_linetype_manual(labels = classlabs, values = c(1:2), name = "Risk", guide = guide_legend(reverse = TRUE)) +
+  scale_shape_manual(labels = classlabs, values = c(15, 17), name = "Risk", guide = guide_legend(reverse = TRUE)) +
+  scale_x_continuous(breaks = 0:4, labels = 14:18, expand = c(0, 0)) +
[…]

### 2022-02-27 traverse_light.R

@@ -1,0 +1,114 @@
+# forest <- readRDS("results/full_forest_all_2.RData")
+# tree = forest$forest[[1]]
+# parameters = c("meani", "means", "meanq")
+tree_light <- function(tree, parameters){
+  if(tree$caption == "TERMINAL"){
+    sums <- summary(tree$model)
+    return(list(parameters = sums$parameters$Estimate[match(parameters, sums$parameters$name)]))
+  } else {
+    return(list(rule = tree$rule,
+                left_child = tree_light(tree = tree$left_child, parameters = parameters),
+                right_child = tree_light(tree = tree$right_child, parameters = parameters)))
+  }  
+}
+#tree_light(tree, parameters)
+
+traverse_light <- function(row, tree){
+  if(!is.null(tree[["parameters"]])){
+    return(tree[["parameters"]])
+  } else {
+    traverse_light(row = row,
+                   tree = tree[c("left_child", "right_child")][[(do.call(tree$rule$relation, list(row[tree$rule$name], tree$rule$value))+1)]])
+  }
+}
+forest_light <- function(forest, parameters){
+  out <- lapply(forest$forest, tree_light, parameters = parameters)
+  class(out) <- c("semtree_light", class(out))
+  out
+}
+
+predict_pars <- function(forest, data, parameters){
+  UseMethod("predict_pars", forest)
+}
+predict_pars.semtree_light <- function(forest, data, parameters){
+  t(apply(data, 1, function(r){
+    apply(sapply(forest, function(t){
+      traverse_light(r, t)
+    }), 1, median)
+  }))
+}
+shape_fun <- function(data){
+  apply(data, 1, function(i){
+    #i = df_plot[1,]
+    sns <- sign(diff(i))
+    if(all(sns == -1)){
+      return("Decreasing")
+    }
+    if(all(sns == 1)){
+      return("Increasing")
+    }
+    if(sns[1] == -1){
+      return("Temporary decline")
+    }
+    "Temporary increase"
+  })
+}
+
+
+# Analyses ----------------------------------------------------------------
+forest <- readRDS("results/forest_light.RData")
+param_per_person <- predict_pars(forest, df_anal, parameters)
+
+# Get max per person
+# # First, calculate apex of parabola
+# i = param_per_person[1,]
[…]

### 2022-02-27 manuscript/manuscript.Rmd
update 2022-02-27 12:40:47

@@
 knit              : worcs::cite_all
+urlcolor: blue
 header-includes:
@@
 
-Preliminary analyses were conducted to ensure the appropriateness of growth curve modeling.
-To this end, trajectories of emotion regulation development were modeled using a curvilinear latent growth curve model with an intercept (I), linear slope (S), and quadratic slope (Q).
-The full sample fit of this model was excellent, $\text{RMSEA} < .05, \text{SRMR} = 0.02, \text{CFI} = 1.00$.
+Preliminary analyses were conducted to ensure the appropriateness of growth curve modeling, and determine the presence of heterogeneity in developmental trajectories.
+First, trajectories of emotion regulation development were modeled using a curvilinear latent growth curve model with an intercept (I), linear slope (S), and quadratic slope (Q).
+The full sample model fit was excellent, $\text{RMSEA} < .05, \text{SRMR} = 0.02, \text{CFI} = 1.00$.
 Second, latent class growth analyses were conducted to determine the presence of youth with trajectories that indicated elevated risk of emotion regulation difficulties.
-We used the R-package `tidySEM` to estimate and compare 1-6 class solutions of the aforementioned curvilinear growth curve model;
-analysis results are displayed in [Supplemental Table 2]().
+We used the R-package `tidySEM` [@tidysem] to estimate and compare 1-6 class solutions of the aforementioned curvilinear growth curve model;
+analysis results are displayed in [Supplemental Table 2](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_2.csv).
 To ensure model convergence in small classes, the latent variable covariance matrix was restricted to zero,
@@
 As displayed in Figure \@ref(fig:figclass),
-this analysis indicated a similar proportion of youth at-risk for difficulties in emotion regulation to what was found in previous studies [cf @leeAdolescentMentalHealth2014].
+about 25% of youth were at-risk for emotion dysregulation,
+a similar proportion to what was found in previous studies [cf @leeAdolescentMentalHealth2014].
 
@@
 
+As we set out to identifying important predictors of emotion regulation development,
+this latent class analysis is not sufficient to answer our research question.
 In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
@@
 They thus answer the question: "Why do some adolescents have similar growth trajectories"?
+This method is suitable to meet the study goals.
 
 For the SEM-forests analysis,
-we imposed an additional restriction on the diagonal of the residual covariance matrix, reflecting an assumption of homoscedastic residual variance over time.
+we imposed an additional restriction on the diagonal of the residual covariance matrix,
+reflecting an assumption of homoscedastic residuals over time.
 The SEM-forests algorithm operates as follows:

### 2022-02-27 mixture_model.R

@@ -43,3 +43,12 @@ p <- p + scale_linetype_manual(labels = classlabs, values = c(1:2), name = "Risk
   theme(legend.position = c(0.88, 0.85))
 saveRDS(p, "classplot.RData")
 ggsave("classplot.pdf", p, device = "pdf", units = "in", width = 7, height = 5)
+
+p <- plotGrowthMixtures(res_mp[[2]], rawdata = TRUE, alpha_range = c(0, 0.05))
+p <- p + scale_linetype_manual(labels = classlabs, values = c(1:2), name = "Risk", guide = guide_legend(reverse = TRUE)) +
+  scale_shape_manual(labels = classlabs, values = c(15, 17), name = "Risk", guide = guide_legend(reverse = TRUE)) +
+  scale_colour_manual(labels = classlabs, values = c("green", "red"), name = "Risk", guide = guide_legend(reverse = TRUE)) +
+  scale_x_continuous(breaks = 0:4, labels = 14:18, expand = c(0, 0)) +
+  labs(x = "Age", y = "Difficulties in emotion regulation") +
+  theme(legend.position = c(0.88, 0.85))
+saveRDS(p, "classplot_color.RData")

### 2022-02-27 scripts/plantingforest.R

@@ -30,7 +30,19 @@ df_anal <- as.data.frame(df_anal)
 df_anal <- df_anal[, !names(df_anal) %in% c("awareness", "clarity", "impulsivity", 
                         "goals", "accept", "strategies")]
 
+# Prepare matrices for shiny ap -------------------------------------------
 
+tmp <- df_anal[, !names(df_anal) %in% c(facs, dich_vars)]
+num_mat <- sapply(tmp, quantile, probs = pnorm(c(-2, -1, 0, 1, 2)))
+num_mat <- data.frame(name = colnames(num_mat), t(num_mat))
+names(num_mat)[-1] <- paste0("Q", c(1:5))
+write.csv(num_mat, "shiny_num_mat.csv", row.names = FALSE)
+
+tmp <- df_anal[, names(df_anal) %in% c(facs, dich_vars)]
+num_mat <- sapply(tmp, quantile, probs = pnorm(c(-2, -1, 0, 1, 2)))
+num_mat <- data.frame(name = colnames(num_mat), t(num_mat))
+names(num_mat)[-1] <- paste0("Q", c(1:5))
+write.csv(num_mat, "shiny_num_mat.csv", row.names = FALSE)
 # Preliminary variable selection ------------------------------------------
 
 library(psych)

### 2022-03-05 manuscript/manuscript.Rmd
1st comments
First (minor) comments to the manuscript.

@@
 Seeking out experiences required to develop mature socio-emotional competences.
-At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014],
+At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
 <!-- exposes youth to a risk of emotion dysregulation, -->
-for example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
+For example, they might not yet be sufficiently equipped to regulate the pursuit of unhealthy incentives,
 or persevere in pursuing adaptive goals when facing obstacles.
@@
 Typically, phenomena are identified using narrative literature reviews.
-However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008]
+However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008].
 Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
@@
 In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
-More exploratory research is thus needed to complete the cycle.
+More exploratory research is thus needed to not only complete the cycle, but to enable confirmatory prediction testing growing from exploratory approaches.
 Secondly, and paradoxically given the first argument,
@@
 from 1 (Almost never) to 5 (Almost always).
-Reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
+Score reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
 We used latent variable scores (instead of mean scores) for further analyses.
@@
 The RMSEA and SRMR of the invariant model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
-As reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
+As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
 we considered it acceptable to proceed with individual factor scores for further analysis.
@@
 meaning it returns realistic values within the range of the observed data.
-The percentage of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
+The proportion of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
 and from $`r report(min((495-desc_num[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for continuous data (see Table \@ref(tab:tabnum)).
@@
 but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
-They thus answer the question: "Why do some adolescents have similar growth trajectories"?
+They thus answer the question: "Why do some adolescents have similar growth trajectories?"
 This method is suitable to meet the study goals.
@@
 Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting behaviors.
-By contrast, promoting positive parenting practices may have limited potential for impact.
+By contrast, promoting positive parenting practices may have limited potential for impact or even been detrimental.
 With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
@@
 
-The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development, where prior work has taken more piecewise approaches, considering only a few predictors at a time.
+The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development.
+Prior work has taken more piecewise approaches, considering only a few predictors at a time.
 <!-- The present study set out to identify the most important early indicators of developmental trajectories -->
@@
 Specifically, this was a non-random, ethnically and socially relatively homogenous sample from the Netherlands.
+The WEIRD-nature of the sample does not allow to generalize this study's findings beyond similar populations.
 This raises some specific concerns, for example, that the low importance of SES might be due to a restriction of range.

### 2022-03-05 manuscript/manuscript.Rmd
update 2022-03-05 11:43:29

--- 2022-02-27 manuscript/manuscript.Rmd
@@
 authornote: |
+  CVL conceived of the study, conducted the final analyses, and wrote the draft paper.
+LB conducted preliminary analyses and gave feedback on the draft.
+SB and WM provided data and gave feedback on the draft.
+This work was funded by a NWO Veni Grant (NWO Grant Number VI.Veni.191G.090), awarded to the lead author.
 Data of the RADAR (Research on Adolescent Development And Relationships) study were used (https://doi.org/10.17026/dans-zrb-v5wp).
@@
 abstract: |
-  https://www.srcd.org/news/child-development-special-section-formalizing-theories-child-development
-  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
+  This paper set out to identify the most important early early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests.
+Candidate predictors were selected based on a prior text mining systematic review of 6305 papers.
+Participants were 497 Dutch adolescents.
+Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model.
+SEM-forests were used to group adolescents with similar trajectories, based on their scores on the predictors.
+This person-centered approach not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories.
+The results indicated that individual differences (e.g., in personality) and conflict behaviors in relationships with parents and friends, and internalizing and externalizing were important early predictors.
+Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use.
+The implications of these findings for theory formation and intervention are discussed, and we present an open source risk assessment tool, ERRATA.
 
-  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
-
-  One sentence clearly stating the **general problem** being addressed by  this particular study.
-
-  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
-
-  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
-
-  One or two sentences to put the results into a more **general context**.
-
-  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
-
-  <!-- https://tinyurl.com/ybremelq -->
-
-keywords          : "keywords"
-wordcount         : "X"
+keywords          : "emotion regulation, adolescence, machine learning, theory formation, random forests"
+wordcount         : "6199"
 
@@
 
+<!-- https://www.srcd.org/news/child-development-special-section-formalizing-theories-child-development -->
 <!-- TS: Emotion regulation development is a critical challenge in adolescence. -->
@@
 Relating these results to prior theory,
-it is noteworthy that predictive importance of different themes appears to correspond to the levels of the bioecological model:
+it is noteworthy that predictive importance of different themes appears to correspond to the levels of the bioecological model [@bronfenbrennerBioecologicalModelHuman2007]:
 Predictors at the individual level were most important,
@@
 and the macrosystem indicator religiosity was among the least important predictors.
+Future theory might thus emphasize proximal factors over distal ones [see also @sameroffUnifiedTheoryDevelopment2010].
 
-* Emphasis on parenting (in literature and theory) might not be justified
-  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
-
-Correspondence between research interest vs empirical importance.
-one of the most commonly studied correlates of emotion regulation in adolescence is parenting [@vanlissaMappingPhenomenaRelevant2021]
-
+Another relevant insight for theory formation is that,
+with the exception of autonomy support and negative parenting,
+other parenting behaviors ranked low in predictive importance.
+This might seem counter intuitive in light of prior theory and empirical research.
 As parents have a legal and moral imperative to care for their children,
-much of developmental research has understandably focused on parents' role in child and adolescent development.
-It is thus noteworthy that, with the exception of autonomy support and negative parenting,
-other parenting behaviors were rated low in terms of predictive importance.
-This finding seems paradoxical in light of prior theory and empirical research.
-At the same time, these findings are consistent with an increasing number of recent studies - particularly those employing innovative methods to study processes within families - that have found little evidence for parent-directed processes in adolescence.
+much of developmental research has focused on parents' role.
+One of the few highly specific theories of emotion regulation development focuses on the family context shapes adolescent emotion regulation development [@morrisRoleFamilyContext2007].
+At the same time, however, this finding is consistent with an increasing number of recent studies - particularly those employing innovative methods to study processes within families - that have found little evidence for parent-directed processes in adolescence [@vanlissaRoleFathersMothers2019; @mastrotheodorosFamilyFunctioningAdolescent2020].
 Of course, an alternative explanation might be that parenting effects take place earlier in development,
 and are engrained in individual differences by the time youth reach adolescence.
-A recent study on the role of parents in younger children's emotion regulation development casts doubt on this hypothesis however, as it similarly found little evidence for parent-directed processes.
+However, a recent study tested this hypothesis in a younger sample and similarly found little evidence for parent-directed processes [@vanlissaMothersFathersQuantitative2020].
 <!-- aren’t individual differences the result of the environment where those people were growing up, hence parental style as well? -->
 We must thus consider the possibility that parents' influence, at least in adolescence, is less prominent than once was thought.
-Based on present findings, the best advice to give parents might be to support their individuating adolescents' autonomy,
-refrain from overly invasive or negative parenting practices,
-and model problem solving behavior when conflicts arise.
+Future theories of adolescent emotion regulation development should thus at least go beyond parents' role.
 
-Another paradoxical finding was that substance use and other "problem behaviors",
+Another finding that challenges contemporary discourse is that substance use and other "problem behaviors",
 like time spent with delinquent peers, had very low variable importance ratings in predicting emotion dysregulation.
@@
 It might be the case, for example, that for some youth, drug use contributes to self-exploration and social relationships, whereas for others, it may be a form of escapism, self-medication, or even self-harm.
-This is merely a hypothesis, and the present study cannot conclusively identify what interactions are at play.
-Future research should explore this further.
+This is merely a hypothesis, and further research is required to identify what interactions are at play.
+These results imply that what are commonly thought of as risk factors might not be decisive in adolescents' emotion regulation development;
+such factors ought to be included in future theory, but should not take center stage.
 
-One emergent theme in the developmental literature is attention for fathers' role in child development [ref Keizer].
-Authors have lamented the lack of attention to fathers, and the equation of parenting with mothering.
-With this in mind, it is worth noteworthy that the predictive importance of mothers' parenting did not in any way outrank that of fathers' parenting.
+One emergent theme in the developmental literature is attention for fathers' role in child development [@vanlissaMappingPhenomenaRelevant2021].
+Scholars have lamented the lack of attention to fathers,
+and the equation of parenting with mothering [@keizerProspectiveStudyFather2014a].
+With this in mind, it is worth noteworthy that the predictive importance of mothers' parenting did not outrank that of fathers' parenting in any way.
 In fact, constructive conflict resolution with fathers was the highest-ranked construct related to conflict with parents;
-paternal power was the highest-ranked parenting construct,
+paternal power was the highest-ranked parenting construct;
 and maternal and paternal support and socio-economic status were both ranked approximately equally, despite both constructs being historically gendered.
-This highlights the importance of considering both parents as equally relevant in research and in daily life.
+This highlights the importance of considering both parents as equally relevant in future theory development.
+
+A final consideration relevant for future theory development is to what extent 
+research interest corresponds to the empirical importance of phenomena relevant for adolescents' emotion regulation.
+A prior text mining systematic review comprehensively mapped the phenomena studied in this field [@vanlissaMappingPhenomenaRelevant2021].
+It showed that among the most commonly studied phenomena were parenting, substance use, and bullying - all phenomena that ranked low in empirical importance in the present study.
+Conversely, phenomena found to be important in the present study, such as personality and other individual differences, parental autonomy support, and conflict-related behavior were less prevalent in the literature.
+Of course, the relative importance of predictors may vary across studies,
+and replications of the present findings are required.
+Nonetheless, this is an important reminder that the salience of a construct is not the same as its empirical importance.
+As many theorists base their work on literature reviews and expert opinion,
+focal topics in the literature are likely to garner undue theoretical weight.
+Considering the empirical importance of phenomena may provide guidance in focusing future theory development in this area.
 
@@
 focusing on these constructs.
+
+The method used in the present study is particularly well-suited to risk assessment instruments,
+because machine learning models maximize predictive performance, whereas traditional methods like linear regression maximize performance within the study sample.
+In fact, the trained model from the present study can be used to make predictions for the emotion regulation development of new adolescents.
+To illustrate the potential for risk assessment, we developed ERRATA,
+an Emotion Regulation Risk Assessment Tool for Adolescents (see Figure \@ref(fig:figerrata)).
+Users can select values of categorical predictors,
+and dial in low- to high values on a five point scale for each continuous predictor,
+which are mapped to the percentiles of the observed data distribution corresponding to -2SD, -1SD, the median, +1SD, and +2SD of a standard normal distribution.
+By default, all inputs are set to the median and mode values.
+This means that users can use informative input for the most important predictors,
+while keeping other less important predictors constant at the most common value.
+The ERRATA application is open source, and can thus be customized for clinical use.
+
+```{r figerrata, fig.cap="Screen captures of the ERRATA app."}
+knitr::include_graphics("errata.png")
+```
 
@@
 and including children's outcomes, including emotion regulation, into consideration.
+Simply put, the present results suggest that the best advice for parents might be to support their individuating adolescents' autonomy,
+refrain from overly invasive or negative parenting practices,
+and model problem solving behavior when conflicts arise.
 
@@
 The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development, where prior work has taken more piecewise approaches, considering only a few predictors at a time.
+This allows us, for the first time, to assess the relative importance of different (themes of) predictors, where prior work has taken a more fragmented approach that does not allow for such comparisons.
+Another important contribution of the present study is that, in contrast to many other approaches,
+our method accommodates non-linear associations and interactions.
+Results indeed showed consistent evidence for non-linear effects,
+and some indications for interactions. 
+The present study answers recent calls for more person-centered research that explores heterogeneity in developmental processes.
+Crucially, whereas other approaches have merely *explored* heterogeneity in outcomes (e.g., using latent class analyses),
+the present study *explained* heterogeneity using a comprehensive battery of predictors.
+This answers the question why some adolescents experience more difficulties in emotion regulation than others.
+Another contribution is that the results have clear implications for theory, assessment, and intervention.
 <!-- The present study set out to identify the most important early indicators of developmental trajectories -->
@@
 
-The study also has some limitations, most notably that the sample was selective,
+The study also has limitations, most notably that the sample was selective,
 which may limit the generalizability of the findings.
-Specifically, this was a non-random, ethnically and socially relatively homogenous  sample from the Netherlands.
-This raises some specific concerns, for example, that the low importance of SES might be due to a restriction of range.
+Specifically, this was a non-random, ethnically and socially relatively homogeneous  sample from the Netherlands.
+This raises specific concerns, for example, that the low importance of SES might be due to a restriction of range.
 The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
 The only way to address these concerns is to replicate these findings in other samples.
-Note that all code is available in a reproducible archive, which lowers the threshold to replication.
+Note that all code is made available in a reproducible archive to enable easy replication.
 Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.

### 2022-03-05 scripts/creatingscales.R

@@ -1,15 +1,7 @@
-### Emotional Dysregulation # Creating Scales from Radar ###
-
-#############################
-# Lukas Beinhauer
-# 08/07/20
-#############################
-
-########################################################################################################
+####################
 # Over the following lines, code can be found in order to create scales from the synthetic RADAR data. #
 # Important: run <missRanger.R> first!                                                                 #
 # Scale scores are averages of each individual's item scores per scale/subscale                        #
-########################################################################################################
 
 library(tidySEM)
 library(lavaan)
@@ -136,8 +128,10 @@ scales_list_desc <- lapply(scales_list, function(i){
 scales_list_desc <- scales_list_desc[!sapply(scales_list_desc, is.null)]
 # Descriptive table -------------------------------------------------------
 renm <- read.csv("scale_rename.csv", stringsAsFactors = FALSE, header = F)
+vim <- readRDS("variable_importance.RData")
 # Scales
-desc_scales <- tidySEM:::create_scales.data.frame(reldata, keys.list = scales_list_desc)$descriptives
+scales <- tidySEM:::create_scales.data.frame(reldata, keys.list = scales_list_desc)
+desc_scales <- scales$descriptives
 pc1 <- pc_mat[,1]
 names(pc1) <- rownames(pc_mat)
 desc_scales$PC1 <- formatC(pc1[desc_scales$Subscale], digits = 2, format = "f")
@@ -145,7 +139,9 @@ desc_scales[c("n", "Items")] <- lapply(desc_scales[c("n", "Items")], as.integer)
 
 # Individual items
 data_noscale2 <- reldata[, which(!names(reldata) %in% unlist(scales_list))]
-data_noscale2$cigarettes <- as.numeric(reldata$su11aa04)
+data_noscale2$cigarettes <- as.factor(!(reldata$su11aa04 == min(reldata$su11aa04, na.rm = TRUE)))
+data_noscale2$alcohol <- as.factor(!(scales$scores$alcohol == min(scales$scores$alcohol, na.rm = TRUE)))
[…]

### 2022-03-06 manuscript/manuscript.Rmd
minor comments
minor comments after update

@@
 abstract: |
-  This paper set out to identify the most important early early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests.
+  This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests.
 Candidate predictors were selected based on a prior text mining systematic review of 6305 papers.
@@
 Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model.
-SEM-forests were used to group adolescents with similar trajectories, based on their scores on the predictors.
-This person-centered approach not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories.
+SEM-forests were used to group adolescents with similar trajectories, based on their predictor scores.
+This person-centered approach not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents may follow similar trajectories.
 The results indicated that individual differences (e.g., in personality) and conflict behaviors in relationships with parents and friends, and internalizing and externalizing were important early predictors.
@@
 In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
-More exploratory research is thus needed to not only complete the cycle, but to enable confirmatory prediction testing growing from exploratory approaches.
+More exploratory research is thus needed to not only complete the cycle, but to enable tests confirmatory prediction following from exploratory approaches.
 Secondly, and paradoxically given the first argument,
@@
 other parenting behaviors ranked low in predictive importance.
-This might seem counter intuitive in light of prior theory and empirical research.
+This might seem counter-intuitive in light of prior theory and empirical research.
 As parents have a legal and moral imperative to care for their children,
@@
 
-The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development, where prior work has taken more piecewise approaches, considering only a few predictors at a time.
-This allows us, for the first time, to assess the relative importance of different (themes of) predictors, where prior work has taken a more fragmented approach that does not allow for such comparisons.
+The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development.
+In contrast, prior work has taken more piecewise approaches, considering only a few predictors at a time.
+This allows us, for the first time, to assess the relative importance of different (themes of) predictors, where the more fragmented approach used in prior work does not allow for such comparisons.
 Another important contribution of the present study is that, in contrast to many other approaches,
@@
 The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
+The WEIRD nature of the sample constrict us in generalizing beyond such populations from these findings.
 The only way to address these concerns is to replicate these findings in other samples.

### 2022-03-08 manuscript/manuscript.Rmd
reviewed changes manuscript
Updated manuscript, according to review of prior changes.

@@
 In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
-More exploratory research is thus needed to not only complete the cycle, but to enable tests confirmatory prediction following from exploratory approaches.
+More exploratory research is thus needed to complete the cycle.
 Secondly, and paradoxically given the first argument,
@@
 Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting behaviors.
-By contrast, promoting positive parenting practices may have limited potential for impact or even been detrimental.
+By contrast, promoting positive parenting practices may have limited potential for impact.
 With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
@@
 The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
-The WEIRD nature of the sample constrict us in generalizing beyond such populations from these findings.
 The only way to address these concerns is to replicate these findings in other samples.

### 2022-03-14 manuscript/manuscript.Rmd
update 2022-03-14 08:25:05

--- 2022-03-06 manuscript/manuscript.Rmd
@@
 authornote: |
+  This is a preprint paper, generated from Git Commit # `r gert::git_commit_id()`.
 CVL conceived of the study, conducted the final analyses, and wrote the draft paper.
@@
 Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model.
-SEM-forests were used to group adolescents with similar trajectories, based on their scores on the predictors.
-This person-centered approach not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories.
-The results indicated that individual differences (e.g., in personality) and conflict behaviors in relationships with parents and friends, and internalizing and externalizing were important early predictors.
+SEM-forests were used to group adolescents based on their scores on the predictors, while maximizing the similarity of developmental trajectories within each group.
+Compared to other person-centered approaches, this method not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories.
+The results indicated that the most important early predictors (at age 13) were individual differences (e.g., in personality), conflict behaviors towards parents and friends, and internalizing and externalizing behavior.
 Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use.
@@
 the SEM-forests algorithm was used to
-identify the most important early indicators of developmental trajectories
-of emotion regulation across adolescence.
+identify the most important early indicators, at age 13, of developmental trajectories
+of emotion regulation across adolescence (age 14 through 18).
 The resulting insights indicate which factors are most predictive of adolescents' diverging destinies in terms of emotion regulation development.
@@
 Seeking out experiences required to develop mature socio-emotional competences.
-At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
+At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014],
 <!-- exposes youth to a risk of emotion dysregulation, -->
-For example, they might not yet be sufficiently equipped to regulate the pursuit of unhealthy incentives,
+for example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
 or persevere in pursuing adaptive goals when facing obstacles.
@@
 Typically, phenomena are identified using narrative literature reviews.
-However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008].
+However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008]
 Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
@@
 In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
-More exploratory research is thus needed to not only complete the cycle, but to enable confirmatory prediction testing growing from exploratory approaches.
+More exploratory research is thus needed to complete the cycle.
 Secondly, and paradoxically given the first argument,
@@
 from 1 (Almost never) to 5 (Almost always).
-Score reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
+Reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
 We used latent variable scores (instead of mean scores) for further analyses.
@@
 The RMSEA and SRMR of the invariant model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
-As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
+As reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
 we considered it acceptable to proceed with individual factor scores for further analysis.
@@
 meaning it returns realistic values within the range of the observed data.
-The proportion of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
+The percentage of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
 and from $`r report(min((495-desc_num[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for continuous data (see Table \@ref(tab:tabnum)).
@@
 but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
-They thus answer the question: "Why do some adolescents have similar growth trajectories?"
+They thus answer the question: "Why do some adolescents have similar growth trajectories"?
 This method is suitable to meet the study goals.
@@
 
-**Note that a number of predictors were reverse coded; re-computing the analyses that generate the Figures requires access to a supercomputer, which we have not been able to obtain at this point.
-These reverse-coded plots have been highlighted in red.
-Reverse the + and - 1SD lines to interpret these.**
-
 Figures \@ref(fig:figpdp1)-\@ref(fig:figpdp4) depict the marginal association of the predictors with the outcome,
@@
 
-```{r figpdp1, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 1-12. Red highlights indicate reverse-coded predictors."}
-knitr::include_graphics("../plots/pdp1-12r.png")
+```{r figpdp1, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 1-20."}
+knitr::include_graphics("../plots/pdp1-20.png")
 ```
-```{r figpdp2, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 13-24. Red highlights indicate reverse-coded predictors."}
-knitr::include_graphics("../plots/pdp13-24r.png")
+```{r figpdp2, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 21-40."}
+knitr::include_graphics("../plots/pdp21-40.png")
 ```
-```{r figpdp3, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 25-36. Red highlights indicate reverse-coded predictors."}
-knitr::include_graphics("../plots/pdp25-36r.png")
+```{r figpdp3, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 41-60."}
+knitr::include_graphics("../plots/pdp41-60.png")
 ```
-```{r figpdp4, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 37-44. Red highlights indicate reverse-coded predictors."}
-knitr::include_graphics("../plots/pdp37-44r.png")
+```{r figpdp4, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 61-87."}
+knitr::include_graphics("../plots/pdp61-87.png")
 ```
@@
 Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
+It is also important to consider that, in our efforts to identify important early predictors of emotion dysregulation, we relied on predictors at a single timepoint (age 13).
+It is possible that the relative importance of these predictors changes over time,
+or even dynamically, in interplay with emotion regulation.
 One final limitation related to the data is the relatively small sample size.

### 2022-03-14 pdp_light.R

@@ -1,15 +1,16 @@
-f <- readRDS('c:/git_repositories/veni_forest/results/forest_light.RData')
-f_full <- readRDS('c:/git_repositories/veni_forest/results/full_forest_all_2.RData')
+library(ggplot2)
+library(gtable)
+library(grid)
+f <- readRDS('results/forest_light.RData')
 class(f) <- c("semforest_light", "list")
 attr(f, "parameters") <- c("meani", "means", "meanq")
-dat <- readRDS("c:/tmp/data.RData")
-renm <- read.csv("c:/git_repositories/veni_forest/scale_rename.csv", stringsAsFactors = FALSE, header = FALSE)
-vim <- readRDS('c:/git_repositories/veni_forest/variable_importance.RData')
-catlevs <- read.csv("c:/git_repositories/veni_forest/desc_cat.csv", stringsAsFactors = FALSE)
-vim
-times <- 
+dat <- readRDS("data.RData")
+renm <- read.csv("scale_rename.csv", stringsAsFactors = FALSE, header = FALSE)
+vim <- readRDS('variable_importance.RData')
+catlevs <- read.csv("desc_cat.csv", stringsAsFactors = FALSE)
+times <- matrix(c(rep(1, 5), 0:4, c(0:4)^2), nrow = 5)
 lvls <- c("-1SD", "mean", "+1SD")
-for(i in 30:length(vim)){
+for(i in 1:length(vim)){
   v <- renm$V1[match(names(vim)[i], renm$V2)]
   lbl <- paste0(i, ". ", names(vim)[i])
   col <- dat[[v]]
@@ -30,11 +31,104 @@ for(i in 30:length(vim)){
   }
   tmp <- pd_growth(x = f, data = dat, reference.var = v, points = points, times = times, FUN = "mean")
   tmp[[1]] <- factor(tmp[[1]], labels = l)
-  # ggplot(tmp, aes_string(x = "Time", y = "value", linetype = names(tmp)[1], shape = names(tmp)[1])) + geom_path() + geom_point()
   saveRDS(tmp, file = paste0("pdpdat_", v, ".RData"))
 }
 
-times <- matrix(c(rep(1, 5), 0:4, c(0:4)^2), nrow = 5)
-.trajectory(df_plot[1, -1], t(times))
 
-pd_full <- semtree::partialDependence(f_full, "neuroticism", "means", 2)
+# Merge plots -------------------------------------------------------------
[…]

### 2022-03-14 manuscript/manuscript.Rmd
Incorporate Lucas' comments

--- 2022-03-08 manuscript/manuscript.Rmd
@@
 Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model.
-SEM-forests were used to group adolescents with similar trajectories, based on their predictor scores.
-This person-centered approach not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents may follow similar trajectories.
-The results indicated that individual differences (e.g., in personality) and conflict behaviors in relationships with parents and friends, and internalizing and externalizing were important early predictors.
+SEM-forests were used to group adolescents based on their predictor scores, while maximizing the similarity of developmental trajectories within each group.
+Compared to other person-centered approaches, this method not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories.
+The results indicated that the most important early predictors (at age 13) were individual differences (e.g., in personality), conflict behaviors towards parents and friends, and internalizing and externalizing behavior.
 Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use.
@@
 <!-- exposes youth to a risk of emotion dysregulation, -->
-For example, they might not yet be sufficiently equipped to regulate the pursuit of unhealthy incentives,
+For example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
 or persevere in pursuing adaptive goals when facing obstacles.
@@
 In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
-It thus answers the question: "Which adolescents have similar growth trajectories?"
+It thus answers the question: "Which adolescents have similar growth trajectories"?
 SEM-forests similarly aim to ensure maximum homogeneity of growth trajectories within clusters,
 but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
-They thus answer the question: "Why do some adolescents have similar growth trajectories?"
+They thus answer the question: "Why do some adolescents have similar growth trajectories"?
 This method is suitable to meet the study goals.

### 2022-03-15 manuscript/manuscript.Rmd
work on Susan's comments

@@
 defined as the ability to manage emotional experience and expression,
-is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010, @schaferEmotionRegulationStrategies2017, @braetEmotionRegulationChildren2014]
+is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010; @schaferEmotionRegulationStrategies2017; @braetEmotionRegulationChildren2014]
 and social competence [@zimmermannEmotionRegulationEarly2014] throughout life.
@@
 staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
-The emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
-This imbalance motivates most adolescents to engage in adaptive exploration:
-Seeking out experiences required to develop mature socio-emotional competences.
-At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
+The emotional salience of social stimuli is intensified, while the development of cognitive capacities lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
+This imbalance might help motivate adolescents to engage in adaptive exploration:
+Seeking out experiences that challenge their socio-emotional competences and stimulate growth.
+At the same time, this imbalance may render some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
 <!-- exposes youth to a risk of emotion dysregulation, -->
@@
 
-Several theories are commonly cited in research on adolescent emotion regulation.
+Several theories are commonly cited in research on adolescent emotion regulation [for a review, see @vanlissaMappingPhenomenaRelevant2021].
 Notably, there appears to be a trade-off between these theories' scope and specificity.
@@
 Other theories are narrow in scope, and are sufficiently specific to derive testable predictions, but lack broader context.
-Two examples of macro-theories are the bioecological model [@bronfenbrennerBioecologicalModelHuman2007] and the transactional model [@sameroffUnifiedTheoryDevelopment2010].
+Two macro-theories that have been influential in research on emotion regulation development are the bioecological model [@bronfenbrennerBioecologicalModelHuman2007; e.g., see @augustEmotionRegulationAtRisk2017] and the transactional model [@sameroffUnifiedTheoryDevelopment2010].
 The bioecological model describes how individual development is shaped by factors at different levels of analysis:
@@
 With increasing age, distal influences are thought to become more influential.
-More specific theories primarily focus on cognitive maturation and the role of primary caregivers,
-including the theory of normative emotional development [@sroufeAttachmentDyadicRegulation1996]
+
+Other theories are more specific, focusing on cognitive maturation and the role of primary caregivers.
+This includes the theory of normative emotional development [@sroufeAttachmentDyadicRegulation1996]
 and the tripartite model of emotion regulation socialization by parents [@morrisRoleFamilyContext2007] and peers [@reindlSocializationEmotionRegulation2016].
-Note that the tripartite model is one of the few theories that explicitly addresses the life phase of adolescence.
+Note that the tripartite model explicitly addresses the life phase of adolescence.<!--is one of the few theories that -->
+<!-- This seems a bit selective. For example, also discuss the work of Hollenstein and of Nicholas Allen. -->
+<!-- I think you do insufficient right to the theories that exist, both in breadth and in depth.
+Perhaps just stay closer to the contribution of the current paper: empirically examine multiple factors simultaneously rather than selectively examining a few?
+-->
 It would be beneficial to work towards a more comprehensive theory of emotion regulation development,
@@
 Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
-<!-- reflection and meaning-making-->
+They identify relevant phenomena by grouping related keywords,
+and ranking these by number of studies those keywords occur in.<!-- reflection and meaning-making-->
+This review offers a relatively comprehensive overview of potentially relevant phenomena.
+The present study used it as a starting point for an empirical investigation of potential predictors of adolescents' emotion regulation development.
 
-The results of that review indicated that internalizing and externalizing problems and mental health-related outcomes were most frequently studied in conjunction with emotion regulation. 
-Some relevant phenomena reflected themes also identified in prior narrative reviews,
+According to this review, the phenomena that received most attention in the literature were related to internalizing and externalizing problems, and mental health.
+Although these phenomena are likely to be, at least in part, outcomes of emotion dysregulation - that does not diminish their usefulness as predictors.
+Some of the phenomena that were identified using text mining reflect what has been discussed in prior narrative reviews,
 including cognitive development, parenting, and external stressors [cf. @coe-odessEmergentEmotionsAdolescence2019].
-Other relevant phenomena were underrepresented in prior theory and reviews,
+Other identified phenomena had received less attention in prior theory and reviews,
 such as substance use, structural disadvantage,
 identity, morality, sex and sexual development.
-This relatively comprehensive overview offers a starting point for empirical investigations of predictors of adolescents' emotion regulation development, such as the present study.
+
 <!-- To this end, the present study sought to identify operationalizations of these relevant phenomena in a large-scale panel study. -->
@@
 
+<!-- SB: you have convincingly argued why machine learning in general is important, yet I miss the details related to ER.
+I would recommend to add a paragraph summarizing theories of ER and the factors they consider important.
+Also, describe the candidate predictors that were derived from the literature search, and how they relate to theories and /or empirical evidence -->
 The present study set out to identify the most important early indicators of developmental trajectories
@@
 ```
-Participants were 497 Dutch adolescents and their parents enrolled in the longitudinal RADAR study  [@branjeResearchAdolescentDevelopment2018].
-The RADAR sample was recruited from randomly selected elementary schools in the province of Utrecht, and four main cities in The Netherlands.
-Of 1,544 randomly selected families, 497 produced informed consent and were included in the study.
-From 2006 to 2012, trained interviewers conducted annual home visits to collect questionnaire data from adolescents and parents.
+Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal RADAR study  [@branjeResearchAdolescentDevelopment2018].
+The RADAR sample was recruited from over 200 randomly selected elementary schools in the province of Utrecht, and four main cities in The Netherlands.
+Of 1,544 eligible families, 497 produced informed consent and were included in the study.
+From 2006 to 2011, trained interviewers conducted annual home visits to collect questionnaire data from adolescents and parents.
 All independent variables were assessed in the first measurement wave.
@@
 We set out to identify operationalizations of all relevant phenomena identified in @vanlissaMappingPhenomenaRelevant2021 that were available in the RADAR data.
-This yielded 87 candidate predictors, of which `r nrow(desc_cat)` categorical (see Table \@ref(tab:tabcat)), `r sum(is.na(desc_num$Items))` numerical, and `r sum(!is.na(desc_num$Items))` multi-item scales, see \@ref(tab:tabnum).
+This yielded 87 candidate predictors, of which `r nrow(desc_cat)` categorical (see Table \@ref(tab:tabcat)), `r sum(is.na(desc_num$Items))` numerical, and `r sum(!is.na(desc_num$Items))` multi-item scales, see Table \@ref(tab:tabnum).
+<!-- SB: I miss a reflection on the extent to which the available data cover all relevant predictors -->
 For multi-item scales, we used principal components analysis for dimension reduction.
@@
 The variable importance is given by averaging the ratio of these two log-likelihoods across all trees.
-Our secondary outcome of interest, marginal associations,
-were derived using a custom function developed for this paper.
-For three values of each continuous predictor - namely the percentiles corresponding to the median, +/- 1SD of a standard normal distribution - 
-model-implied growth trajectories were predicted for all other values of all other predictors.
-To average across all other predictors, the median of these predicted growth trajectories was taken and plotted.
+Our secondary outcome of interest were marginal association plots,
+which visualize the partial dependence of each predictor with trajectories of emotion regulation development.
+New functions for computing partial dependence for growth curve SEM-forests were developed for this paper, which are now available in the `semtree` R-package.
+Partial dependence for a predictor $p$ is defined as following:
+let $\hat{\boldsymbol{g}}_s$ be a vector representing the expected growth trajectory for value $s$ on predictor $p$,
+let $\hat{f}$ be the function implied by a SEM-forest trained on predictor matrix $\boldsymbol{X}$ with $n$ rows,
+and $\boldsymbol{X}_{-p}$ be the predictor matrix excluding $p$.
+Partial dependence is then computed using Monte Carlo integration as follows:
+
+$$
+\hat{\boldsymbol{g}}_s = \frac{1}{n}\sum^{n}_{i = 1}\hat{f}(s, \boldsymbol{X}_{-p})
+$$
+
+For continuous predictors, three values of $s$ were used - namely the percentiles corresponding to the median, $+/- 1 SD$ of a standard normal distribution.
+For categorical predictors, all unique values were used.
 
@@
 and supportive criticism (33).
-Positively valenced parenting dimensions like maternal (54) and paternal (55) support were ranked low,
+Positively valenced dimensions of parent-adolescent relationship quality
+like maternal (54) and paternal (55) support were ranked low,
 and neutrally valenced aspects like parental monitoring (sollicitation 71, adolescent disclosure 72) were ranked very low.
@@
 For instance, indicators of maternal (86) and paternal (87) socio-economic status were ranked the lowest,
-as were other family demographic indicators, like religion (81, 82). 
+as were other family demographic indicators, like religion (81, 82).<!--SB: Why is this important?
+CJ: Religion is part of moral development, SES, and is directly related to emotion regulation: https://link.springer.com/chapter/10.1007/978-94-017-8950-9_13-->
 Aspects of adolescents' peer networks,
@@
 there were differences in relative importance.
-Individual differences in personality, motivation, self-concept clarity, and personal distress were among the most important predictors.
+Individual differences in personality, motivational system, self-concept clarity, and personal distress were among the most important predictors.
 Conflict frequency and behavior also ranked highly.
@@
 Predictors at the individual level were most important,
+<!-- SB: this is not very surprising given that many of the behaviors assessed require ER (e.g.
+becoming angry in a conflict is an indication of low ER, and individuals high in Neuroticism also have fluctuating emotions that they find difficult to control/regulate) -->
 followed by those at the microsystem level, such as aspects of the relationship with parents,
@@
 Future theory might thus emphasize proximal factors over distal ones [see also @sameroffUnifiedTheoryDevelopment2010].
+<!-- SB: So in a way this is support for the specific theories that tend to focus on proximal factors such as personality and parenting? -->
+<!-- Yet perhaps there was too little variation in relevant macro factors (ethnicity, SES/poverty, cultural differences in ER) -->
 
-Another relevant insight for theory formation is that,
-with the exception of autonomy support and negative parenting,
-other parenting behaviors ranked low in predictive importance.
-This might seem counter-intuitive in light of prior theory and empirical research.
-As parents have a legal and moral imperative to care for their children,
-much of developmental research has focused on parents' role.
-One of the few highly specific theories of emotion regulation development focuses on the family context shapes adolescent emotion regulation development [@morrisRoleFamilyContext2007].
-At the same time, however, this finding is consistent with an increasing number of recent studies - particularly those employing innovative methods to study processes within families - that have found little evidence for parent-directed processes in adolescence [@vanlissaRoleFathersMothers2019; @mastrotheodorosFamilyFunctioningAdolescent2020].
-Of course, an alternative explanation might be that parenting effects take place earlier in development,
-and are engrained in individual differences by the time youth reach adolescence.
-However, a recent study tested this hypothesis in a younger sample and similarly found little evidence for parent-directed processes [@vanlissaMothersFathersQuantitative2020].
+Another relevant insight for theory formation is that
+results were somewhat consistent with one of the few theories specific to (adolescent) emotion regulation development: the tripartite model,
+which focuses on the family context [@morrisRoleFamilyContext2007].
+In particular,
+child characteristics and
+aspects of the emotional family climate ranked highly (e.g., autonomy support and interparental conflict).
+Parent-child conflict resolution also ranked highly, but is not explicitly mentioned in the tripartite model.
+Perhaps the definition of the emotional family climate could be expanded to include it.
+Counter to this theory, however, parent characteristics (e.g., parental emotion regulation) ranked low in importance.
+Parenting practices also ranked relatively low - with the notable exception of negative parenting behaviors.
+<!-- SB: here you should really go into more depth.
+you do find evidence for several parenting behaviors, yet you seem to dismiss the theory because some other dimensions were not significant.
+It makes sense that the dimensions that spring out in this analysis are the most relevant ones, particularly in adolescence when autonomy support is one of the most important parenting dimensions and parenting behaviors undermining autonomy development might result in most problems.
+-->
+<!-- Also, conflict resolution with parents was important and reflects the quality of the parent-adolescent relationship as well -->
+This finding is consistent with recent studies that, using innovative methods to study processes within families,<!--The finding that positive parenting practices ranked low -->
+have found limited evidence for effects of parenting behaviors on emotional development in adolescence [@vanlissaRoleFathersMothers2019; @mastrotheodorosFamilyFunctioningAdolescent2020] and childhood [@vanlissaMothersFathersQuantitative2020].
 <!-- aren’t individual differences the result of the environment where those people were growing up, hence parental style as well? -->
-We must thus consider the possibility that parents' influence, at least in adolescence, is less prominent than once was thought.
-Future theories of adolescent emotion regulation development should thus at least go beyond parents' role.
+We must thus consider the possibility that parenting behavior, at least in adolescence, is less influential than once was thought.
+The absence of negative behavior and constructive communication in conflict situations might be more important.
+<!-- Future theories of adolescent emotion regulation development should thus at least go beyond parents' role. -->
 
-Another finding that challenges contemporary discourse is that substance use and other "problem behaviors",
-like time spent with delinquent peers, had very low variable importance ratings in predicting emotion dysregulation.
+Given recent interest in fathers' role in child development [@vanlissaMothersFathersQuantitative2020],
+it is worth examining whether differences exist between the importance of predictors pertaining to mothers versus fathers.
+Some scholars have lamented the lack of research attention to fathers,
+and the equation of parenting with "mothering" [@pleckPaternalInvolvementRevised2004].
+Our results showed no systematic differences between mothers and fathers.
+Anecdotally however, out all parent-related constructs, constructive conflict resolution with fathers ranked highest, and out of all parenting behaviors, paternal power ranked highest.
+These observations validate the notion that fathers matter.
+Also note that two historically gendered constructs, support and socio-economic status, were both ranked approximately equally for mothers and fathers.
+Prior research has found support to be more important in mother-adolescent relationships [@vanlissaRoleFathersMothers2019], and family socio-economic status is often defined by reference to the father [@pleckPaternalInvolvementRevised2004].
+Our results highlight the importance of considering both parents as equally relevant in future empirical research and theory development.
+
+A finding that challenges contemporary discourse is that certain phenomena often conceptualized as "problem behaviors" had very low - for example, substance use [cf @willsEmotionRegulationSubstance2017], bullying, and time spent with delinquent peers.
 Even the highest ranked problem behavior, drug use, had no distinguishable marginal effect.
 This suggests that drug use derives its high importance from higher-order interactions with other predictors.
-It might be the case, for example, that for some youth, drug use contributes to self-exploration and social relationships, whereas for others, it may be a form of escapism, self-medication, or even self-harm.
+It might be the case, for example, that for some youth, drug use contributes to self-exploration and social relationships, whereas for others, it may be a form of escapism, self-medication, or even self-harm [see also @nelemansLongitudinalAssociationsSocial2016].
 This is merely a hypothesis, and further research is required to identify what interactions are at play.
@@
 such factors ought to be included in future theory, but should not take center stage.
-
-One emergent theme in the developmental literature is attention for fathers' role in child development [@vanlissaMappingPhenomenaRelevant2021].
-Scholars have lamented the lack of attention to fathers,
-and the equation of parenting with mothering [@keizerProspectiveStudyFather2014a].
-With this in mind, it is worth noteworthy that the predictive importance of mothers' parenting did not outrank that of fathers' parenting in any way.
-In fact, constructive conflict resolution with fathers was the highest-ranked construct related to conflict with parents;
-paternal power was the highest-ranked parenting construct;
-and maternal and paternal support and socio-economic status were both ranked approximately equally, despite both constructs being historically gendered.
-This highlights the importance of considering both parents as equally relevant in future theory development.
 

### 2022-03-23 manuscript/manuscript.Rmd
reverse code difficulties in emotion regulation as per susan's feedback

@@
 ---
-title             : "Complementing theory on adolescent emotion regulation using data-driven insights from machine learning"
+title             : "Using machine learning to identify early predictors of adolescent emotion regulation development"
 shorttitle        : "Emotion Regulation and Machine Learning"
@@
 Compared to other person-centered approaches, this method not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories.
-The results indicated that the most important early predictors (at age 13) were individual differences (e.g., in personality), conflict behaviors towards parents and friends, and internalizing and externalizing behavior.
+The results indicated that the most important early predictors (at age 13) were individual differences (e.g., in personality), conflict behaviors with parents and friends, and internalizing and externalizing behavior.
 Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use.
@@
 
-Several theories are commonly cited in research on adolescent emotion regulation [for a review, see @vanlissaMappingPhenomenaRelevant2021].
-Notably, there appears to be a trade-off between these theories' scope and specificity.
-Some theories are broad in scope, considering many potential risk factors, but are non-specific.
-This limits their usefulness in generating testable predictions, and renders them impervious to falsification.
-Other theories are narrow in scope, and are sufficiently specific to derive testable predictions, but lack broader context.
-Two macro-theories that have been influential in research on emotion regulation development are the bioecological model [@bronfenbrennerBioecologicalModelHuman2007; e.g., see @augustEmotionRegulationAtRisk2017] and the transactional model [@sameroffUnifiedTheoryDevelopment2010].
+A recent review noted that the literature on potential predictors of emotion regulation development is somewhat fragmented,
+with most studies considering only a handful of variables within the scope of a particular sub-discipline [@vanlissaMappingPhenomenaRelevant2021].
+<!-- At the same time, this literature is relatively fragmented, -->
+<!-- as most studies have examined only a limited number of predictors [@vanlissaMappingPhenomenaRelevant2021]. -->
+This fragmentation may be due, in part, to the fact that few theories are tailored specifically to adolescence,
+and none consistently guide research in the field [@bussTheoriesEmotionalDevelopment2019].
+Theories commonly cited in this field vary in scope and specificity:
+Some are broad in scope, considering many potential risk factors,
+but are somewhat non-specific.
+This limits their usefulness in generating testable hypotheses,
+and renders them impervious to falsification or improvement.
+Other theories are narrower in scope, and are sufficiently specific to derive testable predictions, but lack broader context.
+For a more detailed overview, see @vanlissaMappingPhenomenaRelevant2021.
+Two of the most influential broad theories are the bioecological model [@bronfenbrennerBioecologicalModelHuman2007; e.g., see @augustEmotionRegulationAtRisk2017] and the transactional model [@sameroffUnifiedTheoryDevelopment2010].
 The bioecological model describes how individual development is shaped by factors at different levels of analysis:
@@
 and the macrosystem, consisting of cultural norms and values.
-The transactional model is compatible with the bioecological model;
-it describes development as an interplay between the individual child and the environment.
+The transactional model is compatible with the bioecological model,
+and describes development as an interplay between the individual child and the environment.
 It distinguishes proximal environmental influences, roughly corresponding to the microsystem,
@@
 
-Other theories are more specific, focusing on cognitive maturation and the role of primary caregivers.
-This includes the theory of normative emotional development [@sroufeAttachmentDyadicRegulation1996]
-and the tripartite model of emotion regulation socialization by parents [@morrisRoleFamilyContext2007] and peers [@reindlSocializationEmotionRegulation2016].
-Note that the tripartite model explicitly addresses the life phase of adolescence.<!--is one of the few theories that -->
+More specific theories tend to focus on cognitive maturation and the role of primary caregivers.
+Examples include the theory of normative emotional development [@sroufeAttachmentDyadicRegulation1996]
+and the tripartite model of emotion regulation socialization in families [@morrisRoleFamilyContext2007], which has also been applied to the peer context [@reindlSocializationEmotionRegulation2016].
+This tripartite model is one of the few theories that explicitly addresses the life phase of adolescence.<!---->
 <!-- This seems a bit selective. For example, also discuss the work of Hollenstein and of Nicholas Allen. -->
@@
 -->
-It would be beneficial to work towards a more comprehensive theory of emotion regulation development,
-specific to adolescence,
-that is broad in scope and still specific enough to derive testable predictions.
+<!-- Despite this abundance of relevant theory,  -->
+<!-- several reviews have noted the relatively fragmented nature of the literature  -->
+<!-- that  -->
+<!-- few theories are tailored specifically to emotion regulation development in adolescence, -->
+<!-- and none are used consistently to guide research in this field. -->
+Given that few theories are both comprehensive and sufficiently specific to derive testable hypotheses,
+an important challenge for the field is to work towards an integrating theory of emotion regulation development in adolescence.
 
@@
 <!-- * Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000] -->
-
 <!-- Maybe review theory briefly? -->
-The first step towards theory formation or revision is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
+The first step towards theory formation is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
 <!--Thus, offering a starting point for this endeavor,<!-- empirical investigations of predictors of adolescents' emotion regulation development -->
@@
 Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
-They identify relevant phenomena by grouping related keywords,
-and ranking these by number of studies those keywords occur in.<!-- reflection and meaning-making-->
-This review offers a relatively comprehensive overview of potentially relevant phenomena.
-The present study used it as a starting point for an empirical investigation of potential predictors of adolescents' emotion regulation development.
+Specifically, related keywords are grouped into clusters,
+with each cluster representing a phenomenon.
+These phenomena are then ranked by the number of studies those keywords occur in, and potential relations between phenomena are mapped by co-occurrence within studies.<!-- reflection and meaning-making-->
+This review thus offers a relatively comprehensive overview of potentially relevant phenomena.
+The present study used it as a starting point for an empirical investigation of the relative importance and direction of effect of these phenomena.
 
-According to this review, the phenomena that received most attention in the literature were related to internalizing and externalizing problems, and mental health.
-Although these phenomena are likely to be, at least in part, outcomes of emotion dysregulation - that does not diminish their usefulness as predictors.
-Some of the phenomena that were identified using text mining reflect what has been discussed in prior narrative reviews,
+According to this prior review,
+the phenomena that have received most research attention are internalizing and externalizing problems, and mental health.
+These are likely to be, at least in part, outcomes of emotion dysregulation.
+<!-- - but they that does not diminish their usefulness as predictors. -->
+Other important phenomena reflected what has been discussed in prior narrative reviews,
 including cognitive development, parenting, and external stressors [cf. @coe-odessEmergentEmotionsAdolescence2019].
-Other identified phenomena had received less attention in prior theory and reviews,
+Finally, the method identified some phenomena that have received less attention in prior theory and reviews,
 such as substance use, structural disadvantage,
 identity, morality, sex and sexual development.
+<!-- To this end, the present study sought to identify operationalizations of these relevant phenomena in a large-scale panel study. -->
+Prior theory and empirical research thus suggests many phenomena potentially related to heterogeneity in adolescents' emotion regulation development.
 
-<!-- To this end, the present study sought to identify operationalizations of these relevant phenomena in a large-scale panel study. -->
-
-## Complementing theory using machine learning
-
-According to models of knowledge production within the quantitative positivist paradigm,
-a field progresses through the cyclical alternation of exploratory and confirmatory research, also known as the "empirical cycle" [@degrootMethodologieGrondslagenVan1961, @wagenmakersCreativityVerificationCyclePsychological2018].
-Exploratory research gives rise to testable predictions,
-and confirmatory research evaluates testable predictions on new data.
-Crucially, these phases are linked through evaluation of existing theory in light of empirical findings [@degrootMethodologieGrondslagenVan1961].
-The fact that our field still lacks overarching theory despite decades of productive empirical research
-suggests that the empirical cycle is not routinely completed.
-In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
-More exploratory research is thus needed to complete the cycle.
-Secondly, and paradoxically given the first argument,
-the link between confirmatory research and theory is often tenuous.
-For instance, many studies derive their predictions from prior empirical research, not from theory [e.g., @vanlissaCostEmpathyParentadolescent2017].
-Thirdly, theories are rarely falsified or revised in light of empirical findings.
-This failure to evaluate theory leads to flexible, "invulnerable" frameworks that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
-To advance theory formation, our field might benefit from "closing the empirical cycle" through increased exploratory research and reflection on theory.
-
-There are preliminary indications of a move towards exploratory research:
-Developmental scientists are turning to increasingly complex structural equation models.
-Consider, for example, the burgeoning interest in disentangling between-family differences from within-family processes using the innovative random-intercept cross-lagged panel model [@hamakerCritiqueCrosslaggedPanel2015] and dynamic structural equation modeling [@asparouhovDynamicStructuralEquation2018].
-Many authors using these methods have realized that existing theory is not sufficiently precise to derive distinct hypotheses at these two levels [see @nevilleBidirectionalAssociationsScreen2021; @vanlissaMothersFathersQuantitative2020].
-By necessity, these models are thus used in at least a partly exploratory capacity.
-This approach has two pitfalls.
-Firstly, complex models are by definition more flexible and fit data better then simpler models, at the expense of decreased generalizability [@hastieElementsStatisticalLearning2009].
-Flexibility is a hallmark of exploratory research [@wagenmakersCreativityVerificationCyclePsychological2018],
-but increases the risk of false-positive results and questionable research practices in confirmatory research [@wichertsDegreesFreedomPlanning2016, @simmonsFalsePositivePsychologyUndisclosed2011].
-It is therefore essential that flexibility is disclosed, and that exploratory and confirmatory research are clearly demarcated.
-Secondly, structural equation models are inherently a confirmatory method,
-and they can only include a limited number of predictors due to concerns of statistical power [@klinePrinciplesPracticeStructural2005].
-Their use for exploration is limited, effortful, and potentially fraught (e.g., when using p-values for discovery instead of inference).
-We argue that machine learning offers a suitable tool for rigorous exploration:
-These methods discover patterns in data while maximizing generalizability,
-and incorporate checks and balances to curtail false-positive results [@hastieElementsStatisticalLearning2009].
-In the context of the present study, we use a specific machine learning algorithm uniquely suited for theory-guided exploration: SEM-forests [@brandmaierTheoryguidedExplorationStructural2016].
+What is still missing is an understanding of the relative utility of different early indicators in predicting adolescents' developmental outcomes.
+<!-- we do not know the relative importance of  with respect to others not included in the same study. -->
+Put simply, each study has examined a piece of the puzzle,
+but no single study has captured the complete picture.
+This knowledge gap cannot be addressed by comparing the relative importance of different predictors across studies,
+as different studies tend to sample from different populations,
+use different research designs,
+different control variables and statistical models.
+Furthermore, conventional methods are ill-suited to overcome address this research question:
+<!-- In part, this limitation is a consequence of conventional methods: -->
+linear models limit the total number of predictors that can be included due to concerns of power, overfitting, and multicolinearity [REF].
+By contrast, some machine learning methods offer a more holistic approach,
+as they can incorporate many candidate predictors and perform variable selection,
+while incorporating checks and balances to ensure generalizable results [@hastieElementsStatisticalLearning2009].
+The present study used a specific machine learning algorithm uniquely suited for theory-guided exploration: SEM-forests [@brandmaierTheoryguidedExplorationStructural2016].
 
@@
 Also, describe the candidate predictors that were derived from the literature search, and how they relate to theories and /or empirical evidence -->
-The present study set out to identify the most important early indicators of developmental trajectories
-of emotion regulation across adolescence.
-This involves three important challenges:
-First, the number of potentially relevant risk factors, as identified in a prior text mining systematic review, is large.
-Second, there is a paucity of relevant theory to narrow down the candidate predictors.
-Third, methods commonly used in the field are not well suited for large numbers of predictors.
-Machine learning methods that perform variable selection can overcome these challenges.
+The present study set out to obtain a relatively comprehensive understanding of the predictors of emotion regulation development across adolescence.
+<!-- identify the most important early indicators of developmental trajectories -->
+<!-- of  -->
+The primary aim was to identify the most important predictors of adolescents' developmental trajectories of emotion regulation development.
+The secondary aim was to probe the nature of the relationship of each predictor with these trajectories.
+This involves an important challenges, as
+the number of potentially relevant risk factors is too large for conventional methods.
+At the same time, there is a paucity of relevant theory to narrow down the candidate predictors a priori.
+<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
+This challenge can be overcome by machine learning methods that perform variable selection.
 We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence.
-The primary aim is to identify the most important predictors of adolescents' trajectories of emotion regulation development.
-The secondary aim is to probe the nature of the marginal association of the predictors with these trajectories.
 
@@
 Of 1,544 eligible families, 497 produced informed consent and were included in the study.
-From 2006 to 2011, trained interviewers conducted annual home visits to collect questionnaire data from adolescents and parents.
+From 2006 to 2011, trained interviewers conducted annual home visits to collect questionnaire data.
 All independent variables were assessed in the first measurement wave.
 The dependent variable was measured from the second to the sixth wave.
-Adolescents received financial compensation for their participation in annual measurements (approximately \$17).
+Families received €100 financial compensation for their participation in annual measurements.
 
@@
 
-From wave 2-6, adolescents completed the difficulty in emotion regulation scale [DERS, @gratzMultidimensionalAssessmentEmotion2004].
-Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2018],
+From wave 2-6, adolescents completed the Difficulties in Emotion Regulation Scale, which was reverse coded to represent emotion regulation [DERS, @gratzMultidimensionalAssessmentEmotion2004].
+Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2019],
 we used 24 items of this 32-item scale
@@
 We used latent variable scores (instead of mean scores) for further analyses.
-To this end, we imposed a five-factor CFA (one for each wave) on the indicators of the dependent variable.
-Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the overall trajectory.
+To this end, we estimated a CFA with all DERS items loading on one latent variable per measurement wave.
+Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the sample average trajectory.
 As measurement invariance must be imposed in order to compare latent scores across waves,
@@
 names(desc_cat) <- c("Variable", "n", "%", "Category")
-papaja::apa_table(desc_cat, caption = "Sample descriptive statistics for categorical predictors.", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
+papaja::apa_table(desc_cat, caption = "Sample descriptive statistics for categorical predictors, ranked by variable importance.
+Note that variable importance can be interpreted only in relative terms.", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
 ```
@@
 names(desc_num) <- c("Variable", "n", "i", "M", "SD", "min", "max", "alpha", "PC1")
-papaja::apa_table(desc_num, caption = "Sample descriptive statistics for continuous variables and scales.
+papaja::apa_table(desc_num, caption = "Sample descriptive statistics for continuous variables and scales, ranked by variable importance.
 n: valid sample size, i: number of items, M: mean, SD: standard deviation, alpha: Cronbach's alpha", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
@@
 The full sample model fit was excellent, $\text{RMSEA} < .05, \text{SRMR} = 0.02, \text{CFI} = 1.00$.
-Second, latent class growth analyses were conducted to determine the presence of youth with trajectories that indicated elevated risk of emotion regulation difficulties.
+Second, latent class growth analyses were conducted to determine the presence of youth with trajectories that indicated difficulties in emotion regulation (e.g. lower levels or a temporary dip).
 We used the R-package `tidySEM` [@tidysem] to estimate and compare 1-6 class solutions of the aforementioned curvilinear growth curve model;
@@
 The variable importance is given by averaging the ratio of these two log-likelihoods across all trees.
+Note that variable importance has no meaningful scale and can only be interpreted in relative terms (i.e., one variable is more important than another). 
 Our secondary outcome of interest were marginal association plots,
@@
 
-Figures \@ref(fig:figpdp1)-\@ref(fig:figpdp4) depict the marginal association of the predictors with the outcome,
+Figures \@ref(fig:figpdp1)-\@ref(fig:figpdp4) depict the marginal association plots,
 also in order of importance.
+Each plot shows the expected trajectory of emotion regulation development at different values of each predictor.
+For continuous variables, quantiles corresponding to +/- 1SD and the mean of a standard normal distribution were used.
 This provides an indication of whether the effect is generally positive, negative, or other.
-Note that the dependent variable measures *difficulties* in emotion regulation;
-thus, higher values represent poorer emotion regulation.
+For categorical variables, all values are plotted.
 
-```{r figpdp1, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 1-20."}
+```{r figpdp1, fig.cap="Predicted trajectory of emotion regulation development for predictors 1-20."}
 knitr::include_graphics("../plots/pdp1-20.png")
 ```
-```{r figpdp2, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 21-40."}
+```{r figpdp2, fig.cap="Predicted trajectory of emotion regulation development for predictors 21-40."}
 knitr::include_graphics("../plots/pdp21-40.png")
 ```
-```{r figpdp3, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 41-60."}
+```{r figpdp3, fig.cap="Predicted trajectory of emotion regulation development for predictors 41-60."}
 knitr::include_graphics("../plots/pdp41-60.png")
 ```
-```{r figpdp4, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 61-87."}
+```{r figpdp4, fig.cap="Predicted trajectory of emotion regulation development for predictors 61-87."}
 knitr::include_graphics("../plots/pdp61-87.png")
@@
 All associations were in a direction consistent with prior literature.
-Among these were several factors associated with less difficulties in emotion regulation (i.e., potential protective factors), including
+Among these were several factors associated with better emotion regulation (i.e., potential protective factors), including
 parents' autonomy support and tolerance,
@@
 
-Other factors were associated with more difficulties in emotion regulation (i.e., potential risk factors or diagnostic aids),
+Other factors were associated with poorer emotion regulation (i.e., potential risk factors or diagnostic aids),
 including big five neuroticism, 
@@
 age and pubertal development,
-and sex (girls experienced greater difficulties in emotion regulation).
+and sex (boys girls reported better emotion regulation).
 
@@
 A prior text mining systematic review comprehensively mapped the phenomena studied in this field [@vanlissaMappingPhenomenaRelevant2021].
-It showed that among the most commonly studied phenomena were parenting, substance use, and bullying - all phenomena that ranked low in empirical importance in the present study.
+It showed that parenting, substance use, and bullying were among the phenomena most studied in research on emotion regulation.
+These phenomena all ranked relatively low in empirical importance in the present study.
 Conversely, phenomena found to be important in the present study, such as personality and other individual differences, parental autonomy support, and conflict-related behavior were less prevalent in the literature.
@@
 In particular, they suggest a notion of "good-enough" preconditions for the development of mature emotion regulation.
-Most adolescents seem to follow trajectories characterized by low difficulties in emotion regulation - except those with the most disadvantage.
+Most adolescents seem to follow trajectories characterized by high emotion regulation - except those with the most disadvantage.
 Some prior research has considered this notion of a "good-enough" environment [@vanlissaCostEmpathyParentadolescent2017; @poonGoodEnoughParental2017];

### 2022-03-23 mixture_model.R

@@ -1,7 +1,10 @@
 library(tidySEM)
+library(worcs)
+load_data()
 df <- df_anal[1:5]
 class(df) <- "data.frame"
 names(df) <- paste0("de", 1:5)
+df[] <- lapply(df, `*`, -1)
 res <- mx_growth_mixture(model = "i =~ 1*de1 + 1*de2 + 1*de3 +1*de4 +1*de5
                            s =~ 0*de1 + 1*de2 + 2*de3 +3*de4 +4*de5
                            q =~ 0*de1 + 1*de2 + 4*de3 +9*de4 +16*de5
@@ -17,7 +20,7 @@ res <- mx_growth_mixture(model = "i =~ 1*de1 + 1*de2 + 1*de3 +1*de4 +1*de5
                   data = df,
                   auto.cov.lv.x = FALSE
                   )
-saveRDS(res, "mixmod.RData")
+saveRDS(res, "mixmodrev.RData")
 class(res) <- c("mixture_list", class(res))
 table_fit(res)
 # blrts <- BLRT(res, replications = 4)
@@ -31,10 +34,10 @@ res_mp <- createMixtures(classes = 1:6,
 i-q@0;",
 rdata = df,
 usevariables = names(df), run = 1L)
-saveRDS(res_mp, "mix_mplus.RData")
+saveRDS(res_mp, "mix_mplusrev.RData")
 desc <- mixtureSummaryTable(res_mp)
 write.csv(desc, "supplemental_table_2.csv", row.names = FALSE)
-classlabs <- paste0(c("Low", "High"), ": ", round(res_mp$mix_2_class.out$results$class_counts$modelEstimated$proportion*100), "%")
+classlabs <- paste0(c("High", "Low"), ": ", round(res_mp$mix_2_class.out$results$class_counts$modelEstimated$proportion*100), "%")
 p <- plotGrowthMixtures(res_mp[[2]], rawdata = TRUE, bw = TRUE)
 p <- p + scale_linetype_manual(labels = classlabs, values = c(1:2), name = "Risk", guide = guide_legend(reverse = TRUE)) +
   scale_shape_manual(labels = classlabs, values = c(15, 17), name = "Risk", guide = guide_legend(reverse = TRUE)) +

### 2022-03-23 pdp_light.R

@@ -36,13 +36,14 @@ for(i in 1:length(vim)){
 
 
 # Merge plots -------------------------------------------------------------
-scls <- c(-.06, .01)
+scls <- c(-.01, .06)
 revcod <- readRDS('revcod.RData')
 p_list <- vector("list", length = length(vim))
 for(i in 1:length(vim)){
   n <- names(vim)[i]
-  v <- renm$V1[match(names(vim)[i], renm$V2)]
+  v <- renm$V1[pmatch(names(vim)[i], renm$V2)]
   tmp <- readRDS(file = paste0("pdpdat_", v, ".RData"))
+  tmp$value <- -1 * tmp$value
   if("mean" %in% tmp[[1]]){
     if(n %in% revcod){
       levels(tmp[[1]]) <- levels(tmp[[1]])[c(3,2,1)]
@@ -66,7 +67,7 @@ chunks <- list(
 )
 for(thischunk in chunks){
   pdps <- p_list[thischunk]
-  ylab <- "Difficulties in Emotion Regulation"
+  ylab <- "Emotion Regulation"
   catvars <- which(names(vim[thischunk]) %in% catlevs$name)
   n_grobs <- length(pdps)
   grob_rows <- ceiling(n_grobs/4)

### 2022-03-25 manuscript/manuscript.Rmd
update 2022-03-25 12:28:18

@@
 abstract: |
-  This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests.
+  | **Research Highlights**
+  |
+  | * Individual differences (e.g., in personality, motivation, and empathy) are among the most important predictors of emotion regulation development.
+  | * Parenting may be less relevant for emotion regulation development than previously thought, and negative parenting may be more impactful than positive parenting.
+  | * Important predictors most suitable for early risk assessment include regularly administered personality inventories, and readily observable indicators like externalizing problems and conflict behavior.
+  | * Future theory should emphasize proximal- over distal processes, focus on good-enough parenting, and reassess common perceptions about “risky behaviors” like delinquency, substance use, and bullying.
+  |
+  | This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests.
 Candidate predictors were selected based on a prior text mining systematic review of 6305 papers.
@@
 
-bibliography      : ["r-references.bib", "references.bib"]
+bibliography      : ["r-references.bib", "references.bib", "Veni.bib"]
 
@@
 papaja::apa_table(desc_num, caption = "Sample descriptive statistics for continuous variables and scales, ranked by variable importance.
-n: valid sample size, i: number of items, M: mean, SD: standard deviation, alpha: Cronbach's alpha", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
+n: valid sample size, i: number of items, M: mean, SD: standard deviation, alpha: Cronbach's alpha, PC1: variance explained by first principal component", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
 ```
@@
 
+Finally, an emerging theme is that internalizing- and externalizing symptoms
+often associated with emotion regulation ranked moderately in terms of predictive importance.
+Of these, parent-reported externalizing behavior (5) ranked highest, followed by adolescent-reported anxiety (15) and depression (17).
+Adolescent-reported externalizing behavior ranked substantially lower (30).
+
 It is further of note that some themes that are considered to be theoretically relevant ranked low.
@@
 another meta-analysis called these findings into question,
-citing mixed evidence, a lack of adequate blinding, risk of bias, and the substantial cost of such programs [@wilsonHowEvidencebasedEvidencebased2012].
+citing mixed evidence, a lack of adequate blinding, and risk of bias [@wilsonHowEvidencebasedEvidencebased2012].
 <!-- Another study Even when negative parenting has been assessed as outcomes, -->
@@
 
-The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development.
-In contrast, prior work has taken more piecewise approaches, considering only a few predictors at a time.
-This allows us, for the first time, to assess the relative importance of different (themes of) predictors, where the more fragmented approach used in prior work does not allow for such comparisons.
-Another important contribution of the present study is that, in contrast to many other approaches,
-our method accommodates non-linear associations and interactions.
-Results indeed showed consistent evidence for non-linear effects,
-and some indications for interactions. 
+The present study had some unique strengths.
+Most notably, it included a much larger number of independent variables than prior research,
+which has taken more piecewise approaches.
+The present study thus offers a more holistic understanding potential predictors of adolescent emotion regulation development.
+This allows us, for the first time, to assess the relative importance of different predictors.
+<!-- , where the more fragmented approach used in prior work does not allow for such comparisons. -->
+Another important contribution is that, in contrast to many other approaches,
+the method used in the present study accommodates non-linear associations and interactions.
+The value of this approach is supported by our results, which showed consistent evidence for non-linear effects,
+and some indications for the existence of interactions. 
 The present study answers recent calls for more person-centered research that explores heterogeneity in developmental processes.
@@
 The only way to address these concerns is to replicate these findings in other samples.
-Note that all code is made available in a reproducible archive to enable easy replication.
-Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
+Note that all code is made available in a reproducible archive to facilitate replication.
+Another limitation is that the data source was a panel study with mostly questionnaire measures.
+All conclusions are conditional on the available data; thus, for example,
+our study cannot speak to the potential importance of biological (except pubertal development),
+cognitive and neurological factors.
 It is also important to consider that, in our efforts to identify important early predictors of emotion dysregulation, we relied on predictors at a single timepoint (age 13).
-It is possible that the relative importance of these predictors changes over time,
-or even dynamically, in interplay with emotion regulation.
+This does not account for over time change in the relative importance of predictors.
+Another potential limitation is common method bias.
+Many predictors were assessed using adolescent-reported questionnaire measures,
+which raises the question whether individual-level predictors ranked highly due to shared method variance.
+This concern is assuaged somewhat by the observation that parent-reported externalizing behavior (5) was much more predictive than adolescent self-reported externalizing behavior (30).
+This indicates that predictors less susceptible to common method bias nonetheless explained substantial variance.
 One final limitation related to the data is the relatively small sample size.

### 2022-04-29 manuscript/manuscript.Rmd
update 2022-04-29 19:05:32

@@
 This work was funded by a NWO Veni Grant (NWO Grant Number VI.Veni.191G.090), awarded to the lead author.
-Data of the RADAR (Research on Adolescent Development And Relationships) study were used (https://doi.org/10.17026/dans-zrb-v5wp).
+Data of the RADAR (Research on Adolescent Development And Relationships) study were used, and is available under restricted access (https://doi.org/10.17026/dans-zrb-v5wp).
 RADAR has been financially supported by main grants from the Netherlands Organisation for Scientific Research (GB-MAGW 480-03-005, GB-MAGW 480-08-006), Stichting Achmea Slachtoffer en Samenleving (SASS), the Netherlands Organisation for Scientific Research to the Consortium Individual Development (CID; 024.001.003), a grant of the European Research Council (ERC-2017-CoG - 773023 INTRANSITION), and various other grants from the Netherlands Organisation for Scientific Research, VU University Amsterdam, and Utrecht University.
@@
   | **Research Highlights**
-  |
-  | * Individual differences (e.g., in personality, motivation, and empathy) are among the most important predictors of emotion regulation development.
-  | * Parenting may be less relevant for emotion regulation development than previously thought, and negative parenting may be more impactful than positive parenting.
-  | * Important predictors most suitable for early risk assessment include regularly administered personality inventories, and readily observable indicators like externalizing problems and conflict behavior.
-  | * Future theory should emphasize proximal- over distal processes, focus on good-enough parenting, and reassess common perceptions about “risky behaviors” like delinquency, substance use, and bullying.
-  |
+  | • Individual differences (e.g., in personality, motivation, and empathy) are among the most important predictors of emotion regulation development.
+  | • Parenting may be less relevant for emotion regulation development than previously thought, and negative parenting may be more impactful than positive parenting.
+  | • Important predictors most suitable for early risk assessment include regularly administered personality inventories, and readily observable indicators like externalizing problems and conflict behavior.
+  | • Future theory should emphasize proximal- over distal processes, focus on good-enough parenting, and reassess common perceptions about “risky behaviors” like delinquency, substance use, and bullying.
   | This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests.
@@
 classoption       : "man"
-output            : papaja::apa6_pdf
+output            : papaja::apa6_docx
 knit              : worcs::cite_all
@@
 ```{r setup, include = FALSE}
+mask = TRUE
 library("papaja")
@@
 
-The Workflow for Open Reproducible Code in Science (WORCS) was used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020].
-All analysis code, supplemental materials and synthetic data are available at <https://github.com/cjvanlissa/veni_forest>.
-The original data (restricted) and all study documentation are available via <https://doi.org/10.17026/DANS-ZRB-V5WP>.
+`r ifelse(mask, "The Workflow for Open Reproducible Code in Science (WORCS)", "**Masked**")` was used to make analyses reproducible  `r ifelse(mask, "[@vanlissaWORCSWorkflowOpen2020]", "**(masked reference)**")`.
+All analysis code, supplemental materials and synthetic data are available at `r ifelse(mask, "<https://github.com/cjvanlissa/veni_forest>", "**masked URL**")`.
+The original data (controlled access) and all study documentation are available via `r ifelse(mask, "<https://doi.org/10.17026/DANS-ZRB-V5WP>", "**masked URL**")`.
 
@@
 ```
-Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal RADAR study [@branjeResearchAdolescentDevelopment2018].
-The RADAR sample was recruited from over 200 randomly selected elementary schools in the province of Utrecht, and four main cities in The Netherlands.
+Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal study `r ifelse(mask, "RADAR [@branjeResearchAdolescentDevelopment2018]", "**(masked reference)**")`, which
+received ethical approval from the ethical committee of the University Medical Centre `r ifelse(mask, "Utrecht", "**masked**")`.
+The `r ifelse(mask, "RADAR", "**masked**")` sample was recruited from over 200 randomly selected elementary schools in the province of `r ifelse(mask, "Utrecht", "**masked**")`, and four main cities in `r ifelse(mask, "The Netherlands", "**masked**")`.
 Of 1,544 eligible families, 497 produced informed consent and were included in the study.

### 2022-04-29 presentation_uw2022.Rmd
update 2022-04-29 19:05:32

--- 2022-02-13 presentation_sra2022.Rmd
@@
 ---
-title: "Complementing theory on adolescent emotion regulation using  machine learning"
-author: "Caspar J. Van Lissa<sup>1</sup>, Lukas Beinhauer, Susan Branje, Wim H. J. Meeus"
+title: "Developmental Data Science"
+author: "Caspar J. Van Lissa<sup>1</sup>"
 date: "<p><sup>1</sup>: Utrecht University, dept. Methodology & Statistics | Open Science Community Utrecht<br>Funded by NWO Veni grant  VI.Veni.191G.090"
@@
     css: slide.css
-bibliography      : ["manuscript/references.bib"]
+bibliography      : ["manuscript/references.bib", "../rigorous_exploration/manuscript/r-references.bib", "../rigorous_exploration/manuscript/references.bib", "../rigorous_exploration/manuscript/Rigorous_exploration.bib", "../rigorous_exploration/manuscript/Veni.bib", "../rigorous_exploration/manuscript/Veni_sysrev.bib"]
 ---
@@
 #css: https://maxcdn.bootstrapcdn.com/bootstrap/3.3.7/css/bootstrap.min.css
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
 ```
+
+# Conceptual introduction
+
+## Central thesis
+
+_Machine learning_ can help advance _theory formation_ in developmental psychology
+
+* Replication crisis [@scheelWhyMostPsychological2022; @lavelleWhenCrisisBecomes2021]
+* Solutions improve deductive (theory-testing) research practices
+    + questionable research practices (QRPs)
+    + replication
+    + preregistration
+
+Many preregistered hypotheses are not supported [@scheelExcessPositiveResults2021]
+
+* Lack of "good theories" another explanation for replication crisis
+
+## Defining theory
+
+**Theory**: A model that describes the nature of relations between several phenomena in sufficient detail that it allows for the derivation of quantifiable hypotheses that can be subject to (severe) testing.
+
+* most psychological theories fall short of this definition
+    + insufficiently precise
+    + too flexible
+* Compare to @walasekEvolutionaryModelSensitive2022
+
+Why do we need theory?
+
+* Useful as an instrument of cumulative knowledge acquisition
+* Communicating social scientific research to its consumers
+
+## Deduction vs induction
+
+```{r, out.width="60%"}
+library(tidySEM)
+lay <- get_layout("the1", "the2",
+                  "hyp1", "hyp2",
+                  "pat1", "obs2",
+                  "obs1", "con2", rows = 4)
+nodes <- data.frame(
+  name = as.vector(lay),
+  label = c("Theory", "Working hypothesis", "Pattern", "Observation", "Theory", "Hypothesis", "Observation", "Test"),
+  shape = rep("rect", length(lay)),
+  group = rep(c("Inductive", "Deductive"), each = 4)
+  )
+edges <- data.frame(
+  from = c("the1", "hyp1", "pat1", "the2", "hyp2", "obs2"),
+  to = c("hyp1", "pat1", "obs1", "hyp2", "obs2", "con2"),
+  arrow = rep(c("first", "last"), each = 3),
+  group = rep(c("Inductive", "Deductive"), each = 3)
+)
+g <- prepare_graph(edges = edges, layout = lay, nodes = nodes)
+plot(g)
+```
+
+* Lack of theory cannot be overcome by improving deductive research
+* Theory formation requires inductive (exploratory) research [@creswellDesigningConductingMixed2017]
+
+## Empirical cycle
+
+```{r, out.width="60%"}
+library(tidySEM)
+lay <- get_layout("",     "the1", "", 
+                  "pat1", "",     "hyp1",
+                  "",     "obs1", "",
+                  "",     "tes",  "", rows = 4)
+nodes <- data.frame(
+  name = c("pat1", "the1", "obs1", "tes", "hyp1"),
+  label = c("Patterns", "Theory", "Observation", "Test", "Hypothesis"),
+  shape = "rect")
+edges <- data.frame(
+  from = c("the1", "hyp1", "obs1", "pat1", "obs1"),
+  to = c("hyp1", "obs1", "pat1", "the1", "tes"),
+  arrow = "last"
+)
+g <- prepare_graph(edges = edges, layout = lay, nodes = nodes, angle = 5)
+g$edges$connect_from <- c("right", "bottom", "left", "top", "bottom")
+g$edges$connect_to <- c("top", "right", "bottom", "left", "top")
+
+plot(g)
+```
+
+De Groot’s empirical cycle: a model of knowledge production through scientific research [@degrootMethodologieGrondslagenVan1961]
+
+## Rigorous exploration
+
+How can we improve exploratory research?
+
+Role of flexibility:
+
+* Confirmatory methods poorly suited to exploratory research
+    + p-values: Type I error is 5%, but false discovery rate (FDR) is 14-50% [@vidgenPValuesMisunderstoodMisused2016].
+    + Model fit indices: manually specify models, no guarantee that best model is included
+
+"unguided exploration" is effortful and inflates the risk of spurious results
+
+## Machine learning
+
+**Machine learning**:
+
+* automated model building
+* checks and balances to prevent overfitting
+* maximize predictive performance and generalizability of results
+
+Psychology can benefit from its superior predictive performance [@yarkoniChoosingPredictionExplanation2017]
+
+**NEW**: Implications for the theory crisis have not yet been discussed
+
+## Machine learning vs unguided exploration
+
+```{r}
+knitr::include_graphics("Figure Machine learning 4.png")
+```
+
+## Phenomena detection
+
+First step in *Theory Construction Methods* is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
+
+* Phenomena: stable and general features of the world
+* Few quantitative methods for phenomena detection (aside from expert opinion)
+
+**Text mining** may be suitable for phenomena detection [@vanlissaMappingPhenomenaRelevant2021]
+
+* Unsupervised learning method
+* Clusters in keywords and abstracts
+
+## Holistic approach
+
+Every study examines only a piece of the puzzle; we never see the complete picture
+
+Machine learning accommodates more predictors than classical methods
+
+* Regularization 
+* Variable selection
+
+Including all relevant predictors is important:
+
+* Assumption for causal interpretation
+    + Multicollinearity
+* Good theory incorporates most important causes
+    + Cannot assess the relative utility across studies
+* Include potential predictors from various theories, and undertheorized factors
+
+## Complex effects
+
+Many machine learning methods accommodate:
+
+* non-linear effects 
+* higher-order interactions, without having to specify the nature of these effects a-priori.
+* Automatic: tree-based methods
+* Manual: penalized methods
+
+Psychological theories rarely account for complex effects
+
+* Machine learning can provide nuance by revealing them
+
+## Theoretical elements
+
+Some machine learning methods incorporate theoretical elements
+
+* E.g.: assumption that development follows a latent growth curve (LGC).
+* SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
+* regularized SEM [@jacobucciRegularizedStructuralEquation2016]
+
+When the theory is a (nomological) network:
+
+* LASSO-penalized Gaussian graphical model (GGM) [@epskampGeneralizedNetworkPsychometrics2017]
+* E.g.: network theory of major depression [@cramerMajorDepressionComplex2016]
+
+All of these methods allow for theory guided exploration using machine learning
+
+## Person-centered approaches
+
+Explain heterogeneity at a more fine-grained level than the whole sample
+
+* E.g., latent class analyses
+    + "Which individuals are similar?"
+    + Other *unsupervised learning* methods exist
+* RI-CLPM and DSEM
+    + Rarely explain heterogeneity in within-person effects
+    + Regularization
+* Tree-based models
+    + Group individuals based on predictors to maximize homogeneity of the outcome
+    + "Why are these individuals similar?"
+
+## Generalizability
+
+When using data models to guide theory formation, generalizability is essential
+
+Checks and balances to curtail *overfitting*
+
+* cross-validation to estimate predictive accuracy
+
+<!-- Any model captures both *signal* and *noise* -->
+
+<!-- **Overfitting**: model captures so much noise that it generalizes poorly to new samples -->
+
+```{r, out.width="40%"}
+library(ggplot2)
+colors <- c("Variance" = "red", "Bias" = "green", "Total error" = "black")
+
+ggplot() +
+  xlim(-3, 3) +
+geom_function(
+  data = NULL,
+  stat = "function",
+  position = "identity",
+  fun = ~.x^2,
+  na.rm = FALSE,
+  inherit.aes = TRUE,
+  size = 1,
+  aes(colour = "Total error")
+) +
+  geom_function(
+    data = NULL,
+    stat = "function",
+    position = "identity",
+    fun = ~2^.x -2,
+    na.rm = FALSE,
+    inherit.aes = TRUE,
+    aes(colour = "Bias"),
+    size = 1
+  )+
+  geom_function(
+    data = NULL,
+    stat = "function",
+    position = "identity",
+    fun = ~2^-.x -2,
+    na.rm = FALSE,
+    inherit.aes = TRUE,
+    aes(colour = "Variance"),
+    size = 1
+  ) +
+  labs(x = "Model complexity",
+       y = "Optimal model complexity",
+       color = "Legend") +
+  scale_color_manual(values = colors) +
+  geom_vline(xintercept = 0, linetype = 2) +
+  theme_bw() +
+  theme(axis.text = element_blank(), axis.ticks = element_blank())
+```
+
+## From models to theory
+
+Naive interpretative approach
+
+* Variable importance metrics
+    + congruent with theoretical assumptions about important predictors?
+    + any theoretically important predictors rank low?
+    + any undertheorized factors rank high?
+
+* Marginal associations
+    + non-linear effects?
+    + high importance but flat marginal association?
+
+Comparing predictive performance of simpler parametric model that represents these 'insights' to machine learning model
+
+## From models to theory 2
+
+Data models -> formal theory [@haslbeckModelingPsychopathologyData2021]
+
+* abductive formal theory construction (AFTC) framework
+    + construct formal theory, generate data, mathematically compare to empirical data
+    + Discrepancies: amend formal theory
+
+## Summary
+
+* There is a paucity of good theory
+* Need for exploratory research for theory formation
+* Machine learning for rigorous exploration
+    + automates model building
+    + incorporates checks and balances for generalizable results
+* Unsupervised learning can assist in phenomenon detection
+* Supervised learning to identify important predictors
+* Some algorithms incorporate basic theoretical elements
+
+# Applied examples
 
@@
 
-## Relevant theory
-
-* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
-* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
-* Hall's notion of "storm and stress"
-* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
-* Tripartite model [@morrisRoleFamilyContext2007]
-* Internalization model [@holodynskiDevelopmentEmotionsEmotion2006]
-* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
-* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
-* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
-* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]
-
 ## Limitations of relevant theory
@@
 
+## Open science
+
+All code and data available at <https://github.com/cjvanlissa/veni_sysrev>
+
+Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]
+
+## Baseline
+
+Manual selection of key constructs from relevant theory and prior reviews [ @bariolaChildAdolescentEmotion2011; @coe-odessEmergentEmotionsAdolescence2019]:
+
+* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
+* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
+* Hall's notion of "storm and stress"
+* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
+* Tripartite model [@morrisRoleFamilyContext2007]
+* Internalization model [@holodynskiDevelopmentEmotionsEmotion2006]
+* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
+* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
+* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
+* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]
+
+## Baseline network
+
+```{r figbaseline, warning=FALSE, eval = TRUE, results='asis', out.width = "90%", fig.cap="Phenomena relevant to adolescents' emotion regulation according to theory (a) and narrative reviews (b; transparent nodes indicate constructs also present in the theory)."}
+knitr::include_graphics("../veni_sysrev/baseline_network.png") 
+```
+
+(a) Theory (b) narrative reviews; transparent nodes indicate constructs also present in theory.
+
+## {.fullslide}
+
+![](../veni_sysrev/varimps_small.png)
+
 ## Co-occurrence graph
@@
 ```{r networks1, eval = TRUE, results='asis', out.width = "90%"}
-knitr::include_graphics("tmnetworks_small.png")
+knitr::include_graphics("../veni_sysrev/tmnetworks_small.png")
 ```
@@
 Author keywords (a) and abstracts (b)
+
+## Results
+
+* Both analyses reflected some constructs from theoretical literature
+    + Particularly related to neurodevelopment and socialization
+* Mental health-related outcomes feature prominently
+    + Emotion regulation implicated in mental health problems [@leeAdolescentMentalHealth2014],
+    + This underlines the importance of research in this area
+* Substantial correspondence between keywords and abstracts, suggesting validity of method
+* Networks are sparse; few connections among constructs: fragmented literature
+
+## Undertheorized themes
+
+* Developmental disorders
+* Physical health (sic)
+* External stressors
+* Structural disadvantage
+* Addictive behavior
+* Identity and moral development
+* Sexual development
+
+## Implications
+
+* Empirical research relies on theory; under-theorized phenomena may be overlooked
+    + This study offers guidance for phenomena to consider as confounders/causes
+* Unembedded terms indicate promising areas of future research
+* Present study serves as starting point for theory development
 
@@
 
-## Marginal association 2
-
-```{r, out.width="40%"}
-knitr::include_graphics("plots/pdp13-24.png")
-```
-
-## Marginal association 3
-
-```{r, out.width="40%"}
-knitr::include_graphics("plots/pdp25-36.png")
-```
-
-## Marginal association 4
-
-```{r, out.width="40%"}
-knitr::include_graphics("plots/pdp37-44.png")
-```
-
 ## Summary
@@
 
-## {.fullslide}
-
-![](polarplot2.png)
-
 ## Reflections form
@@
   + This suggests they might be important in **interactions**
-
-## Limitations
-
-* Selective sample
-  + WEIRD, Dutch, mostly mid- high-SES (restriction of range)
-* Panel study, so few biological / cognitive / neurological measures
-* Relatively small sample size
-  + If you have access to 1000+ sample, please get in touch
-* Computational constraints prevented us from tuning the model beyond default settings, and from exploring marginal effects for higher order interactions
 

### 2022-06-09 presentation_vvsor.Rmd
add vvs_or presentation

--- 2022-04-29 presentation_uw2022.Rmd
@@
 ---
-title: "Developmental Data Science"
-author: "Caspar J. Van Lissa<sup>1</sup>"
-date: "<p><sup>1</sup>: Utrecht University, dept. Methodology & Statistics | Open Science Community Utrecht<br>Funded by NWO Veni grant  VI.Veni.191G.090"
+title: "Closing the empirical cycle"
+author: "Using machine learning for rigorous exploration"
+date: "Caspar J. Van Lissa<p>Utrecht University | Open Science Community Utrecht<br>Tilburg University<br>Funded by NWO Veni grant  VI.Veni.191G.090</p>"
 knit: (function(inputFile, encoding) {
@@
 
-Why do we need theory?
+Why should statisticians care about theory?
 
+* It dictates what models are sensible
 * Useful as an instrument of cumulative knowledge acquisition
@@
 
+$$
+FDR = \frac{FP}{TP + FP}
+$$
+
 ## Machine learning
@@
 
+* E.g., personalized (mental) health care, automated assessment aids
+
 **NEW**: Implications for the theory crisis have not yet been discussed
 
-## Machine learning vs unguided exploration
+## Unguided vs rigorous exploration
 
-```{r}
-knitr::include_graphics("Figure Machine learning 4.png")
+```{r, out.width="100%"}
+knitr::include_graphics("unguidedvrigorous.png")
 ```
@@
 
+Substantial empirical research, but no overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]
+
 ```{r, out.height="20%"}
@@
 ```
-
-## Knowledge gap
-
-Substantial empirical research, but:
-
-* Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014]
-* No consistent terminology [@bariolaChildAdolescentEmotion2011]
-* No conceptual frameworks [@stifterEmotionRegulation2019]
-* No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]
-
-Knowledge must be consolidated into overarching theory
-
-## Limitations of relevant theory
-
-* Few explicitly address adolescence
-* Few comprehensively address predictors of emotion regulation
-* None directly guide contemporary research 
-* Theories vary widely in scope: Some are broad and non-specific; others describe a specific phenomenon in detail, but lack a broader perspective.
-    + Broad theories can frame any research, specific theories generate hypotheses.
-    + It would be beneficial to bridge these levels of analysis.
 
@@
 
-## Baseline
-
-Manual selection of key constructs from relevant theory and prior reviews [ @bariolaChildAdolescentEmotion2011; @coe-odessEmergentEmotionsAdolescence2019]:
-
-* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
-* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
-* Hall's notion of "storm and stress"
-* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
-* Tripartite model [@morrisRoleFamilyContext2007]
-* Internalization model [@holodynskiDevelopmentEmotionsEmotion2006]
-* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
-* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
-* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
-* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]
-
 ## Baseline network
@@
 
-## {.fullslide}
-
-![](../veni_sysrev/varimps_small.png)
-
-## Co-occurrence graph
+## TMSR results
 
@@
 
-## Results
+## Interpretation
 
@@
 * Developmental disorders
-* Physical health (sic)
+* Physical health
 * External stressors
@@
 
-## Implications
-
-* Empirical research relies on theory; under-theorized phenomena may be overlooked
-    + This study offers guidance for phenomena to consider as confounders/causes
-* Unembedded terms indicate promising areas of future research
-* Present study serves as starting point for theory development
-
-## Next step
+## Next step: Identifying individual risk factors
 
@@
 
-```{r}
+```{r, out.width = "65%"}
 knitr::include_graphics("tree.png")
@@
 * Most families were  medium- to high-SES (10% low-SES)
-
-## Candidate predictors
-
-Predictors: 87
-
-Categories:
-
-* Demographics (e.g., SES)
-* Biological (age, sex, pubertal development)
-* Individual differences (big5, BIS/BAS, empathy)
-* Risk behavior (in/externalizing symptoms, substance use, delinquency, school performance, victimization)
-* Relationship quality (parents/peers)
-* Parenting (perceived and parent-reported)
-* Conflict resolution styles
-
-## Dependent variable
-
-Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
-
-* 24 items (omitted 'emotional awareness' subscale)
-* E.g.: “When I’m upset, I have difficulty thinking about anything else”
-* 1 (“Almost never”) to 5 (“Almost always”)
-* Excellent reliability, $(.94 > \alpha > .95)$.
-* Metric invariance across waves
-* Quadratic growth model had good fit
-  + CFI = .996, TLI = .993, RMSEA = .047, SRMR = 0.020
+* 87 candidate predictors
+* DV: quadratic growth model of Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
 
@@
 
-```{r, out.width="40%"}
+```{r, out.width="37%"}
 knitr::include_graphics("plots/pdp1-12.png")
@@
 
-## Summary
+## Interpretation
 
-Important:
+Important predictors:
 
-* Best predictors: Personality and related constructs
+* Personality and related constructs
 * Best parental predictor is autonomy support (bal. rel.)
-  + Negative/intrusive parenting more predictive than other positive practices
+  + Negative parenting more predictive than positive parenting
 * Conflict frequency and behavior

### 2022-11-02 action_letter.Rmd
update 2022-11-02 15:34:25

---
title       : "Using machine learning to identify early predictors of adolescent emotion regulation development"
authors     : "masked"
journal     : "Journal of Research on Adolescence"
editor      : "Raval"
manuscript  : "2022-182"
class       : "draft"
bibliography: ["manuscript/veni.bib"]
output      : papaja::revision_letter_pdf
---

```{r}
library(revise)
man <- revise::read_manuscript("manuscript/manuscript.Rmd")
```

Dear Dr. `r rmarkdown::metadata$editor`,

Thank you for considering my manuscript for publication in _`r rmarkdown::metadata$journal`_.
I appreciate the thorough and critical comments you solicited,
which helped strengthen the argumentation, clarify the structure, and add nuance.
I have made a good-faith effort to address all comments as thoroughly as possible.
All changes are explained below, and the difference between the current version and previous one can be seen at <https://github.com/cjvanlissa/rigorous_exploration/compare/d8ba218..48d2b17>.
I hope that you will find the quality of the work to be sufficiently improved.

Yours sincerely,
Author

<!-- ` get_revision(man, "test")` -->

# Editor {-}

\RC{In the introduction section, please provide a comprehensive review of the literature. As reviewer 3 points out, there is existing literature investigating risk factors associated with emotion dysregulation. This research needs to be cited.}

\RC{Also, please specify the demographics of the samples of cited studies. A bulk of the literature in our field, including related to emotion dysregulation, is based on samples from Western countries (e.g., Australia, those in North America, and those in Europe). If the studies you are citing are from these world regions, please specify that.}

\RC{Also, you describe two major theories of development including the bioecological theory (Bronfenbrenner \& Morris, 2007). Note that there is a revision to Bronfenbrenner’s model that places culture as existing in different settings and not only as a distal influence as a macrosystem (see Velez-Agosto et al., 2017). According to this revision, cultural processes are not distal influences, instead cultural and individual processes interact in a transactional fashion and this revised model is actually much more aligned with the Transactional model (Sameroff, 2010) that you reference.}

Note that a recent perspective on the bioecological theory emphasizes that distal factors like culture are often enacted proximally by close others.
This claim might help explain why we found that distal factors were relatively less important predictors;
if distal effects are largely mediated by proximal actors,
their importance might be attenuated by conditioning on proximal effects.

\RC{In the method section, it would be helpful to include when (months and year) and where the data are collected (geographic location/country). Given cultural differences in emotion regulation, the context of the sample is really critical to include.}

We respectfully believe that this may have been an oversight,
as this information was already provided. As per the submission guidelines, some specific details had been masked for review, which may have caused the oversight. To address this comment, we have unmasked these details in the revision:

Timing of data collection (we do not have information about specific months):

`r get_revision(man, "waves")`

Geographic location:

`r get_revision(man, "dutch")`

\RC{Also, as reviewer 1 suggests, please explain the methodology used to extract predictors in more detail.}

As requested, we have addressed this comment by providing more detail and an illustrative example, please refer to our response to Reviewer 1.

\RC{In the results and discussion, as reviewer 2 and 3 highlight, please be careful about interpreting your findings. For example, to refer to parenting as unimportant based on your findings seems inaccurate given that balanced relatedness was the 3rd ranked predictor, so rather than say that parenting was unimportant or underrepresented, state that some dimensions of parenting were important, and name those dimensions.}

\RC{In the discussion, I encourage you to think about predictors from the revised bioecological theory perspective (Velez-Agosto et al., 2017) that does not conceptualize individual level, micro- and macro as influences that move along the proximal to distal continuum and conceptualizes them in a transactional framework. Factors such as religiosity and social class are reflected in relationships with parents and peers and thus to state that they are less important because they are ranked lower may not capture the complexity of their relevance.}

\RC{Also, as you mention in the limitations, it is possible that SES was ranked low because of the restricted range of SES that was represented in your sample.}

If we understand correctly, this statement requires no further action.

\RC{In the discussion, it is stated, “Predictors at the individual level were most important, followed by those at the microsystem level, such as aspects of the relationship with parents, followed by the mesosystem, including aspects of the peer network (e.g., bullying/victimization and delinquency), school performance, and socio-economic status, and the macrosystem indicator religiosity was among the least important predictors.”  This statement does not seem to fit the findings. If you look at the top 10 predictors listed in table 2, they include individual characteristics (big 5) and relational aspects (balanced relatedness, conflict). Please revise. Also, there are frequent references to individual level predictors being the most important and again looking at the top 10 predictors in table 2, there are both individual characteristics and relational dimensions so please revise the language.}

While we are willing to adjust the language as requested,
we do disagree that the statement does not fit the findings.
First of all, note that the top 10 is an arbitrary criterion;
placing the 'cutoff' for important predictors in different places lends more versus less support to the statement.

Variable importance follows an approximately exponential distribution,
which means that the first variables are more important than later ones.
At least the first two most important variables are both individual characteristics.
If we interpret the variable importance like a scree plot (which is also an exponential distribution), then we observe a drop after the first 6 variables, and another drop after the 12th.
Of the first 6, only one relates to the adolescent-reported *relationship with* parents, not *parenting behaviors*. The rest of these top 6 are all individual differences.
Of variables 7-12, two relate to characteristics of the parents - mother's age and interparental conflict. Again, not parenting behavior. The rest of variables in this group relates to individual differences and adolescents' conflict behaviors.
Variables 16 and 25 are the first parenting behaviors.
I think that's a surprisingly low score.

REF

# Reviewer: 1

\RC{It was a delight to read this manuscript that applied multiple novel machine learning methods to identify the most important early predictors of adolescent emotion regulation development. I found this manuscript to be one of the most innovative, potentially revolutionary manuscripts I have read in a long time. I absolutely think it should be published eventually. I have several suggestions and clarifying questions, provided by section below, that, if addressed, would further strengthen the manuscript. Specifically, I think addressing many of my suggestions would allow the manuscript to be even more translatable and understandable by traditional hypothesis-testing longitudinal researchers like me.

However, again, I want to emphasize that overall I found this manuscript to be absolutely enthralling, and I look forward to seeing it published. Bravo to the authors!}

We thank the Reviewer sincerely for these kind words, and for their constructive feedback. We have attempted to address all comments to the best of our ability.

\RC{It looks like a reference needs to be added to this line: “linear models limit the total number of predictors that can be included due to concerns of power, overfitting, and multicolinearity [REF].”}

We thank the Reviewer for pointing this out. We have inserted the correct reference.

\RC{When the authors state that “The RMSEA and SRMR of the invariant model indicated acceptable fit..”; I am assuming they are talking about the metric invariant model correct? It might be important to add the word “metric” just to specify that.}

We thank the Reviewer for pointing this out. We have clarified this, as requested.

\RC{This is not something that needs to be answered in the text, but what do the authors make of the relatively poor CFI model fit, compared to RMSEA and SRMR. Any idea what accounts for that? I think the authors make a convincing enough argument as is for retaining the chosen model despite the low CFI test statistic, but I was just wondering how they interpreted that divergence in omnibus fit statistics.}

One potential reason is that the performance of these fit statistics has been validated in the context of confirmatory factor analysis, and their performance is less well understood for complex models.
This may affect the CFI more than, e.g., RMSEA, because CFI compares the fit of the theoretical model to a null model that may be inappropriate for longitudinal data.
For this reason, it has been suggested to rely more on RMSEA for confirmatory studies (in this case, the growth model part of the analyses is confirmatory), and on CFI for exploratory purposes [@rigdonCFIRMSEAComparison1996].
In line with the above, other research, which we now cite in the revision, has also found the RMSEA to be better at detecting misspecifications of growth curve models than CFI [@leiteDetectingGrowthShape2011]:

`r get_revision(man, "cfi")`

\RC{The Link to Supplementary Table S1 was broken; I couldn’t access it.}

We apologize; the Repository was still set to private at time of submission. This was intended to guarantee anonymous review. For this revision, it has been made public. The links should now work.

\RC{I have a couple of items about the approach to missing data used (that, if there is not room in the text, can certainly be provided in a supplement). It mentions that this method is valid when missingness is either random or contingent on observed data. Can you present some evidence that missingness was MAR or MCAR here?}

We respectfully believe that this comment might be based on a misunderstanding. There are three types of missingness: MNAR, where missingness depends on unobserved data; MCAR, where missingness depends on observed data; and MAR, where missingness is truly random. It is not possible to test for MNAR, and the present methods are valid for both MCAR and MAR - so there is no need to provide evidence for either of those.

Also note the unfeasibility of performing an MCAR test when the number of variables is higher than the number of participants: The raw data, prior to imputation and scale construction, is 495 (participants) x 750 (variables) big. I am not aware of any MCAR test that works with N < p data.

\RC{Additionally, and relatedly, were there any systematic differences between those with missing and complete data? That would also be good for the reader to know.}

To address this comment, and a comment by another Reviewer, we tested for differences on demographic characteristics and the dependent variable between participants with missing and complete data. We now mention this in the revised manuscript:

`r get_revision(man, "missing")`

\RC{I think that the methodology used to extract predictors here is ground-breaking and pretty revolutionary. However, that also means that the procedure might need to be explained in a little more depth for “old methodologists” like me. I actually think because this procedure is so innovative in the field of psychology, I would humbly be in favor of expanding the page limits of the manuscript to provide clarification on some of these questions in further detail, only because as a “tried and true” frequentist longitudinal modeler who is beginning to dabble in machine learning, I (and I suspect many like me) would love to learn even more about this procedure. Here are a few questions I had about it.
o The authors note that for multi-item scales, they used principal components analysis for dimension reduction and took the first component for further analysis. Could, perhaps, the authors provide a specific example to make this process even more “concrete” in the minds of readers. Perhaps the authors could provide an example of how this was done on a single scale, what items were included as part of the first component that represents large amounts of shared variance, etc.}

We appreciate the Reviewer's kind words, but don't give us too much credit - PCA is the oldest dimension reduction technique (invented in 1901), and the main reason we use it here is because it makes fewer assumptions about the measurement model than, e.g., CFA.

We do appreciate the excellent suggestion of providing a specific example to help readers understand this approach. To address this comment, we have thus made the following changes:

1. We added the requested illustrative example, using neuroticism. We could use any scale, but as Neuroticism later turns out to be the most important predictor, it seems a suitable candidate.
2. We have attempted to clarify our description of the dimension reduction procedure
3. We have added further statistics that, hopefully, give the reader a better sense of whether the use of these principal components is defensible. Specifically, we no longer just provide the explained variance by each first component - but we also standardize this explained variance by the number of items in the scale to see how much more variance the first component explains than expected by chance. We see that, across all scales, the first component explains between 1.6 - 17.46 times more variance than expected by chance.

We hope this addresses the comment.

\RC{I see that in the SEM-forest analysis, homoscedastic residuals over time were assumed. I am guessing the reason for doing so is because estimating heteroscedastic residuals over time just makes the model hopelessly complex and unestimable. So my question here is two-fold: First, is that true? Second, can the homoscedasticity versus heteroscedasticity of residuals be tested in similar ways as it can in traditional growth curve modeling here? I don’t at all mind if homoscedastic residuals need to be assumed to ensure the model runs (I think that is the implication from the limitation mentioned in the discussion).}

The Reviewer is correct; homoscedastic residuals were assumed because otherwise the model would be "hopelessly complex". We would have more parameters for the residuals than for the rest of the model, and the SEM-forest analysis would thus be primarily driven to explain differences in residuals, instead of in growth curves.
Unfortunately there is no way to test this assumption. The SEM-forest method estimates tens, or hundreds of thousands of SEM models in its different nodes. It is not possible to test model assumptions in each of those nodes.

We have attempted to address this comment as follows: By explaining our decision to assume homoscedasticity:

`r get_revision(man, "homoscedastic")`

And by addressing it further in the Limitations section:

`r get_revision(man, "homogeneity")`

\RC{Could you speak a little more (maybe just a sentence or so) on the importance of step 2 substep iii. In the SEM forest algorithm? It was hard for me to follow why identifying the value of preditor k that maximizes the likelihood ratio of a post-split two-group model indicates best model fit. I suspect just another sentence or so of clarification is needed.}

We have simply rewritten this step in an attempt to address this comment. It now reads:

> Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).

\RC{Could you speak more to the importance of why steps 3 and 4 need to be repeated until all post-split groups reach a minimum size? What would that minimum size be and why is it important?}

To address this comment, we have rewritten this step as follows:

> Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then fully grown.

Additionally, we have rewritten the introduction to the SEM-forests algorithm to provide more context, thereby hopefully making it easier for readers to grasp:

> The SEM-forests algorithm is based on the random forests algorithm, which in turn is based on regression trees [see @stroblIntroductionRecursivePartitioning2009].
A single regression tree splits the sample repeatedly in two by whichever predictor $p$ makes both post-split groups maximally internally homogenous, and maximally different from one another.
A regression forest consists of many regression trees grown on bootstrap samples of the data. To ensure that each tree learns something different from the data, it may only consider a random subsample of $K$ candidates out of all $p$ predictors at each split.
With this in mind, the SEM-forest algorithm operates as follows... [etc]

\RC{How are predictions aggregated from all SEM-trees? If you could offer even just a sentence or two on that process in a way a “traditional” “applied” longitudinal researcher (like me) would understand, I think such readers would be very grateful.}

We have added the following explanation:

> Given an individual's scores on the predictor variables $p$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters). Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.

\RC{Could you provide a quick applied example to accompany this statement “This approach intrinsically accommodates non-linear effects as subsequent splits on the same predictor, and complex interactions as subsequent splits on interacting variables. This offers greater flexibility than linear parametric models.”. Again, I think providing an example would allow non-machine learning researchers to follow what you are doing here even more clearly.}

We have added the following example:

> As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
In regression analysis, this could be represented with a linear and quadratic term, and a plot might show that the effect of autonomy support on the intercept increases to a value $x_{knee}$, and then tapers off.
A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.

\RC{I think the results clearly presented a complex set of findings. I really appreciated the grouping by theme.
• I have seen many ML applications split data sets into “train” and “test” sub-samples, to validate the findings in the “train” data set in the “test” data set. As far as I can tell, that was not done here. Did the authors consider this approach? I feel like having “test” and “train” sub-samples wherein results are replicated would provide further robustness and validity to study findings. I also understand if that procedure is not possible with the SEM forests approach.}

Thank you for bringing up this point. This is indeed common practice in "standard" machine learning, but unfortunately not possible for SEM-forests because there is (currently?) no error metric that could be used to describe fit in a "test" sample. For a standard regression tree, one can calculate mean squared error, or even $R^2$ in the training data. For SEM-trees, it's not clear what one would calculate.

We had mentioned this in the Limitations section, but evidently that statement was not sufficiently clear. To address this comment, we have rewritten that section as follows:

> Other limitations related to SEM-forests include the fact that,
unlike in traditional random forests, there is no error metric that could be used to calculate fit in a hold-out testing sample.
By extension, at the time of writing, there are not yet methods for cross-validation, model tuning, or evaluating the overall fit of a SEM-forest, as these too require an error metric.

\RC{I know ensuring the article remains inside page limits is always a battle, but I had a suggestion for one additional, hopefully simplistic analysis that might even further bridge the “gap” between these machine learning analyses and “traditional, frequentist” analyses. Perhaps the authors could take only the most important predictors, say the top 10 (or really whatever number they want), and run regression analyses that predict trajectory intercepts and slopes? Basically, I know that people like me really like explanations like “oh yeah, these top 10 machine learning predictors predict XX% of variance in emotion regulation scores over time”, and so I was trying to think of a way to generate those sorts of tangible, “traditional” effect sizes. Could the authors potentially provide such effect size estimates? I also understand if there simply isn’t time or space for this, but figured it was worth asking.}

Thank you for this suggestion; we agree that many people would be interested in such an overview. One major caution is that such post-hoc analyses should not be over-interpreted; they are in no way a substitute for the (much more flexible) machine learning models. Still, if people e.g. want to take guidance from our list of important predictors, having an overview of univariate linear effects might be useful to help plan follow-up studies.

To address this comment, we have added the following paragraph and an additional Supplemental Table:

> A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
Regressing the parameters (intercept, slope and quadratic slope) of a latent growth model on each predictor in turn.
Such analyses might be helpful to plan hypotheses and perform power analysis for follow-up research within the SEM framework.
However, we strongly caution against over-interpretation.
Unlike the machine learning analyses, these analyses do not accommodate non-linear effects or interactions, nor are effects controlled for other predictors.
The results are merely descriptive and should not be used to perform inference.
[Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
All models had good fit ($RMSEA < .07$, $CFI > .99$, $SRMR < .04$).
Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
This indicates that our machine learning analyses indeed recovered relevant patterns in the data.

\RC{I appreciated the brave and nuanced discussion of how these analyses conform to, and also challenge, traditionally held norms in the field of emotion regulation. I especially appreciated the identification that parenting, substance use, and bullying might not be the risk factors we think they are.
• The ERRATA tool seems amazing; I couldn’t find it online. Could a link be provided directly in the text?}

To address this comment, we have added the following information:

> The source code is available at <https://github.com/cjvanlissa/shiny_errata>;
users can run the app from a local R instance,
or view an online demo using a link in the README file of this repository (as long hosting is available).

\RC{I love the notion of “good enough” parenting environment and preconditions, and its’ application here; I thought the link between non-linear effects and “good enough” parenting environments was such an incredible example of translating a statistical insight into “on the ground” wisdom that can be applied by clinicians.}

Thank you for these kind words!

\RC{The only part of the results that were left relatively untouched by the Discussion was the relatively large importance of Big 5 personality traits as predictors of adolescent emotion regulation. Could the authors comment on what they thought about those findings? I am especially interested to hear the authors thoughts about those results, because I feel like investigation of personality traits is so often “siloed” away from investigation of parenting, substance use, and other phenomena when it comes to adolescent emotion regulation. So I think it would be especially important to also highlight those findings in the Discussion.}

REF

\RC{I think there are some broken links to supplemental materials in the manuscript.}

This is correct; at the time of submission, the repository on GitHub was still set to private. For the revision, this has been resolved.

\RC{I think the authors should just re-read the manuscript again before final publication; there were missing words and sentences throughout in some places.}

We have done as suggested and corrected any errors found.

\RC{Pages 43 to 65 of the manuscript materials appeared cut-off and unreadable in many areas.}

We apologize. The manuscript only has 42 pages; pages 43-65 are copies of the online supplemental materials. The submission portal has trouble correctly presenting these materials, but they will not be part of the paper.

# Reviewer: 2

\RC{I reviewed this manuscript from the perspective of a clinical psychologist with expertise in quantitative methods and adult psychopathology. I have years of experience with and expertise in applied predictive modeling and supervised machine learning in general but am not that familiar with the SEM forest algorithm, which seems to blend supervised and unsupervised learning. Thus, I defer to other reviewers to evaluate the substantive aspects of the manuscript related to adolescent emotion regulation development and the highly technical aspects of the manuscript specific to the SEM forest algorithm.

Overall, I found the manuscript to be ambitious and interesting. The size and longitudinal nature of the dataset is impressive, even if it the number of participants is a bit lower than one might ideally want to support complex machine learning. I also valued the manuscript’s integration of theory and methods, transparency of the methods, the addition of variable importance functions to the semTrees package, and the creation of the Shiny app as a demo/first step towards implementation. My concerns are mainly around the lack of external cross-validation (I realize that the out-of-bag approach is like an internal cross-validation) and hyperparameter tuning (which may be related to differences between the supervised approach I am used to and the hybrid approach of SEM forest) as well as some apparent discrepancies between the results as I understand them and the conclusions in the manuscript.}

\RC{1. I think that the number of predictors used should be tuned during cross-validation rather than set to 9 or, at least, multiple values should be compared. Similarly, I think that the complexity parameter should be tuned or compared. Please comment on why this wasn’t done.}

\RC{2. On page 15, there seem to be semantic themes or groupings of predictors. In addition to textual interpretation, I think it would be helpful to create a visualization (even if it were supplemental) that grouped the variable importance scores per theme. Perhaps each theme could be represented by a boxplot or jittered points. This would help communicate the relative importance of each theme (as the text is currently doing informally). A similar approach could be employed for predictors assigned to different levels of the bioecological model.}

\RC{3. The manuscript’s conclusion that parenting behaviors (especially positive ones) are low in importance (e.g., page 18) is hard for me to reconcile with the third-most important predictor being a positive parenting behavior. Overall, there seems to be an interpretive bias against the parenting behaviors being important. Or perhaps I am misunderstanding something?}

\RC{4. On page 12, it seems to me that the interpretation of SEM-forests answering, “why do some adolescents have similar growth trajectories” has some causal connotations that may or may not be intended. Please speak to this directly in the manuscript. Can SEM-forests enable causal inferences, or would it be more accurate to rephrase this sentence to remove the connotation?}

\RC{5. My understanding is that the importance of specific interactions can be interrogated using an generalization of partial dependence profiles (Apley \& Zhu, 2020). I’m not saying that this necessarily has to be added to the current paper, but it might be worth citing this approach in the “next steps” part of the manuscript and considering it in future work.}

In principle, this is possible.
In practice, it is not, for the following reason:
1) Computing univariate partial dependence profiles is already so computationally expensive that I cannot repeat it without additional funding
2) There is no way of determining what bivariate interactions might be important in SEM-forests analysis (because there is no applicable analogy for Friedman's H statistic)
3) So one would have to examine all bivariate interactions by hand, for all model parameters. This comes down to interpreting 84*84*3 = 21168 plots.

\RC{6. That said, my understanding is that partial dependence profiles can be somewhat problematic when applied to correlated predictors (as these predictors surely are) and that a better approach in such cases would be accumulated local profiles (Apley, 2018).}

Again, in principle this is correct, but in practice, no method exists to obtain accumulated local effects for SEM-forests. For the purpose of this paper, I developed a novel method to obtain marginal dependencies for growth curves, but I am not familiar with Apley's method so I cannot implement it (and either way, that seems out of scope for a developmental paper).

\RC{1. On page 10, please clarify whether the CFAs were single factor models.}

\RC{2. On page 10, please provide more explanation of the discrepancy between absolute and comparative fit of the metric invariant model.}

\RC{3. In figure 1, clarify whether the y-axis is reversed. I ask because it seems strange to me that someone with lower difficulties with emotion regulation would be “high risk” and Figure 1 seems to show the opposite pattern as Figure 7.}

\RC{4. On page 12, please speak to the reasonableness of the assumption of homoscedastic residuals over time.}

\RC{5. On page 14, the reference to Figure 1 should be to Figure 2.}

\RC{6. I think Figures 5 and 6 could be moved to supplemental if space is tight.}

\RC{7. On page 18, is it true that “out [of] all parent-related constructs, constructive conflict resolution with fathers ranked highest and, and out of all parenting behaviors, paternal power ranked highest”? This seems to only be true if you look at only the mother/father-specific parenting measures and omit the general ones like balanced relatedness and conflict interparental.}

\RC{8. The supplemental table 1 didn’t render well in the PDF.}

# Reviewer: 3

\RC{Comments to the Author
The current study used machine learning to identify importance of predictors of adolescent emotion regulation trajectories. The study used a large sample of adolescents and included multiple reporters (although adolescent report was featured most heavily). Despite these strengths, I have a number of concerns about the manuscript in its present form that make its contribution unclear at this point. I list these concerns in roughly the order of appearance in the manuscript.}

\RC{The abstract could be strengthened by providing theoretical rationale for the study. The abstract also seems to exceed the journal’s word limit, as several sentences are cut off on the version included in the pdf cover page. In describing the results for parenting behaviors, it seems that autonomy support and invasive parenting were important if I am reading the statement correctly, so please remove these from the “less important” statement in order not to undermine the importance of parenting.}

\RC{Please provide citations for the statement in the first paragraph on page 4, “These difficulties in emotion regulation can cause…” and provide examples of why these incur societal cost.}

\RC{I would suggest removing reference to “destinies” on page 4 and elsewhere, as this word connotes fixedness in developmental outcomes, which does not reflect contemporary developmental views.}

\RC{The introduction would benefit from proofreading for parallel structure (last sentence in first paragraph on page 4), fragments (e.g., “For example, because they are not…”), and other typos (e.g., “an important challenges” on page 8). These are just several examples. The entire manuscript requires proofreading. The Discussion has sentences that are fragments or are missing words.}

\RC{At the bottom of page 5, it is noted that the next step in this area of research is to identify risk factors associated with emotion dysregulation. There is a wealth of research already identifying various risk factors. It should be acknowledged that this work exists (and what these identified factors are – some of which are reviewed beginning at the bottom of page 7) before noting the problems with existing research and theory. I would suggest expanding beyond the small number of labs currently heavily relied upon for background on emotion regulation. It seems that the actual next step for this literature is what is defined at the top of the first full paragraph on page 8 – “What is still missing is an understanding of the relative utility…” In short, the introduction could better acknowledge what we already know and more precisely describe the contribution of this study. I expect this would require a fairly extensive revision to move forward.}

We appreciate this comment. There are two issues here, which we will both address in turn.

The first issue is that the offending paragraph evidently failed to convey that we set out to identify predictors of adolescents' diverging developmental trajectories using *person-centered* methods.
This is certainly different from preceding work, which has nearly exclusively looked at whole sample average effects.
This paragraph thus doesn't argue that "few have examined predictors of emotion regulation in adolescence", as the Reviewer interpreted it, but instead, argues that "few have used person-centered methods to examined predictors of adolescents' diverging developmental trajectories of emotion regulation."
In the Revision, we have attempted to clarify this by more precisely describing the contribution of the present study, in several places:

`r get_revision(man, "personcentered1")`

The second issue is whether our writing does sufficient justice to prior publications on predictors of emotion regulation.
We could not agree more with the Reviewer that there is a wealth of prior research in this area.
In fact, there is so much work on this topic that it would be a tall order to do it all justice.
We cite a recent (2021) comprehensive systematic review of 6305 studies on predictors of adolescent emotion regulation.
From those 6305 studies, we selected the ones that seemed most relevant for the present paper.
If the Reviewer feels that *specific* studies are missing here, we are happy to include them of course.
At the same time, we cannot strive to recreate a comprehensive review here. This also seems redundant, given that one was just recently published, and overly ambitious, given the page limits and attention that must be devoted to explaining the novel methods.
The present study builds upon prior reviews of the literature,
and those reviews are cited.

To address this comment, we now emphasize more strongly that there is an abundance of prior literature on predictors of adolescent emotion regulation, e.g.:

`r get_revision(man, "abundantlit")`

\RC{Please insert the reference listed as [REF] on page 8.}

We have inserted the correct reference.

\RC{Please provide more of an introduction to SEM-forests prior to The Present Study section. The manuscript appears to assume much more knowledge about machine learning than many readers have.}

As requested, we now provide a more extensive introduction to SEM-forests prior to The Present Study. This section reads:

`r get_revision(man, "semforests")`

\RC{Please provide more information about the participants (mean and range of age of adolescents and parents at time of recruitment, race/ethnicity, income, and parental education). This will help contextualize the findings related to demographic characteristics that were predictive of emotion regulation trajectories.}

We respectfully believe that this may be an oversight on the Reviewer's part, as comprehensive descriptive statistics - including those requested by the reviewer - are reported in Tables 1 and 2.

\RC{How far apart were “waves” of the study?}

In the original manuscript, we stated that:

>  From 2006 to 2011, trained interviewers
conducted annual home visits to collect questionnaire data. All independent variables were
assessed in the first measurement wave.

To resolve any remaining ambiguity, in the revision, we rephrased this as follows:

`r get_revision(man, "waves")`

\RC{Did missingeness relate to demographic or emotion regulation variables in the dataset?}

As two Reviewers have requested this information, we understand that this is information readers will want to know. Note, however, that the methods we use are valid regardless of whether missingness is related to other variables in the dataset or not. To address this comment, we performed the suggested check, and report:

`r get_revision(man, "missing")`

\RC{Was the statement, “Class solutions with more than two classes were excluded…” an a priori decision, or a statement of what resulted from the attempt to estimate them? Please be clear in the manuscript.}

We thank the Reviewer for pointing out this unfortunate choice of words. 
Entropy is not a criterion for class enumeration, but for evaluating the quality of the chosen solution (see Jung \& Wickrama, 2008). In the Revision, we have attempted to clarify this as follows:

`r get_revision(man, "entropy")`

\RC{I was curious why the median and plus/minus 1 SD were used for “s” for continuous variables. I would expect the median to be accompanied by percentiles (e.g., 15th and 85th percentiles) or +/- 1 SD to go along with the mean.}

In line with the Reviewer's expectation, we indeed used the median and 16th and 84th percentile. Percentiles do not assume a normal distribution of the predictor, and neither does the model, so this is a good fit. As we explain in the text, the 16th and 84th percentiles correspond to +/- 1SD of a standard normal distribution. We then use this label, "+/- 1SD", in the plots, in an attempt to make the results more understandable. Perhaps this was a misguided decision, but we were not sure whether the readership would be familiar with percentiles.

In the Revision, we have attempted to resolve this issue by stating:

`r get_revision(man, "percentiles")`

\RC{Please correct the finding for sex – it currently reads, “boys girls reported better emotion regulation.”}

Thank you for pointing out this oversight. We have corrected it, along with other typesetting and grammatical mistakes.

\RC{In the first paragraph of the Discussion, where most important predictors are listed, please list parenting behaviors that ranked as highly as aspects currently listed. Throughout the paper, parenting is listed as unimportant, when there are important exceptions that should be more explicit, rather than parenthetical within a negative statement.}

REF

\RC{I do not believe it is accurate to state, “The fact that individual-level predictors were most important suggests that individual assessments might be sufficient.” Just because a factor is not the most important does not mean it is not playing a significant role. Perhaps the most important factors should be prioritized, but it would be contrary to much of developmental theory and many clinical interventions to suggest that it would be best to consider the adolescent as separate from their context. If anything, developmentalists and clinicians are moving towards greater contextualization of individuals and their identities.}

REF

\RC{It is rightfully noted that this machine learning application was performed within one sample within one nation. There is limited reason to believe it would be universally applicable, especially because it seemed to draw on variables that happened to be available in a dataset, rather than designing the study for this purpose. The paper could provide more information about the cultural values that may impact emotion regulation in this sample to provide readers more information with which to evaluate generalizability to other populations.}

\RC{It is not clear how the information on pages 43-65 are to be used. This is not a labeled or formatted table.}

We understand that pages 43-65 are not very useful in their current form, and apologize for the inconvenience. We were required to attach hard copies of the online supplemental tables during submission, and these were evidently not correctly formatted by the submission portal. They are just Excel spreadsheets, and will not be part of the manuscript.

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::

### 2022-11-02 manuscript/action_letter.Rmd
update 2022-11-02 15:34:25

---
title       : "Using machine learning to identify early predictors of adolescent emotion regulation development"
authors     : "masked"
journal     : "Journal of Research on Adolescence"
editor      : "Raval"
manuscript  : "2022-182"
class       : "draft"
bibliography: ["manuscript/veni.bib"]
output      : papaja::revision_letter_pdf
---

```{r}
library(revise)
```

Dear Dr. `r rmarkdown::metadata$editor`,

Thank you for considering my manuscript for publication in _`r rmarkdown::metadata$journal`_.
I appreciate the thorough and critical comments you solicited,
which helped strengthen the argumentation, clarify the structure, and add nuance.
I have made a good-faith effort to address all comments as thoroughly as possible.
All changes are explained below, and the difference between the current version and previous one can be seen at <https://github.com/cjvanlissa/rigorous_exploration/compare/d8ba218..48d2b17>.
I hope that you will find the quality of the work to be sufficiently improved.

Yours sincerely,
Author

`r get_revision("manuscript.Rmd", "test")`

# Editor {-}

# Reviewer \#1

Dear Authors:

Thank you for submitting your manuscript to the Journal of Research on Adolescence.  I have received feedback from external reviewers who are experts in this area of research.  The detailed comments from these reviewers are included at the bottom of this letter.

Overall, the reviewers indicate, and I agree, that there are several strengths of your manuscript including longitudinal design, a reasonably sized data set, multiple reporters, and the use of machine learning to identify importance of a large number of predictors of adolescent emotion regulation trajectories.

However, the reviewers also indicated a number of substantial limitations.  These limitations preclude publication of the manuscript in its current form, but are potentially addressable through revision.  I will not try to summarize all of the reviewers’ comments here, but I trust that you will consider the reviewers’ comments in detail if you decide to revise and resubmit this manuscript.  Below are some of the key critiques raised by the reviewers, along with some further suggestions I am offering for revising the manuscript.

In the introduction section, please provide a comprehensive review of the literature. As reviewer 3 points out, there is existing literature investigating risk factors associated with emotion dysregulation. This research needs to be cited. Also, please specify the demographics of the samples of cited studies. A bulk of the literature in our field, including related to emotion dysregulation, is based on samples from Western countries (e.g., Australia, those in North America, and those in Europe). If the studies you are citing are from these world regions, please specify that.

Also, you describe two major theories of development including the bioecological theory (Bronfenbrenner & Morris, 2007). Note that there is a revision to Bronfenbrenner’s model that places culture as existing in different settings and not only as a distal influence as a macrosystem (see Velez-Agosto et al., 2017). According to this revision, cultural processes are not distal influences, instead cultural and individual processes interact in a transactional fashion and this revised model is actually much more aligned with the Transactional model (Sameroff, 2010) that you reference.

In the method section, it would be helpful to include when (months and year) and where the data are collected (geographic location/country). Given cultural differences in emotion regulation, the context of the sample is really critical to include. Also, as reviewer 1 suggests, please explain the methodology used to extract predictors in more detail.
In the results and discussion, as reviewer 2 and 3 highlight, please be careful about interpreting your findings. For example, to refer to parenting as unimportant based on your findings seems inaccurate given that balanced relatedness was the 3rd ranked predictor, so rather than say that parenting was unimportant or underrepresented, state that some dimensions of parenting were important, and name those dimensions.

In the discussion, I encourage you to think about predictors from the revised bioecological theory perspective (Velez-Agosto et al., 2017) that does not conceptualize individual level, micro- and macro as influences that move along the proximal to distal continuum and conceptualizes them in a transactional framework. Factors such as religiosity and social class are reflected in relationships with parents and peers and thus to state that they are less important because they are ranked lower may not capture the complexity of their relevance. Also, as you mention in the limitations, it is possible that SES was ranked low because of the restricted range of SES that was represented in your sample.

In the discussion, it is stated, “Predictors at the individual level were most important, followed by those at the microsystem level, such as aspects of the relationship with parents, followed by the mesosystem, including aspects of the peer network (e.g., bullying/victimization and delinquency), school performance, and socio-economic status, and the macrosystem indicator religiosity was among the least important predictors.”  This statement does not seem to fit the findings. If you look at the top 10 predictors listed in table 2, they include individual characteristics (big 5) and relational aspects (balanced relatedness, conflict). Please revise. Also, there are frequent references to individual level predictors being the most important and again looking at the top 10 predictors in table 2, there are both individual characteristics and relational dimensions so please revise the language.

Given these limitations, I cannot accept the current manuscript for publication in the Journal of Research on Adolescence.  However, I am willing to consider a revised resubmission if you believe that you can successfully address the reviewers’ concerns.  I must point out that revising a manuscript in no way guarantees publication.  The journal currently has a rejection rate of about 85%, and even revised manuscripts might be rejected.

If you decide to revise and resubmit your manuscript for further consideration, please resubmit your manuscript by 14-Nov-2022. If you are not able to submit your revision within this time frame, you may request an extension. Please contact the editorial office with a reasonable expectation of when your revision would be ready. If you use an electronic calendar (e.g., Outlook, iCal), you can open the email attachment to put a reminder of your due date into your calendar.

To submit your revision, log into https://mc.manuscriptcentral.com/jra and enter your Author Center. You will find your manuscript title listed under "Manuscripts with Decisions."  Under "Actions," click on "Create a Revision."  Your manuscript number has been appended to denote a revision.  Please DO NOT upload your revised manuscripts as a new submission.

My plan is to send this manuscript back to external reviewers before reaching a decision.  Please provide a point-by-point summary of your responses to reviewers’ comments in the “Response to Decision Letter” field.

Wiley Editing Services Available to All Authors
Should you be interested, Wiley Editing Services offers expert help with manuscript, language, and format editing, along with other article preparation services. You can learn more about this service option at www.wileyauthors.com/eeo/preparation. You can also check out Wiley’s collection of free article preparation resources for general guidance about writing and preparing your manuscript at www.wileyauthors.com/eeo/prepresources.

Thank you for submitting to the Journal of Research on Adolescence.  I look forward to reading a revised version of your manuscript.

Sincerely,

Dr. Vaishali Raval
Associate Editor, Journal of Research on Adolescence

Reviewer(s)' Comments to Author:

Reviewer: 1

Comments to the Author
It was a delight to read this manuscript that applied multiple novel machine learning methods to identify the most important early predictors of adolescent emotion regulation development. I found this manuscript to be one of the most innovative, potentially revolutionary manuscripts I have read in a long time. I absolutely think it should be published eventually. I have several suggestions and clarifying questions, provided by section below, that, if addressed, would further strengthen the manuscript. Specifically, I think addressing many of my suggestions would allow the manuscript to be even more translatable and understandable by traditional hypothesis-testing longitudinal researchers like me.

However, again, I want to emphasize that overall I found this manuscript to be absolutely enthralling, and I look forward to seeing it published. Bravo to the authors!

Introduction:
• This was a tour-de-force of an introduction in my opinion. It was well thought out, well-written, and presented compelling arguments that made me enthralled to read the rest of the paper. The authors did fantastic work here.
• It looks like a reference needs to be added to this line: “linear models limit the total number of predictors that can be included due to concerns of power, overfitting, and multicolinearity [REF].”

Method:
• When the authors state that “The RMSEA and SRMR of the invariant model indicated acceptable fit..”; I am assuming they are talking about the metric invariant model correct? It might be important to add the word “metric” just to specify that.
• This is not something that needs to be answered in the text, but what do the authors make of the relatively poor CFI model fit, compared to RMSEA and SRMR. Any idea what accounts for that? I think the authors make a convincing enough argument as is for retaining the chosen model despite the low CFI test statistic, but I was just wondering how they interpreted that divergence in omnibus fit statistics.
• The Link to Supplementary Table S1 was broken; I couldn’t access it.
• I have a couple of items about the approach to missing data used (that, if there is not room in the text, can certainly be provided in a supplement)
o It mentions that this method is valid when missingness is either random or contingent on observed data. Can you present some evidence that missingness was MAR or MCAR here?
o Additionally, and relatedly, were there any systematic differences between those with missing and complete data? That would also be good for the reader to know.
• I think that the methodology used to extract predictors here is ground-breaking and pretty revolutionary. However, that also means that the procedure might need to be explained in a little more depth for “old methodologists” like me. I actually think because this procedure is so innovative in the field of psychology, I would humbly be in favor of expanding the page limits of the manuscript to provide clarification on some of these questions in further detail, only because as a “tried and true” frequentist longitudinal modeler who is beginning to dabble in machine learning, I (and I suspect many like me) would love to learn even more about this procedure. Here are a few questions I had about it.
o The authors note that for multi-item scales, they used principal components analysis for dimension reduction and took the first component for further analysis. Could, perhaps, the authors provide a specific example to make this process even more “concrete” in the minds of readers. Perhaps the authors could provide an example of how this was done on a single scale, what items were included as part of the first component that represents large amounts of shared variance, etc.
o I see that in the SEM-forest analysis, homoscedastic residuals over time were assumed. I am guessing the reason for doing so is because estimating heteroscedastic residuals over time just makes the model hopelessly complex and unestimable. So my question here is two-fold: First, is that true? Second, can the homoscedasticity versus heteroscedasticity of residuals be tested in similar ways as it can in traditional growth curve modeling here? I don’t at all mind if homoscedastic residuals need to be assumed to ensure the model runs (I think that is the implication from the limitation mentioned in the discussion).
o Could you speak a little more (maybe just a sentence or so) on the importance of step 2 substep iii. In the SEM forest algorithm? It was hard for me to follow why identifying the value of preditor k that maximizes the likelihood ratio of a post-split two-group model indicates best model fit. I suspect just another sentence or so of clarification is needed.
o Could you speak more to the importance of why steps 3 and 4 need to be repeated until all post-split groups reach a minimum size? What would that minimum size be and why is it important?
o How are predictions aggregated from all SEM-trees? If you could offer even just a sentence or two on that process in a way a “traditional” “applied” longitudinal researcher (like me) would understand, I think such readers would be very grateful.
o Could you provide a quick applied example to accompany this statement “This approach intrinsically accommodates non-linear effects as subsequent splits on the same predictor, and complex interactions as subsequent splits on interacting variables. This offers greater flexibility than linear parametric models.”. Again, I think providing an example would allow non-machine learning researchers to follow what you are doing here even more clearly.

Results:
• I think the results clearly presented a complex set of findings. I really appreciated the grouping by theme.
• I have seen many ML applications split data sets into “train” and “test” sub-samples, to validate the findings in the “train” data set in the “test” data set. As far as I can tell, that was not done here. Did the authors consider this approach? I feel like having “test” and “train” sub-samples wherein results are replicated would provide further robustness and validity to study findings. I also understand if that procedure is not possible with the SEM forests approach.
• I know ensuring the article remains inside page limits is always a battle, but I had a suggestion for one additional, hopefully simplistic analysis that might even further bridge the “gap” between these machine learning analyses and “traditional, frequentist” analyses. Perhaps the authors could take only the most important predictors, say the top 10 (or really whatever number they want), and run regression analyses that predict trajectory intercepts and slopes? Basically, I know that people like me really like explanations like “oh yeah, these top 10 machine learning predictors predict XX% of variance in emotion regulation scores over time”, and so I was trying to think of a way to generate those sorts of tangible, “traditional” effect sizes. Could the authors potentially provide such effect size estimates? I also understand if there simply isn’t time or space for this, but figured it was worth asking.

Discussion:
• I appreciated the brave and nuanced discussion of how these analyses conform to, and also challenge, traditionally held norms in the field of emotion regulation. I especially appreciated the identification that parenting, substance use, and bullying might not be the risk factors we think they are.
• The ERRATA tool seems amazing; I couldn’t find it online. Could a link be provided directly in the text?
• I love the notion of “good enough” parenting environment and preconditions, and its’ application here; I thought the link between non-linear effects and “good enough” parenting environments was such an incredible example of translating a statistical insight into “on the ground” wisdom that can be applied by clinicians.  
• The only part of the results that were left relatively untouched by the Discussion was the relatively large importance of Big 5 personality traits as predictors of adolescent emotion regulation. Could the authors comment on what they thought about those findings? I am especially interested to hear the authors thoughts about those results, because I feel like investigation of personality traits is so often “siloed” away from investigation of parenting, substance use, and other phenomena when it comes to adolescent emotion regulation. So I think it would be especially important to also highlight those findings in the Discussion.

Miscellaneous:
• I think there are some broken links to supplemental materials in the manuscript.
• I think the authors should just re-read the manuscript again before final publication; there were missing words and sentences throughout in some places.
• Pages 43 to 65 of the manuscript materials appeared cut-off and unreadable in many areas.

Reviewer: 2

Comments to the Author
REVIEWER STATEMENT

I reviewed this manuscript from the perspective of a clinical psychologist with expertise in quantitative methods and adult psychopathology. I have years of experience with and expertise in applied predictive modeling and supervised machine learning in general but am not that familiar with the SEM forest algorithm, which seems to blend supervised and unsupervised learning. Thus, I defer to other reviewers to evaluate the substantive aspects of the manuscript related to adolescent emotion regulation development and the highly technical aspects of the manuscript specific to the SEM forest algorithm.

SUMMARY

Overall, I found the manuscript to be ambitious and interesting. The size and longitudinal nature of the dataset is impressive, even if it the number of participants is a bit lower than one might ideally want to support complex machine learning. I also valued the manuscript’s integration of theory and methods, transparency of the methods, the addition of variable importance functions to the semTrees package, and the creation of the Shiny app as a demo/first step towards implementation. My concerns are mainly around the lack of external cross-validation (I realize that the out-of-bag approach is like an internal cross-validation) and hyperparameter tuning (which may be related to differences between the supervised approach I am used to and the hybrid approach of SEM forest) as well as some apparent discrepancies between the results as I understand them and the conclusions in the manuscript.

CONCERNS

1. I think that the number of predictors used should be tuned during cross-validation rather than set to 9 or, at least, multiple values should be compared. Similarly, I think that the complexity parameter should be tuned or compared. Please comment on why this wasn’t done.

2. On page 15, there seem to be semantic themes or groupings of predictors. In addition to textual interpretation, I think it would be helpful to create a visualization (even if it were supplemental) that grouped the variable importance scores per theme. Perhaps each theme could be represented by a boxplot or jittered points. This would help communicate the relative importance of each theme (as the text is currently doing informally). A similar approach could be employed for predictors assigned to different levels of the bioecological model.

3. The manuscript’s conclusion that parenting behaviors (especially positive ones) are low in importance (e.g., page 18) is hard for me to reconcile with the third-most important predictor being a positive parenting behavior. Overall, there seems to be an interpretive bias against the parenting behaviors being important. Or perhaps I am misunderstanding something?

4. On page 12, it seems to me that the interpretation of SEM-forests answering, “why do some adolescents have similar growth trajectories” has some causal connotations that may or may not be intended. Please speak to this directly in the manuscript. Can SEM-forests enable causal inferences, or would it be more accurate to rephrase this sentence to remove the connotation?

5. My understanding is that the importance of specific interactions can be interrogated using an generalization of partial dependence profiles (Apley & Zhu, 2020). I’m not saying that this necessarily has to be added to the current paper, but it might be worth citing this approach in the “next steps” part of the manuscript and considering it in future work.

6. That said, my understanding is that partial dependence profiles can be somewhat problematic when applied to correlated predictors (as these predictors surely are) and that a better approach in such cases would be accumulated local profiles (Apley, 2018).

COMMENTS

1. On page 10, please clarify whether the CFAs were single factor models.

2. On page 10, please provide more explanation of the discrepancy between absolute and comparative fit of the metric invariant model.

3. In figure 1, clarify whether the y-axis is reversed. I ask because it seems strange to me that someone with lower difficulties with emotion regulation would be “high risk” and Figure 1 seems to show the opposite pattern as Figure 7.

4. On page 12, please speak to the reasonableness of the assumption of homoscedastic residuals over time.

5. On page 14, the reference to Figure 1 should be to Figure 2.

6. I think Figures 5 and 6 could be moved to supplemental if space is tight.

7. On page 18, is it true that “out [of] all parent-related constructs, constructive conflict resolution with fathers ranked highest and, and out of all parenting behaviors, paternal power ranked highest”? This seems to only be true if you look at only the mother/father-specific parenting measures and omit the general ones like balanced relatedness and conflict interparental.

8. The supplemental table 1 didn’t render well in the PDF.

REFERENCES

Apley, D. W. (2018). ALEPlot: Accumulated Local Effects (ALE) Plots and Partial Dependence (PD) Plots (1.1) [Computer software]. https://CRAN.R-project.org/package=ALEPlot

Apley, D. W., & Zhu, J. (2020). Visualizing the effects of predictor variables in black box supervised learning models. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 82(4), 1059–1086. https://doi.org/10.1111/rssb.12377

Reviewer: 3

Comments to the Author
The current study used machine learning to identify importance of predictors of adolescent emotion regulation trajectories. The study used a large sample of adolescents and included multiple reporters (although adolescent report was featured most heavily). Despite these strengths, I have a number of concerns about the manuscript in its present form that make its contribution unclear at this point. I list these concerns in roughly the order of appearance in the manuscript.

The abstract could be strengthened by providing theoretical rationale for the study. The abstract also seems to exceed the journal’s word limit, as several sentences are cut off on the version included in the pdf cover page. In describing the results for parenting behaviors, it seems that autonomy support and invasive parenting were important if I am reading the statement correctly, so please remove these from the “less important” statement in order not to undermine the importance of parenting.

Please provide citations for the statement in the first paragraph on page 4, “These difficulties in emotion regulation can cause…” and provide examples of why these incur societal cost.

I would suggest removing reference to “destinies” on page 4 and elsewhere, as this word connotes fixedness in developmental outcomes, which does not reflect contemporary developmental views.

The introduction would benefit from proofreading for parallel structure (last sentence in first paragraph on page 4), fragments (e.g., “For example, because they are not…”), and other typos (e.g., “an important challenges” on page 8). These are just several examples. The entire manuscript requires proofreading. The Discussion has sentences that are fragments or are missing words.

At the bottom of page 5, it is noted that the next step in this area of research is to identify risk factors associated with emotion dysregulation. There is a wealth of research already identifying various risk factors. It should be acknowledged that this work exists (and what these identified factors are – some of which are reviewed beginning at the bottom of page 7) before noting the problems with existing research and theory. I would suggest expanding beyond the small number of labs currently heavily relied upon for background on emotion regulation. It seems that the actual next step for this literature is what is defined at the top of the first full paragraph on page 8 – “What is still missing is an understanding of the relative utility…” In short, the introduction could better acknowledge what we already know and more precisely describe the contribution of this study. I expect this would require a fairly extensive revision to move forward.

Please insert the reference listed as [REF] on page 8.

Please provide more of an introduction to SEM-forests prior to The Present Study section. The manuscript appears to assume much more knowledge about machine learning than many readers have.

Please provide more information about the participants (mean and range of age of adolescents and parents at time of recruitment, race/ethnicity, income, and parental education). This will help contextualize the findings related to demographic characteristics that were predictive of emotion regulation trajectories.

How far apart were “waves” of the study?

Did missingeness relate to demographic or emotion regulation variables in the dataset?

Was the statement, “Class solutions with more than two classes were excluded…” an a priori decision, or a statement of what resulted from the attempt to estimate them? Please be clear in the manuscript.

I was curious why the median and plus/minus 1 SD were used for “s” for continuous variables. I would expect the median to be accompanied by percentiles (e.g., 15th and 85th percentiles) or +/- 1 SD to go along with the mean.

Please correct the finding for sex – it currently reads, “boys girls reported better emotion regulation.”

In the first paragraph of the Discussion, where most important predictors are listed, please list parenting behaviors that ranked as highly as aspects currently listed. Throughout the paper, parenting is listed as unimportant, when there are important exceptions that should be more explicit, rather than parenthetical within a negative statement.

I do not believe it is accurate to state, “The fact that individual-level predictors were most important suggests that individual assessments might be sufficient.” Just because a factor is not the most important does not mean it is not playing a significant role. Perhaps the most important factors should be prioritized, but it would be contrary to much of developmental theory and many clinical interventions to suggest that it would be best to consider the adolescent as separate from their context. If anything, developmentalists and clinicians are moving towards greater contextualization of individuals and their identities.

It is rightfully noted that this machine learning application was performed within one sample within one nation. There is limited reason to believe it would be universally applicable, especially because it seemed to draw on variables that happened to be available in a dataset, rather than designing the study for this purpose. The paper could provide more information about the cultural values that may impact emotion regulation in this sample to provide readers more information with which to evaluate generalizability to other populations.

It is not clear how the information on pages 43-65 are to be used. This is not a labeled or formatted table.

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::

### 2022-11-02 manuscript/manuscript.Rmd
update 2022-11-02 15:34:25

@@
   | • Future theory should emphasize proximal- over distal processes, focus on good-enough parenting, and reassess common perceptions about “risky behaviors” like delinquency, substance use, and bullying.
-  | This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests.
+  | This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM-forests.
 Candidate predictors were selected based on a prior text mining systematic review of 6305 papers.
@@
 A comprehensive understanding of the risk factors that predispose some adolescents to emotion dysregulation could aid in the development of effective screening and interventions.
-However, despite an abundance of empirical research,
-many studies are narrow in scope, considering only a few predictors,
-the literature is fragmented across subdisciplines, and lacks a unifying theoretical framework
-<!-- that is both specific to adolescence and comprehensive in scope  -->
-[@vanlissaMappingPhenomenaRelevant2021].
+<span id = "abundantlit">There is already an abundance of empirical research on emotion regulation and its predictors; a recent systematic review identified 6305 relevant publications [@vanlissaMappingPhenomenaRelevant2021].
+Still, that review identified specific limitations of the existing literature: many studies are narrow in scope, considering only a few predictors and thus missing the relative importance compared to other predictors.</span>
+This problem is exacerbated because the literature is fragmented across sub-disciplines, each of which examines mostly predictors within its scope.
+And finally, the literature lacks a unifying theoretical framework that is both specific to adolescence and comprehensive in scope [see @bariolaChildAdolescentEmotion2011].
 
@@
 and ranked in terms of predictive importance.
-<!-- These predictors may include both risk factors that render some youth susceptible to emotion regulation difficulties, and correlates that can be used to identify at-risk youth. -->
 Specifically,
-the SEM-forests algorithm was used to
-identify the most important early indicators, at age 13, of developmental trajectories
+<span id = "personcentered1">the SEM-forests algorithm [@brandmaierTheoryguidedExplorationStructural2016] was used as a person-centered analysis technique,
+to identify the most important early indicators (at age 13) of developmental trajectories
 of emotion regulation across adolescence (age 14 through 18).
-The resulting insights indicate which factors are most predictive of adolescents' diverging destinies in terms of emotion regulation development.
+The resulting insights indicate which factors are most predictive of adolescents' diverging trajectories of emotion regulation development, contributing to our understanding of why some youth flourish while others struggle [@croneUnderstandingAdolescencePeriod2012].</span>
 Armed with this knowledge, future research and interventions can focus on the most important predictors.
@@
 At the same time, this imbalance may render some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
-<!-- exposes youth to a risk of emotion dysregulation, -->
 For example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
@@
 
-Empirical support for this notion of diverging destinies is evident from studies that examine developmental trajectories of emotion regulation in adolescence: 
+Longitudinal studies of emotion regulation development in adolescence provide preliminary empirical support for this notion of diverging trajectories: 
 Such studies consistently reveal large between-individual heterogeneity [see @craccoEmotionRegulationChildhood2017; @vanlissaRoleFathersMothers2019; @zimmermannEmotionRegulationEarly2014; @reindlSocializationEmotionRegulation2016].
+Although some studies have examined, on average, how strongly particular risk factors are associated with developmental trajectories
+
 An important next step is to identify which risk factors render some adolescents susceptible to emotion dysregulation.
@@
 <!-- In part, this limitation is a consequence of conventional methods: -->
-linear models limit the total number of predictors that can be included due to concerns of power, overfitting, and multicolinearity [REF].
+linear models limit the total number of predictors that can be included due to concerns of power, overfitting, and multicolinearity [@hastieElementsStatisticalLearning2009].
 By contrast, some machine learning methods offer a more holistic approach,
@@
 The present study used a specific machine learning algorithm uniquely suited for theory-guided exploration: SEM-forests [@brandmaierTheoryguidedExplorationStructural2016].
+
+## Introduction to SEM-forests
+
+<span id = "semforests">The SEM-forests algorithm is based on the random forests algorithm, which in turn is based on regression trees [see @stroblIntroductionRecursivePartitioning2009].
+A single regression tree splits the sample repeatedly in two by whichever predictor $p$ makes both post-split groups maximally internally homogenous, and maximally different from one another.
+A random forest consists of many regression trees grown on bootstrapped samples from the original dataset.
+To ensure that each tree learns something different from the data, it may only consider a random subsample of $K$ candidates out of all $p$ predictors at each split.
+With this in mind, the SEM-forest algorithm operates as follows:
+
+1. Bootstrap the data with replacement; we used 1000 bootstrap replications
+2. For each bootstrap sample, construct a SEM tree:
+    i. Estimate the latent growth model on the bootstrap sample
+    ii. Consider a random selection of $K$ candidates out of all $p$ predictors; we used $k = 9$ based on the standard practice to use $K = \lfloor\sqrt{p}\rfloor$
+    iii.
+Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).
+    iv. Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then fully grown.
+3.
+Given an individual's scores on the predictor variables $p$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters).
+Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.
+
+Each tree thus maps relationships between the predictors and outcome as a series of recursive splits of the data.
+This approach intrinsically accommodates non-linear effects and (complex) interactions, which is an advantage over parametric models that assume linearity.
+As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
+In regression analysis, this could be represented with a linear and quadratic term, and a plot might show that the effect of autonomy support on the intercept increases to a value $x_{knee}$, and then tapers off.
+A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.
+The algorithm additionally performs variable selection;
+unimportant variables are not selected in step 2.iii.
+Thus, the number of predictors can be large without risking model non-identification.</span>
 
@@
 `r ifelse(mask, "The Workflow for Open Reproducible Code in Science (WORCS)", "**Masked**")` was used to make analyses reproducible  `r ifelse(mask, "[@vanlissaWORCSWorkflowOpen2020]", "**(masked reference)**")`.
-All analysis code, supplemental materials and synthetic data are available at `r ifelse(mask, "<https://github.com/cjvanlissa/veni_forest>", "**masked URL**")`.
+All analysis code, Supplemental materials and synthetic data are available at `r ifelse(mask, "<https://github.com/cjvanlissa/veni_forest>", "**masked URL**")`.
 The original data (controlled access) and all study documentation are available via `r ifelse(mask, "<https://doi.org/10.17026/DANS-ZRB-V5WP>", "**masked URL**")`.
@@
 ```
-Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal study `r ifelse(mask, "RADAR [@branjeResearchAdolescentDevelopment2018]", "**(masked reference)**")`, which
+<span id = "dutch">Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal study `r ifelse(mask, "RADAR [@branjeResearchAdolescentDevelopment2018]", "**(masked reference)**")`, which
 received ethical approval from the ethical committee of the University Medical Centre `r ifelse(mask, "Utrecht", "**masked**")`.
-The `r ifelse(mask, "RADAR", "**masked**")` sample was recruited from over 200 randomly selected elementary schools in the province of `r ifelse(mask, "Utrecht", "**masked**")`, and four main cities in `r ifelse(mask, "The Netherlands", "**masked**")`.
+The `r ifelse(mask, "RADAR", "**masked**")` sample was recruited from over 200 randomly selected elementary schools in the province of `r ifelse(mask, "Utrecht", "**masked**")`, and four main cities in `r ifelse(mask, "The Netherlands", "**masked**")`.</span>
 Of 1,544 eligible families, 497 produced informed consent and were included in the study.
-From 2006 to 2011, trained interviewers conducted annual home visits to collect questionnaire data.
+[From 2006 to 2011, trained interviewers conducted six annual measurement waves to collect questionnaire data during home visits.]{#waves}
 All independent variables were assessed in the first measurement wave.
@@
 The lower BIC indicates support for metric invariance.
-The RMSEA and SRMR of the invariant model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
+The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
+[Note that the RMSEA has been found to be more sensitive to misspecifications of growth curve models [@leiteDetectingGrowthShape2011]]{#cfi}.
 As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
@@
 <!-- SB: I miss a reflection on the extent to which the available data cover all relevant predictors -->
-For multi-item scales, we used principal components analysis for dimension reduction.
-Across all scales, the explained variance of the first component ranged from $R^2 \in [0.18, 0.90]$.
-This first component represents the largest amount of shared variance among the scale items, and was used for further analysis.
-[Supplementary Table S1](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_1.csv) additionally reports skew and kurtosis,
+We used principal components analysis to obtain a single score for multi-item scales without making additional assumptions about the underlying measurement model.
+For each scale, we retained the first component for further analysis, which represents the largest amount of shared variance among the scale items.
+For example, consider the construct *neuroticism*, which was measured on a seven point Likert scale.
+Participants were asked to indicate "Please indicate to what extent you posess these traits" for six adjectives associated with neuroticism.
+Principal components analysis indicated that the first component was the only component with an Eigenvalue greater than one (3.22, other Eigenvalues [0.99 - .33].
+This component explained the majority of variance in all six adjectives, $R^2 = .54$, other components $R^2 = [.16 - .05]$.
+All adjectives loaded highly on this first component, $L = [.54 - .83]$.
+We thus used this first component as operationalization of the construct neuroticism.
+Across all other scales, the explained variance of the first component ranged from $R^2 \in [0.18, 0.90]$, which corresponded to 1.60 - 17.46 times as much variance as could be expected if each dimension of the scale explained equal variance.
+We thus considered these principal components to be adequate operationalizations of the scales.
+[Supplemental Table S1](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_1.csv) additionally reports skew and kurtosis,
 minimum and maximum factor loadings,
@@
 a random forests based approach with comparable performance to multiple imputation and FIML estimation [@stekhovenMissForestNonparametricMissing2012].
-This method is valid when missingness is either random or contingent on observed data (MAR or MCAR).
+This method is valid, regardless of whether missingness is random or contingent on observed data (MAR or MCAR).
 It does not make any distributional assumptions and does not extrapolate,
@@
 and from $`r report(min((495-desc_num[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for continuous data (see Table \@ref(tab:tabnum)).
+[We compared participants with missing- versus complete data on demographic characteristics (sex, age, socio-economic status, religion and religiosity) and scores of emotion regulation in all waves, and found no significant differences, $p = [0.06, 0.48]$.]{#missing}
 <!-- Original publication had up to 30% missing, Wolputte, E.
@@
 with remaining heterogeneity attributed to error variance.
-Class solutions with more than two classes were excluded from consideration due to low class separability (entropy < .9).
 The LMR likelihood ratio test indicated that two classes significantly improved fit ($p < .001$); more classes did not ($p = .26$).
+[Furthermore, results indicated that solutions with more than two classes had low class separability (entropy < .9),
+which means that these solutions are less interpretable [@jungIntroductionLatentClass2008].]{#entropy}
 As displayed in Figure \@ref(fig:figclass),
@@
 
-As we set out to identifying important predictors of emotion regulation development,
-this latent class analysis is not sufficient to answer our research question.
-In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
+As our research question was to identifying important predictors of emotion regulation development,
+this latent class analysis does not yet provide an adequate answer.
+Latent class analysis clusters individuals only with respect to the outcome (growth trajectories), and ensures maximum homogeneity of growth trajectories in each class.
 It thus answers the question: "Which adolescents have similar growth trajectories"?
-SEM-forests similarly aim to ensure maximum homogeneity of growth trajectories within clusters,
-but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
+SEM-forests similarly aim for maximum homogeneity of growth trajectories within clusters,
+but unlike latent class analysis, they cluster individuals by reference to a set of *predictors*.
 They thus answer the question: "Why do some adolescents have similar growth trajectories"?
@@
 
-For the SEM-forests analysis,
+<span id = "homogeneity">For the SEM-forests analysis,
 we imposed an additional restriction on the diagonal of the residual covariance matrix,
 reflecting an assumption of homoscedastic residuals over time.
-The SEM-forests algorithm operates as follows:
-
-1. Bootstrap the data with replacement; we used 1000 bootstrap replications
-2. For each bootstrap sample, construct a SEM tree:
-    i. Estimate the latent growth model on the bootstrap sample
-    ii. Consider a random selection of $K$ candidates out of all $p$ predictors; we used $k = 9$ based on the standard practice to use $K = \lfloor\sqrt{p}\rfloor$
-    iii.
-Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the likelihood ratio $LR$ of a post-split multigroup model
-    iv. Repeat steps 3 and 4 until all post-split groups reach a minimum size
-3. Aggregate predictions from all SEM-trees
-
-Each tree thus maps relationships between the predictors and outcome as a series of recursive splits of the data.
-This approach intrinsically accommodates non-linear effects and (complex) interactions, which is an advantage over parametric models that assume linearity.
-The algorithm additionally performs variable selection;
-unimportant variables are not selected in step 2.iii.
-Thus, the number of predictors can be large without risking model non-identification.
+We deemed this assumption to be necessary because the dataset is relatively small, and we wanted the model to be as simple as possible.</span>
 
@@
 
-For continuous predictors, three values of $s$ were used - namely the percentiles corresponding to the median, $+/- 1 SD$ of a standard normal distribution.
+[For continuous predictors, three values of $s$ were used - namely the percentiles, computed on the observed data, corresponding to the median, $+/- 1 SD$ of a standard normal distribution (i.e., the $16^{th}$ and $84^{th}$ percentile).
+For convenience, these percentiles are labeled as $+/- 1 SD$ in Figure \@ref(fig:figpdp1)-\@ref(fig:figpdp4).]{#percentiles}
 For categorical predictors, all unique values were used.
@@
 Candidate predictors were rank-ordered by their relative variable importance,
-as illustrated in Figure 1.
+as illustrated in Figure \@ref(fig:figvarimp).
 <!-- Table \@ref(tab:tablegend) provides a brief legend of the 30 most important variables, and  -->
-Supplementary Table S1 provides a full legend of all variables.
+Supplemental Table S1 provides a full legend of all variables.
 
@@
 age and pubertal development,
-and sex (boys girls reported better emotion regulation).
+and sex (boys reported better emotion regulation).
 
@@
 <!-- * Conflict resolution styles -->
+
+## Post-hoc frequentist analyses
+
+A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
+Regressing the parameters (intercept, slope and quadratic slope) of a latent growth model on each predictor in turn.
+Such analyses might be helpful to plan hypotheses and perform power analysis for follow-up research within the SEM framework.
+However, we strongly caution against over-interpretation.
+Unlike the machine learning analyses, these analyses do not accommodate non-linear effects or interactions, nor are effects controlled for other predictors.
+The results are merely descriptive and should not be used to perform inference.
+[Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
+All models had good fit ($RMSEA < .07$, $CFI > .99$, $SRMR < .04$).
+Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
+This indicates that our machine learning analyses indeed recovered relevant patterns in the data.
 
@@
 among all remaining parenting constructs, negative and intrusive dimensions were more predictive (as risk factors) than positive or neutral parenting practices.
-Themes that ranked notably low were family demographics (SES and religion),
-bullying, victimization and delinquency, and substance use.
+Themes that ranked notably low were family demographics (SES and religion) and bullying, victimization and delinquency, and substance use.
+The relatively low importance of demographics might be attributable to a restriction of range, as the sample was relatively homogenous.
+Still, other empirical studies have also found little evidence for a direct effect of, for example, SES on adolescent emotion regulation [e.g., @defranceExpandingContextRole2021].
+It is thus also possible that the population effect is small, or that it is fully mediated by more proximal process variables.
 
@@
 The ERRATA application is open source, and can thus be customized for clinical use.
+The source code is available at <https://github.com/cjvanlissa/shiny_errata>;
+users can run the app from a local R instance,
+or view an online demo using a link in the README file of this repository (as long hosting is available).
 
@@
 Specifically, this was a non-random, ethnically and socially relatively homogeneous sample from the Netherlands.
-This raises specific concerns, for example, that the low importance of SES might be due to a restriction of range.
+This raises specific concerns, for example, that the low importance of the demograpic factors SES and religion might be due to a restriction of range.
 The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
@@
 Although random forests are relatively robust to overfitting,
-larger samples will likely yield more generalizable results,
-and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
-Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
-Computational constraints thus prevented us from tuning the model beyond default settings,
-and from exploring marginal effects for higher order interactions.
-Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.
+larger samples will likely yield more generalizable results.
+<span i = "homogeneity2">Larger samples also afford more complicated models.
+We saw the need, for example, to assume homogeneity of residuals within each node of the SEM-forest.
+This assumption is likely violated in some nodes; a larger sample would allow us to freely estimate residuals.
+Note, however, that many other assumptions are also likely violated in some nodes (e.g., the shape of the growth trajectory).
+This limitation is inherent to the method.</span>
+Other limitations related to SEM-forests include the fact that,
+unlike in traditional random forests, there is no error metric that could be used to calculate fit in a hold-out testing sample.
+By extension, at the time of writing, there are not yet methods for cross-validation, model tuning, or evaluating the overall fit of a SEM-forest, as these too require an error metric.
+Finally, SEM-forests require substantial computational resources,
+which also limit the ability to compute different iterations of the model beyond default settings,
+or to exploring marginal effects for higher order interactions.
 

### 2022-11-02 posthocmodels.R

@@ -1,0 +1,61 @@
+library(tidySEM)
+vim <- readRDS("results/vim_all_22021-08-21_17_23_27.RData")
+VI <- list(variable.importance = semtree:::aggregateVarimp(vim, aggregate = "median", scale = "absolute", TRUE))
+preds <- names(sort(VI$variable.importance, decreasing = TRUE))
+
+tab <- lapply(preds, function(p){
+  model <- "i =~ 1*de2 + 1*de3 + 1*de4 + 1*de5 + 1*de6
+s =~ 0*de2 + 1*de3 + 2*de4 + 3*de5 + 4*de6
+q =~ 0*de2 + 1*de3 + 4*de4 + 9*de5 + 16*de6\n"
+  if(inherits(df_anal[[p]], "factor")){
+    if(length(table(df_anal[[p]])) == 2){
+      levels(df_anal[[p]]) <- c("0", "1")
+      df_anal[[p]] <- as.numeric(as.character(df_anal[[p]]))
+    }
+  }
+    
+  if(!inherits(df_anal[[p]], "factor")){
+    model <- paste0(model,
+      paste0("i ~ ", p, "\n", "s ~ ", p, "\n", "q ~ ", p))
+    res_preds <- lavaan::growth(model, data = df_anal)
+    tab <- table_results(res_preds, columns = NULL)
+    tab <- tab[(tab$lhs %in% c("i", "s", "q") & tab$rhs %in% preds), ]
+    tab_coef <- unlist(tab[["est_sig_std"]])
+    names(tab_coef) <- unlist(tab$lhs)
+    tab_coef <- tab_coef[c("i", "s", "q")]
+  } else {
+    grpsize <- table(df_anal[[p]])
+    grpnams <- names(grpsize)[grpsize > 50]
+    tmp <- df_anal
+    tmp <- tmp[tmp[[p]] %in% grpnams, ]
+    model <- model <- paste0(model,
+                             paste0("i ~ c(",
+                                    paste0("ci", grpnams, collapse = ","),
+                                    ") *1\n",
+                                    "s ~ c(",
+                                    paste0("cs", grpnams, collapse = ","),
+                                    ") *1\n",
+                                    "q ~ c(",
+                                    paste0("cq", grpnams, collapse = ","),
+                                    ") *1\n"))
+    
[…]

### 2022-11-02 scripts/creatingscales.R

@@ -111,6 +111,13 @@ pc_mat <- t(sapply(pred_list, function(x){
 #pc_mat <- pc_mat[!is.na(pc_mat[, 2]), ]
 range(pc_mat[,1][!pc_mat[,1]==1])
 
+# How much more variance does the first component explain than expected by chance?
+range(sapply(pred_list, function(x){
+  tst <- summary(x)
+  if(ncol(x$x) == 1) return(NA)
+  tst$importance[2,1]/(1/ncol(x$x))
+}), na.rm = TRUE)
+
 #library(psych)
 #fa.parallel(data[scales_list$schoolprestatie_aa], fa = "pc", plot = FALSE)
 #summary(prcomp(data[scales_list$schoolprestatie_aa]))

### 2022-11-02 scripts/missRanger.R

@@ -152,6 +152,17 @@ intvars <- intvars[!intvars %in% numvars]
 reldata[intvars] <- lapply(reldata[intvars], drop_floats)
 table(drop_floats(reldata$de21aa03))
 saveRDS(reldata, "reldata.RData")
+
+miss<- is.na(reldata)
+anymiss <- apply(miss, 1, any)
+saveRDS(anymiss, "anymiss.RData")
+checkthese <- reldata[c("geslacht", "brpmoe_lmh", "brpvad_lmh", "sesgez_low", "bg11aa04", "reli")]
+range(c(sapply(names(checkthese), function(n){
+  chisq.test(factor(anymiss), factor(reldata[[n]]))$p.value
+}), leeftijd = t.test(reldata$leeftijd_target_11~ anymiss)$p.value))
+
+# reldata <- readRDS("reldata.RData")
+# misstest <- mice::mcar(reldata, replications = 1000)
 #!
 # @Caspar - make sure to adjust the num.trees parameter! 
 #!

### 2022-11-02 scripts/plantingforest.R

@@ -37,6 +37,14 @@ fa.parallel(df_anal[, grepl("^de[2-6]", names(df_anal))])
 dv <- principal(df_anal[, grepl("^de[2-6]", names(df_anal))])
 dv$scores
 df_pres <- data.frame(dv$scores, df_anal[, !grepl("^de[2-6]", names(df_anal))])
+
+anymiss <- readRDS("anymiss.RData")
+checkthese <- df_anal[, grepl("^de[2-6]", names(df_anal))]
+range(sapply(names(checkthese), function(n){
+  tmp <- lm(checkthese[[n]] ~ anymiss)
+  summary(tmp)$coefficients[2, 4]
+  }))
+
 library(ranger)
 library(tuneRanger)
 tunetask <- makeRegrTask(data = df_pres, target = "PC1")
@@ -57,6 +65,10 @@ p <- p +
   scale_fill_manual(values = c("FALSE" = "white", "TRUE" = "black"))+
   theme(legend.position = "none")
 
+
+# Post hoc linear models --------------------------------------------------
+
+source("posthocmodels.R")
 
 # SEM forest --------------------------------------------------------------
 basicgrowth <- lavaan::growth("i =~ 1*de2 + 1*de3 + 1*de4 + 1*de5 + 1*de6

### 2022-11-03 action_letter.Rmd
update 2022-11-03 12:39:10

@@
 
-> Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).
+`r get_revision(man, "stepiii")`
 
@@
 
-> Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then fully grown.
+`r get_revision(man, "fullygrown")`
 
@@
 
-> The SEM-forests algorithm is based on the random forests algorithm, which in turn is based on regression trees [see @stroblIntroductionRecursivePartitioning2009].
-A single regression tree splits the sample repeatedly in two by whichever predictor $p$ makes both post-split groups maximally internally homogenous, and maximally different from one another.
-A regression forest consists of many regression trees grown on bootstrap samples of the data.
-To ensure that each tree learns something different from the data, it may only consider a random subsample of $K$ candidates out of all $p$ predictors at each split.
-With this in mind, the SEM-forest algorithm operates as follows... [etc]
+`r get_revision(man, "semforests")`
 
@@
 
-> Given an individual's scores on the predictor variables $p$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters).
-Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.
+`r get_revision(man, "predictions")`
 
@@
 
-> As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
-In regression analysis, this could be represented with a linear and quadratic term, and a plot might show that the effect of autonomy support on the intercept increases to a value $x_{knee}$, and then tapers off.
-A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.
+`r get_revision(man, "examplenonlinear")`
 
@@
 
-> Other limitations related to SEM-forests include the fact that,
-unlike in traditional random forests, there is no error metric that could be used to calculate fit in a hold-out testing sample.
-By extension, at the time of writing, there are not yet methods for cross-validation, model tuning, or evaluating the overall fit of a SEM-forest, as these too require an error metric.
+`r get_revision(man, "limitations")`
 
@@
 
-> A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
-Regressing the parameters (intercept, slope and quadratic slope) of a latent growth model on each predictor in turn.
-Such analyses might be helpful to plan hypotheses and perform power analysis for follow-up research within the SEM framework.
-However, we strongly caution against over-interpretation.
-Unlike the machine learning analyses, these analyses do not accommodate non-linear effects or interactions, nor are effects controlled for other predictors.
-The results are merely descriptive and should not be used to perform inference.
-[Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
-All models had good fit ($RMSEA < .07$, $CFI > .99$, $SRMR < .04$).
-Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
-This indicates that our machine learning analyses indeed recovered relevant patterns in the data.
+`r get_revision(man, "posthoc")`
 
@@
 
-> The source code is available at <https://github.com/cjvanlissa/shiny_errata>;
-users can run the app from a local R instance,
-or view an online demo using a link in the README file of this repository (as long hosting is available).
+`r get_revision(man, "errata")`
 
@@
 
-REF
+We thank the Reviewer for this hepful suggestion, and have written the following to address it:
+
+`r get_revision(man, "includepersonality")`
 
@@
 We apologize.
-The manuscript only has 42 pages; pages 43-65 are copies of the online supplemental materials.
-The submission portal has trouble correctly presenting these materials, but they will not be part of the paper.
+The manuscript only has 42 pages; pages 43-65 are copies of the online supplemental materials (which are supposed to be accessed via the broken links, which have now been fixed).
+For the sake of blind review, we also submitted copies of these tables through the submission portal, but it evidently cannot correctly present these materials.
+The tables will not be part of the paper in this form.
 
@@
 
+We understand why the Reviewer expects the analysis to include model tuning and cross-validation; these are standard procedures in machine learning.
+However, these methods do not (yet?) exist for SEM-forests.
+This is because there is no single error metric for SEM-forests that can be optimized.
+For standard random forests, this error metric is the root mean squared error (difference between predicted and observed values for all cases).
+The "outcome" of each regression tree in the SEM-forest is an entire SEM-model, for which there is no clear-cut case-specific error metric.
+It is important to note, however, that unlike other techniques, random forests typically achieve competitive performance even with default settings [@probstHyperparametersTuningStrategies2019].
+Thus, even when a model isn't tuned - or cannot be tuned, as is the case here - it should perform well.
+Although we did attempt to explain these points in the previous version of the manuscript,
+we have attempted to address this comment by further emphasizing this in the Revision:
+
+`r get_revision(man, "limitations")`
+
 \RC{2.
@@
 A similar approach could be employed for predictors assigned to different levels of the bioecological model.}
+
+We thank the Reviewer for this excellent suggestion, and now provide this figure
 
@@
 
-REF
+We appreciate this point, and the writing here should certainly be more nuanced (that was an oversight).
+We have rephrased this as follows:
+
+`r get_revision(man, "riskassessment")`
 
@@
 
+As requested, we now provide more information about cultural values that may impact the generalizability of our findings:
+
+`r get_revision(man, "culture")`
+
 \RC{It is not clear how the information on pages 43-65 are to be used. This is not a labeled or formatted table.}
 
-We understand that pages 43-65 are not very useful in their current form, and apologize for the inconvenience.
-We were required to attach hard copies of the online supplemental tables during submission, and these were evidently not correctly formatted by the submission portal.
-They are just Excel spreadsheets, and will not be part of the manuscript.
+We apologize for the inconvenience.
+Pages 43-65 are not intended to be used; we were required to attach hard copies of the online supplemental tables in the submission portal, and these were evidently not formatted correctly.
+They are just spreadsheets, available online, and will not be part of the printed manuscript.
 

### 2022-11-03 manuscript/manuscript.Rmd
update 2022-11-03 12:39:10

@@
     iii.
-Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).
-    iv. Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then fully grown.
+[Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).]{#stepiii}
+    iv. [Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then fully grown.]{#fullygrown}
 3.
-Given an individual's scores on the predictor variables $p$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters).
-Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.
+[Given an individual's scores on the predictor variables $p$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters).
+Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.]{#predictions}
 
@@
 This approach intrinsically accommodates non-linear effects and (complex) interactions, which is an advantage over parametric models that assume linearity.
-As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
+<span id = "examplenonlinear">As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
 In regression analysis, this could be represented with a linear and quadratic term, and a plot might show that the effect of autonomy support on the intercept increases to a value $x_{knee}$, and then tapers off.
-A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.
+A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.</span>
 The algorithm additionally performs variable selection;
@@
 
-A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
+<span id = "posthoc">A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
 Regressing the parameters (intercept, slope and quadratic slope) of a latent growth model on each predictor in turn.
@@
 Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
-This indicates that our machine learning analyses indeed recovered relevant patterns in the data.
+This indicates that our machine learning analyses indeed recovered relevant patterns in the data.</span>
 
@@
 <!-- Yet perhaps there was too little variation in relevant macro factors (ethnicity, SES/poverty, cultural differences in ER) -->
+
+<span id = "includepersonality">Several personality traits emerged as particularly important proximal predictors.
+According to a recent review on personality and emotion regulation (in adults),
+personality is a broad framework for describing psychological capacities,
+of which emotion regulation is one [@hughesPersonalityTraitsEmotion2020].
+This view is certainly supported by our findings,
+and yet research on adolescents does not routinely assess the role of personality alongside factors of particular interest in our discipline, such as parenting, peer relationships, and problem behavior [@vanlissaMappingPhenomenaRelevant2021].
+This is certainly a promising area for future research.</span>
 
@@
 
-These findings have several implications for the early identification of youth at-risk for emotion regulation difficulties.
-The fact that individual-level predictors were most important
-suggests that individual assessments might be sufficient.
-Holistic approaches that involve the family system might not be necessary for preliminary risk assessment.
+[These findings have several implications for the early identification of youth at-risk for emotion regulation difficulties.
+The fact that individual-level predictors were most important in predicting adolescents' diverging developmental trajectories suggests that it might be possible to perform preliminary risk assessments for emotion dysregulation using only adolescent-reported individual differences.
+Although the focus in clinical practice is shifting towards holistic approaches that involve the family system,
+this may not always be feasible.
+For instance, a family member may refuse to cooperate, or holistic assessment may be too costly.
+In such cases, it is useful to know that individual differences carry substantial informational value.]{#riskassessment}
 Moreover, some of the most important predictors are already routinely assessed (for example, big five personality).
@@
 while keeping other less important predictors constant at the most common value.
-The ERRATA application is open source, and can thus be customized for clinical use.
+<span id = "errata">The ERRATA application is open source, and can thus be customized for clinical use.
 The source code is available at <https://github.com/cjvanlissa/shiny_errata>;
 users can run the app from a local R instance,
-or view an online demo using a link in the README file of this repository (as long hosting is available).
+or view an online demo using a link in the README file of this repository (as long hosting is available).</span>
 
@@
 Specifically, this was a non-random, ethnically and socially relatively homogeneous sample from the Netherlands.
-This raises specific concerns, for example, that the low importance of the demograpic factors SES and religion might be due to a restriction of range.
+[Cultural values, which differ across and within countries,
+have been found to influence people's motivation to regulate emotions, as well as the strategies they use and their adaptiveness [@fordCultureEmotionRegulation2015].
+In cross-cultural comparisons, the Netherlands has been found (on average) to ascribe to similar cultural values as large Anglophonic countries (e.g., the United States, Canada, and Australia) [@terlutterGLOBEStudyApplicability2006].
+Whether our findings also generalize to populations that adhere to different cultural values remains an open question.]{#culture}
+Another concern related to generalizability is that the relative homogeneity of our sample in terms of demograpic factors, like SES and religion, might have contributed to their lower variable importance.
 The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
 The only way to address these concerns is to replicate these findings in other samples.
-Note that all code is made available in a reproducible archive to facilitate replication.
+To facilitate such replications, all code is made available in a reproducible archive.
 Another limitation is that the data source was a panel study with mostly questionnaire measures.
@@
 larger samples will likely yield more generalizable results.
-<span i = "homogeneity2">Larger samples also afford more complicated models.
+<span id = "homogeneity2">Larger samples also afford more complicated models.
 We saw the need, for example, to assume homogeneity of residuals within each node of the SEM-forest.
@@
 This limitation is inherent to the method.</span>
-Other limitations related to SEM-forests include the fact that,
+<span id = "limitations">Other limitations related to SEM-forests include the fact that,
 unlike in traditional random forests, there is no error metric that could be used to calculate fit in a hold-out testing sample.
 By extension, at the time of writing, there are not yet methods for cross-validation, model tuning, or evaluating the overall fit of a SEM-forest, as these too require an error metric.
+Note, however, that even without model tuning, random forests tend to achieve competitive performance [@probstHyperparametersTuningStrategies2019].</span>
 Finally, SEM-forests require substantial computational resources,

### 2022-11-09 action_letter.Rmd
Add supplemental tables

@@
 
-```{r}
+```{r, message=FALSE, warning=FALSE}
 library(revise)
-man <- revise::read_manuscript("manuscript/manuscript.Rmd")
+man <- read_manuscript("manuscript/manuscript.Rmd")
+letter <- read_manuscript("action_letter.Rmd")
+mask = TRUE
 ```
@@
 
-Thank you for considering my manuscript for publication in _`r rmarkdown::metadata$journal`_.
-I appreciate the thorough and critical comments you solicited,
+Thank you for considering our manuscript for publication in _`r rmarkdown::metadata$journal`_.
+We appreciate the thorough and critical comments you solicited,
 which helped strengthen the argumentation, clarify the structure, and add nuance.
-I have made a good-faith effort to address all comments as thoroughly as possible.
-All changes are explained below, and the difference between the current version and previous one can be seen at <https://github.com/cjvanlissa/rigorous_exploration/compare/d8ba218..48d2b17>.
-I hope that you will find the quality of the work to be sufficiently improved.
+We have attempted to address all comments as thoroughly as possible.
+All changes are explained below.
+We hope that you will find the quality of the work to be sufficiently improved.
 
@@
 Author
-
-<!-- ` get_revision(man, "test")` -->
 
@@
 
+We completely agree with Reviewer 3 that there is already a wealth of prior research in this area.
+In fact, there is so much work on this topic that it would be a tall order to review the literature on an individual study basis.
+Given the large prior literature,
+we rely primarily on prior review articles.
+We currently cite 17 review articles in the introduction,
+including a recent (2021) comprehensive systematic review of 6305 studies on factors associated with adolescent emotion regulation.
+
+We cannot strive to recreate a comprehensive review here - 
+and this also seems redundant, given that many of the cited review articles are very recent (i.e., from 2019-2021),
+and seems overly ambitious, given the page limits and attention that must be devoted to explaining the novel methods,
+and spurious, given that this is an exploratory study and thus has no hypotheses that must be supported with prior literature.
+The present study builds upon prior reviews of the literature,
+and those reviews are cited.
+
+Of course we are willing to include specific omitted literature, but please note that Reviewer 3 does not mention which references they feel are missing.
+
+To address this comment, we now emphasize more strongly that there is an abundance of prior literature on predictors of adolescent emotion regulation, e.g.:
+
+`r get_revision(manuscript = man, "abundantlit")`
+
 \RC{Also, please specify the demographics of the samples of cited studies.
@@
 If the studies you are citing are from these world regions, please specify that.}
+
+As mentioned above, the vast majority of cited studies are reviews (narrative reviews, meta-analyses, and book chapters) or theoretical publications.
+Although one would hope that such publications cover all available literature without selecting for country of origin,
+some bias is likely present.
+Of the few primary empirical studies we cited, all originated in the "Global North" (if one also considers Israel and Greece to be part of the Global North).
+To address this comment, we have extracted information about the country of origin of the samples of all primary data sources cited.
+This information is summarized in a new Supplemental Table,
+which also indicates what references are *not* primary data sources but instead theoretical, review, or methodological articles.
+
+To address this comment, we now state the following in the Limitations section:
+
+`r get_revision(manuscript = man, "globalnorth")`
 
@@
 
-Note that a recent perspective on the bioecological theory emphasizes that distal factors like culture are often enacted proximally by close others.
-This claim might help explain why we found that distal factors were relatively less important predictors;
-if distal effects are largely mediated by proximal actors,
-their importance might be attenuated by conditioning on proximal effects.
+To address this comment, we have incorporated the following in our analysis of the results in light of the bioecological model:
+
+`r get_revision(manuscript = man, "velezagosto")`
 
@@
 We respectfully believe that this may have been an oversight,
-as this information was already provided.
-As per the submission guidelines, some specific details had been masked for review, which may have caused the oversight.
+as the requested information was already provided.
+As per the submission guidelines, some specific details had been masked for review, which may have caused the confusion.
 To address this comment, we have unmasked these details in the revision:
@@
 
-`r get_revision(man, "waves")`
+`r get_revision(manuscript = man, "waves")`
 
@@
 
-`r get_revision(man, "dutch")`
+`r get_revision(manuscript = man, "dutch")`
 
@@
 
-As requested, we have addressed this comment by providing more detail and an illustrative example, please refer to our response to Reviewer 1.
+`r get_revision(manuscript = letter, "extractpreds", quote = F)`
 
@@
 
+<span id = "parents">We now realize that our imprecise use of language caused confusion about the relative importance of what we previously called "parenting" - by which we meant specific parenting practices.
+
+In the revision, we have attempted to avoid confusion by distinguishing "parenting practices" from other "parent-related constructs".
+Parenting practices are defined as specific parental actions related to child-rearing,
+for example, support and dimensions of control [@petersonFamilySocialCompetence1990].
+The highest-ranking parenting practice is parental control of friendships (16).
+
+By contrast, parent-related constructs refer to any variable pertaining to a parent but that is not a parenting practice.
+The highest ranking parent-related construct is balanced relatedness (3),
+which measures whether the adolescent perceives his or her parents to be tolerant of the adolescent's differing beliefs.
+Balanced relatedness is an aspect of the emotional family climate [@morrisRoleFamilyContext2007],
+not a parenting practice, because it does not relate to specific parental actions.
+The specific parenting practices associated with instilling a sense of balanced relatedness are maternal (54) and paternal (55) support, which rank very low.
+
+The observation that *parenting practices* are relatively less important than individual predictors (and certainly much less important than we would have thought a-priori) is thus strongly supported by the data;
+the first such behaviors are (16) and (25).
+This is consistent with the finding that some parent-related constructs indeed score highly.
+
+To address this comment, we have:
+
+- Clarified the distinction between parenting practices and other parent-related constructs in the revision
+- Used consistent terminology throughout 
+- Provided a definition of parenting practices
+
+We have rewritten the following paragraphs:
+
+`r get_revision(manuscript = man, "parenting")`
+
+And:
+
+`r get_revision(manuscript = man, "parentrelated")`
+</span>
+
 \RC{In the discussion, I encourage you to think about predictors from the revised bioecological theory perspective (Velez-Agosto et al., 2017) that does not conceptualize individual level, micro- and macro as influences that move along the proximal to distal continuum and conceptualizes them in a transactional framework.
@@
 
+We thank the Editor for this excellent suggestion; we have addressed this comment by adding the following paragraph:
+
+`r get_revision(manuscript = man, "velezagosto")`
+
 \RC{Also, as you mention in the limitations, it is possible that SES was ranked low because of the restricted range of SES that was represented in your sample.}
 
-If we understand correctly, this statement requires no further action.
+If we understand correctly, this is a statement and requires no further action.
 
@@
 
-While we are willing to adjust the language as requested,
-we do disagree that the statement does not fit the findings.
-First of all, note that the top 10 is an arbitrary criterion;
-placing the 'cutoff' for important predictors in different places lends more versus less support to the statement.
+Reviewer 2 suggested that we include a Figure to illustrate the distribution of variable importances by level of the bioecological model and by theme, to illustrate the two ways we interpret the pattern of findings (see Figure 3).
+We believe that this figure substantially clarifies our interpretation of the results.
 
-Variable importance follows an approximately exponential distribution,
-which means that the first variables are more important than later ones.
-At least the first two most important variables are both individual characteristics.
-If we interpret the variable importance like a scree plot (which is also an exponential distribution), then we observe a drop after the first 6 variables, and another drop after the 12th.
-Of the first 6, only one relates to the adolescent-reported *relationship with* parents, not *parenting behaviors*. The rest of these top 6 are all individual differences.
-Of variables 7-12, two relate to characteristics of the parents - mother's age and interparental conflict.
-Again, not parenting behavior.
-The rest of variables in this group relates to individual differences and adolescents' conflict behaviors.
-Variables 16 and 25 are the first parenting behaviors.
-I think that's a surprisingly low score.
+```{r figviolin, results='asis', out.width="20%", fig.cap="Variable importances clustered by theme (A) and level of the bioecological model (B).", eval = FALSE}
+knitr::include_graphics("violin.png")
+```
 
-REF
+Furthermore, to address the specific comment that "Predictors at the individual level were most important [etc]", we now refer to Figure 3 so readers can see the pattern we are describing.
+At the same time, we now explicitly acknowledge that there is substantial overlap in importance of predictors at the different levels:
+
+`r get_revision(manuscript = man, "overlap")`
 
@@
 
-We thank the Reviewer for pointing this out. We have clarified this, as requested.
+We thank the Reviewer for pointing this out. We have made the requested change.
 
@@
 
-One potential reason is that the performance of these fit statistics has been validated in the context of confirmatory factor analysis, and their performance is less well understood for complex models.
-This may affect the CFI more than, e.g., RMSEA, because CFI compares the fit of the theoretical model to a null model that may be inappropriate for longitudinal data.
-For this reason, it has been suggested to rely more on RMSEA for confirmatory studies (in this case, the growth model part of the analyses is confirmatory), and on CFI for exploratory purposes [@rigdonCFIRMSEAComparison1996].
-In line with the above, other research, which we now cite in the revision, has also found the RMSEA to be better at detecting misspecifications of growth curve models than CFI [@leiteDetectingGrowthShape2011]:
+The performance of these fit statistics has been validated mostly in the context of cross-sectional data, and their performance is less well understood for longitudinal data.
+This may affect the CFI more than, e.g., RMSEA, because CFI compares the fit of the theoretical model to a null model that may be less appropriate for repeated measures.
+We now address this point in the revision:
 
-`r get_revision(man, "cfi")`
+`r get_revision(manuscript = man, "cfi")`
 
@@
 
-We apologize; the Repository was still set to private at time of submission. This was intended to guarantee anonymous review. For this revision, it has been made public. The links should now work.
+We apologize; the Repository hosting these Supplementary Tables was still set to private at time of submission.
+This was intended to guarantee anonymous review.
+For this revision, it has been made public.
+The links now work.
 
@@
 
-To address this comment, and a comment by another Reviewer, we tested for differences on demographic characteristics and the dependent variable between participants with missing and complete data.
+<span id = "missingness">To address this comment, and a comment by another Reviewer, we tested for differences on demographic characteristics and the dependent variable between participants with missing and complete data.
 We now mention this in the revised manuscript:
 
-`r get_revision(man, "missing")`
+`r get_revision(manuscript = man, "missing")`
+</span>
 
@@
 
-We appreciate the Reviewer's kind words, but don't give us too much credit - PCA is the oldest dimension reduction technique (invented in 1901), and the main reason we use it here is because it makes fewer assumptions about the measurement model than, e.g., CFA.
+We appreciate the Reviewer's kind words.
+<span id = "extractpreds">
+The method used to extract predictors is less groundbreaking than the Reviewer gives us credit for.
+PCA is the oldest dimension reduction technique (invented in 1901), and the main reason we use it here is because it makes fewer assumptions about the measurement model than, e.g., CFA.
 
-We do appreciate the excellent suggestion of providing a specific example to help readers understand this approach. To address this comment, we have thus made the following changes:
+We appreciate the Reviewer's excellent suggestion of providing a specific example to help readers understand this approach. To address this comment, we have thus made the following changes:
 
@@
 We see that, across all scales, the first component explains between 1.6 - 17.46 times more variance than expected by chance.
-
-We hope this addresses the comment.
+</span>
 
@@
 
-`r get_revision(man, "homoscedastic")`
+`r get_revision(manuscript = man, "homogeneity")`
 
@@
 
-`r get_revision(man, "homogeneity")`
+`r get_revision(manuscript = man, "homogeneity2")`
 
@@
 
-We have simply rewritten this step in an attempt to address this comment. It now reads:
+We have rewritten this step in an attempt to address this comment. It now reads:
 
-`r get_revision(man, "stepiii")`
+`r get_revision(manuscript = man, "stepiii")`
 
@@
 
-`r get_revision(man, "fullygrown")`
+`r get_revision(manuscript = man, "fullygrown")`
 
@@
 
-`r get_revision(man, "semforests")`
+`r get_revision(manuscript = man, "semforests")`
 
@@
 
-`r get_revision(man, "predictions")`
+`r get_revision(manuscript = man, "predictions")`
 
@@
 
-`r get_revision(man, "examplenonlinear")`
+`r get_revision(manuscript = man, "examplenonlinear")`
 
@@
 
-`r get_revision(man, "limitations")`
+`r get_revision(manuscript = man, "limitations")`
 
@@
 One major caution is that such post-hoc analyses should not be over-interpreted; they are in no way a substitute for the (much more flexible) machine learning models.
-Still, if people e.g.
-want to take guidance from our list of important predictors, having an overview of univariate linear effects might be useful to help plan follow-up studies.
+Still, we see that having an overview of univariate linear effects might be useful, for example, to determine expected effect sizes if people want to take guidance from our list of important predictors to help plan follow-up studies.
 
-To address this comment, we have added the following paragraph and an additional Supplemental Table:
+To address this comment, we have therefore added the following paragraph and an additional Supplemental Table:
 
-`r get_revision(man, "posthoc")`
+`r get_revision(manuscript = man, "posthoc")`
 
@@
 
-`r get_revision(man, "errata")`
+`r get_revision(manuscript = man, "errata")`
 
@@
 
-`r get_revision(man, "includepersonality")`
+`r get_revision(manuscript = man, "includepersonality")`
 
@@
 
-This is correct; at the time of submission, the repository on GitHub was still set to private. For the revision, this has been resolved.
+This is correct; at the time of submission, the repository on GitHub was still set to private to safeguard blind review. For the revision, this has been resolved.
 
@@
 
-We apologize.
+We apologize for the inconvenience.
 The manuscript only has 42 pages; pages 43-65 are copies of the online supplemental materials (which are supposed to be accessed via the broken links, which have now been fixed).
@@
 
+We thank the Reviewer for their kind words and thoughtful comments, which have been instrumental in improving the manuscript.
+Please see our detailed responses to the specific comments below.
+
 \RC{1.
@@
 
-We understand why the Reviewer expects the analysis to include model tuning and cross-validation; these are standard procedures in machine learning.
+Model tuning and cross-validation are indeed industry standard procedures in machine learning.
 However, these methods do not (yet?) exist for SEM-forests.
@@
 The "outcome" of each regression tree in the SEM-forest is an entire SEM-model, for which there is no clear-cut case-specific error metric.
-It is important to note, however, that unlike other techniques, random forests typically achieve competitive performance even with default settings [@probstHyperparametersTuningStrategies2019].
-Thus, even when a model isn't tuned - or cannot be tuned, as is the case here - it should perform well.
+It is important to note, however, that unlike other machine learning techniques, random forests typically achieve competitive performance even with default settings [@probstHyperparametersTuningStrategies2019].
+Thus, even when a random forest isn't tuned - or cannot be tuned, as is the case here - it should perform well.
 Although we did attempt to explain these points in the previous version of the manuscript,
@@
 
-We thank the Reviewer for this excellent suggestion, and now provide this figure
+We thank the Reviewer for this excellent suggestion, and have now included a new Figure (Figure 3) that clusters variable importances by (inductive) theme as discussed in the results section, as well as by level of the bioecological model.
+We now refer to this Figure when discussing the relative importance of each theme / level.
+
+```{r}
+knitr::include_graphics("violin.png")
+```
 
@@
 
+We thank the Reviewer for pointing out the apparent inconsistency between the conclusion and results.
+
+`r get_revision(manuscript = letter, "parents", quote = F)`
+
+We also believe that the Figure above helps clarify these results.
+
 \RC{4.
@@
 
+We agree with the Reviewer that is is prudent to caution against undue causal interpretation of the results - which we have done in the Limitations section:
+
+`r get_revision(manuscript = man, "causality")`
+
+We do believe that the quoted statement is correct and does not require further modification, however, as it does not relate to causal interpretation of the results, but to the affordances of the method: latent class analysis *cannot* answer the question "Why?" some adolescents follow similar causal trajectories,
+whereas SEM-forests *can* answer this question, because cases are clustered based on predictor variables (of course with the caveat that causal interpretation is contingent on study design or theoretical justification).
+
 \RC{5.
@@
 
-In principle, this is possible.
-In practice, it is not, for the following reason:
-1) Computing univariate partial dependence profiles is already so computationally expensive that I cannot repeat it without additional funding
-2) There is no way of determining what bivariate interactions might be important in SEM-forests analysis (because there is no applicable analogy for Friedman's H statistic)
-3) So one would have to examine all bivariate interactions by hand, for all model parameters. This comes down to interpreting 84*84*3 = 21168 plots.
+We appreciate the suggestion and had considered a similar approach - but it is not practically feasible for the following reasons:
+
+1. Computing univariate partial dependence profiles is extremely computationally expensive; bivariate partial dependence is prohibitively so.
+2. There is no a-priori way to determine which bivariate interactions might be important in SEM-forests analysis (because there is no applicable analogy for Friedman's H statistic).
+3. Consequently, one would have to examine all bivariate interactions by hand. This comes down to interpreting 84*84*3 = 21168 plots. This is not possible.
+
+With this in mind, we think it would be preferable *not* to muddy the waters by citing literature on interaction effects for random forests; not until suitable methods have been developed for SEM-forests as well.
 
@@
 
-Again, in principle this is correct, but in practice, no method exists to obtain accumulated local effects for SEM-forests.
-For the purpose of this paper, I developed a novel method to obtain marginal dependencies for growth curves, but I am not familiar with Apley's method so I cannot implement it (and either way, that seems out of scope for a developmental paper).
+Again, we thank the Reviewer for the suggestion;
+in principle this is correct, but in practice, no method (currently?) exists to obtain accumulated local effects for SEM-forests.
+The Reviewer is citing cutting-edge developments in the area of classic random forests; it will take a while before these developments spill over into the SEM-forests literature, if ever.
 
@@
 
+We respectfully believe that this may be an oversight on the Reviewer's part, because the manuscript says:
+
+`r get_revision(manuscript = man, "singlefactor")`
+
+We do not believe further action is required on this point - but if we missed something, please let us know.
+
 \RC{2. On page 10, please provide more explanation of the discrepancy between absolute and comparative fit of the metric invariant model.}
+
+The performance of these fit statistics has been validated mostly in the context of cross-sectional data, and their performance is less well understood for longitudinal data.
+This may affect the CFI more than, e.g., RMSEA, because CFI compares the fit of the theoretical model to a null model that may be less appropriate for repeated measures.
+We now address this point in the revision:
+
+`r get_revision(manuscript = man, "cfi")`
 
@@
 
+We apologize for the oversight; the y-axis was indeed reversed in this Figure. We have corrected this mistake.
+
 \RC{4. On page 12, please speak to the reasonableness of the assumption of homoscedastic residuals over time.}
+
+We assumed homoscedastic residuals because otherwise the model would be too complex to estimate.
+We would have more parameters for the residuals than for the rest of the model, and the SEM-forest analysis would thus be primarily driven to explain differences in residuals, instead of in growth curves.
+
+A SEM-forest method estimates tens, or hundreds of thousands of SEM models in its different nodes. It is not possible to test model assumptions in each of those nodes.
+This applies not only to the assumption of homogeneity of residuals, but to all other assumptions, for example, about normality of residuals, or the shape of the growth trajectories.
+
+We have attempted to address this comment as follows: By explaining our decision to assume homoscedasticity:
+
+`r get_revision(manuscript = man, "homogeneity")`
+
+And by addressing it further in the Limitations section:
+
+`r get_revision(manuscript = man, "homogeneity2")`
 
@@
 
+We have corrected this mistake.
+
 \RC{6. I think Figures 5 and 6 could be moved to supplemental if space is tight.}
+
+We thank the Reviewer for the suggestion,
+and defer to the Editor on this point.
 
@@
 
+We have clarified this by writing:
+
+`r get_revision(manuscript = man, "motherfather")`
+
 \RC{8. The supplemental table 1 didn’t render well in the PDF.}
 
+We apologize for the inconvenience.
+These Supplemental Tables will not be part of the printed manuscript; we were required to attach hard copies of the online supplemental tables in the submission portal, and these were evidently not formatted correctly.
+They are just spreadsheets that can be opened in a web-browser or excel.
+
 # Reviewer: 3
-
-\RC{Comments to the Author
-The current study used machine learning to identify importance of predictors of adolescent emotion regulation trajectories.
-The study used a large sample of adolescents and included multiple reporters (although adolescent report was featured most heavily).
-Despite these strengths, I have a number of concerns about the manuscript in its present form that make its contribution unclear at this point.
-I list these concerns in roughly the order of appearance in the manuscript.}
 
@@
 
+We appreciate these suggestions and the caution about abstract length; we have cut the abstract by half and attempted to incorporate the Reviewer's suggestions. The Abstract now reads:
+
+`r get_revision(manuscript = man, "abstract")`
+
 \RC{Please provide citations for the statement in the first paragraph on page 4, “These difficulties in emotion regulation can cause…” and provide examples of why these incur societal cost.}
 
+The statement is a continuation of the preceding sentence,
+which is supported by two references. 
+To address this comment, we now also added one of these references to the second sentence, and emphasize the connection between the sentences by re-iterating that the societal cost is incurred via emotion regulation-related mental health problems:
+
+`r get_revision("cost", man)`
+
 \RC{I would suggest removing reference to “destinies” on page 4 and elsewhere, as this word connotes fixedness in developmental outcomes, which does not reflect contemporary developmental views.}
+
+We agree with the Reviewer that the word "destinies" has unintended connotations of fixedness in developmental outcomes.
+We certainly want to avoid this implication, and have rephrased all occurrences of "diverging destinies" with "diverging trajectories", which is a statistically accurate description of our outcome of interest.
 
@@
 The Discussion has sentences that are fragments or are missing words.}
+
+We have attempted to correct all such errors.
 
@@
 
-`r get_revision(man, "personcentered1")`
+`r get_revision(manuscript = man, "personcentered1")`
 
@@
 In fact, there is so much work on this topic that it would be a tall order to do it all justice.
-We cite a recent (2021) comprehensive systematic review of 6305 studies on predictors of adolescent emotion regulation.
-From those 6305 studies, we selected the ones that seemed most relevant for the present paper.
-If the Reviewer feels that *specific* studies are missing here, we are happy to include them of course.
-At the same time, we cannot strive to recreate a comprehensive review here.
-This also seems redundant, given that one was just recently published, and overly ambitious, given the page limits and attention that must be devoted to explaining the novel methods.
+At present, we cite a total of 17 review articles in the introduction,
+including a recent (2021) comprehensive systematic review of 6305 studies on factors associated with adolescent emotion regulation.
+
+Given the large prior literature,
+we found it necessary to rely primarily on review articles to do justice to prior work.
+We cannot strive to recreate a comprehensive review here - 
+and this also seems redundant, given that many of the cited review articles are very recent (i.e., from 2019-2021),
+and seems overly ambitious, given the page limits and attention that must be devoted to explaining the novel methods,
+and spurious, given that this is an exploratory study and thus has no hypotheses that must be supported with prior literature.
 The present study builds upon prior reviews of the literature,
@@
 
+Of course we are willing to include **specific** omitted literature, but it is not clear from the comment which specific references should be included.
+
 To address this comment, we now emphasize more strongly that there is an abundance of prior literature on predictors of adolescent emotion regulation, e.g.:
 
-`r get_revision(man, "abundantlit")`
+`r get_revision(manuscript = man, "abundantlit")`
 
@@
 
-We respectfully believe that this may be an oversight on the Reviewer's part, as comprehensive descriptive statistics - including those requested by the reviewer - are reported in Tables 1 and 2.
+Comprehensive descriptive statistics about all available demographic characteristics are reported in Tables 1 and 2.
+To address this comment, we now stress this in the Participants section:
+
+`r get_revision("demographics", man)`
 
@@
 
-In the original manuscript, we stated that:
+We respectfully believe that this may be an oversight on the Reviewer's part, as the original manuscript stated that:
 
@@
 
-As two Reviewers have requested this information, we understand that this is information readers will want to know.
-Note, however, that the methods we use are valid regardless of whether missingness is related to other variables in the dataset or not.
-To address this comment, we performed the suggested check, and report:
+`r get_revision("missingness", letter, quote = F)`
 
-`r get_revision(man, "missing")`
+Note, however, that the methods we use are valid regardless of whether missingness is related to other variables in the dataset or not (i.e., valid for MCAR and MAR).
 
@@
 
-We thank the Reviewer for pointing out this unfortunate choice of words. 
+We thank the Reviewer for pointing out this unfortunate choice of words, which should have been flagged during proof reading.
 Entropy is not a criterion for class enumeration, but for evaluating the quality of the chosen solution (see Jung \& Wickrama, 2008). In the Revision, we have attempted to clarify this as follows:
 
-`r get_revision(man, "entropy")`
+`r get_revision(manuscript = man, "entropy")`
 
@@
 
-In line with the Reviewer's expectation, we indeed used the median and 16th and 84th percentile.
-Percentiles do not assume a normal distribution of the predictor, and neither does the model, so this is a good fit.
-As we explain in the text, the 16th and 84th percentiles correspond to +/- 1SD of a standard normal distribution.
-We then use this label, "+/- 1SD", in the plots, in an attempt to make the results more understandable.
-Perhaps this was a misguided decision, but we were not sure whether the readership would be familiar with percentiles.
+In line with the Reviewer's expectation, we did in fact use the median and 16th and 84th percentile.
+The confusion arises from our attempt to explain to the readership what these percentiles mean, by drawing a comparison to +/- 1SD.
 
-In the Revision, we have attempted to resolve this issue by stating:
+We use percentiles is because they do not assume a normal distribution of the predictor, and neither does the model, so this is a good fit.
+As we explain in the text, the 16th and 84th percentiles and the median correspond to +/- 1SD and the mean of a standard normal distribution.
+We then use the labels "+/- 1SD" in the plots, in an attempt to make the results more understandable, and explain that these correspond to percentiles of a normal distribution in the text.
 
-`r get_revision(man, "percentiles")`
+In the Revision, we have attempted to address this issue by further clarifying that :
+
+`r get_revision(manuscript = man, "percentiles")`
 
@@
 
-REF
+`r get_revision(manuscript = letter, "parents", quote = F)`
+
+To address this comment, we now also mention highly ranked parent-related constructs in the first paragraph of the Discussion:
+
+`r get_revision(manuscript = man, "highrankingparent", quote = F)`
 
@@
 
-We appreciate this point, and the writing here should certainly be more nuanced (that was an oversight).
+This is a fair point, and the writing here should have been much more nuanced (that was an oversight).
 We have rephrased this as follows:
 
-`r get_revision(man, "riskassessment")`
+`r get_revision(manuscript = man, "riskassessment")`
 
@@
 
-`r get_revision(man, "culture")`
+`r get_revision(manuscript = man, "culture")`
+
+And also about cultural bias in prior literature:
+
+`r get_revision(manuscript = man, "globalnorth")`
 

### 2022-11-09 manuscript/manuscript.Rmd
Add supplemental tables

@@
   - name: "Caspar J. Van Lissa"
-    affiliation: "1,2"
+    affiliation: "1"
     corresponding: yes
-    address: "Padualaan 14, 3584CH Utrecht, The Netherlands"
-    email: "c.j.vanlissa@uu.nl"
+    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
+    email: "c.j.vanlissa@tilburguniversity.edu"
   - name: "Lukas Beinhauer"
+    affiliation: "2"
+  - name: "Susan Branje"
     affiliation: "3"
-  - name: "Susan Branje"
-    affiliation: "4"
   - name: "Wim H. J. Meeus"
-    affiliation: "4"
+    affiliation: "3"
 
@@
   - id: "1"
-    institution: "Utrecht University, department of Methodology & Statistics"
+    institution: "Tilburg University, department of Methodology & Statistics"
   - id: "2"
-    institution: "Open Science Community Utrecht"
+    institution: "Helmut-Schmidt-Universität, Department of Methodology and Statistics for Psychology"
   - id: "3"
-    institution: "Helmut-Schmidt-Universität, Department of Methodology and Statistics for Psychology"
-  - id: "4"
     institution: "Utrecht University, department of Youth and Family"
@@
 authornote: |
-  This is a preprint paper, generated from Git Commit # `r gert::git_commit_id()`.
-CVL conceived of the study, conducted the final analyses, and wrote the draft paper.
+  This is a preprint paper, generated from Git Commit # `r substring(gert::git_commit_id(), 1, 8)`.
+CVL conceived of the study, conducted the reported analyses, and wrote the draft paper.
 LB conducted preliminary analyses and gave feedback on the draft.
@@
 This work was funded by a NWO Veni Grant (NWO Grant Number VI.Veni.191G.090), awarded to the lead author.
-Data of the RADAR (Research on Adolescent Development And Relationships) study were used, and is available under restricted access (https://doi.org/10.17026/dans-zrb-v5wp).
+Data of the RADAR (Research on Adolescent Development And Relationships) study were used, and are available under restricted access (https://doi.org/10.17026/dans-zrb-v5wp).
 RADAR has been financially supported by main grants from the Netherlands Organisation for Scientific Research (GB-MAGW 480-03-005, GB-MAGW 480-08-006), Stichting Achmea Slachtoffer en Samenleving (SASS), the Netherlands Organisation for Scientific Research to the Consortium Individual Development (CID; 024.001.003), a grant of the European Research Council (ERC-2017-CoG - 773023 INTRANSITION), and various other grants from the Netherlands Organisation for Scientific Research, VU University Amsterdam, and Utrecht University.
 The funders had no role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript.
-If the regulations of the journal allow it, the participating children, parents and schools are to be thanked as well.
+We thank the participating children, parents and schools.
 
@@
   | • Individual differences (e.g., in personality, motivation, and empathy) are among the most important predictors of emotion regulation development.
-  | • Parenting may be less relevant for emotion regulation development than previously thought, and negative parenting may be more impactful than positive parenting.
+  | • Specific parenting practices may be less relevant than previously thought, and negative parenting practices may be more impactful than positive ones.
   | • Important predictors most suitable for early risk assessment include regularly administered personality inventories, and readily observable indicators like externalizing problems and conflict behavior.
   | • Future theory should emphasize proximal- over distal processes, focus on good-enough parenting, and reassess common perceptions about “risky behaviors” like delinquency, substance use, and bullying.
-  | This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM-forests.
-Candidate predictors were selected based on a prior text mining systematic review of 6305 papers.
+  | [As 20% of adolescents develop emotion regulation difficulties, it is important to identify important early predictors thereof.
+Using the machine learning algorithm SEM-forests, we ranked the importance of (87) candidate variables assessed at age 13 in predicting quadratic latent trajectory models of emotion regulation development from age 14-18.
 Participants were 497 Dutch adolescents.
-Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model.
-SEM-forests were used to group adolescents based on their predictor scores, while maximizing the similarity of developmental trajectories within each group.
-Compared to other person-centered approaches, this method not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories.
-The results indicated that the most important early predictors (at age 13) were individual differences (e.g., in personality), conflict behaviors with parents and friends, and internalizing and externalizing behavior.
-Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use.
-The implications of these findings for theory formation and intervention are discussed, and we present an open source risk assessment tool, ERRATA.
+Results indicated that the most important predictors were individual differences (e.g., in personality), aspects of relationship quality and conflict behaviors with parents and peers, and internalizing and externalizing problems.
+Relatively less important were demographics, bullying, delinquency, substance use, and specific parenting practices – and negative parenting practices ranked higher than positive ones.
+We discuss implications for theory and interventions, and present an open source risk assessment tool, ERRATA.]{#abstract}
 
@@
 Adolescents experience biological, cognitive, and social changes that create an opportunity for positive emotion regulation development [@steinbergAgeOpportunityLessons2014].
-Still, as many as 20% of adolescents instead experience severe emotion regulation difficulties [@leeAdolescentMentalHealth2014],
+Still, [as many as 20% of adolescents instead experience severe emotion regulation difficulties [@leeAdolescentMentalHealth2014],
 with potentially lifelong implications for mental health and well-being [@aldaoEmotionregulationStrategiesPsychopathology2010].
 These difficulties in emotion regulation can cause substantial distress to the individual,
-and incur a sizable societal cost.
+and incur a sizable societal cost via their implications for mental health [@leeAdolescentMentalHealth2014].]{#cost}
 A comprehensive understanding of the risk factors that predispose some adolescents to emotion dysregulation could aid in the development of effective screening and interventions.
-<span id = "abundantlit">There is already an abundance of empirical research on emotion regulation and its predictors; a recent systematic review identified 6305 relevant publications [@vanlissaMappingPhenomenaRelevant2021].
-Still, that review identified specific limitations of the existing literature: many studies are narrow in scope, considering only a few predictors and thus missing the relative importance compared to other predictors.</span>
-This problem is exacerbated because the literature is fragmented across sub-disciplines, each of which examines mostly predictors within its scope.
-And finally, the literature lacks a unifying theoretical framework that is both specific to adolescence and comprehensive in scope [see @bariolaChildAdolescentEmotion2011].
+<span id = "abundantlit">There is already an abundance of empirical research on adolescent emotion regulation and its predictors; a recent systematic review identified 6305 relevant publications [@vanlissaMappingPhenomenaRelevant2021].
+Still, that review identified specific limitations of the existing literature:
+many studies narrowly focus on a few predictors within a specific research area,
+and thus fail to provide insight into these predictors' relative importance compared to other predictors, including those in other domains.
+Relatedly,
+the literature lacks a unifying theoretical framework that is both specific to adolescence and comprehensive in scope [see @bariolaChildAdolescentEmotion2011].</span>
 
@@
 Specifically,
-<span id = "personcentered1">the SEM-forests algorithm [@brandmaierTheoryguidedExplorationStructural2016] was used as a person-centered analysis technique,
+<span id = "personcentered1">the SEM-forests algorithm [@brandmaierTheoryguidedExplorationStructural2016], a person-centered analysis technique, was used
 to identify the most important early indicators (at age 13) of developmental trajectories
 of emotion regulation across adolescence (age 14 through 18).
-The resulting insights indicate which factors are most predictive of adolescents' diverging trajectories of emotion regulation development, contributing to our understanding of why some youth flourish while others struggle [@croneUnderstandingAdolescencePeriod2012].</span>
-Armed with this knowledge, future research and interventions can focus on the most important predictors.
-Moreover, these insights can guide the formation of overarching theory that adequately represents essential phenomena.
-<!-- complement existing theory on adolescent emotion regulation using data-driven insights from machine learning. -->
+The results indicate which factors are most predictive of adolescents' diverging trajectories of emotion regulation development, contributing to our understanding of why some youth flourish while others struggle through adolescence [@croneUnderstandingAdolescencePeriod2012].</span>
+These insights can guide development of interventions that focus on the most important predictors,
+and direct the formation of overarching theory that adequately represents essential phenomena.
 
@@
 staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
-The emotional salience of social stimuli is intensified, while the development of cognitive capacities lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
+The emotional salience of social stimuli is intensified, while the development of cognitive capacities lags behind [see @caseyBrakingAcceleratingAdolescent2011; @vanlissaLongitudinalInterplayAffective2014].
 This imbalance might help motivate adolescents to engage in adaptive exploration:
 Seeking out experiences that challenge their socio-emotional competences and stimulate growth.
-At the same time, this imbalance may render some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
-For example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
-or persevere in pursuing adaptive goals when facing obstacles.
+At the same time, this imbalance may render some adolescents vulnerable to emotion dysregulation if they are not yet able to regulate the pursuit of unhealthy incentives,
+or persevere in pursuing adaptive goals when facing obstacles [@leeAdolescentMentalHealth2014].
 Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions [@craccoEmotionRegulationChildhood2017],
 which has been linked to the emergence of several mental disorders [@schaferEmotionRegulationStrategies2017].
-<!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
-<!-- but renders some of them vulnerable to emotion dysregulation1,10.
-Unfortunately, there is little theory about which risk factors and environmental hazards render adolescents susceptible to emotional difficulties11.
--->
-<!-- Acquiring adaptive emotion regulation skills in adolescence plays a key role in this regard7,20, because t -->
 
@@
 Such studies consistently reveal large between-individual heterogeneity [see @craccoEmotionRegulationChildhood2017; @vanlissaRoleFathersMothers2019; @zimmermannEmotionRegulationEarly2014; @reindlSocializationEmotionRegulation2016].
-Although some studies have examined, on average, how strongly particular risk factors are associated with developmental trajectories
-
+<!-- Although some studies have examined, on average, how strongly particular risk factors are associated with developmental trajectories - most have done  REF write something about other studies examining risk factors -->
 An important next step is to identify which risk factors render some adolescents susceptible to emotion dysregulation.
@@
 The secondary aim was to probe the nature of the relationship of each predictor with these trajectories.
-This involves an important challenges, as
+This involves an important challenge, as
 the number of potentially relevant risk factors is too large for conventional methods.
@@
 All independent variables were assessed in the first measurement wave.
+[The sample demographic information is reported in Tables \@ref(tab:tabcat) and \@ref(tab:tabnum).]{#demographics}
 The dependent variable was measured from the second to the sixth wave.
@@
 We used latent variable scores (instead of mean scores) for further analyses.
-To this end, we estimated a CFA with all DERS items loading on one latent variable per measurement wave.
+[To this end, we estimated a CFA with all DERS items loading on one latent variable per measurement wave.]{#singlefactor}
 Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the sample average trajectory.
@@
 The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
-[Note that the RMSEA has been found to be more sensitive to misspecifications of growth curve models [@leiteDetectingGrowthShape2011]]{#cfi}.
+[Note that the RMSEA has been found to be more appropriate for longitudinal data, possibly because the null-model on which the CFI is based is less appropriate for repeated measures [@leiteDetectingGrowthShape2011]]{#cfi}.
 As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
@@
 ```{r figclass, fig.cap="Latent class analysis of risk profiles."}
-knitr::include_graphics("../classplot.pdf")
+knitr::include_graphics("../classplot.png")
 ```
@@
 
-[For continuous predictors, three values of $s$ were used - namely the percentiles, computed on the observed data, corresponding to the median, $+/- 1 SD$ of a standard normal distribution (i.e., the $16^{th}$ and $84^{th}$ percentile).
-For convenience, these percentiles are labeled as $+/- 1 SD$ in Figure \@ref(fig:figpdp1)-\@ref(fig:figpdp4).]{#percentiles}
+[For continuous predictors, three values of $s$ were used - namely the percentiles, computed on the observed data, corresponding to the mean and $+/- 1 SD$ of a standard normal distribution (i.e., the $50^{th}$, $16^{th}$ and $84^{th}$ percentiles).
+For convenience, these percentiles are labeled as *mean* and $+/- 1 SD$ in Figure \@ref(fig:figpdp1)-\@ref(fig:figpdp4).]{#percentiles}
 For categorical predictors, all unique values were used.
@@
 ```{r figvarimp, fig.cap="Predictors ranked in order of relative importance. m: mother, f: father, b: best friend, s: sibling."}
-knitr::include_graphics("../pcomb.pdf")
+#readRDS("../varimp_comb2_03-08-2022.RData")
+knitr::include_graphics("../pcomb2.png")
 ```
@@
 Examining the variable importance rankings,
-several themes emerged.
-Specifically, results indicated that several individual differences - personality and related constructs - were ranked among the most important predictors.
-For example, three of the big five personality dimensions rank among the top six predictors: Neuroticism (1), extraversion (4), and agreeableness (6).
+several themes emerged (see Figure \@ref(fig:figthemes) panel A).
+Specifically, results indicated that among the most important predictors were personality and personality-related constructs.
+Three of the big five personality dimensions ranked among the top six predictors: Neuroticism (1), extraversion (4), and agreeableness (6).
 The Behavioral Inhibition Scale (BIS, 2), which reflects motivation to avoid aversive outcomes, ranked highly - although the BAS (73), its counterpart which measures motivation to approach goal-oriented outcomes,
@@
 
-A second emerging theme is that most constructs related to conflict frequency and behavior ranked highly, and few ranked low.
+```{r figthemes, fig.cap="Variable importances clustered by theme (A) and level of the bioecological model (B)."}
+knitr::include_graphics("../violin.png")
+```
+
+A second emerging theme was that most constructs related to conflict frequency and behavior ranked highly, and few ranked low.
 For instance, adolescent-perceived interparental conflict (8),
@@
 
-A third emerging theme is that parenting was relatively underrepresented among the most important predictors;
-the highest-ranking parental predictor was balanced relatedness (3),
-a construct that represents the extent to which parents support and respect the adolescent's autonomy.
-It further appeared that negative and intrusive aspects of parenting were more predictive than other positive practices:
-Prohibition and control of friendship choice (16),
-perceived intrusiveness and meddling (25),
+<span id = "parenting">A third emerging theme is that specific parenting practices were relatively underrepresented among the most important predictors.
+Parenting practices are defined as specific parental actions related to child-rearing,
+for example, support and dimensions of (behavioral and psychological) control [@petersonFamilySocialCompetence1990].</span>
+Negative and intrusive parenting practices were more predictive than positive ones:
+Of these, prohibition and control of friendship choice (16) ranked highest, followed by intrusiveness and meddling (25),
 psychological control (27),
 and supportive criticism (33).
-Positively valenced dimensions of parent-adolescent relationship quality
-like maternal (54) and paternal (55) support were ranked low,
+Positive parenting practices,
+like maternal (54) and paternal (55) support,
+were ranked low,
 and neutrally valenced aspects like parental monitoring (sollicitation 71, adolescent disclosure 72) were ranked very low.
+This is interesting, given the extensive research attention into the effect of these specific parenting practices.
 
-Finally, an emerging theme is that internalizing- and externalizing symptoms
-often associated with emotion regulation ranked moderately in terms of predictive importance.
+<span id = "parentrelated">Several parent-related constructs scored higher than the aforementioned child rearing behaviors;
+the highest of these was balanced relatedness (3),
+which represents parents' tolerance for adolescents' differing opinions and is related to parent-child intimacy and autonomy [@shuhnanAdolescentIntimacyRevisited1997].
+Other parent-related constructs that scored relatively highly were interparental conflict, which had a negative effect on emotion regulation (8);
+like balanced relatedness, conflict likely contributes to the emotional family climate [@morrisRoleFamilyContext2007].
+Finally, maternal age (7) had a protective effect, which is well-supported by prior literature [@zondervan-zwijnenburgParentalAgeOffspring2020].]</span>
+<!-- Add reference -->
+
+Finally, an emerging theme was that internalizing- and externalizing symptoms
+often associated with emotion regulation ranked relatively highly in terms of predictive importance.
 Of these, parent-reported externalizing behavior (5) ranked highest, followed by adolescent-reported anxiety (15) and depression (17).
@@
 It is further of note that some themes that are considered to be theoretically relevant ranked low.
-For instance, indicators of maternal (86) and paternal (87) socio-economic status were ranked the lowest,
-as were other family demographic indicators, like religion (81, 82).<!--SB: Why is this important?
+This was the case, for example, for demographic variables like maternal (86) and paternal (87) socio-economic status and family religion (81, 82).<!--SB: Why is this important?
 CJ: Religion is part of moral development, SES, and is directly related to emotion regulation: https://link.springer.com/chapter/10.1007/978-94-017-8950-9_13-->
@@
 such as bullying (60, 84) and victimization (51, 62, 83),
-and contact with delinquent peers (85) also ranked low.
+and contact with delinquent peers (85) also ranked unexpectedly low.
 The same applied to substance use;
@@
 and use of the conflict resolutions styles engagement and withdrawal.
-Intrusive parenting behaviors were also negative predictors, including psychological control, control of friendships, paternal power, and constructive criticism.
+Intrusive parenting practices were also negative predictors, including psychological control, control of friendships, paternal power, and constructive criticism.
 All empathy dimensions except perspective taking predicted greater emotion dysregulation,
@@
 [Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
-All models had good fit ($RMSEA < .07$, $CFI > .99$, $SRMR < .04$).
+All models had good fit ($RMSEAs < .07$, $CFIs > .99$, $SRMRs < .04$).
 Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
@@
 there were differences in relative importance.
-Individual differences in personality, motivational system, self-concept clarity, and personal distress were among the most important predictors.
-Conflict frequency and behavior also ranked highly.
-Parenting ranked relatively low, with two notable exceptions:
-First, balanced relatedness, a measure of autonomy support, was an important protective factor, and second,
-among all remaining parenting constructs, negative and intrusive dimensions were more predictive (as risk factors) than positive or neutral parenting practices.
-Themes that ranked notably low were family demographics (SES and religion) and bullying, victimization and delinquency, and substance use.
+Most of the important predictors were individual differences in personality, the motivational system, self-concept clarity, and empathy.
+<span id = "highrankingparent">Although a few parent-related constructs ranked highly,
+these were primarily related to the family emotional climate (balanced relatedness and interparental conflict) and demographics (maternal age).</span>
+Specific parenting practices, however, ranked lower than might have been expected given prior literature [e.g., support, control, psychological control, @morrisRoleFamilyContext2007].
+Of these specific child-rearing behaviors,
+negative and intrusive dimensions were more predictive (as risk factors) than positive or neutral parenting practices.
+Adolescents' conflict frequency and behavior also ranked highly, across relationships with parents and peers, which suggests that these variables capture general aspects of pugnaciousness and conflict resolution skills.
+Themes that ranked lower than might have been expected given prior research were family demographics (SES and religion) and bullying, victimization and delinquency, and substance use.
 The relatively low importance of demographics might be attributable to a restriction of range, as the sample was relatively homogenous.
@@
 Relating these results to prior theory,
-it is noteworthy that predictive importance of different themes appears to correspond to the levels of the bioecological model [@bronfenbrennerBioecologicalModelHuman2007]:
-Predictors at the individual level were most important,
+it is noteworthy that predictive importance of different themes appears to correspond to the levels of the bioecological model [@bronfenbrennerBioecologicalModelHuman2007], see Figure \@ref(fig:figthemes) panel B.
+[Although there was substantial overlap between predictors at different levels of the bioecological models,
+several predictors at the individual level achieved the highest importance rankings]{#overlap},
 <!-- SB: this is not very surprising given that many of the behaviors assessed require ER (e.g.
 becoming angry in a conflict is an indication of low ER, and individuals high in Neuroticism also have fluctuating emotions that they find difficult to control/regulate) -->
-followed by those at the microsystem level, such as aspects of the relationship with parents,
+followed by predictors at the microsystem level, particularly those representing the emotional family climate,
 followed by the mesosystem, including aspects of the peer network (e.g., bullying/victimization and delinquency), school performance, and socio-economic status,
@@
 <!-- Yet perhaps there was too little variation in relevant macro factors (ethnicity, SES/poverty, cultural differences in ER) -->
+<span id = "velezagosto">Also note that a recent perspective on the bioecological theory emphasizes that distal factors like culture are enacted proximally in different domains of the microsystem [@velez-agostoBronfenbrennerBioecologicalTheory2017].
+Recognizing that culture is a product of human activity leaves more room for heterogeneity in cultural heritage and practices.
+This perspective might help explain why we found that distal factors were relatively less important predictors.
+If factors like religiosity and SES are (partly) enacted in relationships with parents and peers,
+their distal effects might be causally screened off by conditioning on proximal effects [@pearlCausalDiagramsEmpirical1995].
+This could explain their relatively lower variable importance.
+</span>
+A possible alternative explanation is that our sample was relatively homogenous in terms of characteristics typically attributed to the macrosystem.
 
@@
 This view is certainly supported by our findings,
-and yet research on adolescents does not routinely assess the role of personality alongside factors of particular interest in our discipline, such as parenting, peer relationships, and problem behavior [@vanlissaMappingPhenomenaRelevant2021].
+and yet research on adolescents does not routinely assess the role of personality alongside factors of particular interest in our discipline, such as parenting practices, peer relationships, and problem behavior [@vanlissaMappingPhenomenaRelevant2021].
 This is certainly a promising area for future research.</span>
@@
 Counter to this theory, however, parent characteristics (e.g., parental emotion regulation) ranked low in importance.
-Parenting practices also ranked relatively low - with the notable exception of negative parenting behaviors.
+Parenting practices also ranked relatively low, and negative practices ranked higher than positive ones.
 <!-- SB: here you should really go into more depth.
@@
 This finding is consistent with recent studies that, using innovative methods to study processes within families,<!--The finding that positive parenting practices ranked low -->
-have found limited evidence for effects of parenting behaviors on emotional development in adolescence [@vanlissaRoleFathersMothers2019; @mastrotheodorosFamilyFunctioningAdolescent2020] and childhood [@vanlissaMothersFathersQuantitative2020].
+have found limited evidence for effects of specific parenting practices on emotional development in adolescence [@vanlissaRoleFathersMothers2019; @mastrotheodorosFamilyFunctioningAdolescent2020] and childhood [@vanlissaMothersFathersQuantitative2020].
 <!-- aren’t individual differences the result of the environment where those people were growing up, hence parental style as well? -->
-We must thus consider the possibility that parenting behavior, at least in adolescence, is less influential than once was thought.
-The absence of negative behavior and constructive communication in conflict situations might be more important.
+We must thus consider the possibility that specific child-rearing behaviors might be less influential than once was thought, at least in adolescence.
+To ensure adaptive emotion regulation development,
+it might be more important that parents avoid negative parenting practices and engage adolescents in constructive communication in conflict situations.
 <!-- Future theories of adolescent emotion regulation development should thus at least go beyond parents' role. -->
@@
 Our results showed no systematic differences between mothers and fathers.
-Anecdotally however, out all parent-related constructs, constructive conflict resolution with fathers ranked highest, and out of all parenting behaviors, paternal power ranked highest.
+[Anecdotally however, out all parent-related constructs that were measured separately for mothers and fathers]{#motherfather}, constructive conflict resolution with fathers ranked highest, and out of all parenting practices, paternal power ranked highest.
 These observations validate the notion that fathers matter.
@@
 A prior text mining systematic review comprehensively mapped the phenomena studied in this field [@vanlissaMappingPhenomenaRelevant2021].
-It showed that parenting, substance use, and bullying were among the phenomena most studied in research on emotion regulation.
+It showed that parenting practices, substance use, and bullying were among the phenomena most studied in research on emotion regulation.
 These phenomena all ranked relatively low in empirical importance in the present study.
@@
 based on the variable importance rankings:
-conflict resolution behavior, and (negative) parenting.
+conflict resolution behavior, and (negative) parenting practices.
 
@@
 The same evidence indicates that conflict resolution training caused a decrease in externalizing problems (another important predictor), suggesting potential compounding benefits.
-Parenting behavior is another promising target for intervention,
+Parenting practices are another promising target for intervention,
 as indeed it has historically been.
-Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting behaviors.
+Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting practices.
 By contrast, promoting positive parenting practices may have limited potential for impact.
 With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
-Although one meta-analysis did report large effects of the Triple P-Positive Parenting Program on parental outcomes - including some negative parenting behaviors [@fletcherImpactBehaviouralParent2011],
+Although one meta-analysis did report large effects of the Triple P-Positive Parenting Program on parental outcomes - including some negative parenting practices [@fletcherImpactBehaviouralParent2011],
 another meta-analysis called these findings into question,
@@
 is less studied.
-The relatively higher importance rankings for negative versus parenting is in line with the well-established *negativity bias*,
-that negative stimuli are more psychologically impactful than positive ones,
+The relatively higher importance rankings for negative versus positive parenting practices is in line with the well-established *negativity bias*,
+that negative stimuli are more psychologically impactful,
 which has also been found in children [@vaishNotAllEmotions2008].
-Future parenting interventions might take this into consideration by devoting sufficient attention to the elimination of negative parenting practices,
-and including children's outcomes, including emotion regulation, into consideration.
+Future parenting interventions might take this into account by devoting sufficient attention to the elimination of negative parenting practices,
+and by considering children's outcomes,
+including emotion regulation.
 Simply put, the present results suggest that the best advice for parents might be to support their individuating adolescents' autonomy,
@@
 This answers the question why some adolescents experience more difficulties in emotion regulation than others.
+[Note however that, as our dataset describes naturally occurring development,
+any causal interpretation of the results must be substantiated by theory or corroborated with experimental research.]{#causality}
 Another contribution is that the results have clear implications for theory, assessment, and intervention.
@@
 have been found to influence people's motivation to regulate emotions, as well as the strategies they use and their adaptiveness [@fordCultureEmotionRegulation2015].
-In cross-cultural comparisons, the Netherlands has been found (on average) to ascribe to similar cultural values as large Anglophonic countries (e.g., the United States, Canada, and Australia) [@terlutterGLOBEStudyApplicability2006].
+In cross-cultural comparisons, cultural values in Dutch samples have been found (on average) to be similar to those in large Anglophonic countries (e.g., the United States, Canada, and Australia) [@terlutterGLOBEStudyApplicability2006].
 Whether our findings also generalize to populations that adhere to different cultural values remains an open question.]{#culture}
+Relatedly, <span id = "globalnorth">although the majority of references cited in the introduction were (systematic) reviews or theoretical articles, all primary sources relied on data from the Global North (see [Supplemental Table 4]() for the countries of origin of all references cited).
+As geographic and cultural considerations were not selection criteria in our literature search,
+a possible explanation for this bias is that research on predictors of adolescent emotion regulation development may have been conducted primarily in the Global North.
+It thus remains to be determined whether these findings generalize to populations from the Global South.</span>
+
 Another concern related to generalizability is that the relative homogeneity of our sample in terms of demograpic factors, like SES and religion, might have contributed to their lower variable importance.
@@
 Results indicated that the importance of different predictors corresponded somewhat with the levels of the bioecological model:
-Proximal factors were more predictive of emotion regulation development than distal predictors; individual differences were most predictive, and mesosystem factors (e.g., SES, religion) were least predictive.
-The results also suggested that parenting may be less relevant for emotion regulation development than previously thought, and that negative parenting may be more impactful than positive parenting.
+Proximal factors were more predictive of emotion regulation development than distal predictors; individual differences were most predictive; macro- and mesysystem factors (e.g., SES, religion) were least predictive.
+The results also suggested that specific parenting practices may be less relevant for emotion regulation development than previously thought,
+and that negative parenting practices may be more impactful than positive ones.
 The results have clear implications for early risk assessment:
@@
 The results further suggest that conflict skills training might be a promising avenue for adolescent-directed interventions.
-Parenting interventions might benefit more from addressing the notion of "good enough" parenting and remedying negative and invasive behaviors than from promoting positive parenting behaviors.
+Parenting interventions might benefit more from addressing the notion of "good enough" parenting and remedying negative and invasive practices than from promoting positive parenting practices.
 The results also provide clear guidance for the ongoing effort of theory generation in this field:
@@
 Parents' roles might be less central than previously thought;
-their support of adolescents' autonomy and the way they navigate conflict might be more important than their parenting behaviors.
+their support of adolescents' autonomy and the way they navigate conflict might be more important than their parenting practices.
 Moreover, there was no evidence of differences between mothers' and fathers' parenting, indicating that both parents should be equally represented in future theory.

### 2022-11-09 mixture_model.R

@@ -39,13 +39,18 @@ desc <- mixtureSummaryTable(res_mp)
 write.csv(desc, "supplemental_table_2.csv", row.names = FALSE)
 classlabs <- paste0(c("High", "Low"), ": ", round(res_mp$mix_2_class.out$results$class_counts$modelEstimated$proportion*100), "%")
 p <- plotGrowthMixtures(res_mp[[2]], rawdata = TRUE, bw = TRUE)
-p <- p + scale_linetype_manual(labels = classlabs, values = c(1:2), name = "Risk", guide = guide_legend(reverse = TRUE)) +
-  scale_shape_manual(labels = classlabs, values = c(15, 17), name = "Risk", guide = guide_legend(reverse = TRUE)) +
+p <- p + 
+  scale_y_reverse(breaks = c(0.5, 0.0, -0.5, -1.0, -1.5), labels = as.character((-1*c(0.5, 0.0, -0.5, -1.0, -1.5)))) +
+  scale_linetype_manual(labels = classlabs, values = c(1:2), name = "Risk") +
+  scale_shape_manual(labels = classlabs, values = c(15, 17), name = "Risk") +
   scale_x_continuous(breaks = 0:4, labels = 14:18, expand = c(0, 0)) +
   labs(x = "Age", y = "Difficulties in emotion regulation") +
   theme(legend.position = c(0.88, 0.85))
+p
 saveRDS(p, "classplot.RData")
 ggsave("classplot.pdf", p, device = "pdf", units = "in", width = 7, height = 5)
+ggsave("classplot.svg", p, device = "svg", units = "in", width = 7, height = 5)
+ggsave("classplot.png", p, device = "png", units = "in", width = 7, height = 5)
 
 p <- plotGrowthMixtures(res_mp[[2]], rawdata = TRUE, alpha_range = c(0, 0.05))
 p <- p + scale_linetype_manual(labels = classlabs, values = c(1:2), name = "Risk", guide = guide_legend(reverse = TRUE)) +

### 2022-11-09 scripts/assessingforest.R

@@ -62,6 +62,99 @@ p <- metaforest::VarImpPlot(VI, length(VI$variable.importance))
 saveRDS(p, "varimp_21-08-2021.RData")
 
 
+# Varimp by type ----------------------------------------------------------
+
+VI <- readRDS("variable_importance.RData")
+names(VI)[names(VI) == "Psychological control"] <- "Psychological control (mf)"
+ren <- read.csv("scale_rename.csv", stringsAsFactors = FALSE, header = FALSE)
+VI <- data.frame(Variable = paste0(1:length(VI), ". ", names(VI)), Importance = VI, 
+                 Level = ren$V4[match(names(VI), ren$V2)],
+                 Theme= ren$V5[match(names(VI), ren$V2)])
+VI$Level <- ordered(VI$Level, levels = rev(c("Individual", "Microsystem", "Mesosystem", "Macrosystem")))
+VI$Theme <- ordered(VI$Theme, levels = rev(c("Personality", "Int/ext", "Conflict", "Parenting", "Substance", 
+                                         "Peers", "Demographics")))
+
+#VI <- VI[order(VI$Level, VI$Importance, decreasing = TRUE), ]
+VI$Variable <- ordered(VI$Variable, levels = rev(VI$Variable))
+p1 <- 
+  ggplot(VI[1:44, ], aes(x = Importance, y = Variable, shape = Level)) +
+    geom_segment(aes(x = 0, xend = Importance, y = Variable, yend = Variable), colour = "grey50", linetype = 2) + 
+    geom_vline(xintercept = 0, colour = "grey50", 
+               linetype = 1) +
+    geom_point(size = 2) +
+    xlab("Variable Importance (Permutation importance)") + 
+    theme_bw() + theme(panel.grid.major.x = element_blank(), 
+                     panel.grid.minor.x = element_blank(), axis.title.y = element_blank())+
+    theme(axis.text.y = element_text(hjust=0))+xlab(NULL)+scale_x_continuous(limits = c(0, max(VI$Importance))) +
+    scale_shape_manual(breaks = c("Individual", "Microsystem"), values = c(16, 17), guide = "none")
+  
+  
+p2 <- 
+  ggplot(VI[45:nrow(VI), ], aes(x = Importance, y = Variable, shape = Level)) +
[…]

### 2022-11-10 action_letter.Rmd
update 2022-11-10 11:47:07

@@
 class       : "draft"
-bibliography: ["manuscript/veni.bib"]
+bibliography      : ["manuscript/r-references.bib", "manuscript/references.bib", "manuscript/Veni.bib"]
 output      : papaja::revision_letter_pdf
@@
 Yours sincerely,
-Author
+
+the authors
 
@@
 We completely agree with Reviewer 3 that there is already a wealth of prior research in this area.
-In fact, there is so much work on this topic that it would be a tall order to review the literature on an individual study basis.
-Given the large prior literature,
-we rely primarily on prior review articles.
-We currently cite 17 review articles in the introduction,
-including a recent (2021) comprehensive systematic review of 6305 studies on factors associated with adolescent emotion regulation.
 
-We cannot strive to recreate a comprehensive review here - 
-and this also seems redundant, given that many of the cited review articles are very recent (i.e., from 2019-2021),
-and seems overly ambitious, given the page limits and attention that must be devoted to explaining the novel methods,
-and spurious, given that this is an exploratory study and thus has no hypotheses that must be supported with prior literature.
-The present study builds upon prior reviews of the literature,
-and those reviews are cited.
-
-Of course we are willing to include specific omitted literature, but please note that Reviewer 3 does not mention which references they feel are missing.
-
-To address this comment, we now emphasize more strongly that there is an abundance of prior literature on predictors of adolescent emotion regulation, e.g.:
-
-`r get_revision(manuscript = man, "abundantlit")`
+`r get_revision("priorlit", letter, quote = F)`
 
@@
 
-As mentioned above, the vast majority of cited studies are reviews (narrative reviews, meta-analyses, and book chapters) or theoretical publications.
+First, as mentioned above, the vast majority of cited studies are reviews (narrative reviews, meta-analyses, and book chapters) or theoretical publications.
 Although one would hope that such publications cover all available literature without selecting for country of origin,
@@
 
+<!-- Second, this is not a hypothesis-testing study.
+It doesn't matter whether findings from previous samples generalize to the present one, because we do not deduce hypotheses from previous work.
+All that matters is whether our findings can be generalized to other samples.
+-->
+
 \RC{Also, you describe two major theories of development including the bioecological theory (Bronfenbrenner \& Morris, 2007).
@@
 
-To address this comment, we have incorporated the following in our analysis of the results in light of the bioecological model:
+To address this comment, we have incorporated the following into the Introduction:
+
+`r get_revision(manuscript = man, "velez1")`
+
+And the following in the Discussion:
 
@@
 not a parenting practice, because it does not relate to specific parental actions.
-The specific parenting practices associated with instilling a sense of balanced relatedness are maternal (54) and paternal (55) support, which rank very low.
+The parenting practices associated with instilling a sense of balanced relatedness are maternal (54) and paternal (55) support, which rank very low.
 
-The observation that *parenting practices* are relatively less important than individual predictors (and certainly much less important than we would have thought a-priori) is thus strongly supported by the data;
+The observation that parenting practices are relatively less important than individual predictors (and certainly much less important than we would have thought a-priori) is thus strongly supported by the data;
 the first such behaviors are (16) and (25).
-This is consistent with the finding that some parent-related constructs indeed score highly.
+This conclusion is not contradicted by the finding that some parent-related constructs score highly.
 
@@
 
-We thank the Editor for this excellent suggestion; we have addressed this comment by adding the following paragraph:
+We thank the Editor for this excellent suggestion; we have addressed this comment by adding the following to the Introduction:
+
+`r get_revision(manuscript = man, "velez1")`
+
+And the following to the Discussion:
 
@@
 We believe that this figure substantially clarifies our interpretation of the results.
-
-```{r figviolin, results='asis', out.width="20%", fig.cap="Variable importances clustered by theme (A) and level of the bioecological model (B).", eval = FALSE}
-knitr::include_graphics("violin.png")
-```
 
@@
 We respectfully believe that this comment might be based on a misunderstanding.
-There are three types of missingness: MNAR, where missingness depends on unobserved data; MCAR, where missingness depends on observed data; and MAR, where missingness is truly random.
-It is not possible to test for MNAR, and the present methods are valid for both MCAR and MAR - so there is no need to provide evidence for either of those.
+There are three types of missingness: MCAR, where missingness depends on observed data and those observed data can be used to impute; MAR, where missingness is truly random and it doesn't matter if observed data is used to impute; and MNAR, where missingness depends on unobserved data and can therefore neither be detected nor correctly imputed.
 
-Also note the unfeasibility of performing an MCAR test when the number of variables is higher than the number of participants: The raw data, prior to imputation and scale construction, is 495 (participants) x 750 (variables) big.
-I am not aware of any MCAR test that works with N < p data.
+No test exists to exclude MNAR.
+For example, Little's MCAR test and non-parametric alternatives like Jamshidian & Jalal's non-parametric test can provide *evidence confirming MAR*, but the presence or absence of MAR does not exclude MNAR.
+
+The present methods are valid for both MCAR and MAR - so there is no need to provide evidence for either of those.
+
+Also note that no MCAR test exists for cases where the number of variables is higher than the number of participants (N < p): In this case, the raw data, prior to imputation and scale construction, is 495 (participants) x 750 (variables) big, so N < p.
 
@@
 1. We added the requested illustrative example, using neuroticism. We could use any scale, but as Neuroticism later turns out to be the most important predictor, it seems a suitable candidate.
-2. We have attempted to clarify our description of the dimension reduction procedure
+2. We have clarified our description of the dimension reduction procedure
 3.
@@
 Unfortunately there is no way to test this assumption.
-The SEM-forest method estimates tens, or hundreds of thousands of SEM models in its different nodes.
+The SEM-forest method estimates tens to hundreds of thousands of SEM models in across all nodes in all trees.
 It is not possible to test model assumptions in each of those nodes.
@@
 
-`r get_revision(manuscript = man, "examplenonlinear")`
+`r get_revision(manuscript = man, "nonlinearexample")`
 
@@
 Perhaps the authors could take only the most important predictors, say the top 10 (or really whatever number they want), and run regression analyses that predict trajectory intercepts and slopes?
-Basically, I know that people like me really like explanations like “oh yeah, these top 10 machine learning predictors predict XX% of variance in emotion regulation scores over time”, and so I was trying to think of a way to generate those sorts of tangible, “traditional” effect sizes.
+Basically, I know that people like me really like explanations like “oh yeah, these top 10 machine learning predictors predict XX\% of variance in emotion regulation scores over time”, and so I was trying to think of a way to generate those sorts of tangible, “traditional” effect sizes.
 Could the authors potentially provide such effect size estimates?
@@
 \RC{I appreciated the brave and nuanced discussion of how these analyses conform to, and also challenge, traditionally held norms in the field of emotion regulation.
-I especially appreciated the identification that parenting, substance use, and bullying might not be the risk factors we think they are.
-• The ERRATA tool seems amazing; I couldn’t find it online. Could a link be provided directly in the text?}
+I especially appreciated the identification that parenting, substance use, and bullying might not be the risk factors we think they are.}
+
+We thank the Reviewer for these kind words.
+We share the Reviewer's enthusiasm for the potential change in perspective implied by these findings - although we understand the importance of tempering that enthusiasm as this is just a single preliminary study, with considerable limitations.
+
+\RC{The ERRATA tool seems amazing; I couldn’t find it online. Could a link be provided directly in the text?}
 
@@
 We apologize for the inconvenience.
-The manuscript only has 42 pages; pages 43-65 are copies of the online supplemental materials (which are supposed to be accessed via the broken links, which have now been fixed).
+The manuscript only has 42 pages; pages 43-65 are copies of the online supplemental materials (the same Tables that were supposed to be accessed via the broken links, which are now fixed).
 For the sake of blind review, we also submitted copies of these tables through the submission portal, but it evidently cannot correctly present these materials.
@@
 
-```{r}
-knitr::include_graphics("violin.png")
-```
-
 \RC{3.
@@
 
+First, we really want to stress that we are not biased against parenting; much of our prior work has focused on the effect of specific parenting practices on child development.
+This notwithstanding, we were legitimately surprised that these same specific parenting behaviors ranked surprisingly low in machine learning analysis - although we understand and acknowledge that this is just a single preliminary study, with considerable limitations.
+
+Second, with regard to the Reviewer's statement that "the third-most important predictor being a positive
+parenting behavior".
+Balanced relatedness is not a parenting practice (i.e., a specific action parents engage in as part of child rearing).
+Rather, it is an aspect of the (adolescent-perceived) emotional family climate.
 `r get_revision(manuscript = letter, "parents", quote = F)`
 
-We also believe that the Figure above helps clarify these results.
+We also believe that Figure 3 helps clarify these results.
 
@@
 We do believe that the quoted statement is correct and does not require further modification, however, as it does not relate to causal interpretation of the results, but to the affordances of the method: latent class analysis *cannot* answer the question "Why?" some adolescents follow similar causal trajectories,
-whereas SEM-forests *can* answer this question, because cases are clustered based on predictor variables (of course with the caveat that causal interpretation is contingent on study design or theoretical justification).
+whereas SEM-forests *can* answer this question, because cases are clustered based on predictor variables.
+As is the case for all statistical analyses, this comes with the caveat that causal interpretation is contingent on study design or theoretical justification.
 
@@
 2. There is no a-priori way to determine which bivariate interactions might be important in SEM-forests analysis (because there is no applicable analogy for Friedman's H statistic).
-3. Consequently, one would have to examine all bivariate interactions by hand. This comes down to interpreting 84*84*3 = 21168 plots. This is not possible.
+3. Consequently, one would have to examine all bivariate interactions by hand. This comes down to interpreting 84\*84\*3 = 21168 plots. This is not possible.
 
@@
 in principle this is correct, but in practice, no method (currently?) exists to obtain accumulated local effects for SEM-forests.
-The Reviewer is citing cutting-edge developments in the area of classic random forests; it will take a while before these developments spill over into the SEM-forests literature, if ever.
+The Reviewer is citing cutting-edge developments in the area of classic random forests, which are not yet widely adopted or implemented for different algorithms;
+it will take a while before these developments spill over into the SEM-forests literature.
 
@@
 
-We have attempted to address this comment as follows: By explaining our decision to assume homoscedasticity:
+We have attempted to address this comment by explaining our decision to assume homoscedasticity:
 
@@
 
-And by addressing it further in the Limitations section:
+And also, by addressing it in the Limitations section:
 
@@
 They are just spreadsheets that can be opened in a web-browser or excel.
+The links to these Supplemental Tables in the manuscript now work.
 
@@
 
-We appreciate these suggestions and the caution about abstract length; we have cut the abstract by half and attempted to incorporate the Reviewer's suggestions. The Abstract now reads:
+We appreciate these suggestions and the caution about abstract length; we have cut the abstract by half and incorporated the Reviewer's suggestions. The Abstract now reads:
 
@@
 
-We appreciate this comment. There are two issues here, which we will both address in turn.
+First, we agree that this choice of words was unfortunate.
+We did not mean to imply that such research has not yet been conducted.
+The purpose of this sentence was to emphasize that it is *important* to examine risk factors and lead into the following 3-page section which addresses what is already known.
+To clarify this, we have rewritten this sentence and the next as follows:
 
-The first issue is that the offending paragraph evidently failed to convey that we set out to identify predictors of adolescents' diverging developmental trajectories using *person-centered* methods.
-This is certainly different from preceding work, which has nearly exclusively looked at whole sample average effects.
-This paragraph thus doesn't argue that "few have examined predictors of emotion regulation in adolescence", as the Reviewer interpreted it, but instead, argues that "few have used person-centered methods to examined predictors of adolescents' diverging developmental trajectories of emotion regulation."
-In the Revision, we have attempted to clarify this by more precisely describing the contribution of the present study, in several places:
+> `r get_revision("nextstep", man, quote = FALSE)` [...] `r get_revision("fragmented", man, quote = FALSE)`
 
-`r get_revision(manuscript = man, "personcentered1")`
+A second issue is whether our writing sufficiently acknowledges prior research on predictors of emotion regulation.
+We could not agree more with the Reviewer that there is a wealth of prior research in this area.
+<span id = "priorlit">In fact, a recent (2021) systematic review identified 6305 studies on factors associated with adolescent emotion regulation.
+Given this exceedingly large volume of prior literature,
+we found it necessary to rely primarily on review articles and theoretical articles to do justice to prior work.
+We currently cite 17 review articles and 12 theoretical pieces,
+none of which were restricted to any one lab.
+We thus respectfully disagree that we only address studies from a "small number of labs".
 
-The second issue is whether our writing does sufficient justice to prior publications on predictors of emotion regulation.
-We could not agree more with the Reviewer that there is a wealth of prior research in this area.
-In fact, there is so much work on this topic that it would be a tall order to do it all justice.
-At present, we cite a total of 17 review articles in the introduction,
-including a recent (2021) comprehensive systematic review of 6305 studies on factors associated with adolescent emotion regulation.
-
-Given the large prior literature,
-we found it necessary to rely primarily on review articles to do justice to prior work.
 We cannot strive to recreate a comprehensive review here - 
 and this also seems redundant, given that many of the cited review articles are very recent (i.e., from 2019-2021),
-and seems overly ambitious, given the page limits and attention that must be devoted to explaining the novel methods,
+and overly ambitious, given the page limits and attention that must be devoted to explaining the novel methods,
 and spurious, given that this is an exploratory study and thus has no hypotheses that must be supported with prior literature.
-The present study builds upon prior reviews of the literature,
-and those reviews are cited.
 
-Of course we are willing to include **specific** omitted literature, but it is not clear from the comment which specific references should be included.
+This notwithstanding, we are willing to include **specific** omitted literature, but it is not clear from the Reviewer's comment which specific references should be included.
 
@@
 `r get_revision(manuscript = man, "abundantlit")`
+</span>
 
@@
 
-Comprehensive descriptive statistics about all available demographic characteristics are reported in Tables 1 and 2.
-To address this comment, we now stress this in the Participants section:
+Comprehensive descriptive statistics about all available demographic characteristics were already reported in Tables 1 and 2.
+However, we see that this was not made sufficiently clear.
+To address this comment, in the revision, we now state the following:
 
@@
 >  From 2006 to 2011, trained interviewers
-conducted annual home visits to collect questionnaire data. All independent variables were
-assessed in the first measurement wave.
+conducted annual home visits to collect questionnaire data.
 
@@
 We thank the Reviewer for pointing out this unfortunate choice of words, which should have been flagged during proof reading.
-Entropy is not a criterion for class enumeration, but for evaluating the quality of the chosen solution (see Jung \& Wickrama, 2008). In the Revision, we have attempted to clarify this as follows:
+In fact, entropy is not a criterion for class enumeration, but for evaluating the quality of the chosen solution (see Jung \& Wickrama, 2008).
+In the Revision, we have attempted to clarify this as follows:
 
@@
 
-In line with the Reviewer's expectation, we did in fact use the median and 16th and 84th percentile.
-The confusion arises from our attempt to explain to the readership what these percentiles mean, by drawing a comparison to +/- 1SD.
+In line with the Reviewer's expectation, we did use the median and 16th and 84th percentile.
+The confusion arises from our attempt to explain to the readership what these percentiles mean, by pointing out that these percentiles correspond to +/- 1SD of a standard normal distribution.
 
@@
 
-`r get_revision(manuscript = man, "highrankingparent", quote = F)`
+`r get_revision(manuscript = man, "highrankingparent")`
 

### 2022-11-10 manuscript/manuscript.Rmd
update 2022-11-10 11:47:07

@@
 Using the machine learning algorithm SEM-forests, we ranked the importance of (87) candidate variables assessed at age 13 in predicting quadratic latent trajectory models of emotion regulation development from age 14-18.
-Participants were 497 Dutch adolescents.
+Participants were 497 Dutch families.
 Results indicated that the most important predictors were individual differences (e.g., in personality), aspects of relationship quality and conflict behaviors with parents and peers, and internalizing and externalizing problems.
-Relatively less important were demographics, bullying, delinquency, substance use, and specific parenting practices – and negative parenting practices ranked higher than positive ones.
+Relatively less important were demographics, bullying, delinquency, substance use, and specific parenting practices – although negative parenting practices ranked higher than positive ones.
 We discuss implications for theory and interventions, and present an open source risk assessment tool, ERRATA.]{#abstract}
@@
 <span id = "abundantlit">There is already an abundance of empirical research on adolescent emotion regulation and its predictors; a recent systematic review identified 6305 relevant publications [@vanlissaMappingPhenomenaRelevant2021].
-Still, that review identified specific limitations of the existing literature:
-many studies narrowly focus on a few predictors within a specific research area,
+An important limitation of the existing literature is that many studies narrowly focus on a few predictors within a specific research area,
 and thus fail to provide insight into these predictors' relative importance compared to other predictors, including those in other domains.
@@
 
-The present study seeks to overcome these limitations
-by using data-driven insights to complement and reflect on existing theory.
-To this end, a broad range of potential risk factors were analyzed using machine learning,
-and ranked in terms of predictive importance.
+This exploratory study seeks to overcome these limitations by casting a broad net among potential risk factors, 
+and using the resulting data-driven insights to complement and reflect on existing theory.
+Machine learning was used to determine the relative utility of a wide range of potential risk factors.
 Specifically,
@@
 of emotion regulation across adolescence (age 14 through 18).
-The results indicate which factors are most predictive of adolescents' diverging trajectories of emotion regulation development, contributing to our understanding of why some youth flourish while others struggle through adolescence [@croneUnderstandingAdolescencePeriod2012].</span>
+The results indicate which factors are most predictive of adolescents' diverging trajectories of emotion regulation development, contributing to our understanding of why some youth flourish while others struggle [@croneUnderstandingAdolescencePeriod2012].</span>
 These insights can guide development of interventions that focus on the most important predictors,
@@
 <!-- Although some studies have examined, on average, how strongly particular risk factors are associated with developmental trajectories - most have done  REF write something about other studies examining risk factors -->
-An important next step is to identify which risk factors render some adolescents susceptible to emotion dysregulation.
+[An important target of research is thus to identify which risk factors render some adolescents susceptible to emotion dysregulation.]{#nextstep}
 <!-- This goal fits within the paradigm shift that has gripped the field of pedagogy, towards "person-centered" research, which focuses on individual developmental trajectories and within-family predictive processes, rather than average patterns across samples16,23.
@@
 
-A recent review noted that the literature on potential predictors of emotion regulation development is somewhat fragmented,
+[Substantial prior research has examined potential risk factors.
+However, a recent review of 6305 relevant publications noted that this literature is somewhat fragmented,
 with most studies considering only a handful of variables within the scope of a particular sub-discipline [@vanlissaMappingPhenomenaRelevant2021].
-<!-- At the same time, this literature is relatively fragmented, -->
-<!-- as most studies have examined only a limited number of predictors [@vanlissaMappingPhenomenaRelevant2021]. -->
-This fragmentation may be due, in part, to the fact that few theories are tailored specifically to adolescence,
+If each study examines only a piece of the puzzle, it is hard to see the big picture.]{#fragmented}
+This fragmentation may be due, in part,
+to the fact that few theories are tailored to emotion regulation in adolescence,
 and none consistently guide research in the field [@bussTheoriesEmotionalDevelopment2019].
-Theories commonly cited in this field vary in scope and specificity:
+Commonly cited theories in this field vary in scope and specificity [see @vanlissaMappingPhenomenaRelevant2021]:
 Some are broad in scope, considering many potential risk factors,
@@
 Other theories are narrower in scope, and are sufficiently specific to derive testable predictions, but lack broader context.
-For a more detailed overview, see @vanlissaMappingPhenomenaRelevant2021.
 Two of the most influential broad theories are the bioecological model [@bronfenbrennerBioecologicalModelHuman2007; e.g., see @augustEmotionRegulationAtRisk2017] and the transactional model [@sameroffUnifiedTheoryDevelopment2010].
@@
 and the macrosystem, consisting of cultural norms and values.
-The transactional model is compatible with the bioecological model,
-and describes development as an interplay between the individual child and the environment.
+<span id = "velez1">Note that a recent perspective on the bioecological model emphasizes that cultural practices are enacted by individuals at different levels of the model, including parents [@velez-agostoBronfenbrennerBioecologicalTheory2017].
+This perspective is consistent with the transactional model,
+which describes development as an interplay between the individual child and the environment.</span>
 It distinguishes proximal environmental influences, roughly corresponding to the microsystem,
@@
 and the tripartite model of emotion regulation socialization in families [@morrisRoleFamilyContext2007], which has also been applied to the peer context [@reindlSocializationEmotionRegulation2016].
-This tripartite model is one of the few theories that explicitly addresses the life phase of adolescence.<!---->
-<!-- This seems a bit selective. For example, also discuss the work of Hollenstein and of Nicholas Allen. -->
-<!-- I think you do insufficient right to the theories that exist, both in breadth and in depth.
-Perhaps just stay closer to the contribution of the current paper: empirically examine multiple factors simultaneously rather than selectively examining a few?
--->
-<!-- Despite this abundance of relevant theory,  -->
-<!-- several reviews have noted the relatively fragmented nature of the literature  -->
-<!-- that  -->
-<!-- few theories are tailored specifically to emotion regulation development in adolescence, -->
-<!-- and none are used consistently to guide research in this field. -->
-Given that few theories are both comprehensive and sufficiently specific to derive testable hypotheses,
-an important challenge for the field is to work towards an integrating theory of emotion regulation development in adolescence.
+The tripartite model is one of the few theories that explicitly addresses the life phase of adolescence.
+It posits that children's emotion regulation is shaped by modeling, specific parenting practices, and the emotional family climate (including the interparental relationship).
+This model is relatively narrow, focusing on parents' role - but also sufficiently specific to derive testable hypotheses.
+Note that few theories are both comprehensive and specific.
+An important challenge for the field is thus to work towards an integrating theory of emotion regulation development in adolescence.
 
-<!-- * Bioecological model [@bronfenbrennerBioecologicalModelHuman2007] -->
-<!-- * Transactional model [@sameroffUnifiedTheoryDevelopment2010] -->
-<!-- * Hall's notion of "storm and stress" -->
-<!-- * Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995] -->
-<!-- * Tripartite model [@morrisRoleFamilyContext2007] -->
-<!-- * Internalization model [@holodynskiDevelopmentEmotionsEmotion2006] -->
-<!-- * Polyvagal theory [@porgesOrientingDefensiveWorld1995]. -->
-<!-- * Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012] -->
-<!-- * Process model of emotion regulation [@grossHandbookEmotionRegulation2013] -->
-<!-- * Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000] -->
-<!-- Maybe review theory briefly? -->
 The first step towards theory formation is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
-<!--Thus, offering a starting point for this endeavor,<!-- empirical investigations of predictors of adolescents' emotion regulation development -->
-A recent literature review did so
-using a text mining systematic review of author keywords and abstracts of 6305 publications on  adolescents’ emotion regulation [@vanlissaMappingPhenomenaRelevant2021].
-Typically, phenomena are identified using narrative literature reviews.
-However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008].
-Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
-Specifically, related keywords are grouped into clusters,
-with each cluster representing a phenomenon.
-These phenomena are then ranked by the number of studies those keywords occur in, and potential relations between phenomena are mapped by co-occurrence within studies.<!-- reflection and meaning-making-->
-This review thus offers a relatively comprehensive overview of potentially relevant phenomena.
-The present study used it as a starting point for an empirical investigation of the relative importance and direction of effect of these phenomena.
+Phenomena considered relevant in a field of research are often summarized using literature reviews.
+A recent review of 6305 publications on adolescents’ emotion regulation used text mining to identify relevant phenomena, including potential risk factors [@vanlissaMappingPhenomenaRelevant2021].
+Narrative reviews are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008].
+Text mining systematic reviews are more comprehensive and derive insights from the literature through a transparent and somewhat objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
+This review thus offers a relatively comprehensive overview of potential predictors of emotion regulation development.
+The present study used it as a starting point for an empirical investigation of the relative importance and direction of effect of these predictors.
 
@@
 the phenomena that have received most research attention are internalizing and externalizing problems, and mental health.
-These are likely to be, at least in part, outcomes of emotion dysregulation.
-<!-- - but they that does not diminish their usefulness as predictors. -->
+Although reciprocal effects cannot be excluded, these phenomena are typically conceptualized as outcomes of emotion dysregulation.
+Note, however, that this does not diminish their usefulness as predictors.
 Other important phenomena reflected what has been discussed in prior narrative reviews,
@@
 identity, morality, sex and sexual development.
-<!-- To this end, the present study sought to identify operationalizations of these relevant phenomena in a large-scale panel study. -->
 Prior theory and empirical research thus suggests many phenomena potentially related to heterogeneity in adolescents' emotion regulation development.
 
+The aforementioned potential risk factors have largely been studied in isolation.
 What is still missing is an understanding of the relative utility of different early indicators in predicting adolescents' developmental outcomes.
-<!-- we do not know the relative importance of  with respect to others not included in the same study. -->
-Put simply, each study has examined a piece of the puzzle,
-but no single study has captured the complete picture.
 This knowledge gap cannot be addressed by comparing the relative importance of different predictors across studies,
@@
 different control variables and statistical models.
-Furthermore, conventional methods are ill-suited to overcome address this research question:
-<!-- In part, this limitation is a consequence of conventional methods: -->
-linear models limit the total number of predictors that can be included due to concerns of power, overfitting, and multicolinearity [@hastieElementsStatisticalLearning2009].
+<!-- Furthermore, conventional methods are ill-suited to address this research question: -->
+In part, this knowledge gap is a consequence of the limitations of conventional methods:
+linear models limit the total number of predictors that can be included due to power concerns, risk of overfitting, and multicolinearity [@hastieElementsStatisticalLearning2009].
 By contrast, some machine learning methods offer a more holistic approach,
@@
 
-<span id = "semforests">The SEM-forests algorithm is based on the random forests algorithm, which in turn is based on regression trees [see @stroblIntroductionRecursivePartitioning2009].
-A single regression tree splits the sample repeatedly in two by whichever predictor $p$ makes both post-split groups maximally internally homogenous, and maximally different from one another.
+<span id = "semforests">The SEM-forests algorithm is based on the random forests algorithm, which in turn is based on regression trees [for an accessible introduction, see @stroblIntroductionRecursivePartitioning2009].
+A single regression tree splits the sample repeatedly in two by whichever predictor makes both post-split groups maximally internally homogenous, and maximally different from one another.
 A random forest consists of many regression trees grown on bootstrapped samples from the original dataset.
-To ensure that each tree learns something different from the data, it may only consider a random subsample of $K$ candidates out of all $p$ predictors at each split.
+To ensure that each tree learns something different from the data, it may only consider a random subsample of $m$ candidates out of all $p$ predictors at each split.
 With this in mind, the SEM-forest algorithm operates as follows:
@@
     i. Estimate the latent growth model on the bootstrap sample
-    ii. Consider a random selection of $K$ candidates out of all $p$ predictors; we used $k = 9$ based on the standard practice to use $K = \lfloor\sqrt{p}\rfloor$
+    ii. Consider a random selection of $m$ candidates out of all $p$ predictors; commonly $m = \lfloor\sqrt{p}\rfloor$
     iii.
-[Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).]{#stepiii}
-    iv. [Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then fully grown.]{#fullygrown}
+[Identify the predictor $p_k$ and value of that predictor $x$ that, when using individual values on that predictor $p_{ki}$ to split the sample in two by the rule $p_{ki} \leq x$, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).]{#stepiii}
+    iv. [Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then considered fully grown.]{#fullygrown}
 3.
-[Given an individual's scores on the predictor variables $p$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters).
+[Given an individual's scores on all predictor variables $p_i$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters).
 Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.]{#predictions}
@@
 This approach intrinsically accommodates non-linear effects and (complex) interactions, which is an advantage over parametric models that assume linearity.
-<span id = "examplenonlinear">As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
+[As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
 In regression analysis, this could be represented with a linear and quadratic term, and a plot might show that the effect of autonomy support on the intercept increases to a value $x_{knee}$, and then tapers off.
-A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.</span>
-The algorithm additionally performs variable selection;
-unimportant variables are not selected in step 2.iii.
+A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.]{#nonlinearexample}
+Note that SEM-forests intrinsically perform variable selection: unimportant variables are not selected in step 2.iii.
 Thus, the number of predictors can be large without risking model non-identification.</span>
@@
 
-<!-- SB: you have convincingly argued why machine learning in general is important, yet I miss the details related to ER.
-I would recommend to add a paragraph summarizing theories of ER and the factors they consider important.
-Also, describe the candidate predictors that were derived from the literature search, and how they relate to theories and /or empirical evidence -->
-The present study set out to obtain a relatively comprehensive understanding of the predictors of emotion regulation development across adolescence.
-<!-- identify the most important early indicators of developmental trajectories -->
-<!-- of  -->
+The present study set out to obtain a comprehensive understanding of the relative utility of predictors of emotion regulation development across adolescence.
 The primary aim was to identify the most important predictors of adolescents' developmental trajectories of emotion regulation development.
 The secondary aim was to probe the nature of the relationship of each predictor with these trajectories.
-This involves an important challenge, as
-the number of potentially relevant risk factors is too large for conventional methods.
+An important challenge of these research aims is that the number of potentially relevant risk factors is too large for conventional methods.
 At the same time, there is a paucity of relevant theory to narrow down the candidate predictors a priori.
-<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
-This challenge can be overcome by machine learning methods that perform variable selection.
-We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence.
+We sought to overcome this challenge by using SEM-forests, 
+a machine learning algorithm methods that performs variable selection,
+to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence.
 
@@
 
-`r ifelse(mask, "The Workflow for Open Reproducible Code in Science (WORCS)", "**Masked**")` was used to make analyses reproducible  `r ifelse(mask, "[@vanlissaWORCSWorkflowOpen2020]", "**(masked reference)**")`.
-All analysis code, Supplemental materials and synthetic data are available at `r ifelse(mask, "<https://github.com/cjvanlissa/veni_forest>", "**masked URL**")`.
-The original data (controlled access) and all study documentation are available via `r ifelse(mask, "<https://doi.org/10.17026/DANS-ZRB-V5WP>", "**masked URL**")`.
+`r ifelse(mask, "The Workflow for Open Reproducible Code in Science (WORCS)", "**Masked**")` was used to create a reproducible research archive  `r ifelse(mask, "[@vanlissaWORCSWorkflowOpen2020]", "**(masked reference)**")`.
+All analysis code, supplemental materials and synthetic data are available at `r ifelse(mask, "<https://github.com/cjvanlissa/veni_forest>", "**masked URL**")`.
+The original data and all study documentation are available under controlled access via `r ifelse(mask, "<https://doi.org/10.17026/DANS-ZRB-V5WP>", "**masked URL**")`.
 
@@
 Of 1,544 eligible families, 497 produced informed consent and were included in the study.
-[From 2006 to 2011, trained interviewers conducted six annual measurement waves to collect questionnaire data during home visits.]{#waves}
+[From 2006 to 2011, trained interviewers completed six annual measurement waves by collecting questionnaire data during home visits.]{#waves}
 All independent variables were assessed in the first measurement wave.
-[The sample demographic information is reported in Tables \@ref(tab:tabcat) and \@ref(tab:tabnum).]{#demographics}
+[Sample demographic characteristics are reported in Tables \@ref(tab:tabcat) and \@ref(tab:tabnum).]{#demographics}
 The dependent variable was measured from the second to the sixth wave.
@@
 
-<!-- **Participants** -->
-
-<!-- * RADAR data (https://www.uu.nl/en/research/radar) -->
-<!-- * 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) -->
-<!-- * Collected between 2006-2011 -->
-<!-- * Most families were  medium- to high-SES (10% low-SES) -->
-
 ## Measures
 
-## Dependent variable
+### Dependent variable
 
@@
 The lower BIC indicates support for metric invariance.
-The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
+The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .05$,  $\text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
 [Note that the RMSEA has been found to be more appropriate for longitudinal data, possibly because the null-model on which the CFI is based is less appropriate for repeated measures [@leiteDetectingGrowthShape2011]]{#cfi}.
 As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
-we considered it acceptable to proceed with individual factor scores for further analysis.
+we proceeded with individual factor scores for further analysis.
 
-## Candidate predictors
+### Candidate predictors
 
 We set out to identify operationalizations of all relevant phenomena identified in @vanlissaMappingPhenomenaRelevant2021 that were available in the RADAR data.
-This yielded 87 candidate predictors, of which `r nrow(desc_cat)` categorical (see Table \@ref(tab:tabcat)), `r sum(is.na(desc_num$Items))` numerical, and `r sum(!is.na(desc_num$Items))` multi-item scales, see Table \@ref(tab:tabnum).
-<!-- SB: I miss a reflection on the extent to which the available data cover all relevant predictors -->
+This yielded 87 candidate predictors, including `r nrow(desc_cat)` categorical (see Table \@ref(tab:tabcat)), `r sum(is.na(desc_num$Items))` numerical, and `r sum(!is.na(desc_num$Items))` multi-item scales, see Table \@ref(tab:tabnum).
 We used principal components analysis to obtain a single score for multi-item scales without making additional assumptions about the underlying measurement model.
@@
 Principal components analysis indicated that the first component was the only component with an Eigenvalue greater than one (3.22, other Eigenvalues [0.99 - .33].
-This component explained the majority of variance in all six adjectives, $R^2 = .54$, other components $R^2 = [.16 - .05]$.
+This component explained the majority of variance in all six adjectives, $R^2 = .54$, or 3.24 times as much variance as could be expected if each dimension of the scale explained equal variance.
+Other components' $R^2 \in [.16 - .05]$.
 All adjectives loaded highly on this first component, $L = [.54 - .83]$.
 We thus used this first component as operationalization of the construct neuroticism.
-Across all other scales, the explained variance of the first component ranged from $R^2 \in [0.18, 0.90]$, which corresponded to 1.60 - 17.46 times as much variance as could be expected if each dimension of the scale explained equal variance.
+Across all other scales, explained variance of the first component ranged from $R^2 \in [0.18, 0.90]$, which corresponded to 1.60 - 17.46 times as much variance as could be expected if scale dimension explained equal variance.
 We thus considered these principal components to be adequate operationalizations of the scales.
-[Supplemental Table S1](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_1.csv) additionally reports skew and kurtosis,
+[Supplemental Table S1](https://github.com/cjvanlissa/veni_forest/blob/2c545ab2195fade5d2be3e4a269c24e57e8921a8/supplemental_table_1.csv) additionally reports skew and kurtosis,
 minimum and maximum factor loadings,
@@
 names(desc_cat) <- c("Variable", "n", "%", "Category")
-papaja::apa_table(desc_cat, caption = "Sample descriptive statistics for categorical predictors, ranked by variable importance.
-Note that variable importance can be interpreted only in relative terms.", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
+desc_cat$Category[desc_cat$Category == "TRUE"] <- "Yes"
+desc_cat$Category[desc_cat$Category == "FALSE"] <- "No"
+papaja::apa_table(desc_cat, caption = "Sample descriptive statistics for categorical predictors, ranked by variable importance.", landscape = FALSE, longtable = TRUE, format.args = list(na_string = ""))
 ```
+
 ```{r tabnum, results='asis'}
@@
 
-Preliminary analyses were conducted to ensure the appropriateness of growth curve modeling, and determine the presence of heterogeneity in developmental trajectories.
+All analyses were conducted in R [@R-base]. 
+Preliminary analyses were used to establish the appropriateness of growth curve modeling, and determine the presence of heterogeneity in developmental trajectories.
 First, trajectories of emotion regulation development were modeled using a curvilinear latent growth curve model with an intercept (I), linear slope (S), and quadratic slope (Q).
@@
 We used the R-package `tidySEM` [@tidysem] to estimate and compare 1-6 class solutions of the aforementioned curvilinear growth curve model;
-analysis results are displayed in [Supplemental Table 2](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_2.csv).
+analysis results are displayed in [Supplemental Table S2](https://github.com/cjvanlissa/veni_forest/blob/2c545ab2195fade5d2be3e4a269c24e57e8921a8/supplemental_table_2.csv).
 To ensure model convergence in small classes, the latent variable covariance matrix was restricted to zero,
@@
 It thus answers the question: "Which adolescents have similar growth trajectories"?
-SEM-forests similarly aim for maximum homogeneity of growth trajectories within clusters,
-but unlike latent class analysis, they cluster individuals by reference to a set of *predictors*.
-They thus answer the question: "Why do some adolescents have similar growth trajectories"?
-This method is suitable to meet the study goals.
+A SEM-forest similarly aims for maximum homogeneity of growth trajectories within clusters,
+but unlike latent class analysis, clusters individuals by reference to a set of *predictors*.
+It thus answers the question: "Why do some adolescents have similar growth trajectories"?
+This matches the present study goals.
 
-<span id = "homogeneity">For the SEM-forests analysis,
-we imposed an additional restriction on the diagonal of the residual covariance matrix,
-reflecting an assumption of homoscedastic residuals over time.
-We deemed this assumption to be necessary because the dataset is relatively small, and we wanted the model to be as simple as possible.</span>
+The main analysis consisted of a SEM-forest with 1000 trees grown on bootstrapped samples of the original data, estimated using the `semtree` package [@brandmaierSemtreeRecursivePartitioning2022].
+At each split, the model could examine a random selection of 9 predictors, based on the standard practice to use $m = \lfloor\sqrt{p}\rfloor$.
+<span id = "homogeneity">As described above, the outcome of the SEM-tree was a quadratic latent growth curve model.
+We additionally assumed homoscedastic residuals over time by imposing an equality restriction on the diagonal of the residual covariance matrix.
+Thus, instead of four residual variances,
+a single residual variance parameter is estimated.
+We deemed this assumption to be necessary because the dataset was relatively small, and we wanted the model to be as simple as possible.
+The resulting model had four parameters in total.</span>
 
 Our primary outcome of interest, variable importance,
-is defined as the average absolute increase in model misfit due to randomization of each variable.
+is defined as the average absolute increase in model misfit resulting from randomization of a particular variable.
 For each tree, the log-likelihood of cases not in that tree's bootstrap sample is computed twice:
@@
 which visualize the partial dependence of each predictor with trajectories of emotion regulation development.
-New functions for computing partial dependence for growth curve SEM-forests were developed for this paper, which are now available in the `semtree` R-package.
+New functions for computing partial dependence for growth curve SEM-forests were developed for this paper, which are now available in the `semtree` R-package [@brandmaierSemtreeRecursivePartitioning2022].
 Partial dependence for a predictor $p$ is defined as following:
@@
 
-[For continuous predictors, three values of $s$ were used - namely the percentiles, computed on the observed data, corresponding to the mean and $+/- 1 SD$ of a standard normal distribution (i.e., the $50^{th}$, $16^{th}$ and $84^{th}$ percentiles).
-For convenience, these percentiles are labeled as *mean* and $+/- 1 SD$ in Figure \@ref(fig:figpdp1)-\@ref(fig:figpdp4).]{#percentiles}
+[For continuous predictors, three values of $s$ were used, namely the $50^{th}$, $16^{th}$ and $84^{th}$ percentiles of the observed data.
+These percentiles were chosen because they correspond to the mean and $+/- 1 SD$ of a standard normal distribution.
+For ease of interpretation, in Figures \@ref(fig:figpdp1)-\@ref(fig:figpdp4), these percentiles are therefore labeled as *mean* and +/- 1 SD.]{#percentiles}
 For categorical predictors, all unique values were used.
-
-<!-- $$ -->
-<!-- \begin{align} -->
-<!-- \bm{\lambda} &=  -->
-<!-- \begin{bmatrix} -->
-<!-- 1 & 0 & 0^2 \\ -->
-<!-- 1 & 1 & 1^2 \\ -->
-<!-- \vdots & \vdots & \vdots \\ -->
-<!-- 1 & 5 & 5^2 \\ -->
-
-<!-- \end{bmatrix} -->
-<!-- \\ -->
-<!-- \bm{\psi} &= -->
-<!-- \end{align} -->
-<!-- $$ -->
-<!-- Across all measurements, 13.3% of data were missing. -->
-<!-- Little’s (1988) Missing Completely at Random (MCAR) -->
-<!-- test produced a statistically significant value for the daily -->
-<!-- identity and self-concept clarity variables, x -->
-<!-- 2 -->
-<!-- (8592) ¼ -->
-<!-- 9228.07, p < .001, although the x -->
-<!-- 2 -->
-<!-- /df ratio was only 1.07. -->
-<!-- Similarly, for the annual assessments, the MCAR test -->
-<!-- produced a statistically significant value, x -->
-<!-- 2 -->
-<!-- (56) ¼ 75.19, -->
-<!-- p < .001, although the x -->
-<!-- 2 -->
-<!-- /df ratio was only 1.34. These -->
-<!-- findings suggest that data were likely missing at random, -->
-<!-- and that maximum likelihood techniques were appropriate -->
-<!-- for use. Moreover, even when the assumption of missing -->
-<!-- completely at random is not met, the Full Information -->
-<!-- Maximum Likelihood (FIML) procedure is still preferable to -->
-<!-- listwise deletion of cases with missing data (McKnight, -->
-<!-- McKnight, Sidani, & Figuredo, 2007). Listwise deletion -->
-<!-- results in a biased sample, and therefore biased results, -->
-<!-- by systematically excluding cases with missing data on -->
-<!-- any of the study variables. We therefore utilized FIML -->
-<!-- estimation using Mplus release 5.1 (Muthe´n & Muthe´n, -->
-<!-- 2007). -->
-<!-- Our analyses were based on an extension of cross-lagged panel modeling (Selig & Little, 2012), with staggered measurement occasions for one of the variables (mood variability).
-Mood variability was measured each time during the year between two measurement waves.
-Thus, for mood variability, no within-wave correlations were estimated, and all links with conflict frequency and difficulties in emotion regulation were specified as (predictive) regression paths.
-We used this staggered cross-lagged panel model to investigate over-time predictive effects between conflict frequency, day-to-day mood variability, and difficulties in emotion regulation.
-First, we estimated a single-group path model with autoregressive and cross-lagged paths, as well as within-time correlations for concurrent measurements (Model 1, see Table 1).
-Then we introduced empathy (high, average, low) as a moderator in a multi-group model, with all parameters free to vary between groups (Model 2).
-To derive the best fitting, most parsimonious model, we constrained parameters that were not significantly different.
-We used Wald \chi^2 tests to evaluate which parameters did not differ significantly, first over time and then between groups.
-We used the resulting final model to test our hypotheses (Model 3, see Figure 2).
-For hypotheses involving moderation, we used Wald tests to investigate whether parameters differed significantly between empathy groups.
-To test hypotheses involving mediation, we inspected indirect effects with bootstrapped standard errors (10,000 resamples).
-Furthermore, although posterior predictive probabilities were high, we conducted additional analyses, using the approach described by Asparouhov and Muthén (2014) to account for classification uncertainty.
-As can be seen in Appendix A, results were in line with the multi-group model reported here.
--->
-<!-- Attrition for the annual measurements ranged from 0.40% at age 13 to 14.50% at age 18, and attrition for the daily mood diaries ranged from 3.40% to 23.40%.
-These data were missing completely at random, Little’s (1988) MCAR test \chi^2(2179) = 2124.07, p = .80.
-Therefore, full information maximum likelihood estimation (FIML) was warranted to make use of all available information without estimating missing data.
-All analyses were conducted in Mplus Version 7 (Muthén & Muthén, 1998–2012).
-Per the developers’ recommendation, we used robust maximum-likelihood estimation, which yields a Satorra-Bentler scaled \chi^2 value to account for potential non-normality (Satorra, 2000).
-Model fit was evaluated using the Comparative Fit Index and Tucker–Lewis Index (CFI and TLI, acceptable fit = 0.90-0.95, Bentler & Bonett, 1980), and Root Mean Square Error of Approximation (RMSEA, close fit = 0.01-0.06, acceptable fit = 0.06-0.08, Browne & Cudeck, 1993).
-According to these criteria, the fit of the final model was acceptable-to-good (Table 1).
-In addition to these measures of objective fit, we also provide two comparative fit indices, the Akaike Information Criterion (AIC) and sample-size adjusted Bayesian Information Criterion (saBIC).
-Lower values on these comparative fit indices indicate a better fitting model.
-According to both relative and absolute fit indices, the final model had the best fit.
--->
 
@@
 as illustrated in Figure \@ref(fig:figvarimp).
-<!-- Table \@ref(tab:tablegend) provides a brief legend of the 30 most important variables, and  -->
-Supplemental Table S1 provides a full legend of all variables.
+[Supplemental Table S1](https://github.com/cjvanlissa/veni_forest/blob/2c545ab2195fade5d2be3e4a269c24e57e8921a8/supplemental_table_1.csv) provides a full legend of all variables.
 
-```{r figvarimp, fig.cap="Predictors ranked in order of relative importance. m: mother, f: father, b: best friend, s: sibling."}
+```{r figvarimp, fig.cap="Predictors ranked in order of relative importance, and categorized by level of the bioecological model. m: mother, f: father, b: best friend, s: sibling."}
 #readRDS("../varimp_comb2_03-08-2022.RData")
@@
 several themes emerged (see Figure \@ref(fig:figthemes) panel A).
-Specifically, results indicated that among the most important predictors were personality and personality-related constructs.
+Specifically, results indicated that several of the highest-ranking predictors were individual differences in personality and related constructs.
 Three of the big five personality dimensions ranked among the top six predictors: Neuroticism (1), extraversion (4), and agreeableness (6).
@@
 
-A second emerging theme was that most constructs related to conflict frequency and behavior ranked highly, and few ranked low.
+A second emerging theme was that internalizing- and externalizing symptoms
+often associated with emotion regulation ranked relatively highly.
+Of these, parent-reported externalizing behavior (5) ranked highest,
+followed by adolescent-reported anxiety (15) and depression (17).
+Adolescent-reported externalizing behavior also ranked somewhat highly (30), but much lower than parent reports.
+
+A third emerging theme was that constructs related to conflict frequency and behavior ranked relatively highly.
 For instance, adolescent-perceived interparental conflict (8),
@@
 
-<span id = "parenting">A third emerging theme is that specific parenting practices were relatively underrepresented among the most important predictors.
+<span id = "parenting">A fourth emerging theme was that specific parenting practices were relatively underrepresented among the most important predictors.
 Parenting practices are defined as specific parental actions related to child-rearing,
@@
 like maternal (54) and paternal (55) support,
-were ranked low,
-and neutrally valenced aspects like parental monitoring (sollicitation 71, adolescent disclosure 72) were ranked very low.
-This is interesting, given the extensive research attention into the effect of these specific parenting practices.
+ranked low,
+and neutrally valenced aspects like parental monitoring ranked very low (sollicitation 71, adolescent disclosure 72).
+This is interesting, given the extensive research attention into the effect of these specific parenting practices - but also consistent with other recent critical accounts [@keijsersParentalMonitoringAdolescent2016; @vanlissaRoleFathersMothers2018].
 
@@
 like balanced relatedness, conflict likely contributes to the emotional family climate [@morrisRoleFamilyContext2007].
-Finally, maternal age (7) had a protective effect, which is well-supported by prior literature [@zondervan-zwijnenburgParentalAgeOffspring2020].]</span>
-<!-- Add reference -->
+Finally, maternal age (7) had a protective effect, which is well-supported by prior literature [@zondervan-zwijnenburgParentalAgeOffspring2020].</span>
 
-Finally, an emerging theme was that internalizing- and externalizing symptoms
-often associated with emotion regulation ranked relatively highly in terms of predictive importance.
-Of these, parent-reported externalizing behavior (5) ranked highest, followed by adolescent-reported anxiety (15) and depression (17).
-Adolescent-reported externalizing behavior ranked substantially lower (30).
-
-It is further of note that some themes that are considered to be theoretically relevant ranked low.
-This was the case, for example, for demographic variables like maternal (86) and paternal (87) socio-economic status and family religion (81, 82).<!--SB: Why is this important?
+Finally, it is noteworthy that some themes that are considered to be theoretically relevant ranked low.
+This was the case, for example, for substance use;
+the most commonly used substances of alcohol (78) and cigarettes (61) were ranked very low; drug use, which has a lower incidence, ranked a bit higher than this (37) - but did not appear to have a marginal negative effect (Figure 5).
+Similarly, aspects of adolescents' peer networks also ranked unexpectedly low,
+such as bullying (60, 84), victimization (51, 62, 83),
+and contact with delinquent peers (85).
+The same applied to demographic variables like maternal (86) and paternal (87) socio-economic status and family religion (81, 82).<!--SB: Why is this important?
 CJ: Religion is part of moral development, SES, and is directly related to emotion regulation: https://link.springer.com/chapter/10.1007/978-94-017-8950-9_13-->
-Aspects of adolescents' peer networks,
-such as bullying (60, 84) and victimization (51, 62, 83),
-and contact with delinquent peers (85) also ranked unexpectedly low.
-The same applied to substance use;
-the most commonly used substances of alcohol (78) and cigarettes (61) were ranked very low; drug use, which is less common, ranked a bit higher than this (37, but see direction of association).
 
@@
 
-Figures \@ref(fig:figpdp1)-\@ref(fig:figpdp4) depict the marginal association plots,
-also in order of importance.
+Figures \@ref(fig:figpdp1)-\@ref(fig:figpdp4) depict marginal association plots of all predictors in order of importance.
 Each plot shows the expected trajectory of emotion regulation development at different values of each predictor.
@@
 
-All associations were in a direction consistent with prior literature.
-Among these were several factors associated with better emotion regulation (i.e., potential protective factors), including
-parents' autonomy support and tolerance,
-big five extraversion and agreeableness, 
-self-concept clarity,
-problem solving in conflict with both parents,
+All directions of association were consistent with prior literature.
+Several potential protective factors were associated with better emotion regulation, including
+individual differences in big five extraversion and agreeableness and self-concept clarity, 
+aspects of the emotional family climate like
+balanced relatedness and interparental conflict,
+and problem solving in conflict with both parents,
 parents' own emotion regulation ability,
@@
 
-<span id = "posthoc">A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
+<span id = "posthoc">A Reviewer suggested following up the planned machine learning analysis with post-hoc analyses that might be more familiar to a developmental readership:
 Regressing the parameters (intercept, slope and quadratic slope) of a latent growth model on each predictor in turn.
-Such analyses might be helpful to plan hypotheses and perform power analysis for follow-up research within the SEM framework.
-However, we strongly caution against over-interpretation.
-Unlike the machine learning analyses, these analyses do not accommodate non-linear effects or interactions, nor are effects controlled for other predictors.
-The results are merely descriptive and should not be used to perform inference.
-[Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
-All models had good fit ($RMSEAs < .07$, $CFIs > .99$, $SRMRs < .04$).
+Such analyses might help formulate hypotheses and perform power analyses for future research within the SEM framework.
+However, we strongly caution against over-interpretation of these post-hoc analyses.
+Unlike the primary analysis, they do not accommodate non-linear effects or interactions, nor are effects controlled for other predictors.
+The results are descriptive and should not be used for inference.
+[Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/2c545ab2195fade5d2be3e4a269c24e57e8921a8/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
+All models had good fit ($\text{RMSEAs} < .07$, $\text{CFIs} > .99$, $\text{SRMRs} < .04$).
 Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
@@
 there were differences in relative importance.
-Most of the important predictors were individual differences in personality, the motivational system, self-concept clarity, and empathy.
-<span id = "highrankingparent">Although a few parent-related constructs ranked highly,
-these were primarily related to the family emotional climate (balanced relatedness and interparental conflict) and demographics (maternal age).</span>
-Specific parenting practices, however, ranked lower than might have been expected given prior literature [e.g., support, control, psychological control, @morrisRoleFamilyContext2007].
+Some of the most important predictors were individual differences in personality, the motivational system, self-concept clarity, and empathy.
+<span id = "highrankingparent">Although a few parent-related constructs also ranked highly,
+these were related to the emotional family climate (balanced relatedness and interparental conflict).</span>
+Specific parenting practices, by contrast, ranked lower than might have been expected given prior literature [e.g., support, control, psychological control, @morrisRoleFamilyContext2007].
 Of these specific child-rearing behaviors,
@@
 Themes that ranked lower than might have been expected given prior research were family demographics (SES and religion) and bullying, victimization and delinquency, and substance use.
-The relatively low importance of demographics might be attributable to a restriction of range, as the sample was relatively homogenous.
+One possible explanation for the relatively low importance of demographics might be restriction of range, as the sample was relatively homogenous.
 Still, other empirical studies have also found little evidence for a direct effect of, for example, SES on adolescent emotion regulation [e.g., @defranceExpandingContextRole2021].
-It is thus also possible that the population effect is small, or that it is fully mediated by more proximal process variables.
+It is thus also possible that the population effect is small.
+Finally, another explanation may be that the effect of demographics is fully mediated by more proximal process variables.
 
@@
 Relating these results to prior theory,
-it is noteworthy that predictive importance of different themes appears to correspond to the levels of the bioecological model [@bronfenbrennerBioecologicalModelHuman2007], see Figure \@ref(fig:figthemes) panel B.
-[Although there was substantial overlap between predictors at different levels of the bioecological models,
-several predictors at the individual level achieved the highest importance rankings]{#overlap},
-<!-- SB: this is not very surprising given that many of the behaviors assessed require ER (e.g.
-becoming angry in a conflict is an indication of low ER, and individuals high in Neuroticism also have fluctuating emotions that they find difficult to control/regulate) -->
+it is noteworthy that there appears to be some correspondence between the relative importance of different predictors and the levels of the bioecological model [@bronfenbrennerBioecologicalModelHuman2007], see Figure \@ref(fig:figthemes) panel B.
+[Although there was substantial overlap between predictors at different levels of the bioecological model,
+several predictors at the individual level achieved the highest importance rankings,
 followed by predictors at the microsystem level, particularly those representing the emotional family climate,
-followed by the mesosystem, including aspects of the peer network (e.g., bullying/victimization and delinquency), school performance, and socio-economic status,
-and the macrosystem indicator religiosity was among the least important predictors.
+followed by mesosystem predictors, including aspects of the peer network (e.g., bullying/victimization and delinquency), school performance, and socio-economic status.
+The only macrosystem-level indicator, religiosity, ranked among the least important predictors.]{#overlap},
 Future theory might thus emphasize proximal factors over distal ones [see also @sameroffUnifiedTheoryDevelopment2010].
@@
 </span>
-A possible alternative explanation is that our sample was relatively homogenous in terms of characteristics typically attributed to the macrosystem.
+A possible alternative explanation is that our sample was relatively homogenous in terms of religion and SES.
 
@@
 Our results showed no systematic differences between mothers and fathers.
-[Anecdotally however, out all parent-related constructs that were measured separately for mothers and fathers]{#motherfather}, constructive conflict resolution with fathers ranked highest, and out of all parenting practices, paternal power ranked highest.
+[Anecdotally however, out of all constructs that were measured separately for mothers and fathers, constructive conflict resolution with fathers was the highest-ranking parent-related construct, and paternal power was the highest-ranking parenting practice.]{#motherfather}
 These observations validate the notion that fathers matter.
@@
 
-A finding that challenges contemporary discourse is that certain phenomena often conceptualized as "problem behaviors" had very low - for example, substance use [cf @willsEmotionRegulationSubstance2017], bullying, and time spent with delinquent peers.
-Even the highest ranked problem behavior, drug use, had no distinguishable marginal effect.
-This suggests that drug use derives its high importance from higher-order interactions with other predictors.
-It might be the case, for example, that for some youth, drug use contributes to self-exploration and social relationships, whereas for others, it may be a form of escapism, self-medication, or even self-harm [see also @nelemansLongitudinalAssociationsSocial2016].
-This is merely a hypothesis, and further research is required to identify what interactions are at play.
-These results imply that what are commonly thought of as risk factors might not be decisive in adolescents' emotion regulation development;
-such factors ought to be included in future theory, but should not take center stage.
+A finding that challenges contemporary discourse is that certain phenomena often conceptualized as "problem behaviors" had very low predictive importance - for example, substance use [cf @willsEmotionRegulationSubstance2017], bullying, and time spent with delinquent peers.
+Even the highest ranked problem behavior, drug use, had no distinguishable marginal effect, see Figure \@ref(fig:figpdp2).
+This suggests that drug use derives its high importance from interactions with other predictors.
+It might be the case, for example, that for some youth, drug use contributes to self-exploration and social relationships, whereas for others, it may be a form of escapism, self-medication, or even self-harm [see @nelemansLongitudinalAssociationsSocial2016].
+This is merely one potential explanation; further research is required to identify what interactions are at play.
+These results imply that what are commonly thought of as risk factors might not be decisive in adolescents' emotion regulation development.
+As we found little evidence of straightforward negative effects,
+future theory should strive to provide a more nuanced account of their potential role in emotion regulation development.
 
@@
 These phenomena all ranked relatively low in empirical importance in the present study.
-Conversely, phenomena found to be important in the present study, such as personality and other individual differences, parental autonomy support, and conflict-related behavior were less prevalent in the literature.
-Of course, the relative importance of predictors may vary across studies,
+Conversely, phenomena found to be important in the present study, such as personality and other individual differences, parental autonomy support, and conflict-related behavior, are less prevalent in existing literature.
+Of course, this study provides only a single estimate of variable importance,
 and replications of the present findings are required.
@@
 focal topics in the literature are likely to garner undue theoretical weight.
-Considering the empirical importance of phenomena may provide guidance in focusing future theory development in this area.
+Taking the empirical importance of phenomena into account may help guide future theory development in this area.
 
@@
 [These findings have several implications for the early identification of youth at-risk for emotion regulation difficulties.
-The fact that individual-level predictors were most important in predicting adolescents' diverging developmental trajectories suggests that it might be possible to perform preliminary risk assessments for emotion dysregulation using only adolescent-reported individual differences.
+The fact that individual-level predictors were most important in predicting adolescents' diverging developmental trajectories underscores the utility of adolescent self-reports for preliminary risk assessments.
 Although the focus in clinical practice is shifting towards holistic approaches that involve the family system,
@@
 For instance, a family member may refuse to cooperate, or holistic assessment may be too costly.
-In such cases, it is useful to know that individual differences carry substantial informational value.]{#riskassessment}
-Moreover, some of the most important predictors are already routinely assessed (for example, big five personality).
-Other important predictors are overt, and thus easily observed by teachers and parents - for example, externalizing problem behavior and conflict frequency and behavior.
+In such cases, it is useful to know that individual differences - which are already routinely assessed - carry substantial informational value.]{#riskassessment}
+Moreover, some of the most important predictors are also among the most commonly administered  scales (e.g., big five personality).
+Other important predictors were overt behaviors, which are easily observed by teachers and parents - for example, externalizing problem behavior and conflict frequency and behavior.
 Future research might thus develop and validate risk assessment instruments
@@
 
-The method used in the present study is particularly well-suited to risk assessment instruments,
+The method used in the present study also has substantial potential for risk assessment instruments,
 because machine learning models maximize predictive performance, whereas traditional methods like linear regression maximize performance within the study sample.
@@
 Intervention can only be effective if this association can additionally be assumed to be causal, based on theory or experimental research.
-Two phenomena appear to be prime candidates for intervention,
-based on the variable importance rankings:
-conflict resolution behavior, and (negative) parenting practices.
+Based on the variable importance rankings,
+and prior experimental research,
+two promising candidates for intervention
+are conflict resolution behavior and (negative) parenting practices.
 
@@
 Parenting practices are another promising target for intervention,
-as indeed it has historically been.
-Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting practices.
+as indeed they have historically been.
+Importantly, our results suggest that such interventions should focus on promoting balanced relatedness and decreasing intrusive, negative, guilt-inducing and overbearing parenting practices.
 By contrast, promoting positive parenting practices may have limited potential for impact.
-With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
+With this in mind, it is important to consider that existing parenting training tends to focus primarily on positive parenting practices.
 Although one meta-analysis did report large effects of the Triple P-Positive Parenting Program on parental outcomes - including some negative parenting practices [@fletcherImpactBehaviouralParent2011],
@@
 Linear statistical models are the default in our field.
-Therefore, an interesting result is that most predictors had non-linear effects:
+With this in mind, it is noteworthy that most predictors had non-linear effects:
 Out of the three values contrasted in the marginal dependence plots,
 only the one corresponding to the least desirable conditions diverged from the others.
-<!-- For instanhigh neuroticism, high interparental conflict, low constructive conflict resolution with fathers. -->
-Our findings send an important signal that many of the associations between predictors and developmental outcomes might be better modeled non-linearly.
-In particular, they suggest a notion of "good-enough" preconditions for the development of mature emotion regulation.
+An important methodological implication is that the effects of many developmental influences ought to be modeled non-linearly.
+In particular, our findings suggest a notion of "good-enough" preconditions for the development of mature emotion regulation.
 Most adolescents seem to follow trajectories characterized by high emotion regulation - except those with the most disadvantage.
-Some prior research has considered this notion of a "good-enough" environment [@vanlissaCostEmpathyParentadolescent2017; @poonGoodEnoughParental2017];
+Although prior research has considered this notion of a "good-enough" environment [@vanlissaCostEmpathyParentadolescent2017; @poonGoodEnoughParental2017];
 the consistency of our findings suggests that it may be more ubiquitous than previously thought.
@@
 but almost no marginal effect (e.g., father's age, drug use, openness to experience).
-Such predictors likely derive their importance from interactions.
+Such predictors likely derive their importance from interactions [@vanlissaDevelopmentalDataScience2022].
 For instance, drug use might be a vehicle for social status or self-harm;
@@
 openness to experience may be an asset if it leads to self-discovery but a liability if it leads to risk behavior.
-The present method is not well-suited to identify what interaction drives a variable's importance,
-but these results might indicate potentially fruitful areas of future research.
+The present method is not well-suited to identify what specific interactions drive a variable's importance,
+but these results do suggest potentially fruitful areas of future research.
 
@@
 which may limit the generalizability of the findings.
-Specifically, this was a non-random, ethnically and socially relatively homogeneous sample from the Netherlands.
+Specifically, this was a non-random, ethnically and socially homogeneous sample from the Netherlands.
 [Cultural values, which differ across and within countries,
 have been found to influence people's motivation to regulate emotions, as well as the strategies they use and their adaptiveness [@fordCultureEmotionRegulation2015].
-In cross-cultural comparisons, cultural values in Dutch samples have been found (on average) to be similar to those in large Anglophonic countries (e.g., the United States, Canada, and Australia) [@terlutterGLOBEStudyApplicability2006].
-Whether our findings also generalize to populations that adhere to different cultural values remains an open question.]{#culture}
-Relatedly, <span id = "globalnorth">although the majority of references cited in the introduction were (systematic) reviews or theoretical articles, all primary sources relied on data from the Global North (see [Supplemental Table 4]() for the countries of origin of all references cited).
+In cross-cultural comparisons, cultural values in Dutch samples have been found to be similar on average to those in large Anglophonic countries (e.g., the United States, Canada, and Australia) [@terlutterGLOBEStudyApplicability2006].
+Whether our findings also generalize to populations that emphasize different cultural values remains an open question.]{#culture}
+Relatedly, <span id = "globalnorth">although the majority of references cited in the introduction were (systematic) reviews or theoretical articles, all primary sources relied on data from the Global North (see [Supplemental Table S4](https://github.com/cjvanlissa/veni_forest/blob/2c545ab2195fade5d2be3e4a269c24e57e8921a8/supplemental_table_4.csv) for the countries of origin of all references cited).
 As geographic and cultural considerations were not selection criteria in our literature search,
-a possible explanation for this bias is that research on predictors of adolescent emotion regulation development may have been conducted primarily in the Global North.
+a possible explanation for this bias is that prior research in this area may have been conducted primarily in the Global North.
 It thus remains to be determined whether these findings generalize to populations from the Global South.</span>
@@
 The only way to address these concerns is to replicate these findings in other samples.
-To facilitate such replications, all code is made available in a reproducible archive.
+To facilitate replications, all analysis code is made available in a reproducible archive.
 Another limitation is that the data source was a panel study with mostly questionnaire measures.
@@
 This concern is assuaged somewhat by the observation that parent-reported externalizing behavior (5) was much more predictive than adolescent self-reported externalizing behavior (30).
-This indicates that predictors less susceptible to common method bias nonetheless explained substantial variance.
+Thus, some predictors not susceptible to common method bias were also found to explain substantial variance.
 One final limitation related to the data is the relatively small sample size.
-Machine learning methods benefit from large samples.
+Machine learning benefits from large samples.
 Although random forests are relatively robust to overfitting,
@@
 Finally, SEM-forests require substantial computational resources,
-which also limit the ability to compute different iterations of the model beyond default settings,
+which limits the ability to compute different iterations of the model beyond default settings,
 or to exploring marginal effects for higher order interactions.
@@
 Results indicated that the importance of different predictors corresponded somewhat with the levels of the bioecological model:
-Proximal factors were more predictive of emotion regulation development than distal predictors; individual differences were most predictive; macro- and mesysystem factors (e.g., SES, religion) were least predictive.
+Proximal factors were more predictive of emotion regulation development than distal predictors; individual differences were most predictive; meso- and macrosystem factors (e.g., SES, religion) were least predictive.
 The results also suggested that specific parenting practices may be less relevant for emotion regulation development than previously thought,
@@
 Such theory should focus on proximal processes over distal ones.
-Parents' roles might be less central than previously thought;
-their support of adolescents' autonomy and the way they navigate conflict might be more important than their parenting practices.
+Parenting practices might be less crucial than previously thought;
+the emotional family climate and the way parents and adolescents navigate conflict might be more important.
 Moreover, there was no evidence of differences between mothers' and fathers' parenting, indicating that both parents should be equally represented in future theory.
@@
 delinquency, substance use, and bullying were all relatively less important predictors.
-Even drug use, the highest-ranking problem behavior, had no distinguishable marginal effect, suggesting that it is more problematic for some adolescents than others.
+Even drug use, the highest-ranking problem behavior, had no distinguishable marginal effect, suggesting that it is problematic for some adolescents but not for others.
 These results illustrate how machine learning can provide a holistic view of a research area,

### 2022-11-10 mixture_model.R

@@ -40,12 +40,12 @@ write.csv(desc, "supplemental_table_2.csv", row.names = FALSE)
 classlabs <- paste0(c("High", "Low"), ": ", round(res_mp$mix_2_class.out$results$class_counts$modelEstimated$proportion*100), "%")
 p <- plotGrowthMixtures(res_mp[[2]], rawdata = TRUE, bw = TRUE)
 p <- p + 
-  scale_y_reverse(breaks = c(0.5, 0.0, -0.5, -1.0, -1.5), labels = as.character((-1*c(0.5, 0.0, -0.5, -1.0, -1.5)))) +
-  scale_linetype_manual(labels = classlabs, values = c(1:2), name = "Risk") +
-  scale_shape_manual(labels = classlabs, values = c(15, 17), name = "Risk") +
+  # scale_y_reverse(breaks = c(0.5, 0.0, -0.5, -1.0, -1.5), labels = as.character((-1*c(0.5, 0.0, -0.5, -1.0, -1.5)))) +
+  scale_linetype_manual(labels = classlabs, values = c(1:2), name = "Risk", guide = guide_legend(reverse = TRUE) ) +
+  scale_shape_manual(labels = classlabs, values = c(15, 17), name = "Risk", guide = guide_legend(reverse = TRUE) ) +
   scale_x_continuous(breaks = 0:4, labels = 14:18, expand = c(0, 0)) +
-  labs(x = "Age", y = "Difficulties in emotion regulation") +
-  theme(legend.position = c(0.88, 0.85))
+  labs(x = "Age", y = "Emotion regulation") +
+  theme(legend.position = c(0.88, 0.15))
 p
 saveRDS(p, "classplot.RData")
 ggsave("classplot.pdf", p, device = "pdf", units = "in", width = 7, height = 5)

### 2023-02-01 action_letter.Rmd
update 2023-02-01 10:12:21

@@
 class       : "draft"
+mask: no
 bibliography      : ["manuscript/r-references.bib", "manuscript/references.bib", "manuscript/Veni.bib"]
@@
 library(revise)
-man <- read_manuscript("manuscript/manuscript.Rmd")
-letter <- read_manuscript("action_letter.Rmd")
+test_env <- new.env()
+read_manuscript("c:/git_repositories/veni_forest/manuscript/manuscript.Rmd", envir = test_env)
+read_manuscript("action_letter.Rmd", envir = test_env)
 mask = TRUE
@@
 
-`r get_revision("priorlit", letter, quote = F)`
+`r get_revision("priorlit", quote = F)`
 
@@
 
-`r get_revision(manuscript = man, "globalnorth")`
+`r get_revision( "globalnorth")`
 
@@
 
-`r get_revision(manuscript = man, "velez1")`
+`r get_revision( "velez1")`
 
@@
 
-`r get_revision(manuscript = man, "velezagosto")`
+`r get_revision( "velezagosto")`
 
@@
 
-`r get_revision(manuscript = man, "waves")`
+`r get_revision( "waves")`
 
@@
 
-`r get_revision(manuscript = man, "dutch")`
+`r get_revision( "dutch")`
 
@@
 
-`r get_revision(manuscript = letter, "extractpreds", quote = F)`
+`r get_revision("extractpreds", quote = F)`
 
@@
 
-`r get_revision(manuscript = man, "parenting")`
+`r get_revision( "parenting")`
 
@@
 
-`r get_revision(manuscript = man, "parentrelated")`
+`r get_revision( "parentrelated")`
 </span>
@@
 
-`r get_revision(manuscript = man, "overlap")`
+`r get_revision( "overlap")`
 
@@
 
-`r get_revision(manuscript = man, "cfi")`
+`r get_revision( "cfi")`
 
@@
 
-`r get_revision(manuscript = man, "missing")`
+`r get_revision( "missing")`
 </span>
@@
 
-`r get_revision(manuscript = man, "homogeneity")`
+`r get_revision( "homogeneity")`
 
@@
 
-`r get_revision(manuscript = man, "homogeneity2")`
+`r get_revision( "homogeneity2")`
 
@@
 
-`r get_revision(manuscript = man, "stepiii")`
+`r get_revision( "stepiii")`
 
@@
 
-`r get_revision(manuscript = man, "fullygrown")`
+`r get_revision( "fullygrown")`
 
@@
 
-`r get_revision(manuscript = man, "semforests")`
+`r get_revision( "semforests")`
 
@@
 
-`r get_revision(manuscript = man, "predictions")`
+`r get_revision( "predictions")`
 
@@
 
-`r get_revision(manuscript = man, "nonlinearexample")`
+`r get_revision( "nonlinearexample")`
 
@@
 
-`r get_revision(manuscript = man, "limitations")`
+`r get_revision( "limitations")`
 
@@
 
-`r get_revision(manuscript = man, "posthoc")`
+`r get_revision( "posthoc")`
 
@@
 
-`r get_revision(manuscript = man, "errata")`
+`r get_revision( "errata")`
 
@@
 
-`r get_revision(manuscript = man, "includepersonality")`
+`r get_revision( "includepersonality")`
 
@@
 Rather, it is an aspect of the (adolescent-perceived) emotional family climate.
-`r get_revision(manuscript = letter, "parents", quote = F)`
+`r get_revision("parents", quote = F)`
 
@@
 
-`r get_revision(manuscript = man, "causality")`
+`r get_revision( "causality")`
 
@@
 
-`r get_revision(manuscript = man, "singlefactor")`
+`r get_revision( "singlefactor")`
 
@@
 
-`r get_revision(manuscript = man, "motherfather")`
+`r get_revision( "motherfather")`
 
@@
 
-`r get_revision(manuscript = man, "abstract")`
+`r get_revision( "abstract")`
 
@@
 
-`r get_revision(manuscript = man, "abundantlit")`
+`r get_revision( "abundantlit")`
 </span>
@@
 
-`r get_revision(manuscript = man, "entropy")`
+`r get_revision( "entropy")`
 
@@
 
-`r get_revision(manuscript = man, "percentiles")`
+`r get_revision( "percentiles")`
 
@@
 
-`r get_revision(manuscript = letter, "parents", quote = F)`
+`r get_revision("parents", quote = F)`
 
@@
 
-`r get_revision(manuscript = man, "highrankingparent")`
+`r get_revision( "highrankingparent")`
 
@@
 
-`r get_revision(manuscript = man, "riskassessment")`
+`r get_revision( "riskassessment")`
 
@@
 
-`r get_revision(manuscript = man, "culture")`
+`r get_revision( "culture")`
 

### 2023-02-01 manuscript/manuscript.Rmd
update 2023-02-01 10:12:21

@@
 Text mining systematic reviews are more comprehensive and derive insights from the literature through a transparent and somewhat objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
-This review thus offers a relatively comprehensive overview of potential predictors of emotion regulation development.
-The present study used it as a starting point for an empirical investigation of the relative importance and direction of effect of these predictors.
+The present study used this prior text mining review by Van Lissa as a starting point for an empirical investigation of the relative importance and direction of effect of these predictors.
+It provides a relatively comprehensive overview of potential predictors of emotion regulation development.
 
@@
 ```
-<span id = "dutch">Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal study `r ifelse(mask, "RADAR [@branjeResearchAdolescentDevelopment2018]", "**(masked reference)**")`, which
-received ethical approval from the ethical committee of the University Medical Centre `r ifelse(mask, "Utrecht", "**masked**")`.
-The `r ifelse(mask, "RADAR", "**masked**")` sample was recruited from over 200 randomly selected elementary schools in the province of `r ifelse(mask, "Utrecht", "**masked**")`, and four main cities in `r ifelse(mask, "The Netherlands", "**masked**")`.</span>
+
+<span id = "dutch">Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal RADAR study  [@branjeResearchAdolescentDevelopment2018], which 
+received ethical approval from the ethical committee of the University Medical Centre Utrecht.
+The RADAR sample was recruited from over 200 randomly selected elementary schools in the province of Utrecht, and four main cities in The Netherlands.</span>
 Of 1,544 eligible families, 497 produced informed consent and were included in the study.
@@
 The lower BIC indicates support for metric invariance.
-The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .05$,  $\text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
+The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .047$,  $\text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
 [Note that the RMSEA has been found to be more appropriate for longitudinal data, possibly because the null-model on which the CFI is based is less appropriate for repeated measures [@leiteDetectingGrowthShape2011]]{#cfi}.
@@
 a random forests based approach with comparable performance to multiple imputation and FIML estimation [@stekhovenMissForestNonparametricMissing2012].
-This method is valid, regardless of whether missingness is random or contingent on observed data (MAR or MCAR).
+This method is valid, regardless of whether missingness is random or contingent on observed data (MCAR or MAR).
 It does not make any distributional assumptions and does not extrapolate,

### 2023-05-23 presentation_mpib2023.Rmd
update 2023-05-23 08:13:32

--- 2022-04-29 presentation_uw2022.Rmd
@@
 ---
-title: "Developmental Data Science"
+title: "How Machine Learning can Advance Theory Formation"
 author: "Caspar J. Van Lissa<sup>1</sup>"
-date: "<p><sup>1</sup>: Utrecht University, dept. Methodology & Statistics | Open Science Community Utrecht<br>Funded by NWO Veni grant  VI.Veni.191G.090"
+date: "<sup>1</sup>: Tilburg University, dept. Methodology & Statistics<br>Funded by NWO Veni grant  VI.Veni.191G.090"
 knit: (function(inputFile, encoding) {
@@
     css: slide.css
-bibliography      : ["manuscript/references.bib", "../rigorous_exploration/manuscript/r-references.bib", "../rigorous_exploration/manuscript/references.bib", "../rigorous_exploration/manuscript/Rigorous_exploration.bib", "../rigorous_exploration/manuscript/Veni.bib", "../rigorous_exploration/manuscript/Veni_sysrev.bib"]
+bibliography      : ["manuscript/references.bib", "presentation/all_refs.bib"]
 ---
@@
 
-_Machine learning_ can help advance _theory formation_ in developmental psychology
+_Machine learning_ can help advance _theory formation_ in the social sciences
 
 * Replication crisis [@scheelWhyMostPsychological2022; @lavelleWhenCrisisBecomes2021]
-* Solutions improve deductive (theory-testing) research practices
-    + questionable research practices (QRPs)
+* Current solutions improve deductive research (theory-testing)
+    + questionable research practices [@wichertsDegreesFreedomPlanning2016]
     + replication
@@
 
-Many preregistered hypotheses are not supported [@scheelExcessPositiveResults2021]
+Many preregistered hypotheses are not supported (56%) [@scheelExcessPositiveResults2021]
 
@@
 
-* most psychological theories fall short of this definition
+* Most social scientific theories are:
     + insufficiently precise
@@
 
-```{r, out.width="60%"}
-library(tidySEM)
-lay <- get_layout("",     "the1", "", 
-                  "pat1", "",     "hyp1",
-                  "",     "obs1", "",
-                  "",     "tes",  "", rows = 4)
-nodes <- data.frame(
-  name = c("pat1", "the1", "obs1", "tes", "hyp1"),
-  label = c("Patterns", "Theory", "Observation", "Test", "Hypothesis"),
-  shape = "rect")
-edges <- data.frame(
-  from = c("the1", "hyp1", "obs1", "pat1", "obs1"),
-  to = c("hyp1", "obs1", "pat1", "the1", "tes"),
-  arrow = "last"
-)
-g <- prepare_graph(edges = edges, layout = lay, nodes = nodes, angle = 5)
-g$edges$connect_from <- c("right", "bottom", "left", "top", "bottom")
-g$edges$connect_to <- c("top", "right", "bottom", "left", "top")
-
-plot(g)
+```{r, out.width="40%"}
+knitr::include_graphics("presentation/empiricalcycle.png")
 ```
@@
 
-Role of flexibility:
-
 * Confirmatory methods poorly suited to exploratory research
@@
     + Model fit indices: manually specify models, no guarantee that best model is included
-
-"unguided exploration" is effortful and inflates the risk of spurious results
+* Role of flexibility: there are infinite ways to explore
+* "unguided exploration" is effortful and inflates the risk of spurious results
 
@@
 
-```{r}
+```{r, out.width="100%"}
 knitr::include_graphics("Figure Machine learning 4.png")
 ```
+
+# Advantages of Machine Learning
 
@@
 
-Psychological theories rarely account for complex effects
+Social scientific theories rarely account for complex effects
 
@@
 * SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
-* regularized SEM [@jacobucciRegularizedStructuralEquation2016]
+* regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)
 
@@
   inherit.aes = TRUE,
-  size = 1,
+  linewidth = 1,
   aes(colour = "Total error")
@@
     aes(colour = "Bias"),
-    size = 1
+    linewidth = 1
   )+
@@
     aes(colour = "Variance"),
-    size = 1
+    linewidth = 1
   ) +
@@
 * abductive formal theory construction (AFTC) framework
-    + construct formal theory, generate data, mathematically compare to empirical data
+    + construct formal theory
+    + Generate data
+    + Mathematically compare to empirical data
+    + like a posterior predictive check in Bayes
     + Discrepancies: amend formal theory
 
-## Summary
+<!-- ## Summary -->
 
-* There is a paucity of good theory
-* Need for exploratory research for theory formation
-* Machine learning for rigorous exploration
-    + automates model building
-    + incorporates checks and balances for generalizable results
-* Unsupervised learning can assist in phenomenon detection
-* Supervised learning to identify important predictors
-* Some algorithms incorporate basic theoretical elements
+<!-- * There is a paucity of good theory -->
+<!-- * Need for exploratory research for theory formation -->
+<!-- * Machine learning for rigorous exploration -->
+<!--     + automates model building -->
+<!--     + incorporates checks and balances for generalizable results -->
+<!-- * Unsupervised learning can assist in phenomenon detection -->
+<!-- * Supervised learning to identify important predictors -->
+<!-- * Some algorithms incorporate basic theoretical elements -->
 
-# Applied examples
+# Example: Text Mining Systematic Review
 
@@
 
-## Knowledge gap
+<!-- ## Knowledge gap -->
 
-Substantial empirical research, but:
+<!-- Substantial empirical research, but: -->
 
-* Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014]
-* No consistent terminology [@bariolaChildAdolescentEmotion2011]
-* No conceptual frameworks [@stifterEmotionRegulation2019]
-* No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]
+<!-- * Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014] -->
+<!-- * No consistent terminology [@bariolaChildAdolescentEmotion2011] -->
+<!-- * No conceptual frameworks [@stifterEmotionRegulation2019] -->
+<!-- * No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019] -->
 
-Knowledge must be consolidated into overarching theory
+<!-- Knowledge must be consolidated into overarching theory -->
 
@@
 
-First step: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
+* First step: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
+    + Often done by systematic review
+* Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
+* Introducing text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]:
+    + Unlimited sample size, transparent, objective, reproducible
 
-We conducted a text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]
+6653 papers on **emotion regulation** in **adolescence** [age 10-24]
 
-  * Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
-  * TMSR: Unlimited sample size, transparent, objective, reproducible
-
-6653 papers on Addresses **emotion regulation** in population overlapping with adolescence [10-24]
-
-<doi.org/10.1007/s40894-021-00160-7>
+<https://doi.org/10.1007/s40894-021-00160-7>
 
@@
 
-## Baseline network
+## Baseline network {.smaller}
 
-```{r figbaseline, warning=FALSE, eval = TRUE, results='asis', out.width = "90%", fig.cap="Phenomena relevant to adolescents' emotion regulation according to theory (a) and narrative reviews (b; transparent nodes indicate constructs also present in the theory)."}
-knitr::include_graphics("../veni_sysrev/baseline_network.png") 
+```{r figbaseline, warning=FALSE, eval = TRUE, results='asis', out.width = "100%", fig.cap="Phenomena relevant to adolescents' emotion regulation according to theory (a) and narrative reviews (b; transparent nodes indicate constructs also present in the theory)."}
+knitr::include_graphics("presentation/baseline_network.png") 
 ```
@@
 
-![](../veni_sysrev/varimps_small.png)
+![](presentation/varimps_small.png)
 
-## Co-occurrence graph
+## Co-occurrence graph {.smaller}
 
-```{r networks1, eval = TRUE, results='asis', out.width = "90%"}
-knitr::include_graphics("../veni_sysrev/tmnetworks_small.png")
+```{r networks1, eval = TRUE, results='asis', out.width = "100%"}
+knitr::include_graphics("presentation/tmnetworks_small.png")
 ```
@@
 * Empirical research relies on theory; under-theorized phenomena may be overlooked
-    + This study offers guidance for phenomena to consider as confounders/causes
+    + TMSR offers guidance for phenomena to consider as confounders/causes
 * Unembedded terms indicate promising areas of future research
-* Present study serves as starting point for theory development
+* TMSR network can serve as proto-theoretical nomological network
 
-## Next step
+# Example: Predictors of emotion regulation development
+
+## Person-centered models
 
@@
 
-## Solution: Machine learning
-
-* Allows non-linear differences between trajectories
-* Performs variable selection
-  + Cast wide net among potentially relevant predictors
-* Checks & balances ensure generalizability
-* Exploratory: No hypothesis
-
-## Method illustrated
-
-```{r}
-knitr::include_graphics("tree.png")
-```
-
 ## Method
@@
 We used 1000 bootstrap samples, $k = 9$
+
+## SEM Forest
+
+```{r, out.width="70%"}
+knitr::include_graphics("tree.png")
+```
 
@@
 
-## Present study
+## Study design
 
@@
 
-```{r, out.height="100%"}
-vim <- readRDS("varimp_comb_21-08-2021.RData")
-vim
+```{r, out.width="90%"}
+knitr::include_graphics("pcomb2.png")
+```
+
+## Variable importance 2
+
+```{r, out.width="100%"}
+knitr::include_graphics("violin.png")
 ```
@@
 
+* Proximal factors more predictive of ER development than distal predictors
+    + Order mirrors bioecological model
 * Most important predictors are routinely assessed (big 5) or overt (conflict)
@@
   + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
-* Order of predictors mirrors bioecological model
 
@@
 
-## Conclusions
+# Limitations and future directions
 
-* Proximal factors more predictive of ER development than distal predictors
-* Parenting may be less relevant than previously thought
-* Personality and conflict behavior are candidates for early diagnosis
-* Conflict skills training might be avenue for intervention
+## Limitations
 
-## References
+1. Theories are still verbal and flexible
+    * Even if ML leads to new insights, it is not clear how theories should be updated
+2. Ignoring causality limits interpretability of machine learning
+    * Incorrect direction of causality: meaningless results
+    * Conditioning on a collider: biased results
+
+## Future direction 1: Formal theoretical object
+
+Formal theories are expressed in mathematical terms
+
+* unambiguous
+* logically sound
+* explicates assumptions
+* reveals ambiguities and gaps in knowledge
+* facilitates conveying ideas to others
+* applying rules of formal logic implies testable hypotheses
+* refutation of hypotheses dictates specific modifications
+
+## Characteristics of formal theory
+
+According to Meehl:
+
+1. phenomena to be explained
+2. putative causal connections between them
+3. direction of the effect
+4. relative importance of predictors
+5. interaction effects
+6. functional form of associations
+
+## Formal theory
+
+* Characteristics 1-2 can be described by a causal network (DAG)
+* Characteristics 3-6 refer to the nature of associations within that network
+
+Augmented causal networks (ACN, see also Crielaard et al., 2022)
+
+* encode both causal relationships and the nature of those relationships
+* imply testable hypotheses
+* shared in plain text, e.g., YAML or RDF
+* version controlled
+* shared conform FAIR software principles,
+to be used and updated through distributed collaboration
+
+## Learning a causal network
+
+* Using TMSR, based on what previous authors have written
+* Using causal discovery, based on large panel studies
+* Using ESM data to leverage precedence in time
+    + Causality requires association, precedence in time, and nonspuriousness
+
+## Future direction 2: Functional form
+
+Previous work: Post-hoc interpretation of marginal associations
+
+Future direction: Use a-priori interpretable machine learning
+
+* Symbolic regression applies deep learning or genetic algorithms to learn a mathematical formula that describes the relationship between an outcome and its predictors
+
+## References {.smaller}
+
+Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. https://doi.org/10.1002/icd.2370
+
+Van Lissa, C.
+J.
+(2022).
+Complementing preregistered confirmatory analyses with rigorous, reproducible exploration using machine learning.
+Religion, Brain & Behavior.
+https://doi.org/10.1080/2153599X.2022.2070254
+
+Van Lissa, C.
+J., Beinhauer, L., Branje, S., & Meeus, W.
+(2022).
+Using machine learning to identify early predictors of adolescent emotion regulation development.
+Journal of Research on Adolescence.
+https://doi.org/10.31234/osf.io/rbqf9
+
+Van Lissa, C. J. (2021). Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-Mining Systematic Review. Adolescent Research Review. https://doi.org/10.1007/s40894-021-00160-7
+
+Van Lissa, C.
+J., et al., (2022).
+Using machine learning to identify important predictors of COVID-19 infection prevention behaviors during the early phase of the pandemic.
+Patterns, 3(4), 100482.
+https://doi.org/10.1016/j.patter.2022.100482
+
+##

### 2023-05-23 presentation_ut2023.Rmd
update 2023-05-23 08:13:32

--- 2023-05-23 presentation_mpib2023.Rmd
@@
     + Generate data
-    + Mathematically compare to empirical data
-    + like a posterior predictive check in Bayes
+    + Mathematically compare to empirical data (like posterior predictive check in Bayes)
     + Discrepancies: amend formal theory
@@
 
-## Study design
+## Present study
 
@@
 
-Augmented causal networks (ACN, see also Crielaard et al., 2022)
+Augmented causal networks (ACN)
 
 * encode both causal relationships and the nature of those relationships
+* can be shared in plain text (YAML-type) data
 * imply testable hypotheses
-* shared in plain text, e.g., YAML or RDF
-* version controlled
-* shared conform FAIR software principles,
-to be used and updated through distributed collaboration
+* can be shared and updated by multiple researchers
 
@@
 
-Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. https://doi.org/10.1002/icd.2370
+Van Lissa, C.
+J.
+(2022).
+Developmental data science: How machine learning can advance theory formation in Developmental Psychology.
+Infant and Child Development, n/a(n/a), e2370.
+https://doi.org/10.1002/icd.2370
 
@@
 Complementing preregistered confirmatory analyses with rigorous, reproducible exploration using machine learning.
-Religion, Brain & Behavior.
+Religion, Brain & Behavior, 0(0), 1–5.
 https://doi.org/10.1080/2153599X.2022.2070254
@@
 Using machine learning to identify early predictors of adolescent emotion regulation development.
-Journal of Research on Adolescence.
+PsyArXiv.
 https://doi.org/10.31234/osf.io/rbqf9

### 2023-07-11 presentation_isdp2023.qmd
update 2023-07-11 11:52:24.053328

--- 2023-05-23 presentation_mpib2023.Rmd
@@
 ---
-title: "How Machine Learning can Advance Theory Formation"
-author: "Caspar J. Van Lissa<sup>1</sup>"
-date: "<sup>1</sup>: Tilburg University, dept. Methodology & Statistics<br>Funded by NWO Veni grant  VI.Veni.191G.090"
+title: "Machine Learning can Advance Theory Formation in the Social Sciences"
+subtitle: "Funded by NWO Veni grant VI.Veni.191G.090<br>[doi:10.1002/icd.2370](https://doi.org/10.1002/icd.2370)"
+author: "Caspar J. Van Lissa"
+institute: "Tilburg University, dept. Methodology & Statistics"
 knit: (function(inputFile, encoding) {
   rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
-output: 
-  ioslides_presentation:
-    widescreen: true
-    css: slide.css
+format: revealjs
 bibliography      : ["manuscript/references.bib", "presentation/all_refs.bib"]
@@
 
-```{r setup, include=FALSE}
+```{r}
+#| label = "setup",
+#| include = FALSE
 #css: https://maxcdn.bootstrapcdn.com/bootstrap/3.3.7/css/bootstrap.min.css
@@
 
-# Conceptual introduction
-
-## Central thesis
+## Central thesis {.smaller}
 
@@
 
-## Defining theory
+## Defining theory {.smaller}
 
@@
 
-```{r, out.width="60%"}
-library(tidySEM)
-lay <- get_layout("the1", "the2",
-                  "hyp1", "hyp2",
-                  "pat1", "obs2",
-                  "obs1", "con2", rows = 4)
-nodes <- data.frame(
-  name = as.vector(lay),
-  label = c("Theory", "Working hypothesis", "Pattern", "Observation", "Theory", "Hypothesis", "Observation", "Test"),
-  shape = rep("rect", length(lay)),
-  group = rep(c("Inductive", "Deductive"), each = 4)
-  )
-edges <- data.frame(
-  from = c("the1", "hyp1", "pat1", "the2", "hyp2", "obs2"),
-  to = c("hyp1", "pat1", "obs1", "hyp2", "obs2", "con2"),
-  arrow = rep(c("first", "last"), each = 3),
-  group = rep(c("Inductive", "Deductive"), each = 3)
-)
-g <- prepare_graph(edges = edges, layout = lay, nodes = nodes)
-plot(g)
-```
-
-* Lack of theory cannot be overcome by improving deductive research
-* Theory formation requires inductive (exploratory) research [@creswellDesigningConductingMixed2017]
-
-## Empirical cycle
-
-```{r, out.width="40%"}
+```{r}
+#| out.width = "40%"
 knitr::include_graphics("presentation/empiricalcycle.png")
@@
 
-De Groot’s empirical cycle: a model of knowledge production through scientific research [@degrootMethodologieGrondslagenVan1961]
+* Empirical cycle [@degrootMethodologieGrondslagenVan1961]
+* Lack of theory cannot be overcome by improving deductive practices
+* Theory formation requires inductive (exploratory) research [@creswellDesigningConductingMixed2017]
 
@@
 
-**NEW**: Implications for the theory crisis have not yet been discussed
-
-## Machine learning vs unguided exploration
-
-```{r, out.width="100%"}
-knitr::include_graphics("Figure Machine learning 4.png")
-```
-
-# Advantages of Machine Learning
+**NEW**: Epistemological implications of ML for theory formation [@vanlissaDevelopmentalDataScience2022]
 
@@
 
-## Holistic approach
+## Holistic approach {.smaller}
 
@@
 
-* Assumption for causal interpretation
+* Good theory incorporates most important causes
+* Cannot assess the relative utility of predictors across studies
+* Include potential predictors from various theories, and undertheorized factors
+* Pitfalls:
     + Multicollinearity
-* Good theory incorporates most important causes
-    + Cannot assess the relative utility across studies
-* Include potential predictors from various theories, and undertheorized factors
+    + Colliders
 
@@
 
-## Person-centered approaches
+## Person-centered approaches {.smaller}
 
@@
     + Other *unsupervised learning* methods exist
+* Tree-based models
+    + "Why are these individuals similar?"
+    + Group individuals based on predictors to maximize homogeneity of the outcome
 * RI-CLPM and DSEM
@@
     + Regularization
-* Tree-based models
-    + Group individuals based on predictors to maximize homogeneity of the outcome
-    + "Why are these individuals similar?"
 
@@
 
-When using data models to guide theory formation, generalizability is essential
+When using data models to guide theory formation, generalizability is essential [@haslbeckModelingPsychopathologyData2021]
 
@@
 
-* cross-validation to estimate predictive accuracy
+* cross-validation to estimate predictive accuracy [@hastieElementsStatisticalLearning2009]
 
@@
 
-```{r, out.width="40%"}
+```{r}
+#| out.width = "40%"
 library(ggplot2)
@@
 
-## From models to theory
+## From models to theory {.smaller}
 
@@
 
-* abductive formal theory construction (AFTC) framework
+* Abductive formal theory construction (AFTC) framework
     + construct formal theory
@@
 
-# Example: Text Mining Systematic Review
+## Text Mining Systematic Review
 
-## Emotion regulation in adolescence
-
-Developmentally sensitive period [@zimmermannEmotionRegulationEarly2014]
-
-20% develop psychopathology [@leeAdolescentMentalHealth2014]
-
-Potentially lifelong implications for mental health and well-being
-
-```{r, out.height="20%"}
-knitr::include_graphics("trajectories.jpg")
-```
-
-<!-- ## Knowledge gap -->
-
-<!-- Substantial empirical research, but: -->
-
-<!-- * Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014] -->
-<!-- * No consistent terminology [@bariolaChildAdolescentEmotion2011] -->
-<!-- * No conceptual frameworks [@stifterEmotionRegulation2019] -->
-<!-- * No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019] -->
-
-<!-- Knowledge must be consolidated into overarching theory -->
-
-## Limitations of relevant theory
-
-* Few explicitly address adolescence
-* Few comprehensively address predictors of emotion regulation
-* None directly guide contemporary research 
-* Theories vary widely in scope: Some are broad and non-specific; others describe a specific phenomenon in detail, but lack a broader perspective.
-    + Broad theories can frame any research, specific theories generate hypotheses.
-    + It would be beneficial to bridge these levels of analysis.
-
-## Towards integrative theory
-
-* First step: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
-    + Often done by systematic review
-* Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
-* Introducing text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]:
-    + Unlimited sample size, transparent, objective, reproducible
-
-6653 papers on **emotion regulation** in **adolescence** [age 10-24]
-
-<https://doi.org/10.1007/s40894-021-00160-7>
-
-## Open science
-
-All code and data available at <https://github.com/cjvanlissa/veni_sysrev>
-
-Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]
-
-## Baseline
-
-Manual selection of key constructs from relevant theory and prior reviews [ @bariolaChildAdolescentEmotion2011; @coe-odessEmergentEmotionsAdolescence2019]:
-
-* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
-* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
-* Hall's notion of "storm and stress"
-* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
-* Tripartite model [@morrisRoleFamilyContext2007]
-* Internalization model [@holodynskiDevelopmentEmotionsEmotion2006]
-* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
-* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
-* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
-* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]
+* Identify which constructs are discussed in 6653 papers on **emotion regulation** in **adolescence** [age 10-24]
+* Identify which constructs co-occur in papers
+* Compare resulting networks to construct networks from existing theories and systematic reviews
+* Paper: <https://doi.org/10.1007/s40894-021-00160-7>
+* Code and data: <https://github.com/cjvanlissa/veni_sysrev>
+* Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]
 
@@
 
-```{r figbaseline, warning=FALSE, eval = TRUE, results='asis', out.width = "100%", fig.cap="Phenomena relevant to adolescents' emotion regulation according to theory (a) and narrative reviews (b; transparent nodes indicate constructs also present in the theory)."}
+```{r}
+#| label = "figbaseline",
+#| warning = FALSE,
+#| eval = TRUE,
+#| results = "asis",
+#| out.width = "100%",
+#| fig.cap = "Phenomena relevant to adolescents' emotion regulation according to theory
+#| (a) and narrative reviews (b; transparent nodes indicate constructs also present in
+#| the theory)."
 knitr::include_graphics("presentation/baseline_network.png") 
@@
 
-## {.fullslide}
-
-![](presentation/varimps_small.png)
-
 ## Co-occurrence graph {.smaller}
 
-```{r networks1, eval = TRUE, results='asis', out.width = "100%"}
+```{r}
+#| label = "networks1",
+#| eval = TRUE,
+#| results = "asis",
+#| out.width = "100%"
 knitr::include_graphics("presentation/tmnetworks_small.png")
@@
 
-## Results
+## Results {.smaller}
 
-* Both analyses reflected some constructs from theoretical literature
-    + Particularly related to neurodevelopment and socialization
+* Both analyses reflected constructs from theoretical literature
+    + Especially pertaining to neurodevelopment and socialization
 * Mental health-related outcomes feature prominently
@@
 
-## Implications
+<!-- ## Implications -->
 
-* Empirical research relies on theory; under-theorized phenomena may be overlooked
-    + TMSR offers guidance for phenomena to consider as confounders/causes
-* Unembedded terms indicate promising areas of future research
-* TMSR network can serve as proto-theoretical nomological network
+<!-- * Empirical research relies on theory; under-theorized phenomena may be overlooked -->
+<!--     + TMSR offers guidance for phenomena to consider as confounders/causes -->
+<!-- * Unembedded terms indicate promising areas of future research -->
+<!-- * TMSR network can serve as proto-theoretical nomological network -->
 
-# Example: Predictors of emotion regulation development
+## Predicting ER development {.smaller}
 
-## Person-centered models
+* What are important predictors of adolescents' ER development?
+* What is the nature of the association of the predictors with the trajectories?
+* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) 
+* Predictors: 87 (Demographics, Biological, Individual differences, Risk behavior, Relationship quality, Parenting, Conflict resolution styles)
+* Outcome: Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
+* Code and synthetic data: <https://github.com/cjvanlissa/veni_forest>
 
-```{r, out.height="50%"}
-knitr::include_graphics("groups.png")
-```
+## SEM Forests
 
-* Predictors of latent growth model?
-  + Only linear differences, few predictors (power)
-* Multigroup latent growth model?
-  + Non-linear differences, but only one moderator
-* Latent class growth model with auxilliary variables?
-  + Groups with regard to trajectory, not predictors
-
-## Method
-
-SEM-forest [@brandmaierTheoryguidedExplorationStructural2016]
-
-1. Bootstrap sample
-2. On each sample, estimate a SEM-model
-3. Consider $k$ candidate predictors
-4. Identify predictor and value that maximizes $LR$ of post-split multi-group model
-5. Average predictions across bootstrap samples
-
-We used 1000 bootstrap samples, $k = 9$
-
-## SEM Forest
-
-```{r, out.width="70%"}
+```{r}
+#| out.width = "70%"
 knitr::include_graphics("tree.png")
@@
 
-<!-- TS: Confirmatory versus exploratory research: Define the distinction. -->
-<!-- TS: Some longitudinal research falls between these categories.
-For example, complex longitudinal studies, where prior research is sufficient to lead us to hypothesize *some* associations between two constructs, but not sufficiently specific to let us hypothesize which parameters will be significant.
--->
+## Results {.smaller}
 
-<!-- ## Results -->
-
-<!-- * Both analyses reflected some constructs from theoretical literature -->
-<!--     + neurodevelopment, socialization -->
-<!-- * Undertheorized themes -->
-<!--     + Developmental disorders -->
-<!--     + Physical health (sic) -->
-<!--     + External stressors -->
-<!--     + Structural disadvantage -->
-<!--     + Addictive behavior -->
-<!--     + Identity and moral development -->
-<!--     + Sexual development -->
-
-<!-- * This offers guidance for phenomena to consider as relevant predictors -->
-
-## Study design
-
-RQ1: What are the most important predictors of adolescents' trajectories of emotion regulation development?
-
-RQ2: What is the nature of the association of the predictors with the trajectories?
-
-**Participants**
-
-* RADAR data (https://www.uu.nl/en/research/radar)
-* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46)
-* Collected between 2006-2011
-* Most families were  medium- to high-SES (10% low-SES)
-
-## Candidate predictors
-
-Predictors: 87
-
-Categories:
-
-* Demographics (e.g., SES)
-* Biological (age, sex, pubertal development)
-* Individual differences (big5, BIS/BAS, empathy)
-* Risk behavior (in/externalizing symptoms, substance use, delinquency, school performance, victimization)
-* Relationship quality (parents/peers)
-* Parenting (perceived and parent-reported)
-* Conflict resolution styles
-
-## Dependent variable
-
-Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
-
-* 24 items (omitted 'emotional awareness' subscale)
-* E.g.: “When I’m upset, I have difficulty thinking about anything else”
-* 1 (“Almost never”) to 5 (“Almost always”)
-* Excellent reliability, $(.94 > \alpha > .95)$.
-* Metric invariance across waves
-* Quadratic growth model had good fit
-  + CFI = .996, TLI = .993, RMSEA = .047, SRMR = 0.020
-
-## Open science
-
-Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]
-
-All code and synthetic data available at
-
-<https://github.com/cjvanlissa/veni_forest>
-
-# Results
-
-## Variable importance
-
-```{r, out.width="90%"}
-knitr::include_graphics("pcomb2.png")
-```
-
-## Variable importance 2
-
-```{r, out.width="100%"}
-knitr::include_graphics("violin.png")
-```
-
-## Marginal association
-
-```{r, out.width="40%"}
-knitr::include_graphics("plots/pdp1-12.png")
-```
-
-## Summary
-
-Important:
-
-* Best predictors: Personality and related constructs
-* Best parental predictor is autonomy support (bal. rel.)
-  + Negative/intrusive parenting more predictive than other positive practices
-* Conflict frequency and behavior
-* Empathy, particularly personal distress
+* Best predictors: Personality and conflict resolution behavior
+* Negative parenting more predictive than other positive practices
 
@@
 
-## Reflection content
+* Many predictors show non-linear effects
 
-* Proximal factors more predictive of ER development than distal predictors
-    + Order mirrors bioecological model
-* Most important predictors are routinely assessed (big 5) or overt (conflict)
-  + Prime candidates for early risk assessment
-  + Conflict resolution behavior can be taught
-    - Target for intervention if association is causal
-* Emphasis on parenting (in literature and theory) might not be justified
-  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
+## Conclusions
 
-## Reflections form
+* Machine learning reveals blind spots in existing theory
+    + Missing constructs, relative importance, functional form
+* ML can detect patterns in published research and primary data
+* "Inductive methods" for consolidating insights into theory remain vague
+    + Theories are still verbal and flexible, how should they be updated?
+* Ignoring causality limits interpretability of machine learning
+    + Incorrect direction of causality: meaningless results
+    + Conditioning on a collider: biased results
 
-* Many predictors show non-linear effects: Emotion dysregulation only for +/-1SD
-* Some predictors show almost no marginal effect (e.g., father's age, drug use)
-  + This suggests they might be important in **interactions**
+<!-- ## Future direction 1: Formal theoretical object -->
 
-# Limitations and future directions
+<!-- Formal theories are expressed in mathematical terms -->
 
-## Limitations
+<!-- * unambiguous -->
+<!-- * logically sound -->
+<!-- * explicates assumptions -->
+<!-- * reveals ambiguities and gaps in knowledge -->
+<!-- * facilitates conveying ideas to others -->
+<!-- * applying rules of formal logic implies testable hypotheses -->
+<!-- * refutation of hypotheses dictates specific modifications -->
 
-1. Theories are still verbal and flexible
-    * Even if ML leads to new insights, it is not clear how theories should be updated
-2. Ignoring causality limits interpretability of machine learning
-    * Incorrect direction of causality: meaningless results
-    * Conditioning on a collider: biased results
+<!-- ## Characteristics of formal theory -->
 
-## Future direction 1: Formal theoretical object
+<!-- According to Meehl: -->
 
-Formal theories are expressed in mathematical terms
+<!-- 1. phenomena to be explained -->
+<!-- 2. putative causal connections between them -->
+<!-- 3. direction of the effect -->
+<!-- 4. relative importance of predictors -->
+<!-- 5. interaction effects -->
+<!-- 6. functional form of associations -->
 
-* unambiguous
-* logically sound
-* explicates assumptions
-* reveals ambiguities and gaps in knowledge
-* facilitates conveying ideas to others
-* applying rules of formal logic implies testable hypotheses
-* refutation of hypotheses dictates specific modifications
+<!-- ## Formal theory -->
 
-## Characteristics of formal theory
+<!-- * Characteristics 1-2 can be described by a causal network (DAG) -->
+<!-- * Characteristics 3-6 refer to the nature of associations within that network -->
 
-According to Meehl:
+<!-- Augmented causal networks (ACN, see also Crielaard et al., 2022) -->
 
-1. phenomena to be explained
-2. putative causal connections between them
-3. direction of the effect
-4. relative importance of predictors
-5. interaction effects
-6. functional form of associations
+<!-- * encode both causal relationships and the nature of those relationships -->
+<!-- * imply testable hypotheses -->
+<!-- * shared in plain text, e.g., YAML or RDF -->
+<!-- * version controlled -->
+<!-- * shared conform FAIR software principles, -->
+<!-- to be used and updated through distributed collaboration -->
 
-## Formal theory
+<!-- ## Learning a causal network -->
 
-* Characteristics 1-2 can be described by a causal network (DAG)
-* Characteristics 3-6 refer to the nature of associations within that network
+<!-- * Using TMSR, based on what previous authors have written -->
+<!-- * Using causal discovery, based on large panel studies -->
+<!-- * Using ESM data to leverage precedence in time -->
+<!--     + Causality requires association, precedence in time, and nonspuriousness -->
 
-Augmented causal networks (ACN, see also Crielaard et al., 2022)
+<!-- ## Future direction 2: Functional form -->
 
-* encode both causal relationships and the nature of those relationships
-* imply testable hypotheses
-* shared in plain text, e.g., YAML or RDF
-* version controlled
-* shared conform FAIR software principles,
-to be used and updated through distributed collaboration
+<!-- Previous work: Post-hoc interpretation of marginal associations -->
 
-## Learning a causal network
+<!-- Future direction: Use a-priori interpretable machine learning -->
 
-* Using TMSR, based on what previous authors have written
-* Using causal discovery, based on large panel studies
-* Using ESM data to leverage precedence in time
-    + Causality requires association, precedence in time, and nonspuriousness
-
-## Future direction 2: Functional form
-
-Previous work: Post-hoc interpretation of marginal associations
-
-Future direction: Use a-priori interpretable machine learning
-
-* Symbolic regression applies deep learning or genetic algorithms to learn a mathematical formula that describes the relationship between an outcome and its predictors
+<!-- * Symbolic regression applies deep learning or genetic algorithms to learn a mathematical formula that describes the relationship between an outcome and its predictors -->
 

### 2023-07-12 presentation_isdp2023.qmd
update 2023-07-12 12:23:25.619832

@@
 title: "Machine Learning can Advance Theory Formation in the Social Sciences"
-subtitle: "Funded by NWO Veni grant VI.Veni.191G.090<br>[doi:10.1002/icd.2370](https://doi.org/10.1002/icd.2370)"
+subtitle: "Funded by NWO Veni grant VI.Veni.191G.090<br>Paper: [doi:10.1002/icd.2370](https://doi.org/10.1002/icd.2370)"
 author: "Caspar J. Van Lissa"
@@
 * Current solutions improve deductive research (theory-testing)
+    + preregistration [@peikertReproducibleResearchTutorial2021]
+    + replication
     + questionable research practices [@wichertsDegreesFreedomPlanning2016]
-    + replication
-    + preregistration
 
@@
 
-**Theory**: A model that describes the nature of relations between several phenomena in sufficient detail that it allows for the derivation of quantifiable hypotheses that can be subject to (severe) testing.
-
-* Most social scientific theories are:
-    + insufficiently precise
-    + too flexible
-* Compare to @walasekEvolutionaryModelSensitive2022
-
 Why do we need theory?
@@
 * Useful as an instrument of cumulative knowledge acquisition
-* Communicating social scientific research to its consumers
+* Communicating social scientific research to its consumers (parents, practitioners, policy makers)
 
@@
 * Confirmatory methods poorly suited to exploratory research
-    + p-values: Type I error is 5%, but false discovery rate (FDR) is 14-50% [@vidgenPValuesMisunderstoodMisused2016].
-    + Model fit indices: manually specify models, no guarantee that best model is included
+    + p-values: 5% Type I error prob is consistent with 14-50% false discovery rate [@vidgenPValuesMisunderstoodMisused2016]
+    + Depending on prior probability of hypotheses being true
+    + Model fit indices: manually specify models, no guarantee that best model is in set
 * Role of flexibility: there are infinite ways to explore
@@
 * automated model building
-* checks and balances to prevent overfitting
-* maximize predictive performance and generalizability of results
 
@@
 
-* Phenomena: stable and general features of the world
-* Few quantitative methods for phenomena detection (aside from expert opinion)
+* Phenomena: stable features of the world reliably evidenced by patterns in data (Woodward, 1989)
 
-**Text mining** may be suitable for phenomena detection [@vanlissaMappingPhenomenaRelevant2021]
+Machine learning  [@vanlissaMappingPhenomenaRelevant2021]
 
-* Unsupervised learning method
-* Clusters in keywords and abstracts
+* Unsupervised learning methods find patterns in data
+* Text mining finds patterns in published literature
 
@@
 * Include potential predictors from various theories, and undertheorized factors
-* Pitfalls:
-    + Multicollinearity
-    + Colliders
 
 ## Complex effects
+
+Social scientific theories rarely account for complex effects
 
@@
 
-Social scientific theories rarely account for complex effects
-
-* Machine learning can provide nuance by revealing them
-
 ## Theoretical elements
@@
 * SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
-* regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)
+* Regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)
 
@@
 
-* E.g., latent class analyses
+* Unsupervised learning methods
+    + Group individuals to maximize homogeneity of some outcomes
     + "Which individuals are similar?"
-    + Other *unsupervised learning* methods exist
 * Tree-based models
+    + Group individuals by predictors to maximize homogeneity of outcomes
     + "Why are these individuals similar?"
-    + Group individuals based on predictors to maximize homogeneity of the outcome
-* RI-CLPM and DSEM
-    + Rarely explain heterogeneity in within-person effects
-    + Regularization
+* Mixed-effects / multilevel models (DSEM, RI-CLPM)
+    + Often model but rarely explain heterogeneity in within-person effects
+    + Regularization (e.g., regularized SEM)
 
@@
 
-Checks and balances to curtail *overfitting*
+Machine learning
 
-* cross-validation to estimate predictive accuracy [@hastieElementsStatisticalLearning2009]
+* Incorporates checks and balances to curtail *overfitting*
+* Maximizes predictive performance and generalizability of results
+* Cross-validation to estimate predictive accuracy [@hastieElementsStatisticalLearning2009]
+* Possible to estimate generalizability to representative subsample (e.g., Douglas, Sutton, Van Lissa et al., 2023)
 
@@
     + any undertheorized factors rank high?
-
 * Marginal associations
@@
     + high importance but flat marginal association?
-
-Comparing predictive performance of simpler parametric model that represents these 'insights' to machine learning model
-
-## From models to theory 2
-
-Data models -> formal theory [@haslbeckModelingPsychopathologyData2021]
-
-* Abductive formal theory construction (AFTC) framework
-    + construct formal theory
-    + Generate data
-    + Mathematically compare to empirical data
-    + like a posterior predictive check in Bayes
+* Abductive formal theory construction (AFTC) framework [@haslbeckModelingPsychopathologyData2021]
+    + Translate insights into formal theory
+    + Conduct "posterior predictive check" to check fit with data
     + Discrepancies: amend formal theory
@@
 
-## Text Mining Systematic Review
+## Example: TMSR {.smaller}
 
-* Identify which constructs are discussed in 6653 papers on **emotion regulation** in **adolescence** [age 10-24]
-* Identify which constructs co-occur in papers
-* Compare resulting networks to construct networks from existing theories and systematic reviews
-* Paper: <https://doi.org/10.1007/s40894-021-00160-7>
-* Code and data: <https://github.com/cjvanlissa/veni_sysrev>
-* Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]
+* Identifying relevant phenomena
+* Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
+* Text Mining Systematic Review [@vanlissaMappingPhenomenaRelevant2021]
+    + Unlimited sample size, transparent, objective, reproducible
+* 6653 papers on **emotion regulation** in **adolescence** [age 10-24]
 
-## Baseline network {.smaller}
+Paper: <https://doi.org/10.1007/s40894-021-00160-7>
+Reproducible code and data: <https://github.com/cjvanlissa/veni_sysrev>
 
-```{r}
-#| label = "figbaseline",
-#| warning = FALSE,
-#| eval = TRUE,
-#| results = "asis",
-#| out.width = "100%",
-#| fig.cap = "Phenomena relevant to adolescents' emotion regulation according to theory
-#| (a) and narrative reviews (b; transparent nodes indicate constructs also present in
-#| the theory)."
-knitr::include_graphics("presentation/baseline_network.png") 
-```
-
-(a) Theory (b) narrative reviews; transparent nodes indicate constructs also present in theory.
+Workflow for Open Reproducible Code in Science (WORCS)  [@vanlissaWORCSWorkflowOpen2020]
 
@@
 
-* Both analyses reflected constructs from theoretical literature
+* TM analyses retrieve constructs from theoretical literature
     + Especially pertaining to neurodevelopment and socialization
-* Mental health-related outcomes feature prominently
-    + Emotion regulation implicated in mental health problems [@leeAdolescentMentalHealth2014],
-    + This underlines the importance of research in this area
-* Substantial correspondence between keywords and abstracts, suggesting validity of method
-* Networks are sparse; few connections among constructs: fragmented literature
+    + This suggests validity
+* Substantial correspondence between keyword and abstract network
+    + This suggests reliability
+* Networks are sparse; few connections among constructs:
+    + Fragmented literature, as noted by prior reviews
 
@@
 
-<!-- ## Implications -->
+## Implications
 
-<!-- * Empirical research relies on theory; under-theorized phenomena may be overlooked -->
-<!--     + TMSR offers guidance for phenomena to consider as confounders/causes -->
-<!-- * Unembedded terms indicate promising areas of future research -->
-<!-- * TMSR network can serve as proto-theoretical nomological network -->
+* Empirical research relies on theory; under-theorized phenomena may be overlooked
+    + TMSR offers guidance for phenomena to consider as confounders/causes
+* Unembedded terms indicate promising areas of future research
+* TMSR network can serve as proto-theoretical nomological network
 
-## Predicting ER development {.smaller}
+## Example: Predictors of Development {.smaller}
 
-* What are important predictors of adolescents' ER development?
-* What is the nature of the association of the predictors with the trajectories?
+* What are important predictors of ER development trajectories?
+* What is the nature of association of predictors with trajectories?
 * 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) 
@@
 
+## Variable importance
+
+```{r, out.width="90%"}
+knitr::include_graphics("pcomb2.png")
+```
+
+## Variable importance 2
+
+```{r, out.width="100%"}
+knitr::include_graphics("violin.png")
+```
+
+## Marginal association
+
+```{r, out.width="40%"}
+knitr::include_graphics("plots/pdp1-12.png")
+```
+
 ## Results {.smaller}
 
-* Best predictors: Personality and conflict resolution behavior
-* Negative parenting more predictive than other positive practices
+* Best predictors: Personality and conflict behaviors
+* Negative parenting more predictive than positive parenting
 
@@
 
-* Many predictors show non-linear effects
+* Most predictors show non-linear effects
+
+## Reflection content
+
+* Proximal factors more predictive of ER development than distal predictors
+    + Order mirrors bioecological model
+* Most important predictors are routinely assessed (big 5) or overt (conflict)
+  + Prime candidates for early risk assessment
+  + Conflict resolution behavior can be taught
+    - Target for intervention if association is causal
+* Emphasis on parenting (in literature and theory) might not be justified
+  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
+
+## Reflections form
+
+* Many predictors show non-linear effects: Emotion dysregulation only for +/-1SD
+* Some predictors show almost no marginal effect (e.g., father's age, drug use)
+  + This suggests they might be important in **interactions**
 
@@
 * ML can detect patterns in published research and primary data
-* "Inductive methods" for consolidating insights into theory remain vague
-    + Theories are still verbal and flexible, how should they be updated?
+* Does not solve the "inductive problem"
+    + Still your responsibility to consolidate insights into theory
 * Ignoring causality limits interpretability of machine learning
@@
 
-Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. https://doi.org/10.1002/icd.2370
+Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. ICD. DOI:10.1002/icd.2370
 
-Van Lissa, C.
-J.
-(2022).
-Complementing preregistered confirmatory analyses with rigorous, reproducible exploration using machine learning.
-Religion, Brain & Behavior.
-https://doi.org/10.1080/2153599X.2022.2070254
+Van Lissa, C. J. (2022). Complementing preregistered confirmatory analyses with rigorous, reproducible exploration using machine learning. Religion, Brain & Behavior. DOI:10.1080/2153599X.2022.2070254
 
@@
 Journal of Research on Adolescence.
-https://doi.org/10.31234/osf.io/rbqf9
+DOI:10.31234/osf.io/rbqf9
 
-Van Lissa, C. J. (2021). Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-Mining Systematic Review. Adolescent Research Review. https://doi.org/10.1007/s40894-021-00160-7
+Van Lissa, C. J. (2021). Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-Mining Systematic Review. ARR. DOI:10.1007/s40894-021-00160-7
 
@@
 Patterns, 3(4), 100482.
-https://doi.org/10.1016/j.patter.2022.100482
+DOI:10.1016/j.patter.2022.100482
+
+Douglas, K., Sutton, R.
+M., Van Lissa, C.
+J., Wolfgang, S., Jannis, K., Maximilian, A., ...
+& N Pontus, L.
+(2023).
+Identifying Important Individual-and Country-Level Predictors of Conspiracy Theorizing.
+EJSP.
+DOI: 10.1002/ejsp.2968
 

### 2023-07-12 presentation_tuebingen2023.qmd
update 2023-07-12 12:23:25.619832

--- 2023-07-12 presentation_isdp2023.qmd
@@
 
+**Theory**: A model that describes the nature of relations between several phenomena in sufficient detail that it allows for the derivation of quantifiable hypotheses that can be subject to (severe) testing.
+
+* Most social scientific theories are:
+    + insufficiently precise
+    + too flexible
+* Compare to @walasekEvolutionaryModelSensitive2022
+
 Why do we need theory?
@@
 * Useful as an instrument of cumulative knowledge acquisition
-* Communicating social scientific research to its consumers (parents, practitioners, policy makers)
+* Communicating social scientific research to its consumers
 
@@
 
+## Machine learning vs unguided exploration
+
+```{r, out.width="100%"}
+knitr::include_graphics("Figure Machine learning 4.png")
+```
+
 ## Phenomena detection
@@
 
-Social scientific theories rarely account for complex effects
-
 Many machine learning methods accommodate:
@@
 
+Social scientific theories rarely account for complex effects
+
+* Machine learning can provide nuance by revealing them
+
 ## Theoretical elements
@@
 * SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
-* Regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)
+* regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)
 
@@
     + "Which individuals are similar?"
+    + Non-parametric (e.g., K-means) and model-based (e.g., LCA)
 * Tree-based models
@@
     + "Why are these individuals similar?"
-* Mixed-effects / multilevel models (DSEM, RI-CLPM)
+* Mixed-effects / multilevel models
     + Often model but rarely explain heterogeneity in within-person effects
@@
 
-Machine learning
+Machine learning...
 
@@
     + any undertheorized factors rank high?
+
 * Marginal associations
@@
     + high importance but flat marginal association?
-* Abductive formal theory construction (AFTC) framework [@haslbeckModelingPsychopathologyData2021]
-    + Translate insights into formal theory
-    + Conduct "posterior predictive check" to check fit with data
+
+Comparing predictive performance of simpler parametric model that represents these 'insights' to machine learning model
+
+## From models to theory 2
+
+Data models -> formal theory [@haslbeckModelingPsychopathologyData2021]
+
+* Abductive formal theory construction (AFTC) framework
+    + construct formal theory
+    + Generate data
+    + Mathematically compare to empirical data
+    + like a posterior predictive check in Bayes
     + Discrepancies: amend formal theory
@@
 
-## Example: TMSR {.smaller}
+## Example: TMSR
 
-* Identifying relevant phenomena
+* Text Mining Systematic Review
+* First step in theory development: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
+    + Often done by systematic review
 * Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
-* Text Mining Systematic Review [@vanlissaMappingPhenomenaRelevant2021]
+* Introducing text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]:
     + Unlimited sample size, transparent, objective, reproducible
-* 6653 papers on **emotion regulation** in **adolescence** [age 10-24]
 
-Paper: <https://doi.org/10.1007/s40894-021-00160-7>
-Reproducible code and data: <https://github.com/cjvanlissa/veni_sysrev>
+6653 papers on **emotion regulation** in **adolescence** [age 10-24]
 
-Workflow for Open Reproducible Code in Science (WORCS)  [@vanlissaWORCSWorkflowOpen2020]
+<https://doi.org/10.1007/s40894-021-00160-7>
+
+## Open science
+
+All code and data available at <https://github.com/cjvanlissa/veni_sysrev>
+
+Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]
+
+## Design
+
+1. Manually construct baseline network from relevant theory and prior reviews
+2. Identify which constructs co-occur in papers
+    + Use POS-tagging to identify nouns and adjectives
+    + Use (Google pagerank-based) TextRank to identify bigrams
+    + Use regex dictionary to cluster terms
+3. Frequency -> Relevance
+4. Co-occurrence in publications -> Potential association
+5. Compare co-occurrence networks to baseline network
+
+## Baseline network {.smaller}
+
+```{r}
+#| label = "figbaseline",
+#| warning = FALSE,
+#| eval = TRUE,
+#| results = "asis",
+#| out.width = "100%",
+#| fig.cap = "Phenomena relevant to adolescents' emotion regulation according to theory
+#| (a) and narrative reviews (b; transparent nodes indicate constructs also present in
+#| the theory)."
+knitr::include_graphics("presentation/baseline_network.png") 
+```
+
+(a) Theory (b) narrative reviews; transparent nodes indicate constructs also present in theory.
 
@@
 
-* What are important predictors of ER development trajectories?
-* What is the nature of association of predictors with trajectories?
+* What are important predictors of adolescents' ER development?
+* What is the nature of the association of the predictors with the trajectories?
 * 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) 
@@
 ```
+
+## Method
+
+SEM-forest [@brandmaierTheoryguidedExplorationStructural2016]
+
+1. Bootstrap sample
+2. On each sample, estimate a SEM-model
+3. Consider $k$ candidate predictors
+4. Identify predictor and value that maximizes $LR$ of post-split multi-group model
+5. Average predictions across bootstrap samples
+
+We used 1000 bootstrap samples, $k = 9$
+
+## Relation to other methods
+
+```{r, out.height="50%"}
+knitr::include_graphics("groups.png")
+```
+
+* Predictors of latent growth model?
+  + Only linear differences, few predictors (power)
+* Multigroup latent growth model?
+  + Non-linear differences, but only one moderator
+* Latent class growth model with auxilliary variables?
+  + Groups with regard to trajectory, not predictors
+* SEM forest
+  + Groups with regard to trajectory, based on predictors
 
@@
 
-## Reflection content
+## Limitations
 
-* Proximal factors more predictive of ER development than distal predictors
-    + Order mirrors bioecological model
-* Most important predictors are routinely assessed (big 5) or overt (conflict)
-  + Prime candidates for early risk assessment
-  + Conflict resolution behavior can be taught
-    - Target for intervention if association is causal
-* Emphasis on parenting (in literature and theory) might not be justified
-  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]
+1. Problem of induction
+2. Causal inference
 
-## Reflections form
+## Induction {.smaller}
 
-* Many predictors show non-linear effects: Emotion dysregulation only for +/-1SD
-* Some predictors show almost no marginal effect (e.g., father's age, drug use)
-  + This suggests they might be important in **interactions**
+It is impossible to logically induce general principles from specific data
+
+* You may have unobserved counterexamples
+* Multiple principles could be consistent with the same data
+
+But it is possible to reason about general principles based on observed patterns in data data
+
+* Popper:
+    + "there is no such thing as a logical method of having new ideas"
+    + "my arguments in this book are quite independent of [the problem of having new ideas]"
+* Machine learning is *one possible method* for generating ideas inductively
+
+## Causal Inference {.smaller}
+
+Machine learning finds patterns in data, irrespective of causality
+
+* If you goal is prediction, this is fine
+* But interpreting those patterns assumes causality
+
+Be particularly mindful of:
+
+* Conditioning on colliders
+* Multicolinearity
+
+See Pearl 2019
 
@@
 * Does not solve the "inductive problem"
-    + Still your responsibility to consolidate insights into theory
+    + Still your job to consolidate insights into theory
 * Ignoring causality limits interpretability of machine learning
@@
 
-Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. ICD. DOI:10.1002/icd.2370
+Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. DOI:10.1002/icd.2370
 
@@
 
-Van Lissa, C. J. (2021). Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-Mining Systematic Review. ARR. DOI:10.1007/s40894-021-00160-7
+Van Lissa, C. J. (2021). Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-Mining Systematic Review. Adolescent Research Review. DOI:10.1007/s40894-021-00160-7
 
@@
 Douglas, K., Sutton, R.
-M., Van Lissa, C.
-J., Wolfgang, S., Jannis, K., Maximilian, A., ...
+M., Caspar, V.
+L., Wolfgang, S., Jannis, K., Maximilian, A., ...
 & N Pontus, L.
 (2023).
-Identifying Important Individual-and Country-Level Predictors of Conspiracy Theorizing.
-EJSP.
+Identifying Important Individual-and Country-Level Predictors of Conspiracy Theorizing: A Machine Learning Analysis.
+European Journal of Social Psychology.
 DOI: 10.1002/ejsp.2968

### 2023-11-24 presentation_copenhagen2023.Rmd
update 2023-11-24 10:41:01.673245

--- 2023-05-23 presentation_mpib2023.Rmd
@@
 ---
-title: "How Machine Learning can Advance Theory Formation"
+title: "Machine Learning can Advance Theory Formation in the Social Sciences"
 author: "Caspar J. Van Lissa<sup>1</sup>"
@@
 
-**Theory**: A model that describes the nature of relations between several phenomena in sufficient detail that it allows for the derivation of quantifiable hypotheses that can be subject to (severe) testing.
+> **Theory**: A (causal?) explanation of phenomena reliably evidenced by data [@bogenSavingPhenomena1988]
 
-* Most social scientific theories are:
-    + insufficiently precise
-    + too flexible
-* Compare to @walasekEvolutionaryModelSensitive2022
+* Allows for derivation of testable hypotheses for deductive research
 
-Why do we need theory?
+## Characteristics of Good Theory
 
+@meehlAppraisingAmendingTheories1990
+
+1. phenomena to be explained
+2. putative causal connections between them
+3. direction of the effect
+4. relative importance of predictors
+5. interaction effects
+6. functional form of associations
+
+Most social scientific theories are [@scheelWhyMostPsychological2022]:
+
+* insufficiently precise
+* too flexible
+* But cf. @walasekEvolutionaryModelSensitive2022
+
+## Why do we need theory?
+
+* Social science strives to understand, not only predict
 * Useful as an instrument of cumulative knowledge acquisition
@@
 
-De Groot’s empirical cycle: a model of knowledge production through scientific research [@degrootMethodologieGrondslagenVan1961]
+De Groot’s empirical cycle
+
+* Model of cumulative knowledge acquisition through scientific research [@degrootMethodologieGrondslagenVan1961]
 
@@
 
-* automated model building
-* checks and balances to prevent overfitting
-* maximize predictive performance and generalizability of results
+* Automated model building
+* Checks and balances to prevent overfitting
+* Maximize predictive performance and generalizability of results
 
@@
 First step in *Theory Construction Methods* is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
-
+*
+* Few quantitative methods for phenomena detection (aside from expert opinion)
 * Phenomena: stable and general features of the world
-* Few quantitative methods for phenomena detection (aside from expert opinion)
-
-**Text mining** may be suitable for phenomena detection [@vanlissaMappingPhenomenaRelevant2021]
-
-* Unsupervised learning method
-* Clusters in keywords and abstracts
+* **Cross-validated patterns of association** may constitute phenomena
+* **Text mining** existing literature may reveal what phenomena other authors have identified [@vanlissaMappingPhenomenaRelevant2021]
 
@@
 * Assumption for causal interpretation
-    + Multicollinearity
+    + But mind multicolinearity and collider bias
 * Good theory incorporates most important causes
-    + Cannot assess the relative utility across studies
 * Include potential predictors from various theories, and undertheorized factors
+    + Assess the relative utility across studies
+    + Cannot do this across studies
 
@@
 
-Some machine learning methods incorporate theoretical elements
+Some methods allow for theory guided exploration using machine learning
 
-* E.g.: assumption that development follows a latent growth curve (LGC).
+When (part of) the theory can be expressed as a structural equation model, and part still remains to be discovered through exploratory research
+
+* E.g., some relevant constructs are latent variables
+* E.g., development follows a latent growth curve (LGC)
 * SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
@@
 
-All of these methods allow for theory guided exploration using machine learning
+## Person-centered approaches
 
-## Person-centered approaches
+Call for person-centered research [@laursenPersonCenteredVariableCenteredApproaches2006]
 
@@
 
-* E.g., latent class analyses
+* Latent class analyses
     + "Which individuals are similar?"
     + Other *unsupervised learning* methods exist
-* RI-CLPM and DSEM
-    + Rarely explain heterogeneity in within-person effects
-    + Regularization
+* Multilevel models
+    + Allow for, but rarely explain heterogeneity
+    + Includes RI-CLPM and DSEM
+    + Could use regularized SEM or SEM forests
 * Tree-based models
@@
 
-When using data models to guide theory formation, generalizability is essential
+When using data models to guide theory formation, generalizability is essential [@haslbeckModelingPsychopathologyData2021]
 
-Checks and balances to curtail *overfitting*
+ML: Checks and balances to curtail *overfitting*
 
@@
 
-* Variable importance metrics
+* Start with "interpretable machine learning" metrics [@henningerInterpretableMachineLearning2022]
+
+* Variable importance
     + congruent with theoretical assumptions about important predictors?
@@
     + high importance but flat marginal association?
-
-Comparing predictive performance of simpler parametric model that represents these 'insights' to machine learning model
+* Exemplar-based interpretation
 
@@
 
-* abductive formal theory construction (AFTC) framework
+* Abductive Formal Theory Construction (AFTC) framework
     + construct formal theory
@@
     + Discrepancies: amend formal theory
+* "Light" version of AFTC: Comparing predictive performance of simpler parametric model based on "interpretation" to more flexible machine learning model
 
@@
 
-<!-- ## Knowledge gap -->
+## Knowledge gap
 
-<!-- Substantial empirical research, but: -->
+Substantial empirical research, but:
 
-<!-- * Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014] -->
-<!-- * No consistent terminology [@bariolaChildAdolescentEmotion2011] -->
-<!-- * No conceptual frameworks [@stifterEmotionRegulation2019] -->
-<!-- * No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019] -->
+* Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014]
+* No consistent terminology [@bariolaChildAdolescentEmotion2011]
+* No conceptual frameworks [@stifterEmotionRegulation2019]
+* No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019]
 
-<!-- Knowledge must be consolidated into overarching theory -->
+Knowledge must be consolidated into overarching theory
 
@@
 * TMSR network can serve as proto-theoretical nomological network
+
+## Next Steps
+
+Rasoul Norouzi is further developing this method
+
+* Goal: Reconstructing a causal loop diagram based on authors' causal claims in literature
+
+Together with Denny Borsboom, applying TMSR v2 to Cooperation Databank
+
+* Using sciBERT vector embeddings and HDBSCAN clustering instead of dictionary
 
@@
 
+# Example: Predictors of Infection Prevention Behavior
+
+## Context
+
+Prior to vaccines, infection prevention behaviors were the only defense against COVID
+
+Behavioral science could thus inform policy
+
+Existing theories suggest a limited set of relevant predictors
+
+We used ML to cast a wider net and identify important predictors at the individual and country level
+
+## Study design
+
+**Participants**
+
+* PsyCorona consortium
+* 56.072 participants across 28 countries
+* Collected between March-May 2020
+
+## Variables
+
+* DV: Self-reported infection prevention behaviors (handwashing, avoiding crowds, isolation)
+* IVs: 80 individual-level variables (16 multi-item composites)
+* Enriched with 35 country-level variables
+    + E.g., political stability, government effectiveness
+    + Some matched to date of participation (pandemic severity, government response)
+
+## Open science
+
+Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]
+
+All code and synthetic data available at
+
+<https://github.com/cjvanlissa/covid19_metadata>
+
+# Results
+
+## Variable importance
+
+```{r, out.height="100%"}
+knitr::include_graphics("presentation/varimp_covid.png")
+```
+
+## Summary
+
+* Injunctive norms most important
+    + Consistent with theory of planned behavior
+* Coping styles were important:
+    + Problem-focused coping, focus on the present/ future, hopefulness
+* Five of most important predictors were country-level variables
+    + E.g., national healthcare infrastructure
+
+## Notable absences
+
+* Perceived likelihood of infection
+    + Contradicts Health Belief Model
+* Psychological and material struggle
+    + E.g., loneliness, boredom, job insecurity
+* No evidence for "intergenerational conflict of interest"
+    + Confirmed by preregistered studies
+
+## Strengths
+
+This sample included a representative sub-sample
+
+* Allowed us to estimate generalizability
+* Predictive $R^2$ in the representative subsample was even higher than in the total hold-out sample ($R^2_{rep} = .59$, $R^2_{test} = .52$)
+
 # Limitations and future directions
@@
 
-## Characteristics of formal theory
+## Recall: Characteristics of Good Theory
 
-According to Meehl:
+@meehlAppraisingAmendingTheories1990
 
@@
 
-## Formal theory
+## Formal Theory
 
-* Characteristics 1-2 can be described by a causal network (DAG)
-* Characteristics 3-6 refer to the nature of associations within that network
-
-Augmented causal networks (ACN, see also Crielaard et al., 2022)
+* 1-2 can be described by a causal network (DAG), 3-6 refer to the nature of associations within that network
+* Augmented causal networks (ACN, see Crielaard et al., 2022)
 
@@
 
+Working on this with Aaron Peikert, Andreas Brandmaier and Felix Schönbrodt
+
 ## Learning a causal network
 
-* Using TMSR, based on what previous authors have written
-* Using causal discovery, based on large panel studies
-* Using ESM data to leverage precedence in time
-    + Causality requires association, precedence in time, and nonspuriousness
+For example, using TMSR, based on what previous authors have written
 
-## Future direction 2: Functional form
+Working on this with my PhD student, Rasoul Norouzi
 
-Previous work: Post-hoc interpretation of marginal associations
+## Epistemology
 
-Future direction: Use a-priori interpretable machine learning
+Working on an epistemology of machine learning-informed theory formation
 
-* Symbolic regression applies deep learning or genetic algorithms to learn a mathematical formula that describes the relationship between an outcome and its predictors
+With Denny Borsboom and James Woodward
 
@@
 
-Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. https://doi.org/10.1002/icd.2370
+<font size = 4>
+**Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. https://doi.org/10.1002/icd.2370**
 
@@
 
-##
+Douglas, K., Sutton, R.
+M., Van Lissa, C.
+J., ...
+& N Pontus, L.
+(2023).
+Identifying Important Individual-and Country-Level Predictors of Conspiracy Theorizing: A Machine Learning Analysis.
+European Journal of Social Psychology.
+DOI: 10.1002/ejsp.2968
+
+</font>
+
+## More refs
